### 2017-05-26 paper/prelims.tex
added a folder with basics of the paper writeup

\vspace{-1em}
\section{Preliminaries}\label{sec:prelim}

\vspace{-1em}
\subsection{Basic Notions, Linear Predicates, Valuations}
\vspace{-0.5em}
For a set $A$ we denote by $|A|$ the cardinality of $A$. We denote by $\Nset$,
$\Nset_0$, $\Zset$, and $\Rset$ the sets of all positive integers, non-negative
integers, integers, and real numbers, respectively. We assume basic knowledge 
of matrix calculus.
We use boldface notation for
vectors, e.g. $\vec{x}$, $\vec{y}$, etc., and we denote an $i$-th component of a
vector $\vec{x}$ by $\vec{x}[i]$. For the purpose of matrix calculations we 
assume that (non-transposed) vectors are row vectors. If $\vec{v},\vec{v}'$ are 
$n$ and $m$ dimensional vectors, respectively, then $(\vec{v},\vec{v}')$ is an 
$(n+m)$-dimensional vector obtained by ``concatenation'' of $\vec{v}$ and 
$\vec{v}'$. We identify 1-dimensional vectors with numbers. For an 
$n$-dimensional vector 
$\vec{x}$, index $1 \leq i\leq n$, and number $a$ we denote by $\vec{x}(i\leftarrow a)$ 
a 
vector $\vec{y}$ such that $\vec{y}[i]=a$ and $\vec{y}[j]=\vec{x}[j]$ for all 
$1\leq j \leq n$, $j\neq i$.
For comparison of vectors (e.g. as in $\vec{x}\leq \vec{y}$), we consider componentwise 
comparison. 
For comparing functions $f,g$ with the same domains, we write $f\leq g$ 
if $f(x)\leq g(x)$ for all $x$ in the domain.

\smallskip\noindent{\em Variables and valuations.}
Throughout the paper we fix a countable set of variables $\vars$. 
We consider some arbitrary but fixed linear order on the set of all variables. 
Hence, given some set of variables $V$ we can enumerate its members in 
ascending order (w.r.t. the fixed ordering) and write 
$V=\{x_1,x_2,x_3,\dots\}$. 
%A \emph{valuation} is a partial function 
%$\val\colon 
%\vars \dashrightarrow \Rset$ assigning a unique number to some of the 
%variables. A \emph{domain} of a valuation $\val$, denoted by %$\dom(\val)$, is 
%the set of all variables $x$ such that $\val(x)$ is defined. Given a %set 
%$\pvars\subseteq \vars$ of variables, we denote by $\vals(\pvars)$ the %set of 
%all valuations $\val$ with $\dom(\val)=\pvars$.

\smallskip\noindent{\em Affine expressions.}
An \emph{affine expression} over the set of variables $\{x_1,\dots,x_n\}$ is an 
expression of the form $d+\sum_{i=1}^{n}a_i
x_i$, where  $d,a_1,\dots,a_n$ are real-valued
constants. Each affine expression $E$ over $\{x_1,\dots,x_n\}$ 
determines a function which for each $m$-dimensional vector $\vec{x}$, where $m\geq n$,  
returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
Slightly abusing our notation, we denote this function also by $E$ and the 
value of this function on argument $\vec{x}$ by $E(\vec{x})$. A function of the form $E(\vec{x})$ for some affine expression $E$ is called affine.
%Following the terminology of~\cite{DBLP:conf/sas/KatoenMMM10} we fix the
%following nomenclature:

\smallskip\noindent{\em Linear constraint, assertion, predicates.}
We use the following nomenclature:
\begin{compactitem}
\item {\em Linear Constraint.} A \emph{linear constraint} is a formula of the
form $\psi$ or $\neg\psi$, where  $\psi$ is a non-strict inequality
between affine expressions.
\item {\em Linear Assertion.} A \emph{linear assertion} is a finite conjunction
of linear constraints.
\item {\em Propositionally Linear Predicate.}
A  \emph{propositionally linear predicate} (PLP) is a finite disjunction of
linear assertions.
\end{compactitem}

\noindent{\em Arity and satisfaction of PLP.}
%The \emph{arity} of a PLP is the number of distinct variables appearing in it. 
For a PLP $\varphi$ we denote by $\vars(\varphi)$ the set of all variables that 
appear in $\varphi$. As noted above, we stipulate that $\vars(\varphi)=\{x_1,\dots,x_{n(\varphi)}\}$ for some $n(\varphi)\in \Nset$. A vector $\vec{x}$ of dimension $m\geq n(\varphi)$ \emph{satisfies} $\varphi$, we write 
$\vec{x}\models\phi$, if 
 the arithmetic formula obtained by 
substituting each occurrence of a variable $x_i$ in $\varphi$ by $\vec{x}[i]$ is 
valid. We denote $\sat{\phi} = \{\vec{x}\in \Rset^m\mid m\geq n(\varphi) \wedge \vec{x} \models\phi\}$ 
and $\exsat{\phi} = \sat{\phi}
\cap \Rset^{n(\varphi)}$.

%In this paper, we deem any linear assertion equivalently as a polyhedron
%defined by the linear assertion (i.e., the set of points satisfying the
%assertion).
%It will be always clear from the context whether a linear assertion is deemed
%as a logical formula or as a polyhedron.

\vspace{-1em}
\subsection{Syntax of Affine Probabilistic Programs (\APP s)}\label{subsec:syntax}
\vspace{-0.5em}

\noindent{\em The Syntax.}
We consider the standard syntax for affine probabilistic programs,
which encompasses basic programming mechanisms such as assignment statement 
(indicated by `:='), while-loop, if-branch. We also consider basic probabilistic mechanisms 
such as probabilistic branch (indicated by `prob') and random sampling (e.g. 
$x:=\textbf{sample(}\mathrm{Uniform}[-2,1]\textbf{)}$ assigns to $x$ a random 
number 
uniformly sampled from interval $[-2,1]$). We also allow constructs for 
(demonic) non-determinism, in particular 
non-deterministic branching indicated by `\textbf{if }$\star$ \textbf{then...}' construct and non-deterministic assignment.  
Variables (or identifiers) of a probabilistic program are of \emph{real} type, i.e., 
values of the variables are real numbers. 
We allow only affine expressions in test statements and in the right-hand sides of assignments.
We also assume that assume that each \APP{} $\program$ is preceded by an initialization preamble 
in 
which 
each variable appearing in $\program$ is assigned some concrete number.
%the program variables are initialized.
Due to space restrictions, details (such as grammar) are relegated to the Appendix.
For an example see Figure~\ref{fig:invariant-running}.
We refer to this class of affine probabilistic programs as \APP s.

\vspace{-1em}
\subsection{Semantics of Affine Probabilistic Programs}\label{subsec:semantics}
\vspace{-0.5em}

We now formally define the semantics of \APP's.
In order to do this, we first recall some fundamental concepts from probability
theory.

\smallskip\noindent{\em Basics of Probability Theory.}
%%We now recall basic definitions of probability theory.
The crucial notion is the one of a probability space. A probability space is a triple
$(\Omega,\mathcal{F},\probm)$, where $\Omega$ is a non-empty set (so called
\emph{sample space}), $\mathcal{F}$ is a \emph{sigma-algebra} over $\Omega$,
i.e. a collection of subsets of $\Omega$ that contains the empty set
$\emptyset$, and that is closed under complementation and countable unions, and
$\probm$ is a \emph{probability measure} on $\mathcal{F}$, i.e., a function
$\probm\colon \mathcal{F}\rightarrow[0,1]$ such that
\begin{compactitem}
\item $\probm(\emptyset)=0$,
\item for all $A\in \mathcal{F}$ it holds $\probm(\Omega\smallsetminus
A)=1-\probm(A)$, and
\item for all pairwise disjoint countable set sequences $A_1,A_2,\dots \in
\mathcal{F}$ (i.e., $A_i \cap A_j = \emptyset$ for all $i\neq j$)
we have $\sum_{i=1}^{\infty}\probm(A_i)=\probm(\bigcup_{i=1}^{\infty} A_i)$.
\end{compactitem}

\noindent{\em Random variables and filtrations.}
A \emph{random variable} in a probability space $(\Omega,\mathcal{F},\probm)$ is
an $\mathcal{F}$-measurable function $R\colon \Omega \rightarrow \Rset \cup
\{\infty\}$, i.e.,
a function such that for every $a\in \Rset \cup \{ \infty\}$ the set
$\{\omega\in \Omega\mid R(\omega)\leq a\}$ belongs to $\mathcal{F}$.
We denote by $\expv[R]$ the \emph{expected value} of a random variable $X$~(see \cite[Chapter 5]{Billingsley:book}
for a formal definition). 
%%i.e.
%%the Lebesgue integral of $X$ with respect to the probability measure $\probm$.
%%The precise definition of the Lebesgue integral of $R$ is somewhat technical and
%%we omit it here, see,
%%e.g., ~\cite[Chapter 4]{Rosenthal:book}, or~\cite[Chapter 5]{Billingsley:book}
%for a formal definition. 
A \emph{random vector} in $(\Omega,\mathcal{F},\probm)$ is a vector whose every component is a random 
variable in this probability space. A \emph{stochastic process} in a 
probability space $(\Omega,\mathcal{F},\probm)$ is an infinite sequence of 
random vectors in this space.
We will also use random variables of the form $R\colon\Omega \rightarrow S$ for some finite 
set $S$, which is easily translated to the variables above.
A \emph{filtration} of a sigma-algebra $\mathcal{F}$ is a
sequence $\{\mathcal{F}_i \}_{i=0}^{\infty}$ of $\sigma$-algebras 
such that $\mathcal{F}_0 \subseteq \mathcal{F}_1 \subseteq \cdots \subseteq
\mathcal{F}_n \subseteq \cdots \subseteq \mathcal{F}$.

\emph{Distributions.} We assume the standard definition of a probability 
distribution specified by a cumulative distribution 
function~\cite{Billingsley:book}. We denote by $\mathcal{D}$ be a set of 
probability distributions on 
real numbers, both discrete and continuous.

\smallskip\noindent{\em Probabilistic Control Flow Graphs.}
%There are several ways in which one can express the semantics of \APP's with
%(angelic and demonic) 
%non-determinism
The semantics can be defined as the semantics of an uncountable state-space
Markov decision process (MDP) (uncountable due to real-valued variables).
We take an operational approach to define the semantics, and associate to each program 
a certain stochastic process~\cite{SriramCAV,HolgerPOPL,Kozen:prob-semantics}.
To define this process, we first define so called 
\emph{probabilistic control flow graphs}~\cite{CFG16:positivstellensatz-arxiv}.

\smallskip
\begin{definition}
\label{def:stochgame}
A \emph{probabilistic control flow graph (pCFG)} is a tuple
$\pCFG=(\locs,\pvars,\locinit,\vecinit,\transitions,\probdist,\guards)$,
where
\begin{compactitem}
\item $\locs$ is a finite set of \emph{locations} partitioned into three 
pairwise
disjoint subsets  $\locs_N$, $\locs_P$, and $\locs_D$ of non-deterministic, 
probabilistic, and deterministic locations;
\item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
variables} (note that $\pvars \subseteq \vars$) ;
\item $\locinit$ is an initial location and $\vecinit$ is an initial 
\emph{assignment vector};
% POKRACOVAT ODSUD
\item $\transitions$ is a transition relation, whose members are tuples of
the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target variable index}, and $u$ 
is an 
\emph{update element}, which can 
be one of the following mathematical objects: %\begin{itemize}
%\item 
(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
%\item 
(b)~a distribution $d\in \mathcal{D}$; or
%%\item 
(c)~a set $R\subseteq \Rset$.
%\end{itemize}
\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locs_P}$ is a collection of
probability distributions, where each $\prob_{\ell}$ is a discrete probability
distribution on the set of all transitions outgoing from~$\ell$.
\item $\guards$ is a function assigning a propositionally linear predicate
(a \emph{guard}) over $\pvars$ to each transition outgoing from a deterministic 
location.
\end{compactitem}

We assume that each location has at least one outgoing transition.
Also, for every deterministic location $\ell$ we assume the following: if
$\tau_1,\dots,\tau_k$ are all transitions outgoing from $\ell$, then $G(\tau_1)
\vee \dots \vee G(\tau_k) \equiv \mathit{true}$ and $G(\tau_i) \wedge G(\tau_j)
\equiv \mathit{false}$ for each $1\leq i < j \leq k$. Moreover, for each 
distribution $d$ appearing in the 
pCFG we assume the following features are known: expected value $\expv[d]$ of 
$d$ and a 
single-variable PLP $\varphi_d$ such that the \emph{support} of $d$ (i.e. the 
smallest 
closed set of real numbers whose complement has probability zero 
under $d$)\footnote{In particular, a support of a \emph{discrete} probability 
	distribution $d$ is simply the at most countable set of all points on a 
	real 
	line that have positive probability under $d$.} satisfies 
$\support(d)\subseteq\exsat{\varphi_d}$. Finally, we assume that for each 
transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
deterministic. This is just a technical assumption yielding no loss of 
generality, and it somewhat simplifies notation.
%And we assume that each coordinate of $\mathcal{D}$ represents an integrable
%random variable (i.e., the expected value of the absolute value of the random
%variable exists).
\end{definition}

%We can naturally transfer the linear ordering of variables in $\vars$ to 
%variables in $X$. Hence, the set of variables $X$ can be uniquely written as a 
%vector $\vec{x}=(x_1,\dots,x_{|X|})$.
%Since $\pvars$ is a linearly ordered set of variables, there is a natural 
%correspondence between valuations in $\vals(\pvars)$ and 
%$|\pvars|$-dimensional 
%vectors. To formalize this correspondence, to any valuation 
%$\val\in\vals(X)$ we assign a vector 
%$\vec{x}_{\val}=(\val(x_1),\dots,\val(x_{|X|}))$, while to an 
%$|X|$-dimensional vector $\vec{x}$ we assign a valuation $\val_{\vec{x}}\in 
%\vals(\pvars)$ such that $\val_{\vec{x}}(x_i) = \vec{x}[i]$ for all $1\leq i 
%\leq |X|$.

%Given a pCFG $\pCFG$ 
%
%For notational convenience we assume that the sets $\pvars$ and $\rvars$ are
%endowed with some fixed linear ordering, which allows us to write $\pvars =
%\{x_1,x_2,\dots,x_{|\pvars|}\}$ and $\rvars = \{r_1,r_2,\dots,r_{|\rvars|}\}$.
%Every update function $f$ in a stochastic game can then be viewed as a tuple
%$(f_1,\dots,f_{|\pvars|})$, where each $f_i$ is of type
%$\Rset^{|\pvars\cup\rvars|}\rightarrow\Rset$. We denote by
%$\mathbf{x}=(\vec{x}[i])_{i=1}^{|\pvars|}$ and
%$\mathbf{r}=(\vec{r}[i])_{i=1}^{|\rvars|}$ the vectors of concrete valuations 
%of
%program and random variables, respectively.
%In particular, we assume that each component of $\vec{r}$ lies within the range
%of the corresponding random variable. %A valuation $\vec{r}$ is admissible if
%%each its component lies within the range of the corresponding random variable.
%We use the following succinct notation for special update functions: by $\id$ 
%we
%denote a function which does not change the program variables at all, i.e. for
%every $1\leq i \leq |\pvars|$ we have $f_i(\vec{x},\vec{r})= \vec{x}[i]$. For a
%function $g$ over the program and random variables we denote by 
%$\assgn{x_j}{g}$
%the update function $f$ such that $f_j(\vec{x},\vec{r})=g(\vec{x},\vec{r})$ and
%$f_i(\vec{x},\vec{r})=\vec{x}[i]$ for all $i\neq j$.

%We say that an SGS $\mathcal{G}$ is \emph{normalized} if all guards of all
%transitions in $\mathcal{G}$ are in a disjunctive normal form.

%\begin{example}\label{ex:sgs}
%Figure~\ref{fig:Q4:SGS} shows an example of stochastic game structure.
%Deterministic locations are represented by boxes, angelic locations by
%triangles, demonic locations by diamonds,
%and stochastic locations by circles. Transitions are labelled with update
%functions, while guards and probabilities
%of transitions outgoing from deterministic and stochastic locations,
%respectively, are given in rounded rectangles
%on these transitions. For the sake of succinctness we do not picture
%tautological guards and identity update functions. Note that the SGS is
%normalized.
%We will describe in Example~\ref{ex:illustrate} how the stochastic game
%structure shown corresponds to the program described in
%Example~\ref{ex:prog}.
%\end{example}

%\paragraph*{Dynamics of Stochastic Games}

\smallskip\noindent{\em Configurations.}
A \emph{configuration} of a pCFG $\pCFG$ is a tuple $(\ell,\vec{x})$,
where $\ell$ is a location of $\pCFG$ and $\vec{x}$ is an 
$|\pvars|$-dimensional vector.
We say that a transition $\tau$ is \emph{enabled} in a configuration
$(\ell,\vec{x})$ if $\ell$ is the source location of $\tau$ and in addition,
${\vec{x}}\models G(\tau)$ provided that $\ell$ is deterministic. A 
configuration 
$(\loc,\vec{x})$ is non-deterministic/probabilistic/deterministic if $\loc$ is 
non-deterministic/probabilistic/deterministic, respectively.

\smallskip\noindent{\em Executions and reachable configurations.}
A \emph{finite path} (or
\emph{execution fragment}) in $\pCFG$ is a finite sequence of
configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
for each
$0 \leq i < k$ there is a transition  $(\ell_i,j,\up,\ell_{i+1})$ enabled in
$(\ell_i,\vec{x}_i)$ such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
satisfies one of the following:
\begin{compactitem}
\item $\up$ is a function $f\colon\Rset^{|X|}\rightarrow\Rset$ and
$a=f(\vec{x}_{i})$;
\item $\up$ is an integrable\footnote{A distribution on some numerical domain 
is integrable if its expected value exists and is finite. In particular, each 
Dirac distribution is integrable.} distribution $d$ and 
$a\in \support(d)$; or
\item $\up$ is a set and $a\in \up$.
\end{compactitem}
A \emph{run} (or \emph{execution}) in
$\pCFG$ is an infinite sequence of configurations whose every finite
prefix is a finite path.
A configuration $(\loc,\vec{x})$ is {\em reachable} from the initial 
configuration
$(\locinit,\vecinit)$
if there is a finite path starting in $(\locinit,\vecinit)$ that ends in
$(\loc,\vec{x})$.

\smallskip\noindent{\em Schedulers.}
Due to the presence of non-determinism and probabilistic choices, a pCFG
$\pCFG$ may represent a multitude of possible behaviours. The probabilistic
behaviour of $\pCFG$ can be captured by constructing a suitable
probability measure over the set of all its runs. Before this can be
done, non-determinism in $\pCFG$ needs to be resolved. This is done using the 
standard notion of a \emph{scheduler}.

\smallskip
\begin{definition}[Schedulers]
\label{def:schedulers}
A scheduler in an pCFG $\pCFG$ is a tuple $\sigma=(\sigmat,\sigmaa)$,
%(the subscripts stand for 'transition' and 'assignment')
where
\begin{compactitem}
\item $\sigmat$ (here '$t$' stands for 'transition') is a function assigning 
to every finite path that ends in a non-deterministic configuration 
$(\loc,\vec{x})$ a probability distribution on transitions outgoing from 
$\loc$; 
and
\item $\sigmaa$ (here '$a$' stands for 'assignment') is a function which takes 
as an argument a finite path ending in a deterministic configuration in which 
some transition 
$(\loc,j,u,\loc')$ with $u$ being a set is enabled, and for such a path it 
returns a probability distribution on $u$.
\end{compactitem}
%function
%which assigns to every finite path in $\mathcal{G}$ that ends in an angelic
%(resp., demonic) configuration $(\ell,\vec{x})$, respectively,
%a discrete probability distribution on transitions outgoing from~$\ell$.
%a transition outgoing from~$\ell$.
\end{definition}

%A scheduler $\sigma$ is \emph{memoryless} if $\sigmat(w)=\sigmat(w')$ whenever 
%the finite paths 
%$w$ and $w'$ end in the same configuration and if $\sigmaa(w)=\sigmaa(w')$ 
%whenever $w$ and $w'$ end with the same transition. A scheduler $\sigma$ is 
%\emph{deterministic} if for all $w,w'$ the distributions $\sigmaa(w)$ and 
%$\sigmat(w')$ are \emph{Dirac}, i.e. they assign probability 1 to a single 
%transition (resp. to a single number) and probability 0 to all the others.

\smallskip\noindent{\em Stochastic process.}
A pCFG $\pCFG$ together with a scheduler $\sigma$ can be seen as a stochastic 
process which produces a random run 
$(\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)(\loc_2,\vec{x}_2)\cdots$. The evolution 
of 
this process can be informally described as follows: we start in the initial 
configuration, i.e. $(\loc_{0},\vec{x}_0)=(\locinit,\vecinit)$. 
Now assume that $i$ 
steps have elapsed, i.e. a finite path 
$(\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i)$ has already 
been produced. Then
\begin{compactitem}
\item
A transition $\tau=(\loc,j,u,\loc')$ enabled in 
$(\lr{}{i},\vr{}{i})$ is chosen as follows:

\begin{compactitem}
\item
If $\ell_i$ is non-deterministic then $\tau$ is chosen randomly
according to the distribution specified
by scheduler
$\sigma$, i.e. according to the distribution 
$\sigmat((\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i))$.
%That is, if $\ell_i$ is angelic (resp., demonic) and $c_0c_1\cdots c_i$ is the
%sequence of configurations observed so far,
%then $\tau$ equals %%the distribution
%$\sigma(c_0c_1\cdots c_i)$ (resp., $\pi(c_0c_1\cdots c_i)$).

\item
If $\ell_i$ is probabilistic, then $\tau$ is chosen randomly according to the
distribution $\probdist_{\ell_i}$.
\item
If $\ell_i$ is deterministic, then by the definition of a pCFG there is exactly
one enabled transition outgoing from $\ell_i$, and this transition is chosen as
$\tau$.
\end{compactitem}
\item
Once $\tau$ is chosen as above, we put $\loc_{i+1}=\loc'$. Next, we put 
$\vec{x}_{i+1}=\vec{x}_{i}(j\leftarrow a)$, where $a$ chosen as follows:
\begin{compactitem}
\item If $u$ is a function $u\colon
\Rset^{|\pvars|}\rightarrow \Rset$, then $a=f(\vec{x}_{i})$.
\item If $\up$ is a distribution $d$, then $a$ 
is sampled from $d$.
\item If $\up$ is a set, then $a$ is sampled from a 
distribution 
$\sigmaa((\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i))$.
\end{compactitem}
\end{compactitem}

The above intuitive explanation can be formalized by showing that
each pCFG $\pCFG$ together with a scheduler $\sigma$ uniquely determines a 
certain 
probabilistic 
space $(\Omega,\natfilt,\probm^{\sigma})$ in which $\Omega$ is a set of all 
runs in $\pCFG$, and a stochastic process 
$\pCFG^{\sigma}=\{\cfg{\sigma}{i}\}_{i=0}^{\infty}$ in this space 
such 
that for each $\run\in \Omega$ we have that $\cfg{\sigma}{i}(\run)$ is the 
$i$-th configuration on run $\run$ (i.e., $\cfg{\sigma}{i}$ is a random vector 
$(\lr{\sigma}{i},\vr{\sigma}{i})$ with  $\lr{\sigma}{i}$ taking 
values in $\locs$ and $\vr{\sigma}{i}$ being a random vector of dimension 
$|\pvars|$ consisting of real-valued random variables). The sigma-algebra 
$\natfilt$ is the smallest (w.r.t. inclusion) sigma algebra under which all 
the functions $\cfg{\sigma}{i}$, for all $i\geq 0$ and all schedulers $\sigma$, 
are $\natfilt$-measurable (a 
function $f$ returning vectors is $\natfilt$-measurable if for all 
real-valued vectors 
$\vec{y}$ of appropriate dimension the set $\{\omega\in \Omega\mid 
f(\omega)\leq \vec{y}\}$ belongs to $\natfilt$). The 
probability 
measure 
$\probm^{\sigma}$ is such that for each $i$, the distribution of 
$\cfg{\sigma}{i}$ reflects the aforementioned way in which runs are randomly 
generated. The formal construction 
of $\natfilt$ and $\probm^{\sigma}$ is standard~\cite{Billingsley:book} and 
somewhat 
technical, hence we omit it. 
We denote by $\E^\sigma$ the expectation operator in probability space 
$(\Omega,\natfilt,\probm^{\sigma})$.
The translation from probabilistic programs to the corresponding pCFG is 
standard~\cite{CFNH16:prob-termination}, and the details are presented 
in~\cite{CNZ16:prob-invariants-arxiv}. We point out that the 
construction produces pCFGs with a property that only transitions outgoing from 
a deterministic state can update program variables. All other transitions are 
assumed to be of the form $(\ell,1,\id_1,\ell')$ for some locations 
$\ell,\ell'$, where $\id_1(\vec{x})=\vec{x}[1]$ for all $\vec{x}$. We use this 
to simplify notation.
An illustration of a pCFG is given in Figure~\ref{fig:invariant-running}.

%and instead describe it informally by describing 
%the evolution of the stochastic process 
%$\pCFG^{\sigma}=(\cfg{\sigma}{i})_{i=0}^{\infty}$:
%%The evolution of 
%%$\pCFG^{\sigma}$
%%can be intuitively described as follows: 
%We have that $\lr{\sigma}{0}$ is the initial 
%location $\loc_0$, and $\vr{\sigma}{0}=\vec{x}_0$, i.e. 
%$\cfg{\sigma}{0}=(\loc_0,\vec{x}_0)$ with probability 1. Now assume that $i$ 
%%steps have elapsed, i.e. the vectors 
%%%$\cfg{\sigma}{0},\cfg{\sigma}{1},\dots,\cfg{\sigma}{i}$ are known. Then
%%\PN{continue from here}
%%the behaviour of $\mathcal{G}$ 
%%under angelic scheduler $\sigma$ and demonic
%%scheduler $\pi$ can then be intuitively described as follows: 
%%%The process starts in
%%the initial configuration $(\ell_0,\vec{x}_0)$. In every step $i$, assuming 
%%the
%%current configuration to be $(\ell_i,\vec{x}_i)$ the following happens:
%%%\PN{Continue from here}
%

%In this way, the players and random choices eventually produce a random run in
%$\mathcal{G}$. The above intuitive explanation can be formalized by showing 
%that
%the schedulers $\sigma$ and $\pi$ induce a unique probability measure
%$\probm^{\sigma,\pi}$ over a suitable $\sigma$-algebra having runs in
%$\mathcal{G}$ as a sample space. If $\mathcal{G}$ does not have any
%angelic/demonic locations, there is only one angelic/demonic scheduler (an 
%empty
%function) that we typically omit from the notation, i.e. if there are no 
%angelic
%locations we write only $\probm^{\pi}$ etc.
%%%The formal construction of this probability space is performed
%%in~\cite{appendix}.
%%%{\bf KRISH: DO WE WANT THE FORMAL CONSTRUCTION. IT IS FINE TO OMIT.}

\vspace{-1em}
\subsection{Almost-Sure and Probabilistic Termination}
\vspace{-0.5em}
We consider computational problems related to the basic liveness properties of 
\APP{s}, namely \emph{termination} and its generalization, \emph{reachability}.

\smallskip\noindent{\em Termination, reachability, and termination time.}
In the following, consider an \APP{} $P$ and its associated pCFG $\pCFG_P$. 
We say that a run $\run$ of $\pCFG_P$ \emph{reaches} a set of configurations 
$C$ if it  contains a configuration from $C$. A run \emph{terminates} if it reaches a 
configuration whose first component (i.e. a location of $\pCFG_P$) is the 
location
$\loc_{P}^{\lout}$ corresponding to the value of the program counter after 
executing $P$. 
To each set of configurations $C$ we can assign a random variable $\treach{C}$ 
such that for each run $\run$ the value $\treach{C}(\run)$ represents the first 
point in time when 
the current configuration on $\run$ is in $C$. If a run $\run$ does \emph{not} 
reach a set $C$, then 
$\treach{C}(\run)=\infty$. We call $\treach{C}$ the 
\emph{reachability time} of $C$. In particular, if $C$ is the set 
of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
terminal location of $\pCFG_P$), then $\treach{C}$ is 
called a \emph{termination time}, as it returns the number of steps after which 
$\run$ terminates. Since termination time is an important concept on its own, 
we use a special notation $\ttime$ for it.
%\begin{example}
%Consider the program $P$ such that \hspace{-0.25cm}\newsavebox{\figa}
%\lstset{language=affprob}
%\lstset{tabsize=1}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$P=$x:=2;y:=1
%\end{lstlisting}
%\end{lrbox}
%\usebox{\figa}. The associated pCFG has three locations: initial location 
%$\loc_{P}^{\lin}$, terminal location $\loc_{P}^{\lout}$, and one 
%``intermediate'' location, say $\loc$. In the pCFG consider a run 
%$\run=(\loc_{P}^{\lin},(0,0))(\loc,(2,0))(\loc_{P}^{\lout},(2,1))(\loc_{P}^{\lout},(2,1))\cdots$.
% Then $\ttime(\run)=2$. If we denote by $C$ the set of configurations 
% $\{(\loc',(x,y))\mid \loc' \in\{\loc_{P}^{\lin},\loc_{P}^{\lout},\loc\} \text{ 
% and } x\geq 2\}$, then $\treach{C}(\run)=1$.
%\end{example}
Since a probabilistic program may exhibit more than one run, we are interested 
in probabilities of runs that terminate or reach some set of configurations. 
This gives rise to the following fundamental computational problems regarding 
termination:

\begin{compactenum}
\item \emph{Almost-sure termination:} A probabilistic program $P$ is 
almost-surely (a.s.) 
terminating if under each scheduler $\sigma$ it holds that 
$\probm^{\sigma}(\{\run\mid \run \text{ terminates}\}) = 1$, or equivalently, 
if for each $\sigma$ it holds $\probm^{\sigma}(\ttime<\infty)=1$. In 
almost-sure termination question for $P$ we aim to prove that $P$ 
is 
almost-surely terminating.
\item \emph{Probabilistic termination:} In probabilistic termination question 
for $P$ we aim to compute a \emph{lower bound on the probability of 
termination}, i.e. a bound $b\in[0,1]$ such that for each scheduler 
$\sigma$ it holds $\probm^{\sigma}(\{\run\mid \run \text{ terminates}\}) \geq b$ 
(or equivalently $\probm^{\sigma}(\ttime<\infty)\geq b$).
\end{compactenum}

We also define corresponding questions for the more general reachability 
concept.

\begin{compactenum}
\item \emph{Almost-sure reachability:} For a set $C$ of configurations of a
probabilistic program $P$, prove (if possible) that under each scheduler 
$\sigma$ it holds that
$\probm^{\sigma}(\treach{C}<\infty)=1$.
\item \emph{Probabilistic reachability:} For a set $C$ of configurations of 
a probabilistic program $P$, compute a bound $b\in[0,1]$ such that for each scheduler
$\sigma$ it holds $\probm^{\sigma}(\treach{C}<\infty)\geq b$.
\end{compactenum}

Since termination is a special case of reachability, each solution to the 
almost-sure or probabilistic reachability questions provides solution for the 
corresponding termination questions.
%\begin{example}\label{ex:illustrate}
%We now illustrate step by step how the SGS of Example~\ref{ex:sgs} corresponds
%to
%the program of Example~\ref{ex:prog}.
%We first consider the statements $Q_1$ and $Q_2$ (Figure~\ref{fig:Q1Q2}), and
%show the
%corresponding SGSs in Figure~\ref{fig:Q1Q2:SGS}.
%Then consider the statement block $Q_3$ which is a probabilistic choice between
%$Q_1$ and $Q_2$  (Figure~\ref{fig:Q3}).
%The corresponding SGS  (Figure~\ref{fig:Q3:SGS}) is obtained from the previous
%two
%SGSs as follows: we consider a probabilistic start location where there is a
%probabilistic
%branch to the start locations of the SGSs of $Q_1$ and $Q_2$, and the SGS ends
%in a location with only self-loop.
%Finally, we consider the whole program as $Q_4$ (Figure~\ref{fig:Q4}),
%and the corresponding SGS in  (Figure~\ref{fig:Q4:SGS}).
%The SGS is obtained from SGS for $Q_3$, with the self-loop replaced by a
%transition
%back to the probabilistic location (with guard $x\geq 0$), and an edge to the
%final location
%(with guard $x<0$).
%The start location of the whole program is a new location, with transition
%labeled
%$x:=0$, to the start of the while loop location.
%We label the locations in Figure~\ref{fig:Q4:SGS} to refer to them later.
%\lstset{language=affprob}
%\lstset{tabsize=3}
%\newsavebox{\figa}
%\begin{lrbox}{\figa}
%\begin{lstlisting}[mathescape]
%$Q_1$:	if angel then $x:=x+1$ else $x:=x-1$ fi	
%$Q_2$:	if demon then $x:=x+1$ else $x:=x-1$ fi
%\end{lstlisting}
%\end{lrbox}
%\begin{figure}[h]
%\centering
%\usebox{\figa}
%\caption{Programs $Q_1$ and $Q_2$}
%\label{fig:Q1Q2}
%\end{figure}
%
%\begin{figure}[h]
%\centering
%\begin{tikzpicture}[x=1.8cm]
%\node[det] (fin1) at (0,-1.5)  {};
%\node[ang] (angel) at (0,0) {};
%\node[det] (fin2) at (2.3,-1.5)  {};
%\node[dem] (demon) at (2.3,0)  {};
%\node[det] (aif) at (-1,0) {};
%\node[det] (aelse) at (1,0) {};
%\node[det] (dif) at (1.3,-1.5) {};
%\node[det] (delse) at (3.3,-1.5) {};
%
%\draw[tran] (angel.155) -- (aif);
%\draw[tran] (angel.25) -- (aelse);
%\draw[tran] (demon) -- (dif);
%\draw[tran] (demon) -- (delse);
%
%\path (fin1) edge [tran, loop left] (fin1);
%\path (fin2) edge [tran, loop below] (fin2);
%\draw[tran] (dif) -- node[auto,swap] {x:=x+1} (fin2);
%\draw[tran] (delse) -- node[auto] {x:=x-1} (fin2);
%\draw[tran] (aif) -- node[auto, swap] {x:=x+1} (fin1);
%%\draw[tran] (demon) to node[auto] {x:=x-1} (fin2);
%\draw[tran] (aelse) to node[auto] {x:=x-1} (fin1);
%\end{tikzpicture}
%\caption{SGSs for $Q_1$ (left) and $Q_2$ (right)}
%\label{fig:Q1Q2:SGS}
%\end{figure}
%
%\newsavebox{\figb}
%\begin{lrbox}{\figb}
%\begin{lstlisting}[mathescape]
%$Q_3$:	if prob(0.6) then $Q_1$ else $Q_2$ fi
%\end{lstlisting}
%\end{lrbox}
%\begin{figure}[h]
%\centering
%\usebox{\figb}
%\caption{Program $Q_3$}
%\label{fig:Q3}
%\end{figure}
%\begin{figure}[h]
%\centering
%\begin{tikzpicture}[x = 1.8cm]
%\node[det] (while) at (1.5,-1.5)  {};
%%%\node[det] (fin) at (0,-3) {};
%\node[ran] (prob) at (3.7,-1.5) {};
%\node[ang] (angel) at (3,0) {};
%\node[dem] at (3,-3) (demon) {};
%\node[det] (aif) at (2,-0.075) {};
%\node[det] (aelse) at (3,-0.8) {};
%\node[det] (dif) at (2,-3) {};
%\node[det] (delse) at (3,-2.2) {};
%
%\draw[tran] (angel) -- (aif);
%\draw[tran] (angel) -- (aelse);
%\draw[tran] (demon) -- (dif);
%\draw[tran] (demon) -- (delse);
%
%\path (while) edge [tran, loop left] (while);
%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3,
%inner sep = 1pt] {$\frac{6}{10}$} (prob|-angel) -- (angel);
%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3,
%inner sep = 1pt] {$\frac{4}{10}$} (prob|-demon) --(demon);
%\draw[tran] (dif) --  (demon-|while) -- node[auto, pos=0.2] {x:=x+1} (while);
%\draw[tran] (aif) --  (angel-|while) -- node[auto, swap, pos=0.2] {x:=x+1}
%(while);
%\draw[tran] (delse) -- node[auto] {x:=x-1} (delse-|while.305) --  (while.-55);
%\draw[tran] (aelse) -- node[auto, swap] {x:=x-1} (aelse-|while.55) -- 
%(while.55);
%\end{tikzpicture}
%\caption{SGS of $Q_3$}
%\label{fig:Q3:SGS}
%\end{figure}
%\newsavebox{\figc}
%\begin{lrbox}{\figc}
%\begin{lstlisting}[mathescape]
%$Q_4$:	$x:=0$; while ($x\ge 0$) do $Q_3$ od
%\end{lstlisting}
%\end{lrbox}
%\begin{figure}[h]
%\centering
%\usebox{\figc}
%\caption{Program $Q_4$}
%\label{fig:Q4}
%\end{figure}
%\begin{figure}[h]
%\centering
%\begin{tikzpicture}[x = 1.8cm]
%%\node[det] (det1) at (0,0) {$\loc_0$};
%%\node[det] (while) at (1.5,-1.5)  {$\loc_1$};
%%\node[det] (fin) at (0,-3) {$\loc_5$};
%%\node[ran] (prob) at (3,-1.5) {$\loc_2$};
%%\node[ang] (angel) at (3,0) {$\loc_3$};
%%\node[dem] (demon) at (3,-3)  {$\loc_4$};
%%
%%\draw[tran] (det1) to node[auto, swap] {x:=0} (while);
%%\draw[tran] (while) to node[font=\scriptsize,draw, fill=white,
%%rectangle,pos=0.3] {$x<0$} (fin);
%%\draw[tran, loop, looseness = 5, in =-65, out = -115] (fin) to (fin);
%%\draw[tran] (while) to node[font=\scriptsize,draw, fill=white,
%%rectangle,pos=0.3] {$x\geq 0$} (prob);
%%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white,
%%rectangle,pos=0.3, inner sep = 1pt] {$\frac{6}{10}$} (angel);
%%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white,
%%rectangle,pos=0.3, inner sep = 1pt] {$\frac{4}{10}$} (demon);
%%\draw[tran] (demon) -- node[auto] {x:=x+1} (demon-|while) -- (while);
%%\draw[tran] (angel) -- node[auto, swap] {x:=x+1} (angel-|while) -- (while);
%%\draw[tran] (demon) to node[auto] {x:=x-1} (while);
%%\draw[tran] (angel) to node[auto, swap] {x:=x-1} (while);
%
%\node[det] (while) at (1.5,-1.5)  {$\loc_1$};
%%%\node[det] (fin) at (0,-3) {};
%\node[ran] (prob) at (3.7,-1.5) {$\loc_2$};
%\node[ang] (angel) at (3,0) {$\loc_3$};
%\node[dem] at (3,-3) (demon) {$\loc_4$};
%\node[det] (aif) at (2,-0.075) {$\loc_5$};
%\node[det] (aelse) at (3,-0.8) {$\loc_6$};
%\node[det] (dif) at (2,-3) {$\loc_7$};
%\node[det] (delse) at (3,-2.2) {$\loc_8$};
%\node[det] (det1) at (0,0) {$\loc_0$};
%%\node[det] (while) at (1.5,-1.5)  {};
%\node[det] (fin) at (0,-3) {$\loc_9$};
%%\node[ran] (prob) at (3,-1.5) {};
%%\node[ang] (angel) at (3,0) {};
%%\node[dem] at (3,-3) (demon) {};
%%
%\draw[tran] (det1) to node[auto, swap] {x:=0} (while);
%\draw[tran] (while) to node[font=\scriptsize,draw, fill=white,
%rectangle,pos=0.4] {$x<0$} (fin);
%\draw[tran, loop, looseness = 5, in =-65, out = -115] (fin) to (fin);
%\draw[tran] (while) to node[font=\scriptsize,draw, fill=white,
%rectangle,pos=0.3] {$x\geq 0$} (prob);
%%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white,
%%rectangle,pos=0.3, inner sep = 1pt] {$\frac{6}{10}$} (angel);
%%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white,
%%rectangle,pos=0.3, inner sep = 1pt] {$\frac{4}{10}$} (demon);
%%\draw[tran] (demon) -- node[auto] {x:=x+1} (demon-|while) -- (while);
%%\draw[tran] (angel) -- node[auto, swap] {x:=x+1} (angel-|while) -- (while);
%%\draw[tran] (demon) to node[auto] {x:=x-1} (while);
%%\draw[tran] (angel) to node[auto, swap] {x:=x-1} (while);
%
%
%\draw[tran] (angel) -- (aif);
%\draw[tran] (angel) -- (aelse);
%\draw[tran] (demon) -- (dif);
%\draw[tran] (demon) -- (delse);
%
%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3,
%inner sep = 1pt] {$\frac{6}{10}$} (prob|-angel) -- (angel);
%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3,
%inner sep = 1pt] {$\frac{4}{10}$} (prob|-demon) --(demon);
%\draw[tran] (dif) --  (demon-|while) -- node[auto, pos=0.2] {x:=x+1} (while);
%\draw[tran] (aif) --  (angel-|while) -- node[auto, swap, pos=0.2] {x:=x+1}
%(while);
%\draw[tran] (delse) -- node[auto] {x:=x-1} (delse-|while.305) --  (while.-55);
%\draw[tran] (aelse) -- node[auto, swap] {x:=x-1} (aelse-|while.55) -- 
%(while.55);
%\end{tikzpicture}
%\caption{SGS of $Q_4$}
%\label{fig:Q4:SGS}
%\end{figure}
%\end{example}

%\subsection{Qualitative and Quantitative Termination Questions}
%We consider the most basic notion of liveness, namely {\em termination},
%for probabilistic programs, and present the relevant {\em qualitative} and
%{\em quantitative} questions.
%
%
%
%\paragraph*{Qualitative question.}
%We consider the two basic qualitative questions, namely, almost-sure 
%termination
%(i.e., termination with probability~$1$) and finite expected termination time.
%We formally define them below.
%
%
%Given a program $P$, let $\mathcal{G}_P$ be the associated SGS.
%A run $\rho$ is \emph{terminating} if it reaches a configuration in which the
%location is $\ell^{\lout}_{P}$.
%Consider the random variable $T$ which to every run $\rho$ in $\mathcal{G}_P$
%assigns the first point in time in which a configuration with the location
%$\ell^{\lout}_{P}$ is encountered,
%and if the run never reaches such a configuration, then the value assigned is
%$\infty$.
%
%
%\begin{definition}[Qualitative termination questions]
%Given a program $P$ and its associated normalized SGS $\mathcal{G}_P$, we
%consider the following
%two questions:
%\begin{compactenum}
%\item {\em Almost-Sure Termination.} The program is {\em almost-surely (a.s.)}
%terminating if
%there exists an angelic scheduler (called {\em a.s terminating}) $\sigma$ such
%that for all demonic
%schedulers $\pi$ we have
%$\probm^{\sigma,\pi}\left( \{\rho\mid\rho \text{ is terminating}\,\}\right) =
%1;$ or equivalently,
%$\probm^{\sigma,\pi}(T<\infty)=1$.
%
%\item {\em Finite Termination.}
%The program $P$ is \emph{finitely terminating} (aka positively almost-sure
%terminating) if there
%exists an angelic scheduler $\sigma$ (called {\em finitely terminating})
%such that for all demonic schedulers $\pi$ it holds that
%$\expv^{\sigma,\pi}[T]<\infty.$
%%%A scheduler $\sigma$ with this property is said to be \emph{positively
%%terminating}.
%\end{compactenum}
%
%\end{definition}
%
%Note that for all angelic schedulers $\sigma$ and demonic schedulers $\pi$ we
%have $\expv^{\sigma,\pi}[T]<\infty$ implies $\probm^{\sigma,\pi}(T<\infty)=1$,
%however, the converse does not hold in general.
%In other words, finitely terminating implies a.s. terminating, but a.s.
%termination does not imply finitely termination.
%
%\begin{definition}[Quantitative termination questions]
%\label{def:quanttq}
%Given a program $P$ and its associated normalized SGS $\mathcal{G}_P$, we
%consider the following
%notions:
%\begin{compactenum}
%\item {\em Expected Termination Time.} The {\em expected termination} time of
%$P$ is
%$\Eval(P) = \inf_{\sigma}\sup_{\pi}\expv^{\sigma,\pi}[T]. $
%
%\item {\em Concentration bound.} A bound $B$ is a {\em concentration bound} if
%there exists two positive constants $c_1$ and $c_2$ such that for all $x\geq 
%B$,
%we have
%$\Thr(P,x) \leq c_1 \cdot \exp(-c_2\cdot x)$, where $\Thr(P,x)=\inf_{\sigma}
%\sup_{\pi} \probm^{\sigma,\pi}(T > x)$ (i.e., the probability that the
%termination time exceeds $x \geq B$ decreases exponentially in $x$).
%\end{compactenum}
%\end{definition}
%
%Note that we assume that an SGS $\mathcal{G}_P$ for a program $P$ is already
%given in a normalized form, as our algorithms assume that all guards in
%$\mathcal{G}_P$ are in DNF.
%%To separate the complexity of normalizing an SGS from the complexity of 
%%solving
%%our qualitative and quantitative question we assume
%In general, converting an SGS into a normalized SGS incurs an exponential
%blow-up as this is the worst-case blowup when converting a formula into DNF.
%However, we note that for programs $\mathcal{P}$ that contain only \emph{simple
%guards}, i.e. guards that are either conjunctions or disjunctions of linear
%constraints, a normalized game $\mathcal{G}_P$ can be easily constructed in
%polynomial time using de Morgan laws. In particular, we stress that all our
%hardness results hold already for programs with simple guards, so they do not
%rely on the requirement that $\mathcal{G}_P$ must be normalized.

\lstset{language=affprob}
\lstset{tabsize=3}
\newsavebox{\invrun}
\begin{lrbox}{\invrun}
\begin{lstlisting}[mathescape]
$x:=10$
while $x\geq 1$ do
	if prob(0.75) then $x:=x-1$	else $x:=x+1$ 
	fi
od
\end{lstlisting}
\end{lrbox}
\begin{figure}[t]
\centering
\usebox{\invrun}

\begin{tikzpicture}[x = 1.8cm]
%\node[det] (det1) at (0,0) {$\loc_0$};
%\node[det] (while) at (1.5,-1.5)  {$\loc_1$};
%\node[det] (fin) at (0,-3) {$\loc_5$};
%\node[ran] (prob) at (3,-1.5) {$\loc_2$};
%\node[ang] (angel) at (3,0) {$\loc_3$};
%\node[dem] (demon) at (3,-3)  {$\loc_4$};
%
%\draw[tran] (det1) to node[auto, swap] {x:=0} (while);
%\draw[tran] (while) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3] {$x<0$} (fin);
%\draw[tran, loop, looseness = 5, in =-65, out = -115] (fin) to (fin);
%\draw[tran] (while) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3] {$x\geq 0$} (prob);
%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3, inner sep = 1pt] {$\frac{6}{10}$} (angel);
%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3, inner sep = 1pt] {$\frac{4}{10}$} (demon);
%\draw[tran] (demon) -- node[auto] {x:=x+1} (demon-|while) -- (while);
%\draw[tran] (angel) -- node[auto, swap] {x:=x+1} (angel-|while) -- (while);
%\draw[tran] (demon) to node[auto] {x:=x-1} (while);
%\draw[tran] (angel) to node[auto, swap] {x:=x-1} (while);

\node[det] (while) at (1.5,0)  {$\loc_0$};
%%\node[det] (fin) at (0,-3) {};
\node[ran] (prob) at (3,0) {$\loc_1$};
%\node[ang] (angel) at (3,0) {$\loc_3$};
%\node[dem] at (3,-3) (demon) {$\loc_4$};
%\node[det] (aif) at (2,-0.075) {$\loc_5$};
%\node[det] (aelse) at (3,-0.8) {$\loc_6$};
%\node[det] (dif) at (2,-3) {$\loc_7$};
%\node[det] (delse) at (3,-2.2) {$\loc_8$};
%\node[det] (det1) at (0,0) {$\loc_0$};
%\node[det] (while) at (1.5,-1.5)  {};
\node[det] (fin) at (0,0) {$\loc_2$};
%\node[ran] (prob) at (3,-1.5) {};
%\node[ang] (angel) at (3,0) {};
%\node[dem] at (3,-3) (demon) {};
%
%%\draw[tran] (det1) to node[auto, swap] {x:=0} (while);
\draw[tran] (while) to node[font=\scriptsize,draw, fill=white, 
rectangle,pos=0.5] {$x<1$} (fin);
\draw[tran, loop, looseness = 5, in =-65, out = -115] (fin) to (fin);
\draw[tran] (while) to node[font=\scriptsize,draw, fill=white, 
rectangle,pos=0.5] {$x\geq 1$} (prob);
%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3, inner sep = 1pt] {$\frac{6}{10}$} (angel);
%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3, inner sep = 1pt] {$\frac{4}{10}$} (demon);
%\draw[tran] (demon) -- node[auto] {x:=x+1} (demon-|while) -- (while);
%\draw[tran] (angel) -- node[auto, swap] {x:=x+1} (angel-|while) -- (while);
%\draw[tran] (demon) to node[auto] {x:=x-1} (while);
%\draw[tran] (angel) to node[auto, swap] {x:=x-1} (while);

%\draw[tran] (angel) -- (aif);
%\draw[tran] (angel) -- (aelse);
%\draw[tran] (demon) -- (dif);
%\draw[tran] (demon) -- (delse);

\node (dum1) at (0,0.8) {};
\node (dum2) at (0,-0.8) {};

\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
rectangle,pos=0.5, inner sep = 1pt] {$\frac{3}{4}$} (prob|-dum1) -- node[auto] 
{x:=x-1}
(while|-dum1)--(while);
\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
rectangle,pos=0.5, inner sep = 1pt] {$\frac{1}{4}$} (prob|-dum2) 
-- node[auto,swap] {x:=x+1} (while|-dum2)--(while);
%\draw[tran] (dif) --  (demon-|while) -- node[auto, pos=0.2] {x:=x+1} (while);
%\draw[tran] (aif) --  (angel-|while) -- node[auto, swap, pos=0.2] {x:=x+1} 
%(while);
%\draw[tran] (delse) -- node[auto] {x:=x-1} (delse-|while.305) --  (while.-55);
%\draw[tran] (aelse) -- node[auto, swap] {x:=x-1} (aelse-|while.55) --  
%(while.55);
\end{tikzpicture}
\caption{An \APP{} modelling an asymmetric 1-D random walk and the associated 
pCFG. Probabilistic locations are depicted by circles, with probabilities given 
on outgoing 
transitions. Transitions are labelled by their effects. Location $\loc_0$ is 
initial and $\loc_2$ is terminal.}
\label{fig:invariant-running}
\end{figure}

### 2017-05-29 implementation/Parser.cpp
Generating equations started

@@ -658,7 +658,7 @@ void CFG_edge::print(){
 	}
 	if(guard!=NULL){
 		cout<<"Guard is: "<<endl;
-		// guard-> print();
+		guard-> print();
 	}
 	cout<<"Probability to occur is"<<probability<<endl<<endl; 
 }
@@ -675,15 +675,13 @@ void CFG_location::print(){
 	if(invariant!=NULL){
 		cout<<"Invariant: ";
 		cout<<invariant;
-		// invariant->print();
+		invariant->print();
 		cout<<endl;
 	}
 
-	// cout<<"Edges: "<<endl;
 	int i=1;
 	for(vector<CFG_edge>::iterator it = edges.begin();it!=edges.end();++it){
 		cout<<"Edge #"<<i<<endl;
-		// edges[i].print();
 		it->print();
 		i++;
 	}

### 2017-05-29 implementation/Parser.h

@@ -54,11 +54,6 @@ public:
 	void print();
 };
 
-struct equation{
-	node* affexpr;
-	node* literal;
-};
-
 void skip_spaces(int &begin, int &end);
 void vcopy(vector<node*> &sink,vector<node*> &tocopy);
 node* negation(node* tonegate);

### 2017-05-29 implementation/main.cpp

@@ -18,30 +18,89 @@ using namespace std;
 #define part(x,a,b) (x.substr((a),((b)-(a))))
 #define pb push_back
 
-stack<equation> equations;
+struct cond{
+	int src;
+	int dest1;
+	int dest2;
+	bool probability;
+	cond(int src,int dest1,int dest2 = 0,double probability = -1.0){
+		this->src = src;
+		this->dest1 = dest1;
+		this->dest2 = dest2;
+		this->probability = probability;
+	}
+};
+
+struct equation{
+	node* affexpr;
+	cond* condition;
+	equation(){
+		affexpr = NULL;
+		condition = NULL;
+	}
+	equation(node* affexpr,cond* condition){
+		this->affexpr = affexpr;
+		this->condition = condition;
+	}
+};
+
+stack<equation*> equations;
 
 void generate_equations(ofstream& equationsfile){ //Would use the ofstream file to write the equations into it later
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
 		if(it->second->type=="det"){
 			//Invariant and guard imply the value decrease
-			// Guard would have been NULL here
 			if(it->second->edges.size()==1){
+				//First make a condition
+				cond* condition = new cond(it->first,it->second->edges[0].next->label);
+				// Guard would have been NULL here
 				if(it->second->invariant==NULL){
-
+					//This should never be the case as this would pose conditions that c==0 and d>0, which are not good
+					cerr<<"No invariant specified here"<<endl;
 				}
 				else{
-					//As te 
+					// Invariant implies the given condition
+					int nEquations = it->second->invariant->children.size();
+					for(int i=0;i<nEquations;i++){
+						equations.push(new equation(it->second->invariant->children[i],condition));
+					}
 				}
 			}
 			else{
-				// Size is 2
+				//First make the 2 conditions required
+				cond* condition1 = new cond(it->first,it->second->edges[0].next->label);
+				cond* condition2 = new cond(it->first,it->second->edges[1].next->label);
+				// Size is 2, so considering the guards is important and guards cannot be NULL
+				if(it->second->invariant==NULL or it->second->edges[0].guard==NULL or it->second->edges[1].guard==NULL){
+					//This should never be the case as this would pose conditions that c==0 and d>0, which are not good
+					cerr<<"No invariant or no guard specified here"<<endl;
+				}
+				else{
+					//First take and of guard and invariant and then make the corresponding equations
+				}
 			}
 		}
 		else if(it->second->type=="ndet"){
[…]

### 2017-05-29 implementation/main.cpp
Equations made

@@ -46,6 +46,28 @@ struct equation{
 
 stack<equation*> equations;
 
+node* and_node(node* one,node* two){
+	if(one->type!="bexpr" or two->type!="bexpr"){
+		cerr<<"Taking and of wrong type of nodes"<<endl;
+		return NULL;
+	}
+	node* node_and = NULL;
+	node* temp_affexpr = NULL;
+	node_and = new node("bexpr");
+	node_and->constant = "or";
+	int a = one->children.size(), b = two->children.size();
+	for(int i = 0;i<a;i++){
+		for(int j = 0;j<b;j++){
+			temp_affexpr = new node("affexpr");
+			temp_affexpr->constant = "and";
+			temp_affexpr->children = one->children[i]->children;
+			temp_affexpr->children.insert(temp_affexpr->children.end(),two->children[j]->children.begin(),two->children[j]->children.end());
+			node_and->children.pb(temp_affexpr);
+		}
+	}
+	return node_and;
+}
+
 void generate_equations(ofstream& equationsfile){ //Would use the ofstream file to write the equations into it later
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
 		if(it->second->type=="det"){
@@ -68,8 +90,8 @@ void generate_equations(ofstream& equationsfile){ //Would use the ofstream file 
 			}
 			else{
 				//First make the 2 conditions required
-				cond* condition1 = new cond(it->first,it->second->edges[0].next->label);
-				cond* condition2 = new cond(it->first,it->second->edges[1].next->label);
+				cond* condition0 = new cond(it->first,it->second->edges[0].next->label);
+				cond* condition1 = new cond(it->first,it->second->edges[1].next->label);
 				// Size is 2, so considering the guards is important and guards cannot be NULL
 				if(it->second->invariant==NULL or it->second->edges[0].guard==NULL or it->second->edges[1].guard==NULL){
 					//This should never be the case as this would pose conditions that c==0 and d>0, which are not good
@@ -77,20 +99,30 @@ void generate_equations(ofstream& equationsfile){ //Would use the ofstream file 
 				}
 				else{
 					//First take and of guard and invariant and then make the corresponding equations
+					node* n0 = and_node(it->second->invariant,it->second->edges[0].guard);
+					node* n1 = and_node(it->second->invariant,it->second->edges[1].guard);
+					int nEquations = n0->children.size();
+					for(int i=0;i<nEquations;i++){
[…]

### 2017-05-29 implementation/Parser.cpp
Expression parsing corrected

@@ -1,5 +1,6 @@
 //Compile with C++11
 //The variable names should be of the form x_i where i is a number and the indexing starts from 1 i.e. the variables in the program are of the form x_1, x_2 ... x_n.
+//Assuming * is for multiplication rather than .
 #include <iostream>
 #include <algorithm>
 #include <cstdio>
@@ -281,7 +282,7 @@ void node::proc_assgn(int s,int l){
 		}
 	}
 	if(pos==-1){
-		cerr<<"Wrong assignment between ["<<begin<<","<<end<<")"<<endl;
+		cerr<<"Wrong assignment between ["<<begin<<","<<end<<") i.e. "<<part(program,begin,end)<<endl;
 	}
 	else{
 		children.resize(2);
@@ -368,47 +369,59 @@ void node::proc_pvar(){
 	constant = part(program,begin,end);
 }
 
-void node::recursively_form_vector(int begin,int end){ //Note that this shadows the original ones
+void node::analyse_expr(int begin,int end,bool negate){
+	skip_spaces(begin,end);
+	int star = -1;
+	for(int i = begin;i<end;++i){ //Assuming that the program variable names don't contain star anywhere
+		if(program[i]=='*'){
+			star = i;
+		}
+	}
+	if(star==-1){
+		if(part(program,begin,begin+2)=="x_"){
+			if(negate){
+				expression[stoi(part(program,begin+2,end))] += -1.0;
+			}
+			else{
+				expression[stoi(part(program,begin+2,end))] += 1.0;
+			}
+		}
+		else{
+			if(negate){
+				expression[0] -= stod(part(program,begin,end));
+			}
+			else{
+				expression[0] += stod(part(program,begin,end));
+			}
+		}
+	}
+	else{
+		int block=star+1;
+		skip_spaces(begin,star);
+		skip_spaces(block,end);
+		if(part(program,block,block+2)!="x_"){
+			cerr<<"Error in the variable in ["<<block<<","<<end<<")"<<"begin = "<<begin<<endl;
+		}
+		expression[stoi(part(program,block+2,end))] += stod(part(program,begin,star));
+	}
+}
+
+void node::form_vector(int begin,int end,bool negate){ //Note that this shadows the original ones
 	skip_spaces(begin,end);
 	int plusminus = -1;
 	// Implemented as a linear but still recursive function
[…]

### 2017-05-29 implementation/Parser.h

@@ -17,12 +17,13 @@ private:
 	void proc_assgn(int s,int l);
 	void proc_affexpr();
 	void proc_pvar();
+	void analyse_expr(int begin,int end,bool negate);
 	void proc_expr();
 	void proc_constant();
 	void proc_literal(bool negate);
 	void proc_bexpr();
 	void proc_ndbexpr();
-	void recursively_form_vector(int begin,int end);
+	void form_vector(int begin,int end,bool negate);
 //main process
 	void process(int s, int l, bool negate);
 public:
@@ -58,7 +59,6 @@ void skip_spaces(int &begin, int &end);
 void vcopy(vector<node*> &sink,vector<node*> &tocopy);
 node* negation(node* tonegate);
 int find_variables();
-void generate_equations(ofstream& equationsfile);
 
 extern string program;
 extern int nVariables;

### 2017-05-29 implementation/main.cpp

@@ -23,7 +23,7 @@ struct cond{
 	int dest1;
 	int dest2;
 	bool probability;
-	cond(int src,int dest1,int dest2 = 0,double probability = -1.0){
+	cond(int src,int dest1,int dest2 = -1,double probability = -1.0){
 		this->src = src;
 		this->dest1 = dest1;
 		this->dest2 = dest2;
@@ -68,7 +68,7 @@ node* and_node(node* one,node* two){
 	return node_and;
 }
 
-void generate_equations(ofstream& equationsfile){ //Would use the ofstream file to write the equations into it later
+void generate_equations(){ //Would use the ofstream file to write the equations into it later
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
 		if(it->second->type=="det"){
 			//Invariant and guard imply the value decrease
@@ -78,7 +78,8 @@ void generate_equations(ofstream& equationsfile){ //Would use the ofstream file 
 				// Guard would have been NULL here
 				if(it->second->invariant==NULL){
 					//This should never be the case as this would pose conditions that c==0 and d>0, which are not good
-					cerr<<"No invariant specified here"<<endl;
+					// cerr<<"No invariant specified here"<<endl;
+					//This would arise for the end states and skip and other special conditions
 				}
 				else{
 					// Invariant implies the given condition
@@ -137,6 +138,16 @@ void generate_equations(ofstream& equationsfile){ //Would use the ofstream file 
 	}
 }
 
+void print_equations(){
+	while(!equations.empty()){
+		// equationsfile<<"Equation"<<endl;
+		equations.top()->affexpr->print();
+		cout<<"implies";
+		cout<<equations.top()->condition->src<<" "<<equations.top()->condition->dest1<<endl;
+		equations.pop();
+	}
+}
+
 int main(){
 	char input[MAXL];
 	// cout<<fixed<<setprecision(10);
@@ -160,17 +171,18 @@ int main(){
 	cout<<"Input Code:"<<endl;
 	cout<<program<<endl;
 	cout<<"Parse Tree:"<<endl;
-	root->print();
-	cout<<"CFG:"<<endl;
-	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
-		cout<<"------------------------"<<endl;
-		cout<<"Node "<<it->first<<endl;
-		it->second->print();
[…]

### 2017-05-29 paper/invariants-martingales.tex
edited prelims

%\vspace{-1em}
\section{Invariants and Ranking Supermartingales}\label{sec:invm}
%\vspace{-0.5em}

In this section we recall known methods and constructs for solving the 
qualitative termination and reachability questions for \APP{s}, namely  
linear invariants and ranking supermartingales. 
We also demonstrate that these methods are not sufficient to address the 
quantitative variants of these questions (i.e., probabilistic termination).
In order to discuss the necessary concepts, we recall the basics of 
martingales, which is relevant for both this and subsequent sections. 
%Since the definitions of this section are from the literature we relegate
%relevant illustrations through example to the Appendix.

%While the core results of this section were already proven in the previous 
%literature, the version we sometimes present slight generalizations of these 
%result, which make them applicable to a larger class of programs. We discuss 
%the nature of these generalizations where appropriate.

%\vspace{-1em}
\subsection{Pure Invariants}
%\vspace{-0.5em}

Invariants are a vital element of many program analysis techniques. 
Intuitively, invariants are maps assigning to each 
program location $\loc$ of some pCFG a predicate which is guaranteed to hold 
whenever $\loc$ is 
entered. To avoid confusion with stochastic invariants, 
that we introduce later, we call these standard invariants \emph{pure invariants}.

\smallskip
\begin{definition}[Linear Predicate Map (LPM) and Pure Invariant] We define the 
following:
\begin{compactenum}
\item
A \emph{linear predicate map (LPM)} for an \APP{} $P$ is a function $I$ 
assigning to each location $\loc$ of the pCFG $\pCFG_P$ a propositionally 
linear predicate $I(\loc)$ over the set of program variables of $P$.
\item
A \emph{pure linear invariant} (or just a pure invariant) for an \APP{} $P$ is 
a linear predicate map $\inv$ for $P$ with
the following property: for each location $\loc$ of $\pCFG_P$ and each finite 
path $(\loc_0,\vec{x}_0),\cdots,(\loc_n,\vec{x}_n)$ such that 
$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and $\loc_n = \loc$ it holds
$\vec{x}_n\models \inv(\loc)$.
\end{compactenum}
%%A 
%%probabilistic LPM for $P$ is a tuple $(\pinv,\pinfunc)$ such that $\pinv$ is 
%%a 
%%pure LPM for $P$ and $\pinfunc$ is a function assigning to each location 
%%$\loc$ 
%%a probability $\pi(\loc)\in[0,1]$.
\end{definition}

%A special class of LPMs are program invariants, i.e. maps assigning to each 
%program location $\loc$ an PLP which is guaranteed to hold whenever $\loc$ is 
%entered.
%Invariants are a standard construct used in automated program analysis. 
%To avoid confusion with probabilistic invariants, we call the 
%standard invariants \emph{pure}.
%
%\begin{definition}[Pure Linear Invariant]
%A \emph{pure linear invariant} (or just a pure invariant) for an \APP{} $P$ is 
%%a function map $\inv$  assigning to each location $\loc$ of the pCFG 
%%$\pCFG_P$ 
%%a propositionally 
%%linear predicate $I(\loc)$ over the set of program variables of $P$ 
%a linear predicate map $\inv$ for $P$ with
%the following property: for each location $\loc$ of $\pCFG_P$ and each finite 
%path $(\loc_0,\vec{x}_0),\cdots,(\loc_n,\vec{x}_n)$ such that 
%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and $\loc_n = \loc$ it holds
%$\val_{\vec{x}_i}\models \inv(\loc)$.
%\end{definition}

%\vspace{-1em}
\subsection{Supermartingales}
%\vspace{-0.5em}

%Submartingales and their counterparts, supermartingales, are a standard tool 
%of 
(Super)martingales, are a standard tool of 
probability theory apt for analyzing probabilistic objects arising in computer 
science, from automata-based models~\cite{BKKNK:pMC-zero-reachability} to 
general probabilistic 
programs~\cite{SriramCAV,HolgerPOPL,CFNH16:prob-termination,CFG16:positivstellensatz-arxiv,BEFH16:doob-decomposition}.

%\PN{Maybe move this sentence to intro.} 
Let us first recall basic
definitions and results related to supermartingales, which we need in our 
analysis.

\smallskip\noindent{\bf Conditional Expectation.} 
%The notion of conditional expectation 
%tightly connected to martingales. We present only a brief informal exposition 
%here, for formal treatment see, e.g.~\cite[Chapter 9]{Williams:book}.
Let $(\Omega,\mathcal{F},\probm)$ be a probability space, 
$X\colon\Omega\rightarrow 
\Rset$ an $\mathcal{F}$-measurable function, and $\mathcal{F}'\subseteq 
\mathcal{F}$ sub-sigma-algebra of $\mathcal{F}$. The \emph{conditional 
expectation} of $X$ given $\mathcal{F}'$ is an $\mathcal{F}'$-measurable random 
variable denoted by $\E[X| \mathcal{F}']$ which satisfies, for each set $A\in 
\mathcal{F}'$, the following: 
\begin{equation}
\label{eq:cond-exp}
\E[X\cdot 1_A] = \E[\E[X|\mathcal{F}]\cdot 1_A],
\end{equation}
where $1_A \colon \Omega\rightarrow \{0,1\}$ is an \emph{indicator function} of 
$A$, i.e. function returning $1$ for 
each $\omega\in A$ and $0$ for each $\omega\in \Omega\setminus A$. Note that 
the left hand-side of~\eqref{eq:cond-exp} intuitively represents the expected 
value 
of $X(\omega)$ with domain restricted to $A$.

Recall that in context of probabilistic programs we work with probability 
spaces of the form $(\Omega,\natfilt,\probm^\sigma)$, where $\Omega$ is a 
set of runs in some $\pCFG$ and $\mathcal{F}$ is (the smallest) sigma-algebra 
such that all the functions $\cfg{\sigma}{i}$, where $i\in \Nset_0$ and 
$\sigma$ is a scheduler, are $\natfilt$-measurable. In such a setting we can 
also consider sub-sigma-algebras $\natfilt_i$, $i\in \Nset_0$, of 
$\natfilt$, where $\natfilt_i$ is the smallest sub-sigma-algebra of 
$\natfilt$ such that all the functions $\cfg{\sigma}{j}$, $0\leq j \leq 
i$, are $\natfilt_i$-measurable. Intuitively, each set $A$ belonging to such 
an $\natfilt_i$ consists of runs whose first $i$ steps satisfy some 
property, and the probability space $(\Omega,\natfilt_i,\probm^\sigma)$ 
allows us to reason about probabilities of certain events happening in the 
first 
$i$ steps of program execution. 
%(note that since $\natfilt_i \subseteq 
%\natfilt$, we can view $\probm^\sigma$ also as a function of type 
%$\natfilt_i\rightarrow [0,1]$). 
Then, for each $A\in \natfilt_i$, the 
value $\E[\E[X|\natfilt_i]\cdot 1_A]$ represents the expected value of 
$X(\run)$ for the randomly generated run $\run$ provided that we restrict to 
runs whose
prefix of length $i$ satisfies the property given by $A$. 
Note that the sequence $\natfilt_0,\natfilt_1,\natfilt_2,\dots$ forms a 
filtration of $\natfilt$, which we call a \emph{canonical filtration}.

%%\paragraph*{Basics of Supermartingales}

\smallskip
\begin{definition}[Supermartingale]
Let $(\Omega,\mathcal{F},\probm)$ be a probability space and 
$\{\mathcal{F}_i\}_{i=0}^{\infty}$ a filtration of $\mathcal{F}$. A sequence of 
random variables $\{X_i\}_{i=0}^{\infty}$ is a \emph{supermartingale} w.r.t. 
filtration $\{\mathcal{F}_i\}_{i=0}^{\infty}$ if it satisfies these conditions:
\begin{compactenum}
\item  The process $\{X_i\}_{i=0}^{\infty}$ is \emph{adapted} to 
$\{\mathcal{F}_i\}_{i=0}^{\infty}$, i.e. for all $i\in \Nset_0$ it holds that 
$X_i$ is $\mathcal{F}_i$-measurable.
\item For all $i\in \Nset_0$ it holds $\E[|X_i|]<\infty$.
\item For all $i\in \Nset_0$ it holds 
\begin{equation}
\label{eq:supermart-def}
\E[X_{i+1}|\mathcal{F}_i] \leq X_i.
\end{equation}

A supermartingale $\{X_i\}_{i=0}^{\infty}$ has $c$-bounded differences, where 
$c\geq 0$, if $|X_{i+1}-X_i|<c$ for all $i\in \Nset_0$
\end{compactenum}
%A sequence of 
%random variables $\{X_i\}_{i=0}^{\infty}$ is a \emph{submartingale} w.r.t.  
%$\{\mathcal{F}_i\}_{i=0}^{\infty}$ if the process $\{-X_i\}_{i=0}^{\infty}$ is 
%a supermartingale w.r.t. the same filtration.
\end{definition}

Intuitively, a supermartingale is a stochastic process whose average value is 
guaranteed not to rise as time evolves, even if some information on the past 
evolution of the process is revealed. 
%Symmetrically, a submartingale is a 
%process whose value does not drop on average. 
We often need to work with supermartingales whose value 
is guaranteed to decrease on average, until a certain condition is 
satisfied. The point in time in which such a condition is satisfied is called a 
\emph{stopping time}.

\smallskip
\begin{definition}[Stopping time]
Let $(\Omega,\mathcal{F},\probm)$ be a probability space and 
$\{\mathcal{F}_i\}_{i=0}^{\infty}$ a filtration. A random variable $\stime 
\colon 
\Omega\rightarrow \Nset_0$ is called a \emph{stopping time} w.r.t. 
$\{\mathcal{F}_i\}_{i=0}^{\infty}$ if %$\stime$ is adapted to this filtration 
%and if 
for all $j\in \Nset_0$ the set $\{\omega\in \Omega\mid \stime(\omega)\leq j\}$ 
belongs to $\mathcal{F}_j$.
\end{definition}

In particular, for each set of configurations $C$ the reachability time 
$\treach{C}$ of $C$ is a stopping time w.r.t. the canonical filtration, since 
at each time $j$ we can decide whether $\treach{C}>j$ or not by looking at the 
prefix of a run of length $j$. 
%When $\Omega$ is a set of runs of some $\pCFG_P$, we can think of 
%$\stime$ as representing a 
%certain condition which can or cannot hold in a given time instance of the 
%execution of $P$, where $\stime(\run)$ represents the first point in time 
%during the execution represented by $\run$ in which $P$ holds. The condition 
%$\{\omega\in \Omega\mid \stime(\omega)\leq j\}\in \mathcal{F}_j$ then states 
%that whether $P$ holds at step $j$ of a run can be determined by looking at 
%the 
%prefix of the run of length $j$. In other words, the stopping condition cannot 
%``talk'' about future events. The typical stopping time we use in the context 
%of \APP{s} is the \emph{reachability} time $\treach{C}$ for some set of 
%configurations $C$, where $\treach{C}(\run)$ is the first point in time when 
%the current configuration on $\run$ is in $C$. In particular, if $C$ is the 
%set 
%of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
%terminal location of $\pCFG_P$), then we denote $\treach{C}$ by $\ttime$ and 
%call it a \emph{termination time} of $P$.
%We combine stopping conditions with martingales to get 
%\emph{$\eps$-decreasing} 
%supermartingales. %
Finally, we recall the fundamental notion of a ranking 
supermartingale.

\smallskip
\begin{definition}[Ranking supermartingale]
\label{def:ranking}
Let $(\Omega,\mathcal{F},\probm)$ be a probability space, 
$\{\mathcal{F}_i\}_{i=0}^{\infty}$ a filtration of $\mathcal{F}$, $\stime$ a 
stopping time w.r.t. that filtration, and $\eps\geq 
0$. 
A supermartingale $\{X_i\}_{i=0}^{\infty}$ (w.r.t. 
$\{\mathcal{F}_i\}_{i=0}^{\infty}$) is \emph{$\eps$-decreasing} until 
$\stime$ 
if it 
satisfies 
the following additional condition: for all $i\in \Nset_0$ it holds
\begin{equation}
\label{eq:ranking-sup}
\E[X_{i+1}|\mathcal{F}_i] \leq X_i - \eps\cdot\vec{1}_{\stime > i}.
\end{equation}

Further, $\{X_i\}_{i=0}^{\infty}$ is an \emph{$\eps$-ranking} supermartingale 
($\eps$-RSM) for $\stime$ if it 
is $\eps$-decreasing until $\stime$ and for 
each $\omega\in\Omega, j\in\Nset_0$ it holds $\stime(\omega)>j \Rightarrow 
X_j(\omega)\geq 0$.
%is a supermartingale w.r.t. $\{\mathcal{F}_i\}_{i=0}^{\infty}$. 
%A supermartingale $\{X_i\}_{i=0}^{\infty}$ is 
%is $\eps$-repulsing if for all $i\in \Nset_0$ it holds
%\begin{equation}
%\label{eq:repulsing-sub}
%\E[X_{i+1}|\mathcal{F}_i] \leq X_i - \eps\cdot\vec{1}_{\stime > i}.
%\end{equation}
\end{definition}
Intuitively, if $\stime$ is the reachability time $\treach{C}$ of some set $C$, 
then the previous definition requires that an $\eps$-ranking supermartingale 
must decrease by at least $\eps$ on average up to the point when $C$ is 
reached for a first time. After that, it must not increase (on average).
The above definition is a bit more general than the standard one in the
literature as we also consider reachability as opposed to only termination.

%Intuitively, until the condition specified by $\stime$ is satisfied, ranking 
%supermartingale is non-negative and at the same time decreasing by at least 
%$\eps$ on average, forming a probabilistic analogue of the classical ranking 
%functions.

%We finish this part by noting that for some applications it is preferable to 
%work with supermartingales that have \emph{bounded differences.}

%\begin{definition}
%A supermartingale $\{X_i\}_{i=0}^{\infty}$ has $c$-bounded differences, where 
%$c\geq 0$, if $|X_{i+1}-X_i|<c$ for all $i\in \Nset_0$.
%\end{definition}
%Note that $\{X_i\}_{i=0}^{\infty}$ being an $\eps$-repulsing supermartingale 
%for 
%$\stime$ is 
%\emph{not} equivalent to $\{-X_i\}_{i=0}^{\infty}$ being an $\eps$-ranking 
%supermartingale for $\stime$.

%We typically work with $\eps$-ranking supermartingales whose initial value is 
%positive and $\eps$-repulsing supermartingales whose initial value is 
%negative. 
%Then, intuitively an $\eps$-ranking supermartingale is driven towards negative 
%values by at least $\eps$ in each step (on average) until it indeed becomes 
%negative, while an $\eps$-repulsing submartingale is driven \emph{away} from 
%positive values by at least $\eps$ per step, as long as it stays non-positive 
%(we cannot rule out that it eventually becomes positive, as the decrease by 
%$\eps$ is guaranteed only on average). 
%%Note that $\{X_i\}_{i=0}^{\infty}$ being an $\eps$-repulsing submartingale is 
%%\emph{not} equivalent to $\{-X_i\}_{i=0}^{\infty}$ being an $\eps$-ranking 
%%supermartingale.

\smallskip\noindent{\bf Martingales in Program Analysis.}
In the context of \APP{} analysis, we consider a special type of 
supermartingales given as functions of the current values of program variables. 
In this paper we focus on the case when these functions are \emph{linear}.

\smallskip
\begin{definition}[Linear Expression Map]
A \emph{linear expression map (LEM)} for an \APP{} $P$ is a function $\lem$ 
assigning to each program location $\loc$ of $\pCFG_P$ an affine expression 
$\lem(\ell)$ over the program variables of $P$.
\end{definition}

Each LEM $\lem$ and location $\loc$ determines an affine function $\lem(\loc)$ 
which takes as an argument an $n$-dimensional vector, where $n$ is the number 
of distinct variables in $P$. We use $\lem(\loc,\vec{x})$ as a shorthand 
notation for $\lem(\loc)(\vec{x})$. 
Martingales for \APP{} analysis are defined via a standard notion of 
pre-expectation~\cite{SriramCAV}. Intuitively, a pre-expectation of $\lem$ is a 
function which for each configuration $(\loc,\vec{x})$ returns the maximal
expected value of $\lem$ after one step is made from this configuration, where 
the maximum is taken over all possible non-deterministic choices.

\smallskip
\begin{definition}[Pre-Expectation]
Let $P$ be an \APP{} such that $\pCFG_P = 
(\locs,\pvars,\locinit,\vecinit,\transitions,\probdist,\guards)$ and let $\lem$ 
a 
linear expression map 
for~$P$. 
%Let $\locs$ 
%be the set of locations of $\pCFG_P$ and $\pvars$ its set of program 
%variables. 
The 
pre-expectation of $\lem$ is a function $\preexp{\lem}\colon \locs\times 
\Rset^{|\pvars|} \rightarrow \Rset$ defined as follows:
\begin{compactitem} %\itemsep1pt \parskip0pt \parsep0pt
\item 
if $\loc$ is a probabilistic location, then
$$\preexp{\lem}(\loc,\vec{x}):=\sum_{(\loc,1,\id_1,\loc')\in\transitions} 
Pr_{\loc}\left((\loc,1,\id_1,\loc')\right)\cdot
 \lem(\loc',\vec{x});$$
\item 
 if $\loc$ is a non-deterministic location, then
$$\preexp{\lem}(\loc,\vec{x}):=
\max_{(\loc,1,\id_1,\loc')\in\transitions}\lem(\loc',\vec{x});$$

%\item 
%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\min_{(\loc,id,\ell')\in\transitions}\eta\left(\loc',\mathbf{x}\right)$
% if $\loc$ is an angelic location;
\item 
if $\loc$ is a deterministic location, then for each $\vec{x}$ the value 
$\preexp{\lem}(\loc,\vec{x})$ is determined as follows: there is exactly one 
transition
$\tau=(\loc,j,\up,\loc')$ such that $\vec{x}\models G(\tau)$. We distinguish 
three cases:
\begin{compactitem}
\item If $\up\colon \Rset^{|\pvars|}\rightarrow \Rset$ is a function, then 
$$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x}(j\leftarrow \up(\vec{x}))).$$
\item If $\up$ is a distribution $d$, then $$ \preexp{\lem}(\loc,\vec{x}):= 
\lem(\loc',\vec{x}(j\leftarrow \E[d])),$$ where $\E[d]$ is the expected value of the 
distribution $d$.
\item 
If $\up$ is a set, then $$ \preexp{\lem}(\loc,\vec{x}):= \max_{a\in\up}
\lem(\loc',\vec{x}(j \leftarrow a)).$$
\end{compactitem}
%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\eta(\loc',\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right))$
% if $\loc$ is a deterministic location, $(\loc,f,\loc')\in\transitions$ and 
%$\mathbf{x}\in G(\loc,f,\loc')$, where 
%$\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right)$ is the expected value of 
%$f(\vec{x},\cdot)$. %on $\rvars$ if $\vec{x}$ is treated as a constant vector.
\end{compactitem}
%\end{definition}
\end{definition}

\smallskip
\begin{definition}(Linear Ranking Supermartingale)
\label{def:lrsm}
Let $P$ be an \APP{} such that $\pCFG_P = 
(\locs,\pvars,\locinit,\vecinit,\transitions,\probdist,\guards)$, let $\inv$ be 
a linear predicate map and let 
$\confset\subseteq \locs \times \Rset^{|\pvars|}$ be some set of 
configurations. 
%\begin{compactenum}
%\item
A linear $\eps$-ranking supermartingale ($\eps$-LRSM) for $\confset$ 
supported by $\inv$ is a 
linear expression map $\lem$ for $P$ such that for all configurations 
$(\loc,\vec{x})$ of $\pCFG_P$ with
$(\loc,\vec{x})\not\in \confset$ and $\vec{x}\models \inv(\loc)$ the following 
two conditions hold:
\begin{compactitem}
%\item For all $(\loc,\vec{x})\in \locs\times\Rset^{|\pvars|}$ it holds
%\item For all  it holds 
\item
$\lem(\loc,\vec{x})\geq 0$
\item 
$\preexp{\lem}(\loc,\vec{x}) \leq \lem(\loc,\vec{x})-\eps$ 
\end{compactitem}
%\item
%\end{compactenum}
A linear $\eps$-ranking supermartingale supported by $\inv$ has 
$c$-bounded differences if for each $(\loc,\vec{x})$ such that $\vec{x}\models 
\inv(\loc)$ and each configuration $(\loc',\vec{x}')$ such that $(\loc,\vec{x}) 
(\loc',\vec{x}')$ is a path in $\pCFG_P$ it holds 
$|\lem(\loc,\vec{x})-\lem(\loc',\vec{x}')|\leq c$. 
\end{definition}

The relationship between $\eps$-LRSM in \APP{s}, (pure) invariants, and almost-sure termination 
is summarized in the following theorem. 
\smallskip
\begin{theorem}[{\cite[Theorem 1]{CFNH16:prob-termination}}]
\label{thm:old-ranking}
Let $P$ be an \APP{}, $\sigma$ a scheduler, and 
$(\Omega,\natfilt,\probm^\sigma)$ the corresponding probability space. Further, 
let $C$ be the set of terminating configurations of $\pCFG_P$ (i.e., the termination
location is reached), such that there exist an $\eps>0$ and an $\eps$-linear ranking 
supermartingale $\lem$ supported by a pure invariant $I$. 
%%for $\treach{C}$. 
Then 
\begin{compactenum}
\item
$\probm^{\sigma}(\ttime<\infty)=1$, i.e. termination is ensured almost-surely.
\item
$\E^{\sigma}[\ttime]<\lem(\locinit,\vecinit)/\eps$.
%\item
%If $\{X_i\}_{i=0}^{\infty}$  has $c$-bounded differences for some $c$, then 
%there exists a concentration bound $B$ for $\treach{C}$ such that $B\leq 
%({E^{\sigma}[X_0]}/{\eps})+1$.
\end{compactenum}
\end{theorem}

The previous result shows that if there exists  an $\eps$-LRSM 
supported by a pure invariant $I$, for $\eps>0$, then under each scheduler 
termination is ensured almost-surely.
We now demonstrate that pure invariants, though effective for almost-sure 
termination, are ineffective to answer probabilistic termination questions.

%\begin{figure}[t]
\lstset{language=affprob}
\lstset{tabsize=3}
\newsavebox{\nonterm}
\begin{lrbox}{\nonterm}
\begin{lstlisting}[mathescape]
$x:=30,y:=20$
while $y\geq 0$ do
	$x:=x+$sample$(\mathrm{Uniform}[-\frac{1}{4},1])$
	$y:=y+$sample$(\mathrm{Uniform}[-1,\frac{1}{4}])$
	while $x \leq 0$ do skip od
od
\end{lstlisting}
\end{lrbox}
\begin{figure}[t]
%%\centering
\usebox{\nonterm}
\begin{tikzpicture}[x = 1.7cm]
%\node[det] (det1) at (0,0) {$\loc_0$};
%\node[det] (while) at (1.5,-1.5)  {$\loc_1$};
%\node[det] (fin) at (0,-3) {$\loc_5$};
%\node[ran] (prob) at (3,-1.5) {$\loc_2$};
%\node[ang] (angel) at (3,0) {$\loc_3$};
%\node[dem] (demon) at (3,-3)  {$\loc_4$};
%
%\draw[tran] (det1) to node[auto, swap] {x:=0} (while);
%\draw[tran] (while) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3] {$x<0$} (fin);
%\draw[tran, loop, looseness = 5, in =-65, out = -115] (fin) to (fin);
%\draw[tran] (while) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3] {$x\geq 0$} (prob);
%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3, inner sep = 1pt] {$\frac{6}{10}$} (angel);
%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3, inner sep = 1pt] {$\frac{4}{10}$} (demon);
%\draw[tran] (demon) -- node[auto] {x:=x+1} (demon-|while) -- (while);
%\draw[tran] (angel) -- node[auto, swap] {x:=x+1} (angel-|while) -- (while);
%\draw[tran] (demon) to node[auto] {x:=x-1} (while);
%\draw[tran] (angel) to node[auto, swap] {x:=x-1} (while);

\node[det] (while) at (1.5,0)  {$\loc_0$};
%%\node[det] (fin) at (0,-3) {};
\node[det] (p1) at (2.5,0) {$\loc_1$};
\node[det] (p2) at (3.5,0) {$\loc_2$};

\node[det] (lp) at (4.5,0) {$\loc_3$};
%\node[ang] (angel) at (3,0) {$\loc_3$};
%\node[dem] at (3,-3) (demon) {$\loc_4$};
%\node[det] (aif) at (2,-0.075) {$\loc_5$};
%\node[det] (aelse) at (3,-0.8) {$\loc_6$};
%\node[det] (dif) at (2,-3) {$\loc_7$};
%\node[det] (delse) at (3,-2.2) {$\loc_8$};
%\node[det] (det1) at (0,0) {$\loc_0$};
%\node[det] (while) at (1.5,-1.5)  {};
\node[det] (fin) at (0,0) {$\loc_4$};
%\node[ran] (prob) at (3,-1.5) {};
%\node[ang] (angel) at (3,0) {};
%\node[dem] at (3,-3) (demon) {};
%
%%\draw[tran] (det1) to node[auto, swap] {x:=0} (while);
\node (dum1) at (0,0.8) {};
\node (dum2) at (0,-0.8) {};
\draw[tran] (while) to node[font=\scriptsize,draw, fill=white, 
rectangle,pos=0.5] {$y<0$} (fin);
\draw[tran, loop, looseness = 5, in =-65, out = -115] (fin) to (fin);
\draw[tran] (while) -- node[font=\scriptsize,draw, fill=white, 
rectangle,pos=0.5, inner sep = 2pt] {$y\geq0$} node[label={[label 
distance=0.12cm]-90:{x:=...}}] {} (p1);
\draw[tran] (p1) to node[label={[label 
distance=0.12cm]-90:{y:=...}}] {} (p2);
\draw[tran] (lp) to  (p2);
\draw[tran] (p2) -- (p2|-dum2) -- node[font=\scriptsize,draw, fill=white, 
rectangle,pos=0.5, inner sep = 2pt] {$x\leq 0$} (lp|-dum2) -- (lp);
\draw[tran] (p2) -- (p2|-dum1) -- node[font=\scriptsize,draw, fill=white, 
rectangle,pos=0.5, inner sep = 2pt] {$x> 0$} (while|-dum1) -- (while);
%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.3, inner sep = 1pt] {$\frac{4}{10}$} (demon);
%\draw[tran] (demon) -- node[auto] {x:=x+1} (demon-|while) -- (while);
%\draw[tran] (angel) -- node[auto, swap] {x:=x+1} (angel-|while) -- (while);
%\draw[tran] (demon) to node[auto] {x:=x-1} (while);
%\draw[tran] (angel) to node[auto, swap] {x:=x-1} (while);

%\draw[tran] (angel) -- (aif);
%\draw[tran] (angel) -- (aelse);
%\draw[tran] (demon) -- (dif);
%\draw[tran] (demon) -- (delse);

%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.5, inner sep = 1pt] {$\frac{3}{4}$} (prob|-dum1) -- node[auto] 
%{x:=x-1}
%(while|-dum1)--(while);
%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
%rectangle,pos=0.5, inner sep = 1pt] {$\frac{1}{4}$} (prob|-dum2) 
%-- node[auto,swap] {x:=x+1} (while|-dum2)--(while);
%\draw[tran] (dif) --  (demon-|while) -- node[auto, pos=0.2] {x:=x+1} (while);
%\draw[tran] (aif) --  (angel-|while) -- node[auto, swap, pos=0.2] {x:=x+1} 
%(while);
%\draw[tran] (delse) -- node[auto] {x:=x-1} (delse-|while.305) --  (while.-55);
%\draw[tran] (aelse) -- node[auto, swap] {x:=x-1} (aelse-|while.55) --  
%(while.55);
\end{tikzpicture}
\caption{A program with infinitely many reachable configurations which 
terminates with high probability, but not almost 
surely, together with a sketch of its pCFG. }\label{fig:nonterm}
\end{figure}

%We now show why this approach does not work for the probabilistic reachability and termination.

\smallskip
\begin{example}\label{ex:nonterm}
Consider the program in Figure~\ref{fig:nonterm}. 
In each 
iteration of the outer loop each of the variables is randomly modified by 
adding a number drawn from some uniform distribution.
%incremented or 
%decremented. The probability of incrementing $x$ is $\frac{3}{8}$, for
%decrementing $x$ it is $\frac{1}{8}$; the probability of incrementing $y$ it 
%is 
%$\frac{1}{8}$ and for decrementing $y$ it is $\frac{3}{8}$. 
Average increase of $x$ in each iteration is $\frac{3}{8}$, while average decrease 
of $y$ is $-\frac{3}{8}$.  It is easy to see that a program does not 
terminate almost-surely: 
there is for instance a tiny but non-zero probability of $x$ being decremented 
by at least $\frac{1}{8}$ in 
each of the first 240 loop iterations, after which we are stuck in the infinite 
inner 
loop. On 
the other hand, the expectations above show that there is a ``trend'' of $y$ 
decreasing and $x$ increasing, and executions that follow this trend eventually 
decrement $y$ below $0$ without entering the inner loop. Hence, the 
probabilistic intuition tells us that the program terminates with a high 
probability. However, the techniques of this section cannot prove 
this high-probability termination, since existence of an $\eps$-LRSM (with 
$\eps>0$) supported by a pure invariant already implies a.s. termination,
and so no such $\eps$-LRSM can exist for the program. 
\end{example}

In the next section we generalize the notion of pure invariants to stochastic invariants
for probabilistic termination to resolve issues like Example~\ref{ex:nonterm}.

%Adirect consequence of Theorems~\ref{thm:pure-supermart} is the following 
%connection 
%between pure invariants, 
%linear ranking supermartingales, and reachability properties. 
%
%\begin{corollary}
%\label{col:linear-ranking-old}
%Let $\lem$ be a linear $\eps$-ranking supermartingale for some set $C$ of 
%configurations of an \APP{} $P$ such that $\lem$ is supported by some pure 
%invariant of $P$. Then under every scheduler $\sigma$ it holds:
%
%\begin{compactenum}
%\item
%$\probm^{\sigma}(\treach{C}<\infty)=1$, i.e. the set $C$ is reached 
%almost-surely
%\item
%$\E^{\sigma}[\treach{C}]<\lem(\locinit,\vecinit)/\eps$.
%\item
%If $\{\lem(\cfg{\sigma}{i})\}_{i=0}^{\infty}$  has $c$-bounded differences for 
%some $c$, then 
%there exists a concentration bound $B$ for $\treach{C}$ such that $B\leq 
%({\lem(\locinit,\vecinit)}/{\eps})+1$.
%\end{compactenum}
%\end{corollary}

%\begin{proof}
%Denote $X_i = \lem(\cfg{\sigma}{i})$. To prove that 
%\end{proof}

%\begin{example}
%If we apply Theorem~\ref{thm:pure-supermart} on the $\frac{1}{4}$-LRSM $\lem$ 
%in Example~\ref{ex:lrsm}, we get exactly the process $\{X'_i\}_{i=0}^{\infty}$ 
%described in Example~\ref{ex:rsm}.
%\end{example}

%Having a supermartingale which decreases, on average, by some constant until 
%$C$ is reached effectively proves that $C$ is reached almost surely, very much like 
%the standard ranking functions prove that non-probabilistic program terminates. 
%Moreover, we can use such a supermartingale to prove additional assertions 
%about the reachability time.

%\begin{theorem}[variant of {\cite[Theorem 1]{CFNH16:prob-termination}}]
%\label{thm:old-ranking}
%Let $P$ be an \APP{}, $\sigma$ a scheduler, and 
%$(\Omega,\natfilt,\probm^\sigma)$ the corresponding probability space. Further, 
%let $C$ be a set of 
%configurations of $\pCFG_P$ such 
%that there exist an $\eps>0$ and an $\eps$-ranking supermartingale  
%$\{X_i\}_{i=0}^{\infty}$
%for $\treach{C}$. Then 
%\begin{compactenum}
%\item
%$\probm^{\sigma}(\treach{C}<\infty)=1$, i.e. the set $C$ is reached 
%almost-surely
%\item
%$\E^{\sigma}[\treach{C}]<\E^{\sigma}[X_0]/\eps$.
%\item
%If $\{X_i\}_{i=0}^{\infty}$  has $c$-bounded differences for some $c$, then 
%there exists a concentration bound $B$ for $\treach{C}$ such that $B\leq 
%({E^{\sigma}[X_0]}/{\eps})+1$.
%\end{compactenum}
%\end{theorem}

%\begin{remark}
%We again somewhat generalize the results of~\cite{CFNH16:prob-termination}, 
%where the 
%supermartingale 
%$\{X_i\}_{i=0}^{\infty}$ was additionally required to satisfy the 
%following: there exists $K>0$ such that with probability $1$ it holds $X_i > 
%-K$ for all $i$. We avoid this requirement by use of \emph{stopped ranking 
%supermartingale property} similar to~\cite[Section 10.9]{Williams:book}.
%\end{remark}

%Note that the requirement~\eqref{eq:ranking-sup} can be equivalently expressed 
%by stipulating that $\E[X_{i+1}|\mathcal{F}_i] \leq X_i - \eps\cdot 
%\vec{1}_{\{X_i \geq 0\}}$

### 2017-05-29 paper/prelims.tex
edited prelims

@@
 $(\Omega,\mathcal{F},\probm)$, where $\Omega$ is a non-empty set (so called
-\emph{sample space}), $\mathcal{F}$ is a \emph{sigma-algebra} over $\Omega$,
+\emph{sample space}), $\mathcal{F}$ is a \emph{sigma-algebra} of measurable 
+sets over $\Omega$,
 i.e. a collection of subsets of $\Omega$ that contains the empty set
@@
 A \emph{probabilistic control flow graph (pCFG)} is a tuple
-$\pCFG=(\locs,\pvars,\locinit,\vecinit,\transitions,\probdist,\guards)$,
+$\pCFG=(\locs,\pvars,\locinit,\vecinit,\transitions,\updates,\probdist,\guards)$,
 where
 \begin{compactitem}
-\item $\locs$ is a finite set of \emph{locations} partitioned into three 
+\item $\locs$ is a finite set of \emph{locations} partitioned into four 
 pairwise
-disjoint subsets  $\locs_N$, $\locs_P$, and $\locs_D$ of non-deterministic, 
-probabilistic, and deterministic locations;
+disjoint subsets  $\locs_N$, $\locs_P$, $\locs_D$, and $\locs_A$ of 
+non-deterministic, 
+probabilistic, deterministic, and assignment locations;
 \item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
@@
 % POKRACOVAT ODSUD
-\item $\transitions$ is a transition relation, whose members are tuples of
-the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
-program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target variable index}, and $u$ 
+\item $\transitions\subseteq\locs\times\locs$ is a transition relation'
+%whose members are tuples of
+%the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
+%program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target 
+%variable index}, and $u$ 
+%is an 
+%\emph{update element}, which can 
+%be one of the following mathematical objects: %\begin{itemize}
+%%\item 
+%(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
+%%\item 
+%(b)~a distribution $d\in \mathcal{D}$; or
+%%%\item 
+%(c)~a set $R\subseteq \Rset$.
+%\end{itemize}
+\item $\updates$ is a function assigning to each transition outgoing from an 
+assignment location a tuple $(i,\up)$, where $1\leq i \leq |\pvars|$ is a 
+\emph{target 
+	variable index} and $\up$ 
 is an 
@@
 (b)~a distribution $d\in \mathcal{D}$; or
-%%\item 
+%\item 
 (c)~a set $R\subseteq \Rset$.
@@
 probability distributions, where each $\prob_{\ell}$ is a discrete probability
-distribution on the set of all transitions outgoing from~$\ell$.
+distribution on the set of all transitions outgoing from~$\ell$;
 \item $\guards$ is a function assigning a propositionally linear predicate
@@
 	line that have positive probability under $d$.} satisfies 
-$\support(d)\subseteq\exsat{\varphi_d}$. Finally, we assume that for each 
-transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
-deterministic. This is just a technical assumption yielding no loss of 
-generality, and it somewhat simplifies notation.
+$\support(d)\subseteq\exsat{\varphi_d}$. Finally, we assume that each 
+assignment location has at most (and thus exactly) one outgoing transition.
+%Finally, we assume that for each 
+%transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
+%deterministic. This is just a technical assumption yielding no loss of 
+%generality, and it somewhat simplifies notation.
 %And we assume that each coordinate of $\mathcal{D}$ represents an integrable
@@
 $(\ell,\vec{x})$ if $\ell$ is the source location of $\tau$ and in addition,
-${\vec{x}}\models G(\tau)$ provided that $\ell$ is deterministic. A 
-configuration 
-$(\loc,\vec{x})$ is non-deterministic/probabilistic/deterministic if $\loc$ is 
-non-deterministic/probabilistic/deterministic, respectively.
+${\vec{x}}\models G(\tau)$ provided that $\ell$ is deterministic. 
+%A 
+%configuration 
+%$(\loc,\vec{x})$ is non-deterministic/probabilistic/deterministic/assignment 
+%one if 
+%$\loc$ is 
+%non-deterministic/probabilistic/deterministic/assignment one, respectively.
 
 \smallskip\noindent{\em Executions and reachable configurations.}
+We say that a configuration $(\loc',\vec{x}')$ is a \emph{successor} of a 
+configuration $(\loc,\vec{x})$ if 
+%A \emph{finite path} (or
+%\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
+%configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
+%%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
+%for each
+%$0 \leq i < k$ 
+there is a transition  $\tau=(\loc,\loc')$ enabled in
+$(\loc,\vec{x})$ 
+%such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
+%satisfies one of the following:
+and $\vec{x}'$ satisfies the following:
+\begin{compactitem}
+\item if $\loc$ is not an assignment location, then $\vec{x}'=\vec{x}$;
+\item if $\loc$ is an assignment location with $\updates(\tau)=(j,\up)$, then 
+$\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ satisfies one of the 
+following 
+depending on the type of $\up$:
+\begin{compactitem}
+\item if $\up$ is an affine function, then $a=\up(\vec{x})$;
+\item if $\up$ is an integrable\footnote{A distribution on some numerical 
+domain 
+is integrable if its expected value exists and is finite. In particular, each 
+Dirac distribution is integrable.} distribution $d$, then 
+$a\in \support(d)$;
+\item if $\up$ is a set, then $a$ is some element of $\up$.
+\end{compactitem}
+\end{compactitem}
 A \emph{finite path} (or
-\emph{execution fragment}) in $\pCFG$ is a finite sequence of
+\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
 configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
@@
 for each
-$0 \leq i < k$ there is a transition  $(\ell_i,j,\up,\ell_{i+1})$ enabled in
-$(\ell_i,\vec{x}_i)$ such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
-satisfies one of the following:
-\begin{compactitem}
-\item $\up$ is a function $f\colon\Rset^{|X|}\rightarrow\Rset$ and
-$a=f(\vec{x}_{i})$;
-\item $\up$ is an integrable\footnote{A distribution on some numerical domain 
-is integrable if its expected value exists and is finite. In particular, each 
-Dirac distribution is integrable.} distribution $d$ and 
-$a\in \support(d)$; or
-\item $\up$ is a set and $a\in \up$.
-\end{compactitem}
+$0 \leq i < k$ the configuration $(\loc_{i+1},\vec{x}_{i+1})$ is a successor of 
+$(\loc_i,\vec{x}_i)$.
 A \emph{run} (or \emph{execution}) in
@@
 if there is a finite path starting in $(\locinit,\vecinit)$ that ends in
-$(\loc,\vec{x})$.
+$(\loc,\vec{x})$. We denote by $\Conf_\pCFG,\Fpath_\pCFG$ and $\Run_\pCFG$ the 
+sets of all configurations, finite paths and runs in $\pCFG$, respectively, 
+dropping the index $\pCFG$ when known from the context.
 
@@
 probability measure over the set of all its runs. Before this can be
-done, non-determinism in $\pCFG$ needs to be resolved. This is done using the 
+done, non-determinism in $\pCFG$ needs to be resolved. This is achieved using 
+the 
 standard notion of a \emph{scheduler}.
+
+We call a configuration $(\loc,\vec{x})$ non-deterministic if $\loc$ is a 
+non-deterministic location or if $\loc$ is an assignment location with a 
+non-deterministic assignment (i.e., the only transition $\tau$ outgoing from 
+$\loc$ has $\updates(\tau)=(j,\up)$ where $\up$ is a set).
 
@@
 \label{def:schedulers}
-A scheduler in an pCFG $\pCFG$ is a tuple $\sigma=(\sigmat,\sigmaa)$,
+A scheduler in a pCFG $\pCFG$ is a function $\sigma$
 %(the subscripts stand for 'transition' and 'assignment')
-where
-\begin{compactitem}
-\item $\sigmat$ (here '$t$' stands for 'transition') is a function assigning 
+%\begin{compactitem}
+ assigning 
 to every finite path that ends in a non-deterministic configuration 
-$(\loc,\vec{x})$ a probability distribution on transitions outgoing from 
-$\loc$; 
-and
-\item $\sigmaa$ (here '$a$' stands for 'assignment') is a function which takes 
-as an argument a finite path ending in a deterministic configuration in which 
-some transition 
-$(\loc,j,u,\loc')$ with $u$ being a set is enabled, and for such a path it 
-returns a probability distribution on $u$.
-\end{compactitem}
+$(\loc,\vec{x})$ a probability distribution on successor configurations of 
+$(\loc,\vec{x})$.
+
+%\end{compactitem}
 %function
@@
 \end{definition}
+
+\smallskip\noindent{\em Measurable schedulers.}
+Note that schedulers can be viewed as partial functions from the set $\Fpath$ 
+to the set of probability distributions over the set $\Conf$.
+Since we deal with programs operating over real-valued variables, both $\Fpath$ 
+and $\Conf$ can be uncountable sets. Hence, we impose an 
+additional \emph{measurability} condition on schedulers, so as to ensure that 
+the semantics of probabilistic non-deterministic programs is defined in a 
+mathematically 
+sound way. First we need to clarify what are measurable sets of configurations 
+and histories. We define a sigma-algebra $\mathcal{F}_{\Conf}$ of measurable 
+sets of configurations to be the sigma-algebra 
+over $\Conf$ 
+generated\footnote{In general, it is known that for each set $\Omega$ and each 
+collection of its subsets $F\subseteq 2^{\Omega}$ there exists at least one 
+sigma-algebra $\mathcal{F}$ s.t. $F\subseteq \mathcal{F}$ and the intersection 
+of all such sigma-algebras is again a sigma algebra --  so called sigma-algebra 
+generated by $F$~\cite{Billingsley:book}.} by 
+all sets of the form $\{\loc\}\times B$, where $\loc$ is a location of $\pCFG$ 
+and  
+$B$ is a Borel-measurable subset 
+of $\Rset^{|\pvars|}$. Next, the set of finite paths $\Fpath$ can be viewed as 
+a subset of $\Conf\cup \Conf\times\Conf \cup \Conf\times\Conf\times\Conf \cup 
+\dots$. Hence, we define the sigma-algebra $\mathcal{H}$ of measurable sets of 
+finite paths 
+to be the sigma algebra generated by all sets of the form $\confset_1\times 
+\confset_2 \times 
+\cdots \times \confset_k \subseteq \Fpath$ such that $k\in \Nset$ and 
+$\confset_i \in 
+\mathcal{F}_{\Conf}$ for 
+all $1\leq i \leq k$. Now we can define the measurability of schedulers. Recall 
+that for each finite path $\genpath$ ending in a non-deterministic 
+configuration 
+we have that $\sigma(\genpath)$ is a probability distribution on $\Conf$, hence 
+for each measurable set of configurations $\confset$ we denote by 
+$\sigma(\genpath)(\confset)$ the probability that the random draw from 
+distribution $\sigma(\genpath)$ selects an element of $\confset$. We say that a 
+scheduler $\sigma$ is measurable if for each $\confset \in \mathcal{F}_{\Conf}$ 
+and each $p\in[0,1]$ the set $\{\pi \in \Fpath \mid 
+\sigma(\genpath)(\confset)\leq p\}$ belongs to $\mathcal{H}$, i.e., it is a 
+measurable set of paths.
+
+While the definition of a measurable scheduler might seem somewhat technical, 
+it is natural from measure-theoretic point of view and analogous definitions 
+inevitably emerge in works dealing with systems that exhibit both probabilistic 
+and non-deterministic behaviour over a continuous state 
+space~\cite{NSK:CTMDP-delayed,NK:CTMDP-bisimulation}. In particular, if all the 
+variables in a program range over a discrete subset (such as the integers), 
+then each scheduler in the associated pCFG is measurable.
+%Formally, let 
+%$\Borel(\Rset^{|\pvars|})$ be the standard sigma-algebra of Borel-measurable 
+%subsets of $\Rset^{|\pvars|}$. Now let $\genPathSet$ be any set of finite 
+%paths 
+%in 
+%$\pCFG$ and denote by $\proj_j (A)$ the set of all variable valuations that 
+%appear on some path in $A$ in step $j$, i.e. $\proj_j(A) = 
+%\{\vec{y}\in\Rset^{|\pvars|} \mid \exists  
+%(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k) \in A \text{ s.t. } k \geq j \} $
+%A \emph{measurable cylinder} of length 
+%$k$ is a 
+%set of paths $\cylinder_n$ in $\pCFG$ satisfying the following condition:
 

### 2017-05-30 paper/prelims.tex
edited prelims

@@
 pairwise
-disjoint subsets  $\locs_N$, $\locs_P$, $\locs_D$, and $\locs_A$ of 
-non-deterministic, 
-probabilistic, deterministic, and assignment locations;
+disjoint subsets  $\locsNB$, $\locsPB$, $\locs_D$, and $\locs_A$ of 
+non-deterministic branching, 
+probabilistic branching, deterministic, and assignment locations;
 \item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
@@
 %\end{itemize}
-\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locs_P}$ is a collection of
+\item $\probdist=\{\prob_{\ell}\}_{\ell \in \locsPB}$ is a collection of
 probability distributions, where each $\prob_{\ell}$ is a discrete probability
@@
 %\paragraph*{Dynamics of Stochastic Games}
+
+The translation from probabilistic programs to the corresponding pCFG is 
+standard~\cite{CFNH16:prob-termination}, and the details are presented 
+in~\AppendixMaterial.
 
@@
 
-\smallskip\noindent{\em Schedulers.}
+\smallskip\noindent{\em Non-determinism and Schedulers.}
 Due to the presence of non-determinism and probabilistic choices, a pCFG
@@
 
-We call a configuration $(\loc,\vec{x})$ non-deterministic if $\loc$ is a 
-non-deterministic location or if $\loc$ is an assignment location with a 
+There are two sources of non-determinism in our programs: one in branching and 
+one in assignments. We call a location $\loc$ non-deterministic if $\loc$ is a 
+non-deterministic branching location or if $\loc$ is an assignment location 
+with a 
 non-deterministic assignment (i.e., the only transition $\tau$ outgoing from 
-$\loc$ has $\updates(\tau)=(j,\up)$ where $\up$ is a set).
+$\loc$ has $\updates(\tau)=(j,\up)$ where $\up$ is a set). A configuration 
+$(\loc,\vec{x})$ is non-deterministic if $\loc$ is non-deterministic.
 
@@
 configuration 
-we have that $\sigma(\genpath)$ is a probability distribution on $\Conf$, hence 
-for each measurable set of configurations $\confset$ we denote by 
+we have that $\sigma(\genpath)$ is a probability distribution on $\Conf$. For 
+each measurable set of configurations $\confset$ we denote by 
 $\sigma(\genpath)(\confset)$ the probability that the random draw from 
@@
 space~\cite{NSK:CTMDP-delayed,NK:CTMDP-bisimulation}. In particular, if all the 
-variables in a program range over a discrete subset (such as the integers), 
+variables in a program range over a discrete set (such as the integers), 
 then each scheduler in the associated pCFG is measurable.
@@
 \smallskip\noindent{\em Stochastic process.}
-A pCFG $\pCFG$ together with a scheduler $\sigma$ can be seen as a stochastic 
+A pCFG $\pCFG$ together with a scheduler $\sigma$ define a stochastic 
 process which produces a random run 
@@
 steps have elapsed, i.e. a finite path 
-$(\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i)$ has already 
-been produced. Then
+$\genpath_i=(\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i)$ has 
+already 
+been produced. Then a successor configuration $(\loc_{i+1},\vec{x}_{i+1})$ is 
+chosen as follows:
 \begin{compactitem}
 \item
-A transition $\tau=(\loc,j,u,\loc')$ enabled in 
-$(\lr{}{i},\vr{}{i})$ is chosen as follows:
-
-\begin{compactitem}
+If $\loc_i$ is a non-deterministic location, then 
+$(\loc_{i+1},\vec{x}_{i+1})$ is sampled according to scheduler $\sigma$, i.e. 
+from 
+the distribution $\sigma(\genpath_i)$.
 \item
-If $\ell_i$ is non-deterministic then $\tau$ is chosen randomly
-according to the distribution specified
-by scheduler
-$\sigma$, i.e. according to the distribution 
-$\sigmat((\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i))$.
-%That is, if $\ell_i$ is angelic (resp., demonic) and $c_0c_1\cdots c_i$ is the
-%sequence of configurations observed so far,
-%then $\tau$ equals %%the distribution
-%$\sigma(c_0c_1\cdots c_i)$ (resp., $\pi(c_0c_1\cdots c_i)$).
-
-\item
-If $\ell_i$ is probabilistic, then $\tau$ is chosen randomly according to the
-distribution $\probdist_{\ell_i}$.
-\item
-If $\ell_i$ is deterministic, then by the definition of a pCFG there is exactly
-one enabled transition outgoing from $\ell_i$, and this transition is chosen as
-$\tau$.
-\end{compactitem}
-\item
-Once $\tau$ is chosen as above, we put $\loc_{i+1}=\loc'$. Next, we put 
-$\vec{x}_{i+1}=\vec{x}_{i}(j\leftarrow a)$, where $a$ chosen as follows:
+If $\loc_i$ is an assignment location (but not a non-deterministic one) with, 
+there 
+is exactly one transition $\tau=(\loc_i,\loc')$ outgoing from it and we put 
+$\loc_{i+1}=\loc'$. Denoting $\updates(\tau)=(j,\up)$, the vector 
+$\vec{x}_{i+1}$ is 
+then defined as 
+$\vec{x}_{i+1}=\vec{x}_{i}(j\leftarrow a)$ where $a$ is chosen depending on 
+$\up$:
 \begin{compactitem}
@@
 is sampled from $d$.
-\item If $\up$ is a set, then $a$ is sampled from a 
-distribution 
-$\sigmaa((\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i))$.
 \end{compactitem}
+\item In all other cases we have $\vec{x}_{i+1}=\vec{x}_i$, and $\loc_{i+1}$ is 
+determined as follows:
+\begin{itemize}
+	\item If $\loc_i$ is a probabilistic branching location, then a transition 
+	$(\loc_i,\loc')$ is sampled from $\probdist_{\loc_i}$ and we put
+	$\loc_{i+1}=\loc'$;
+	\item If $\loc_i$ is deterministic, then there is exactly one transition 
+	$(\loc_i,\loc')$ enabled in $(\loc_i,\vec{x}_i)$, in which case we put 
+	$\loc_{i+1}=\loc'$.
+\end{itemize}
 \end{compactitem}
+
+%\begin{compactitem}
+%\item
+%a transition $\tau=(\loc_i,\loc')$ enabled in 
+%$(\lr{}{i},\vr{}{i})$ is chosen as follows:
+%
+%\begin{compactitem}
+%\item
+%If $\ell_i$ is non-deterministic, then $\tau$ is chosen randomly
+%according to the distribution specified
+%by scheduler
+%$\sigma$, i.e. according to distribution
+%$\sigma((\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i))$ 
+%(formally, this is a distribution over successor configurations of 
+%$(\loc_i,\vec{x}_i)$, but each such successor $(\loc',\vec{x}')$ uniquely 
+%determines a transition $(\loc_i,\loc')$ ).
+%%That is, if $\ell_i$ is angelic (resp., demonic) and $c_0c_1\cdots c_i$ is the
+%%sequence of configurations observed so far,
+%%then $\tau$ equals %%the distribution
+%%$\sigma(c_0c_1\cdots c_i)$ (resp., $\pi(c_0c_1\cdots c_i)$).
+%
+%
+%\item
+%If $\ell_i$ is probabilistic, then $\tau$ is chosen randomly according to the
+%distribution $\probdist_{\ell_i}$.
+%\item
+%If $\ell_i$ is deterministic or assignment location, then by the definition of 
+%a pCFG there is exactly
+%one enabled transition outgoing from $\ell_i$, and this transition is chosen as
+%$\tau$.
+%\end{compactitem}
+%\item
+%Once $\tau$ is chosen as above, we put $\loc_{i+1}=\loc'$. Next, if $\loc_i$ 
+%is 
+%not an assignment location, we put $\vec{x}_{i+1}=\vec{x}_{i}$. Otherwise 
+%$\vec{x}_{i+1}=\vec{x}_{i}(j\leftarrow a)$, where $a$ chosen as follows:
+%\begin{compactitem}
+%\item If $u$ is a function $u\colon
+%\Rset^{|\pvars|}\rightarrow \Rset$, then $a=f(\vec{x}_{i})$.
+%\item If $\up$ is a distribution $d$, then $a$ 
+%is sampled from $d$.
+%\item If $\up$ is a set, then $a$ is sampled from a 
+%distribution 
+%$\sigmaa((\loc_0,\vec{x}_0)(\loc_1,\vec{x}_1)\cdots(\loc_i,\vec{x}_i))$.
+%\end{compactitem}
+%\end{compactitem}
 
@@
 such 
-that for each $\run\in \Omega$ we have that $\cfg{\sigma}{i}(\run)$ is the 
-$i$-th configuration on run $\run$ (i.e., $\cfg{\sigma}{i}$ is a random vector 
+that for each run $\run\in \Omega$ we have that $\cfg{\sigma}{i}(\run)$ is the 
+$i$-th configuration on $\run$ (i.e., $\cfg{\sigma}{i}$ is a random vector 
 $(\lr{\sigma}{i},\vr{\sigma}{i})$ with  $\lr{\sigma}{i}$ taking 
@@
 $\natfilt$ is the smallest (w.r.t. inclusion) sigma algebra under which all 
-the functions $\cfg{\sigma}{i}$, for all $i\geq 0$ and all schedulers $\sigma$, 
-are $\natfilt$-measurable (a 
-function $f$ returning vectors is $\natfilt$-measurable if for all 
-real-valued vectors 
-$\vec{y}$ of appropriate dimension the set $\{\omega\in \Omega\mid 
-f(\omega)\leq \vec{y}\}$ belongs to $\natfilt$). The 
+the functions $\cfg{\sigma}{i}$, for all $i\geq 0$, 
+are $\natfilt$-measurable (i.e., for each $\cfg{\sigma}{i}$ and each measurable 
+set of configurations $\confset\in\mathcal{F}_{\Conf}$ it holds 
+$\{\omega\mid\cfg{\sigma}{i}(\omega)\in \confset\}\in \natfilt$). Equivalently, 
+$\natfilt$ can be defined as a sigma algebra generated by all set of runs of 
+the form $F\times \Conf^{\infty}$, where $F\in \mathcal{H}$ is a measurable set 
+of finite paths. The 
 probability 
@@
 generated. The formal construction 
-of $\natfilt$ and $\probm^{\sigma}$ is standard~\cite{Billingsley:book} and 
+of $\natfilt$ and $\probm^{\sigma}$ proceeds via the standard \emph{cylinder 
+construction}~\cite[Theorem 2.7.2]{Ash:book} and is
 somewhat 
@@
 $(\Omega,\natfilt,\probm^{\sigma})$.
-The translation from probabilistic programs to the corresponding pCFG is 
-standard~\cite{CFNH16:prob-termination}, and the details are presented 
-in~\cite{CNZ16:prob-invariants-arxiv}. We point out that the 
-construction produces pCFGs with a property that only transitions outgoing from 
-a deterministic state can update program variables. All other transitions are 
-assumed to be of the form $(\ell,1,\id_1,\ell')$ for some locations 
-$\ell,\ell'$, where $\id_1(\vec{x})=\vec{x}[1]$ for all $\vec{x}$. We use this 
-to simplify notation.
-An illustration of a pCFG is given in Figure~\ref{fig:invariant-running}.
+
+%We point out that the 
+%construction produces pCFGs with a property that only transitions outgoing 
+%from 
+%a deterministic state can update program variables. All other transitions are 
+%assumed to be of the form $(\ell,1,\id_1,\ell')$ for some locations 
+%$\ell,\ell'$, where $\id_1(\vec{x})=\vec{x}[1]$ for all $\vec{x}$. We use this 
+%to simplify notation.
+%An illustration of a pCFG is given in Figure~\ref{fig:invariant-running}.
 

### 2017-05-30 implementation/Parser.cpp
Equation printing phase1

@@ -24,7 +24,6 @@ string program;
 int nVariables;
 int last_used_label = 0;
 map<int,CFG_location*> label_map;
-node* id1 = new node("expr");
 
 void skip_spaces(int &begin, int &end){ //skips the spaces in the beginning and the end
 	while(begin<program.size() and isspace(program[begin])){
@@ -122,7 +121,7 @@ void node::proc_stmt(int s,int l){
 	skip_spaces(begin,end);
 	if(part(program,begin,end)=="skip"){
 		constant = "skip";
-		CFG_edge temporary_edge(label_map[l],1,id1);
+		CFG_edge temporary_edge(label_map[l],-1,NULL);
 		label_map[s]->edges.pb(temporary_edge);
 		// cout<<"Adding edge from "<<s<<"to"<<l<<endl;
 		return;
@@ -189,8 +188,8 @@ void node::proc_stmt(int s,int l){
 		label_map[mid] = new CFG_location("det",mid);
 		node* temporary_node;
 		temporary_node = negation(children[0]);
-		CFG_edge temporary_edge1(label_map[l],1,id1,children[0]);
-		CFG_edge temporary_edge2(label_map[mid],1,id1,temporary_node);
+		CFG_edge temporary_edge1(label_map[l],-1,NULL,children[0]);
+		CFG_edge temporary_edge2(label_map[mid],-1,NULL,temporary_node);
 		label_map[s]->edges.pb(temporary_edge1);
 		label_map[s]->edges.pb(temporary_edge2);
 		// cout<<"Adding edge from "<<s<<"to"<<l<<endl;
@@ -236,8 +235,8 @@ void node::proc_stmt(int s,int l){
 		int case2 = ++last_used_label;
 		label_map[case1] = new CFG_location("det",case1);
 		label_map[case2] = new CFG_location("det",case2);
-		CFG_edge temporary_edge1(label_map[case1],1,id1);
-		CFG_edge temporary_edge2(label_map[case2],1,id1);
+		CFG_edge temporary_edge1(label_map[case1],-1,NULL);
+		CFG_edge temporary_edge2(label_map[case2],-1,NULL);
 		constant = "if";
 		children.resize(3);
 		children[0] = new node("ndbexpr",begin+2,firstthen);
@@ -474,6 +473,20 @@ void node::proc_literal(bool negate){
 		children[0] = new node("expr",begin,sign);
[…]

### 2017-05-30 implementation/main.cpp

@@ -17,13 +17,19 @@ using namespace std;
 #define MAXL 100000 //Maximum length of the program
 #define part(x,a,b) (x.substr((a),((b)-(a))))
 #define pb push_back
+#define A(i,j) (top->affexpr->children[i]->children[0]->expression[j+1])
+#define b(i) (-1.0*top->affexpr->children[i]->children[0]->expression[0])
 
 struct cond{
+	int toChange;
+	node* change;
 	int src;
 	int dest1;
 	int dest2;
 	bool probability;
-	cond(int src,int dest1,int dest2 = -1,double probability = -1.0){
+	cond(int toChange,node* change,int src,int dest1,int dest2 = -1,double probability = -1.0){
+		this->toChange = toChange;
+		this->change = change;
 		this->src = src;
 		this->dest1 = dest1;
 		this->dest2 = dest2;
@@ -74,7 +80,7 @@ void generate_equations(){ //Would use the ofstream file to write the equations 
 			//Invariant and guard imply the value decrease
 			if(it->second->edges.size()==1){
 				//First make a condition
-				cond* condition = new cond(it->first,it->second->edges[0].next->label);
+				cond* condition = new cond(it->second->edges[0].toChange,it->second->edges[0].change,it->first,it->second->edges[0].next->label);
 				// Guard would have been NULL here
 				if(it->second->invariant==NULL){
 					//This should never be the case as this would pose conditions that c==0 and d>0, which are not good
@@ -91,8 +97,8 @@ void generate_equations(){ //Would use the ofstream file to write the equations 
 			}
 			else{
 				//First make the 2 conditions required
-				cond* condition0 = new cond(it->first,it->second->edges[0].next->label);
-				cond* condition1 = new cond(it->first,it->second->edges[1].next->label);
+				cond* condition0 = new cond(-1,NULL,it->first,it->second->edges[0].next->label);
+				cond* condition1 = new cond(-1,NULL,it->first,it->second->edges[1].next->label);
 				// Size is 2, so considering the guards is important and guards cannot be NULL
 				if(it->second->invariant==NULL or it->second->edges[0].guard==NULL or it->second->edges[1].guard==NULL){
 					//This should never be the case as this would pose conditions that c==0 and d>0, which are not good
[…]

### 2017-05-31 implementation/main.cpp
edited equation making

@@ -148,39 +148,48 @@ int last_used_lambda = 0;
 
 void print_equations(){
 	vector<string> c;
-	string d;
+	string negative_d; //negative component of d
+	// string constant; //cons
 	c.resize(nVariables);
 	while(!equations.empty()){
 		equation* top = equations.top();
-		// equationsfile<<"Equation"<<endl;
-		// top->affexpr->print();
-		// cout<<" implies ";
-		// cout<<top->condition->src<<" "<<top->condition->dest1<<endl;
 		//A(i,j) means affexpr->children[i-1]->children[0]->expression[j] and b(i) translates to -1.0*affexpr->children[i-1]->children[0]->expression[0]
 		//Creating a macro for this
 		//Create c and d now
-		// cout<<"c is"<<endl;
 		if(top->condition->dest2==-1){
 			if(top->condition->change==NULL){
-				d = "-epsilon+f_"+to_string(top->condition->src)+"_0-f_"+to_string(top->condition->dest1)+"_0"; 
+				negative_d = "epsilon-f_"+to_string(top->condition->src)+"_0+f_"+to_string(top->condition->dest1)+"_0"; 
 				for(int i=0;i<nVariables;i++){
 					// c[i] = "";
-					c[i] = "f_"+to_string(top->condition->dest1)+"_"+to_string(i+1)+"-"+"f_"+to_string(top->condition->src)+"_"+to_string(i+1);
+					c[i] = "f_"+to_string(top->condition->dest1)+"_"+to_string(i+1)+"-f_"+to_string(top->condition->src)+"_"+to_string(i+1);
 					// cout<<c[i]<<endl;
 				}
 			}
 			else{
-				for(int i=0;i<nVariables;i++){
+				// negative_d = "not set";
+				negative_d = "epsilon-f_"+to_string(top->condition->src)+"_0+f_"+to_string(top->condition->dest1)+"_0";
+				if(top->condition->change->expression[0]<0.0){
+					negative_d = negative_d+to_string(top->condition->change->expression[0])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
+				}
+				else if(top->condition->change->expression[0]>0.0){
[…]

### 2017-05-31 implementation/main.cpp
generating equations testing left

@@ -149,7 +149,6 @@ int last_used_lambda = 0;
 void print_equations(){
 	vector<string> c;
 	string negative_d; //negative component of d
-	// string constant; //cons
 	c.resize(nVariables);
 	while(!equations.empty()){
 		equation* top = equations.top();
@@ -160,13 +159,10 @@ void print_equations(){
 			if(top->condition->change==NULL){
 				negative_d = "epsilon-f_"+to_string(top->condition->src)+"_0+f_"+to_string(top->condition->dest1)+"_0"; 
 				for(int i=0;i<nVariables;i++){
-					// c[i] = "";
 					c[i] = "f_"+to_string(top->condition->dest1)+"_"+to_string(i+1)+"-f_"+to_string(top->condition->src)+"_"+to_string(i+1);
-					// cout<<c[i]<<endl;
 				}
 			}
 			else{
-				// negative_d = "not set";
 				negative_d = "epsilon-f_"+to_string(top->condition->src)+"_0+f_"+to_string(top->condition->dest1)+"_0";
 				if(top->condition->change->expression[0]<0.0){
 					negative_d = negative_d+to_string(top->condition->change->expression[0])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
@@ -175,13 +171,25 @@ void print_equations(){
 					negative_d = negative_d+"+"+to_string(top->condition->change->expression[0])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
 				}
 				for(int i = 0;i<nVariables;i++){
-					if(i!=top->condition->toChange){
-						c[i] = "f_"+to_string(top->condition->dest1)+"_"+to_string(i+1)+"-"+"f_"+to_string(top->condition->src)+"_"+to_string(i+1);
+					if((i+1)!=top->condition->toChange){
+						c[i] = "f_"+to_string(top->condition->dest1)+"_"+to_string(i+1);
+						if(top->condition->change->expression[i]>0.0){
+							c[i] = c[i]+"+"+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
+						}
+						else if(top->condition->change->expression[i]<0.0){
+							c[i] = c[i]+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
[…]

### 2017-05-31 implementation/main.cpp
progress

@@ -26,7 +26,7 @@ struct cond{
 	int src;
 	int dest1;
 	int dest2;
-	bool probability;
+	double probability;
 	cond(int toChange,node* change,int src,int dest1,int dest2 = -1,double probability = -1.0){
 		this->toChange = toChange;
 		this->change = change;
@@ -174,22 +174,22 @@ void print_equations(){
 					if((i+1)!=top->condition->toChange){
 						c[i] = "f_"+to_string(top->condition->dest1)+"_"+to_string(i+1);
 						if(top->condition->change->expression[i]>0.0){
-							c[i] = c[i]+"+"+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
+							// c[i] = c[i]+"+"+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
 						}
 						else if(top->condition->change->expression[i]<0.0){
-							c[i] = c[i]+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
+							// c[i] = c[i]+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
 						}
 					}
 					else{
 						// c[i] = "f_"+to_string(top->condition->dest1)+"_"+to_string(i+1);
 						if(top->condition->change->expression[i]!=0.0){
-							c[i] = c[i]+"+"+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
+							// c[i] = c[i]+"+"+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
 						}
 						else{
-							c[i] = "";
+							// c[i] = "";
 						}
 					}
-					c[i] = c[i]+"-f_"+to_string(top->condition->src)+"_"+to_string(i+1);
+					// c[i] = c[i]+"-f_"+to_string(top->condition->src)+"_"+to_string(i+1);
 				}
 			}
 		}
@@ -197,7 +197,7 @@ void print_equations(){
 			// It was a stochastic node, means change would have been NULL
[…]

### 2017-05-31 implementation/Parser.cpp
added the end invariant

@@ -119,6 +119,19 @@ node* negation(node* tonegate){
 
 void node::proc_stmt(int s,int l){
 	skip_spaces(begin,end);
+	if(program[begin]=='['){
+		int closed_bracket = -1;
+		for(int j = begin+1;j<end;++j){
+			if(program[j]==']'){
+				closed_bracket = j;
+				break;
+			}
+		}
+		bracket = new node("bexpr",begin+1,closed_bracket);
+		label_map[s]->invariant = bracket;
+		begin = closed_bracket+1;
+	}
+	skip_spaces(begin,end);
 	if(part(program,begin,end)=="skip"){
 		constant = "skip";
 		CFG_edge temporary_edge(label_map[l],-1,NULL);
@@ -149,19 +162,7 @@ void node::proc_stmt(int s,int l){
 	//look for brackets
 	// Assuming the invariants are at the beginning itself
 	skip_spaces(begin,end);
-	if(program[begin]=='['){
-		int closed_bracket = -1;
-		for(int j = begin+1;j<end;++j){
-			if(program[j]==']'){
-				closed_bracket = j;
-				break;
-			}
-		}
-		bracket = new node("bexpr",begin+1,closed_bracket);
-		label_map[s]->invariant = bracket;
-		begin = closed_bracket+1;
-	}
-	skip_spaces(begin,end);
+	
 
 	//check if it is a while
 	if(part(program,begin,begin+5)=="while"){

### 2017-05-31 implementation/main.cpp

@@ -86,6 +86,7 @@ void generate_equations(){ //Would use the ofstream file to write the equations 
 					//This should never be the case as this would pose conditions that c==0 and d>0, which are not good
 					// cerr<<"No invariant specified here"<<endl;
 					//This would arise for the end states and skip and other special conditions
+					cout<<it->first<<endl;
 				}
 				else{
 					// Invariant implies the given condition
@@ -174,22 +175,21 @@ void print_equations(){
 					if((i+1)!=top->condition->toChange){
 						c[i] = "f_"+to_string(top->condition->dest1)+"_"+to_string(i+1);
 						if(top->condition->change->expression[i]>0.0){
-							// c[i] = c[i]+"+"+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
+							c[i] = c[i]+"+"+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
 						}
 						else if(top->condition->change->expression[i]<0.0){
-							// c[i] = c[i]+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
+							c[i] = c[i]+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
 						}
 					}
 					else{
-						// c[i] = "f_"+to_string(top->condition->dest1)+"_"+to_string(i+1);
 						if(top->condition->change->expression[i]!=0.0){
-							// c[i] = c[i]+"+"+to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
+							c[i] = to_string(top->condition->change->expression[i])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
 						}
 						else{
-							// c[i] = "";
+							c[i] = "";
 						}
 					}
-					// c[i] = c[i]+"-f_"+to_string(top->condition->src)+"_"+to_string(i+1);
+					c[i] = c[i]+"-f_"+to_string(top->condition->src)+"_"+to_string(i+1);
 				}
 			}
 		}
@@ -249,7 +249,26 @@ int main(){
[…]

### 2017-06-01 implementation/main.cpp
progress

@@ -1,3 +1,4 @@
+// Do I have to add a self loop on the end CFG location, because then the epsilon equation creates issues
 #include <iostream>
 #include <algorithm>
 #include <cstdio>
@@ -148,6 +149,7 @@ void generate_equations(){ //Would use the ofstream file to write the equations 
 int last_used_lambda = 0;
 
 void print_equations(){
+	cout<<"maximise eps\n\nst\n\neps >= 0"<<endl;
 	vector<string> c;
 	string negative_d; //negative component of d
 	c.resize(nVariables);
@@ -158,13 +160,13 @@ void print_equations(){
 		//Create c and d now
 		if(top->condition->dest2==-1){
 			if(top->condition->change==NULL){
-				negative_d = "epsilon-f_"+to_string(top->condition->src)+"_0+f_"+to_string(top->condition->dest1)+"_0"; 
+				negative_d = "eps-f_"+to_string(top->condition->src)+"_0+f_"+to_string(top->condition->dest1)+"_0"; 
 				for(int i=0;i<nVariables;i++){
 					c[i] = "f_"+to_string(top->condition->dest1)+"_"+to_string(i+1)+"-f_"+to_string(top->condition->src)+"_"+to_string(i+1);
 				}
 			}
 			else{
-				negative_d = "epsilon-f_"+to_string(top->condition->src)+"_0+f_"+to_string(top->condition->dest1)+"_0";
+				negative_d = "eps-f_"+to_string(top->condition->src)+"_0+f_"+to_string(top->condition->dest1)+"_0";
 				if(top->condition->change->expression[0]<0.0){
 					negative_d = negative_d+to_string(top->condition->change->expression[0])+"f_"+to_string(top->condition->dest1)+"_"+to_string(top->condition->toChange);
 				}
@@ -195,7 +197,7 @@ void print_equations(){
 		}
 		else{
 			// It was a stochastic node, means change would have been NULL
-			negative_d = "epsilon-f_"+to_string(top->condition->src)+"_0+"+to_string(top->condition->probability)+"f_"+to_string(top->condition->dest1)+"_0+"+to_string(1.0-top->condition->probability)+"f_"+to_string(top->condition->dest2)+"_0"; 
[…]

### 2017-06-01 paper/lexicographic.tex
added a definition of LexRSM

\section{Lexicographic Supermartingales}
\label{sec:lexicographic}

In this section we introduce the notion of a \emph{lexicographic ranking 
supermartingale}, which generalizes the standard notion of ranking 
supermartingales.

\begin{definition}[Lexicographic Ranking Supermartingale]
Let $(\Omega,\genfilt,\probm)$ be a probability space, 
$\{\genfilt_i\}_{i=0}^{\infty}$ a filtration of $\genfilt$, $\stime$ a stopping 
time w.r.t. that filtration, and 
$\eps\geq 0$. 
An $n$-dimensional stochastic process $\{\vec{X}_{i}\}_{i=0}^{\infty}$ is a 
\emph{lexicographic $\eps$-ranking supermartingale for $\stime$ 
($\eps$-LexRSM)} if the 
following 
conditions hold:
\begin{compactenum}
\item For each $1\leq j \leq n$ the 1-dimensional stochastic process 
$\{\vecseq{X}{i}{j}\}_{i=0}^{\infty}$ is adapted to 
$\{\genfilt_i\}_{i=0}^{\infty}$.
\item For each $i\in \Nset_0$ and each $1\leq j \leq n$ it holds 
$\E[|\vecseq{X}{i}{j}|]<\infty$.
\item For each $i\in \Nset_0$ there exists $1\leq j \leq n$ such that
\begin{compactitem}
	\item $\E[\vecseq{X}{i+1}{j}\mid \genfilt_i]\leq\vecseq{X}{i}{j} - 
	\eps\cdot \vec{1}_{\stime > i} $; and
	\item for all $1 \leq j' < j$ it holds $\E[\vecseq{X}{i+1}{j'}\mid 
	\genfilt_i]\leq\vecseq{X}{i}{j'}$.
\end{compactitem}
\end{compactenum}
\end{definition}

### 2017-06-01 paper/prelims.tex
added a definition of LexRSM

@@
 \vspace{-1em}
-\subsection{Almost-Sure and Probabilistic Termination}
+\subsection{Almost-Sure and Positive Termination}
 \vspace{-0.5em}
@@
 We say that a run $\run$ of $\pCFG_P$ \emph{reaches} a set of configurations 
-$C$ if it  contains a configuration from $C$. A run \emph{terminates} if it reaches a 
+$\confset$ if it  contains a configuration from $\confset$. A run 
+\emph{terminates} if it reaches a 
 configuration whose first component (i.e. a location of $\pCFG_P$) is the 
@@
 executing $P$. 
-To each set of configurations $C$ we can assign a random variable $\treach{C}$ 
-such that for each run $\run$ the value $\treach{C}(\run)$ represents the first 
+To each set of configurations $\confset$ we can assign a random variable 
+$\treach{\confset}$ 
+such that for each run $\run$ the value $\treach{\confset}(\run)$ represents 
+the first 
 point in time when 
-the current configuration on $\run$ is in $C$. If a run $\run$ does \emph{not} 
-reach a set $C$, then 
-$\treach{C}(\run)=\infty$. We call $\treach{C}$ the 
-\emph{reachability time} of $C$. In particular, if $C$ is the set 
+the current configuration on $\run$ is in $\confset$. If a run $\run$ does 
+\emph{not} 
+reach a set $\confset$, then 
+$\treach{\confset}(\run)=\infty$. We call $\treach{\confset}$ the 
+\emph{reachability time} of $\confset$. In particular, if $\confset$ is the set 
 of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
-terminal location of $\pCFG_P$), then $\treach{C}$ is 
+terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
 called a \emph{termination time}, as it returns the number of steps after which 
@@
 almost-surely terminating.
-\item \emph{Probabilistic termination:} In probabilistic termination question 
-for $P$ we aim to compute a \emph{lower bound on the probability of 
-termination}, i.e. a bound $b\in[0,1]$ such that for each scheduler 
-$\sigma$ it holds $\probm^{\sigma}(\{\run\mid \run \text{ terminates}\}) \geq b$ 
-(or equivalently $\probm^{\sigma}(\ttime<\infty)\geq b$).
+\item \emph{Positive termination:} A probabilistic program $P$ is 
+positively
+terminating if under each scheduler $\sigma$ it holds that 
+$\E^{\sigma}(\{\ttime\}) \infty$. In 
+positive termination question for $P$ we aim to prove that $P$ 
+is 
+positively terminating. Note that each positively terminating program is also 
+a.s. terminating, but the converse does not hold.
 \end{compactenum}
@@
 $\probm^{\sigma}(\treach{C}<\infty)=1$.
-\item \emph{Probabilistic reachability:} For a set $C$ of configurations of 
-a probabilistic program $P$, compute a bound $b\in[0,1]$ such that for each scheduler
-$\sigma$ it holds $\probm^{\sigma}(\treach{C}<\infty)\geq b$.
+\item \emph{Positive reachability:} For a set $C$ of configurations of 
+a probabilistic program $P$, prove (if possible) that under each scheduler 
+$\sigma$ it holds that
+$E^{\sigma}(\treach{C})<\infty$.
 \end{compactenum}

### 2017-06-01 paper/lexicographic.tex
started writing the proof of LexRSM soundness

@@
 $\E[|\vecseq{X}{i}{j}|]<\infty$.
-\item For each $i\in \Nset_0$ there exists $1\leq j \leq n$ such that
+\item For each $i\in \Nset_0$ and each $\omega\in \Omega$ with 
+$\stime(\omega)>i$ there exists $1\leq j 
+\leq n$ such that
 \begin{compactitem}
-	\item $\E[\vecseq{X}{i+1}{j}\mid \genfilt_i]\leq\vecseq{X}{i}{j} - 
-	\eps\cdot \vec{1}_{\stime > i} $; and
+	\item $\E[\vecseq{X}{i+1}{j}\mid 
+	\genfilt_i](\omega)\leq\vecseq{X}{i}{j}(\omega) - 
+	\eps$; and
 	\item for all $1 \leq j' < j$ it holds $\E[\vecseq{X}{i+1}{j'}\mid 
-	\genfilt_i]\leq\vecseq{X}{i}{j'}$.
+	\genfilt_i](\omega)\leq\vecseq{X}{i}{j'}(\omega)$.
 \end{compactitem}
+\item For each $\omega \in \Omega$ and $i\in \Nset_0$ it holds 
+$\stime(\omega)>i 
+\Rightarrow \vec{X}_i (\omega)\geq \vec{0}$ (componentwise).
 \end{compactenum}
 \end{definition}
+
+\begin{theorem}
+Let $(\Omega,\genfilt,\probm)$ be a probability space, 
+$\{\genfilt_i\}_{i=0}^{\infty}$ a filtration of $\genfilt$, and $\stime$ a 
+stopping 
+time w.r.t. that filtration. If there exists an $\eps$-LexRSM for $\stime$, 
+then $\probm(\stime<\infty)=1$.
+\end{theorem}
+\begin{proof}
+The proofs proceeds by contradiction, i.e. we assume that an $\eps$-LexRSM for 
+$\stime$ exists and $\probm(\stime=\infty)>0$. For succinctness we denote the 
+set $\{\omega\mid \stime(\omega)=\infty\}$ by $\genRunSet_{\infty}$.
+
+For $\omega\in \Omega$ we define the level of $\omega$ at step $i$ to be the 
+number $\levelrank{\omega}{i}$ equal to either $0$, if $\stime(\omega)\leq i$, 
+or otherwise to the smallest $1\leq j \leq n$ such that 
+$\E[\vecseq{X}{i+1}{j}\mid 
+\genfilt_i](\omega)\leq\vecseq{X}{i}{j}(\omega)-\eps$. By the definition of 
+$\eps$-LexRSM the value $\levelrank{\omega}{i}$ is well-defined for all 
+$\omega$. 
+%\[
+%\levelrank{\omega}{i} = \begin{dcases}
+%0 & \text{if $\stime(\omega)\leq i$} \\
+% 
+%\end{dcases}
+%\]
+\end{proof}

### 2017-06-01 implementation/Parser.cpp
Equations

@@ -482,7 +482,7 @@ void node::proc_literal(bool negate){
 	}
 	else{
 		constant = "<=";
-		for(int i=1;i<=nVariables;i++){
+		for(int i=0;i<=nVariables;i++){
 			children[0]->expression[i] = children[1]->expression[i]-children[0]->expression[i];
 		}
 	}

### 2017-06-01 implementation/main.cpp

@@ -1,4 +1,5 @@
 // Do I have to add a self loop on the end CFG location, because then the epsilon equation creates issues
+// end node will always be 2
 #include <iostream>
 #include <algorithm>
 #include <cstdio>
@@ -21,20 +22,88 @@ using namespace std;
 #define A(i,j) (top->affexpr->children[i]->children[0]->expression[j+1])
 #define b(i) (-1.0*top->affexpr->children[i]->children[0]->expression[0])
 
+stringstream buffer;
+
 struct cond{
-	int toChange;
-	node* change;
-	int src;
-	int dest1;
-	int dest2;
-	double probability;
+	bool strict;
+	vector<string> c;
+	string negative_d; //negative component of d
 	cond(int toChange,node* change,int src,int dest1,int dest2 = -1,double probability = -1.0){
-		this->toChange = toChange;
-		this->change = change;
-		this->src = src;
-		this->dest1 = dest1;
-		this->dest2 = dest2;
-		this->probability = probability;
+		strict = false;
+		c.resize(nVariables);
+		if(dest2==-1){
+			if(change==NULL){
+				buffer.clear();
+				buffer.str(string());
+				buffer<<"eps-f_"<<src<<"_0+f_"<<dest1<<"_0";
+				negative_d = buffer.str();
+				for(int i=0;i<nVariables;i++){
+					buffer.clear();
+					buffer.str(string());
+					buffer<<"f_"<<dest1<<"_"<<i+1<<"-f_"<<src<<"_"<<i+1;
+					c[i] = buffer.str();
+				}
+			}
+			else{
+				buffer.clear();
+				buffer.str(string());
+				buffer<<"eps-f_"<<src<<"_0+f_"<<dest1<<"_0";
+				if(change->expression[0]<0.0){
+					buffer<<change->expression[0]<<"f_"<<dest1<<"_"<<toChange;
+				}
+				else if(change->expression[0]>0.0){
+					buffer<<"+"<<change->expression[0]<<"f_"<<dest1<<"_"<<toChange;
+				}
+				negative_d = buffer.str();
+				for(int i = 0;i<nVariables;i++){
+					buffer.clear();
+					buffer.str(string());
+					if((i+1)!=toChange){
+						buffer<<"f_"<<dest1<<"_"<<i+1;
+						if(change->expression[i]>0.0){
+							buffer<<"+"<<change->expression[i]<<"f_"<<dest1+"_"<<toChange;
+						}
[…]

### 2017-06-01 implementation/Parser.cpp
debugged sign error

@@ -92,16 +92,14 @@ node* negation(node* tonegate){
 			int ANDs = tonegate->children[i]->children.size();
 			for(int j = 0;j<ANDs;++j){
 				node* ch = new node("literal");
-				if(tonegate->children[i]->children[j]->constant == ">="){
-					ch->constant = "<=";
-					ch->children = tonegate->children[i]->children[j]->children;
-				}
-				else if(tonegate->children[i]->children[j]->constant == "<="){
-					ch->constant = ">=";
-					ch->children = tonegate->children[i]->children[j]->children;
-				}
-				else{
-					cerr<<"Something wrong with a literal"<<endl;
+				ch->constant = "<=";
+				ch->children.resize(1);
+				ch->children[0] = new node("expr");
+				ch->children[0]->constant = "expression";
+				ch->children[0]->expression.resize(nVariables+1);
+				for(int k=0;k<=nVariables;++k){
+					// cout<<tonegate->children[i]->children[j]->children[0]->expression[k]<<endl;
+					ch->children[0]->expression[k] = -1.0*tonegate->children[i]->children[j]->children[0]->expression[k];
 				}
 				vcopy(after_multiplication,before_multiplication);
 				for(int k = 0;k<before_size;++k){
@@ -189,8 +187,8 @@ void node::proc_stmt(int s,int l){
 		label_map[mid] = new CFG_location("det",mid);
 		node* temporary_node;
 		temporary_node = negation(children[0]);
-		CFG_edge temporary_edge1(label_map[l],-1,NULL,children[0]);
-		CFG_edge temporary_edge2(label_map[mid],-1,NULL,temporary_node);
+		CFG_edge temporary_edge1(label_map[mid],-1,NULL,children[0]);
+		CFG_edge temporary_edge2(label_map[l],-1,NULL,temporary_node);
 		label_map[s]->edges.pb(temporary_edge1);
 		label_map[s]->edges.pb(temporary_edge2);
 		// cout<<"Adding edge from "<<s<<"to"<<l<<endl;
@@ -418,7 +416,7 @@ void node::form_vector(int begin,int end,bool negate){ //Note that this shadows 
 	}
 	if(plusminus!=-1){
 		analyse_expr(begin,plusminus,negate);
-		form_vector(plusminus+1,end,program[plusminus]=='+');
+		form_vector(plusminus+1,end,program[plusminus]=='-');
 	}
 	else{
 		analyse_expr(begin,end,negate);

### 2017-06-01 implementation/main.cpp

@@ -13,14 +13,14 @@
 #include <deque>
 #include <iomanip>
 #include <fstream>
-#include <stack>
+#include <queue>
 #include "Parser.h"
 using namespace std;
 #define MAXL 100000 //Maximum length of the program
 #define part(x,a,b) (x.substr((a),((b)-(a))))
 #define pb push_back
-#define A(i,j) (top->affexpr->children[i]->children[0]->expression[j+1])
-#define b(i) (-1.0*top->affexpr->children[i]->children[0]->expression[0])
+#define A(i,j) (front->affexpr->children[i]->children[0]->expression[j+1])
+#define b(i) (-1.0*front->affexpr->children[i]->children[0]->expression[0])
 
 stringstream buffer;
 
@@ -45,6 +45,9 @@ struct cond{
 				}
 			}
 			else{
+				cout<<"tochange is "<<toChange<<"change is"<<endl;
+				change->print();
+				cout<<endl;
 				buffer.clear();
 				buffer.str(string());
 				buffer<<"eps-f_"<<src<<"_0+f_"<<dest1<<"_0";
@@ -120,7 +123,7 @@ struct equation{
 	}
 };
 
-stack<equation*> equations;
+queue<equation*> equations;
 
 node* and_node(node* one,node* two){
 	if(one->type!="bexpr" or two->type!="bexpr"){
@@ -145,14 +148,6 @@ node* and_node(node* one,node* two){
 }
 
 void generate_equations(){ //Would use the ofstream file to write the equations into it later
-	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
-		// Invariant implies the ranking function to be positive
-		int nEquations = it->second->invariant->children.size();
-		for(int i=0;i<nEquations;i++){
-			cond* condition = new cond(it->first);
-			equations.push(new equation(it->second->invariant->children[i],condition));
-		}
-	}
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
 		if(it->second->type=="det"){
 			//Invariant and guard imply the value decrease
@@ -224,6 +219,14 @@ void generate_equations(){ //Would use the ofstream file to write the equations 
 			}
 		}
 	}
+	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
+		// Invariant implies the ranking function to be positive
[…]

### 2017-06-02 paper/lexicographic.tex
added smthing to the proof of LexRSM soundness

@@
 $\stime(\omega)>i 
-\Rightarrow \vec{X}_i (\omega)\geq \vec{0}$ (componentwise).
+\Rightarrow \vec{X}_i (\omega)\geq \vec{0}$ (componentwise). Moreover, there 
+exists a constant $K<0$ such that $\stime(\omega)=i 
+\Rightarrow \vec{X}_i (\omega)\geq (K,\dots,K)$.
 \end{compactenum}
@@
 \begin{proof}
-The proofs proceeds by contradiction, i.e. we assume that an $\eps$-LexRSM for 
+The proof proceeds by contradiction, i.e. we assume that an $\eps$-LexRSM for 
 $\stime$ exists and $\probm(\stime=\infty)>0$. For succinctness we denote the 
@@
 $\eps$-LexRSM the value $\levelrank{\omega}{i}$ is well-defined for all 
-$\omega$. 
+$\omega$. We denote by $\minlev(\omega)$ the smallest $0 \leq j \leq n$ such 
+that $j$ is a level of $\omega$ at infinitely many steps. Note that $\omega \in 
+\genRunSet_{\infty}$ if and only if $\minlev(\omega)\neq 0$. We denote by $M_i$ 
+the set of all $\omega$'s with $\minlev(\omega)=i$.
+
+Throughout the proof we use several times the following fundamental fact: if 
+$\probm(A)>0$ for some set $A$ and $A=A_1\cup A_2 \cup A_3\cdots$ for some 
+sequence of sets $A_1,A_2,A_3,\dots$, then there exists $i$ such that 
+$\probm(A_i)>0$.
+
+Now since $\genRunSet_{\infty}=M_1,\dots,M_n$, there must be $1\leq j \leq n$ 
+s.t. $\probm(M_j)>0$. For each $\omega\in M_j$ there is the smallest number
+$i_{\omega,j}\in\Nset_0$ such that for all $i\geq i_{\omega,j}$ it holds 
+$\levelrank{\omega}{i}\geq 
+j$. Denote by $S_{j,i}$ the set of all $\omega$'s in $M_j$ s.t. 
+$i_{\omega,j}=i$. 
+Since $M_j = M_{j,1} \cup M_{j,2} \cup M_{j,3} \cup \cdots$, there is $i\in 
+\Nset_0$ 
+s.t. 
+$\probm(M_{j,i})>0$. Continuing on the same note, for each $B\in \Nset$ we 
+denote by $M_{j,i}^{B}$ the set off all $\omega$'s in $M_{j,i}$ s.t. 
+$\vecseq{X}{i}{j}(\omega)\leq B$. Since $M_{j,i}= M_{j,i}^{1} \cup M_{j,i}^{2} 
+\cup M_{j,i}^{3} \cup \cdots  $, there is $B\in \Nset$ s.t. 
+$\probm(M_{j,i}^{B})>0$.
+
+%\begin{fact}
+
+%\end{fact}
+
 %\[

### 2017-06-02 paper/lexicographic.tex
expanded the proof of LexRSM soundness

@@
 \cup M_{j,i}^{3} \cup \cdots  $, there is $B\in \Nset$ s.t. 
-$\probm(M_{j,i}^{B})>0$.
+$\probm(M_{j,i}^{B})>0$. Denote $p=\probm(M_{j,i}^{B})$.
+
+Let $D$ be the set of all $\omega\in \Omega$ such that 
+$\vecseq{X}{i}{j}(\omega)\leq B$. Note that $M_{j,i}^{B} \subseteq D$ and $D\in 
+\genfilt_i$ (and thus also $D\in \genfilt_{i'}$ for all $i'\geq i$). 
+Define a stopping time $\tilde{\stime}$ as follows: for all $\omega \in \Omega$ 
+we put 
+$\tilde{\stime}(\omega)=\inf\{k\in \Nset_0 \mid k\geq i \text{ and }
+\levelrank{\omega}{k} < j \}$.
+
+Define a 
+(one-dimensional) stochastic process $\{\tilde{X}_k\}_{k=0}^{\infty}$ as 
+follows:
+$$
+\tilde{X}_k(\omega) = \begin{dcases}
+0 & \text{if $\omega \not\in D$}\\
+B & \text{if $\omega \in D$ and $k<i$ }\\
+\vecseq{X}{k}{j}(\omega) & \text{ if $k\geq i$ and $\tilde{\stime}(\omega)>i$ 
+}\\
+\vecseq{X}{\tilde{\stime}(\omega)}{j}(\omega) & \text{ if $k\geq i$ and 
+$\tilde{\stime}(\omega)\leq i$ }.
+\end{dcases}
+$$
+
+We prove several properties of the process $\{\tilde{X}_k\}_{k=0}^{\infty}$. 
+First, for all $k\geq 0$, $\tilde{X}_k(\omega)\geq K$, where $K<0$ is the 
+number 
+such that $\vec{X}_j(\omega)\geq K$ for all $j$ and $\omega$ (such a number 
+exists from the definition of LexRSM). Second, denote by 
+%\begin{enumerate}
+%\
+%\end{itemize}
+%\end{enumerate}
 

### 2017-06-02 implementation/Parser.cpp
removed junk

@@ -66,8 +66,6 @@ void vcopy(vector<node*> &sink,vector<node*> &tocopy){
 		sink[i]->bracket = tocopy[i]->bracket;
 		sink[i]->expression = tocopy[i]->expression;
 		sink[i]->children = tocopy[i]->children;
-		//Copying everything bluntly, though some of the statements are not required
-		//<flag> Can be optimised if need be
 	}
 }
 
@@ -83,8 +81,6 @@ node* negation(node* tonegate){
 		int ORs = tonegate->children.size();
 		for(int i = 0;i<ORs;++i){
 			//Multply this set of AND to the present collection
-			// before_multiplication.clear();
-			// before_multiplication = negative->children;
 			vcopy(before_multiplication,negative->children);
 			vector<node*> after_multiplication;
 			int before_size = before_multiplication.size();
@@ -98,7 +94,6 @@ node* negation(node* tonegate){
 				ch->children[0]->constant = "expression";
 				ch->children[0]->expression.resize(nVariables+1);
 				for(int k=0;k<=nVariables;++k){
-					// cout<<tonegate->children[i]->children[j]->children[0]->expression[k]<<endl;
 					ch->children[0]->expression[k] = -1.0*tonegate->children[i]->children[j]->children[0]->expression[k];
 				}
 				vcopy(after_multiplication,before_multiplication);
@@ -134,7 +129,6 @@ void node::proc_stmt(int s,int l){
 		constant = "skip";
 		CFG_edge temporary_edge(label_map[l],-1,NULL);
 		label_map[s]->edges.pb(temporary_edge);
-		// cout<<"Adding edge from "<<s<<"to"<<l<<endl;
 		return;
 	}
 	//look for semicolons
@@ -191,9 +185,6 @@ void node::proc_stmt(int s,int l){
 		CFG_edge temporary_edge2(label_map[l],-1,NULL,temporary_node);
 		label_map[s]->edges.pb(temporary_edge1);
 		label_map[s]->edges.pb(temporary_edge2);
-		// cout<<"Adding edge from "<<s<<"to"<<l<<endl;
-		// cout<<"Adding edge from "<<s<<"to"<<mid<<endl;
-
 		children[1] = new node("stmt",firstdo+2,lastod,mid,s);
 		return;
 	}
@@ -257,8 +248,6 @@ void node::proc_stmt(int s,int l){
[…]

### 2017-06-02 implementation/main.cpp

@@ -157,9 +157,7 @@ void generate_equations(){ //Would use the ofstream file to write the equations 
 				// Guard would have been NULL here
 				if(it->second->invariant==NULL){
 					//This should never be the case as this would pose conditions that c==0 and d>0, which are not good
-					// cerr<<"No invariant specified here"<<endl;
-					//This would arise for the end states and skip and other special conditions
-					cout<<it->first<<endl;
+					cerr<<"No invariant specified here"<<endl;
 				}
 				else{
 					// Invariant implies the given condition
@@ -230,15 +228,11 @@ int last_used_lambda = 0;
 
 void print_equations(){
 	cout<<"maximise eps\n\nst\n\neps > 0\neps <= 1"<<endl;
-	// c.resize(nVariables);
 	while(!equations.empty()){
 		equation* front = equations.front();
 		// A(i,j) means affexpr->children[i-1]->children[0]->expression[j] and b(i) translates to -1.0*affexpr->children[i-1]->children[0]->expression[0]
 		// Creating a macro for this
 		//Printing equations
-		// cout<<endl;
-		// front->affexpr->print();
-		// cout<<endl;
 		int size = equations.front()->affexpr->children.size();
 		for(int i=0;i<nVariables;++i){
 			// Each iteration, print out a new equation! :)

### 2017-06-02 implementation/main.cpp
minor change

@@ -227,7 +227,7 @@ void generate_equations(){ //Would use the ofstream file to write the equations 
 int last_used_lambda = 0;
 
 void print_equations(){
-	cout<<"maximise eps\n\nst\n\neps > 0\neps <= 1"<<endl;
+	cout<<"maximise eps\n\nst\n\neps >= 0\neps <= 1"<<endl;
 	while(!equations.empty()){
 		equation* front = equations.front();
 		// A(i,j) means affexpr->children[i-1]->children[0]->expression[j] and b(i) translates to -1.0*affexpr->children[i-1]->children[0]->expression[0]

### 2017-06-07 implementation/Parser.cpp
Epsilons given numbers

@@ -1,694 +1,0 @@
-//Compile with C++11
-//The variable names should be of the form x_i where i is a number and the indexing starts from 1 i.e. the variables in the program are of the form x_1, x_2 ... x_n.
-//Assuming * is for multiplication rather than .
-#include <iostream>
-#include <algorithm>
-#include <cstdio>
-#include <vector>
-#include <cstring>
-#include <sstream>
-#include <cctype>
-#include <set>
-#include <map>
-#include <fstream>
-#include <deque>
-#include <iomanip>
-#include <fstream>
-#include <stdlib.h>
-#include "Parser.h"
-using namespace std;
-#define MAXL 100000 //Maximum length of the program
-#define part(x,a,b) (x.substr((a),((b)-(a))))
-#define pb push_back
-
-string program;
-int nVariables;
-int last_used_label = 0;
-map<int,CFG_location*> label_map;
-
-void skip_spaces(int &begin, int &end){ //skips the spaces in the beginning and the end
-	while(begin<program.size() and isspace(program[begin])){
-		begin++;
-	}
-	while(end>0 and isspace(program[end-1])){
-		end--;
-	}
-}
-
-node::node(string t){
-	bracket = NULL;
-	begin = -1;
-	end = -1;
-	type = t;
-	constant = "";
-}
-
-node::node(string t, int b, int e, int s, int l, bool negate){
-	bracket = NULL;
-	type = t;
-	begin = b;
-	end = e;
-	skip_spaces(begin,end);
-	if(begin>end){
-		cerr<<"Error! begin= "<<b<<" end= "<<e<<endl;
-	}
-	process(s,l,negate);
-}
-
-void vcopy(vector<node*> &sink,vector<node*> &tocopy){
-	sink.clear();
-	int size = tocopy.size();
-	sink.resize(size);
-	for(int i=0;i<size;++i){
-		sink[i] = new node(tocopy[i]->type);
-		sink[i]->constant = tocopy[i]->constant;
-		sink[i]->begin = tocopy[i]->begin;
-		sink[i]->end = tocopy[i]->end;
-		sink[i]->bracket = tocopy[i]->bracket;
-		sink[i]->expression = tocopy[i]->expression;
-		sink[i]->children = tocopy[i]->children;
-	}
-}
-
-node* negation(node* tonegate){
-	if(tonegate->type=="bexpr" and tonegate->constant=="or"){
-		vector<node*> before_multiplication;
-		node* negative;
-		negative = new node("bexpr");
[…]

### 2017-06-07 implementation/Parser.h

@@ -1,68 +1,0 @@
-#include <string>
-using namespace std;
-
-struct node{
-	string type; //The type of entity that this node corresponds to
-	string constant; //The constant factor (subtype) of this node
-	int begin,end; //begin and end point in program
-	vector<node *> children; //children in the parse tree
-	node *bracket; //bexpr written in bracket next to this node (for stmt only)
-	vector<double> expression; //for expressions only
-	//constructor
-	node(string t);
-	node(string t, int b, int e, int s = 0, int l = 0, bool negate = false);
-private:
-//procs
-	void proc_stmt(int s,int l);
-	void proc_assgn(int s,int l);
-	void proc_affexpr();
-	void proc_pvar();
-	void analyse_expr(int begin,int end,bool negate);
-	void proc_expr();
-	void proc_constant();
-	void proc_literal(bool negate);
-	void proc_bexpr();
-	void proc_ndbexpr();
-	void form_vector(int begin,int end,bool negate);
-//main process
-	void process(int s, int l, bool negate);
-public:
-	void print();
-};
-
-struct CFG_location;
-
-struct CFG_edge{
-public:
-	CFG_location* next;
-	int toChange;
-	node* change; // An expr node
-	node* guard; // A bexpr node
-	double probability;
-	CFG_edge();
-	CFG_edge(CFG_location* next,int toChange,node* change,node* guard=NULL,double probability=1.0);
-	void print();
-};
-
-struct CFG_location{
-public:
-	int label;
-	string type; // det ndet and prob
-	vector<CFG_edge> edges;
-	CFG_location(string type,int label);
-	vector<int> ranking_function;
-	node* invariant; // A bexpr node
-	void print();
-};
-
-void skip_spaces(int &begin, int &end);
-void vcopy(vector<node*> &sink,vector<node*> &tocopy);
-node* negation(node* tonegate);
-int find_variables();
-
-extern string program;
-extern int nVariables;
-extern int last_used_label;
-extern map<int,CFG_location*> label_map;
-extern node* id1;
-extern node* root;

### 2017-06-07 implementation/files/script.sh

@@ -1,3 +1,5 @@
+# The script would differ based on the installation directory of cplex
+# This is the path to the executable of CPLEX followed by proper arguments needed
 /opt/ibm/ILOG/CPLEX_Studio_Community127/cplex/bin/x86-64_linux/cplex -c "read files/equations.lp" "optimize" "display solution variables *" > files/EquationsOutput
 # Awk script for trimming EquationsOutput
 awk 'BEGIN {x=0} /^CPLEX>/{x++} x==3{print}' < files/EquationsOutput > EquationsOutput.tmp

### 2017-06-07 implementation/main.cpp

@@ -14,7 +14,7 @@
 #include <iomanip>
 #include <fstream>
 #include <queue>
-#include "Parser.h"
+#include "files/Parser.h"
 using namespace std;
 #define MAXL 100000 //Maximum length of the program
 #define part(x,a,b) (x.substr((a),((b)-(a))))
@@ -28,14 +28,14 @@ struct cond{
 	bool strict;
 	vector<string> c;
 	string negative_d; //negative component of d
-	cond(int toChange,node* change,int src,int dest1,int dest2 = -1,double probability = -1.0){
+	cond(int equation_count,int toChange,node* change,int src,int dest1,int dest2 = -1,double probability = -1.0){
 		strict = false;
 		c.resize(nVariables);
 		if(dest2==-1){
 			if(change==NULL){
 				buffer.clear();
 				buffer.str(string());
-				buffer<<"eps-f_"<<src<<"_0+f_"<<dest1<<"_0";
+				buffer<<"eps"<<equation_count<<"-f_"<<src<<"_0+f_"<<dest1<<"_0";
 				negative_d = buffer.str();
 				for(int i=0;i<nVariables;i++){
 					buffer.clear();
@@ -47,7 +47,7 @@ struct cond{
 			else{
 				buffer.clear();
 				buffer.str(string());
-				buffer<<"eps-f_"<<src<<"_0+f_"<<dest1<<"_0";
+				buffer<<"eps"<<equation_count<<"-f_"<<src<<"_0+f_"<<dest1<<"_0";
 				if(change->expression[0]<0.0){
 					buffer<<change->expression[0]<<"f_"<<dest1<<"_"<<toChange;
 				}
@@ -81,7 +81,7 @@ struct cond{
 			// It was a stochastic node, means change would have been NULL
 			buffer.clear();
 			buffer.str(string());
-			buffer<<"eps-f_"<<src<<"_0+"<<probability<<"f_"<<dest1<<"_0+"<<1.0-probability<<"f_"<<dest2<<"_0"; 
+			buffer<<"eps"<<equation_count<<"-f_"<<src<<"_0+"<<probability<<"f_"<<dest1<<"_0+"<<1.0-probability<<"f_"<<dest2<<"_0"; 
 			negative_d = buffer.str();
 			for(int i=0;i<nVariables;i++){
 				buffer.clear();
@@ -120,7 +120,9 @@ struct equation{
 	}
 };
 
-queue<equation*> equations;
+map<int,equation*> equations;
+int equations_counter = 0;
[…]

### 2017-06-07 implementation/main.cpp
Multidimensional adaptation

@@ -297,6 +297,28 @@ void print_equations(ofstream& equationsfile){
 	equationsfile<<"end"<<endl;
 }
 
+bool process_equations_output(){
+	bool to_return = false;
+	ifstream to_process("files/EquationsOutput");
+	string line;
+	getline(to_process,line);
+	if(line=="CPLEX> Variable Name           Solution Value"){
+		string temp1,temp2;
+		while(to_process>>temp1>>temp2){
+			// cout<<temp1<<" "<<temp2<<endl;
+			if(temp1.length()>3){
+				if(temp1.substr(0,3)=="eps"){
+					// cout<<stoi(temp1.substr(3))<<endl;
+					equations.erase(stoi(temp1.substr(3)));
+					to_return = true;
+					//Removing those equations whose epsilon is 1
+				}
+			}
+		}
+	}
+	return to_return;
+}
+
 int main(){
 	int start,end;
 	char input[MAXL];
@@ -350,7 +372,7 @@ int main(){
 	ofstream equationsfile;
 	equationsfile.open("files/equations.lp");
 	for(int loop_counter=0;loop_counter<100;++loop_counter){	
-		cout<<"Iteration"<<loop_counter+1<<endl;
+		cout<<"Iteration"<<loop_counter+1<<"->"<<endl;
 		print_equations(equationsfile);
 		equationsfile.close();
 		// Jugaad for calling cplex from within the code
@@ -358,6 +380,21 @@ int main(){
 			cout<<"Something wrong with the script analysing equations"<<endl;
 		}
 		//Processing the EquationsOutput file
+		bool state = process_equations_output();
+		if(state==false){
+			cout<<"No solution possible"<<endl;
+			break;
+		}
+		else{
+			//Some equation was deleted, some epsilon was 1
+			// cout<<equations.begin()->first<<endl;
+			if(equations.begin()->first>epsilons_used){
+				cout<<"Solution found"<<endl;
+			}
+			else{
+				cout<<"Going into another iteration"<<endl;
+			}
+		}
 		string command;
 		command = "mv files/EquationsOutput files/EquationsOutput" + to_string(loop_counter);
 		system(command.c_str());

### 2017-06-07 implementation/main.cpp
Debugging

@@ -61,10 +61,10 @@ struct cond{
 					if((i+1)!=toChange){
 						buffer<<"f_"<<dest1<<"_"<<i+1;
 						if(change->expression[i]>0.0){
-							buffer<<"+"<<change->expression[i]<<"f_"<<dest1+"_"<<toChange;
+							buffer<<"+"<<change->expression[i]<<"f_"<<dest1<<"_"<<toChange;
 						}
 						else if(change->expression[i]<0.0){
-							buffer<<change->expression[i]<<"f_"<<dest1+"_"<<toChange;
+							buffer<<change->expression[i]<<"f_"<<dest1<<"_"<<toChange;
 						}
 					}
 					else{
@@ -239,13 +239,17 @@ int last_used_lambda = 0;
 void print_equations(ofstream& equationsfile){
 	// <flag> to be changed
 	
-	equationsfile<<"maximize eps1";
-	for(int i=2;i<=epsilons_used;++i){
-		equationsfile<<"+eps"<<i;
+	equationsfile<<"maximize ";
+	for(int i=1;i<=epsilons_used;++i){
+		if(equations.find(i)!=equations.end()){
+			equationsfile<<"+eps"<<i;
+		}
 	}
 	equationsfile<<"\n\nst\n\n";
 	for(int i=1;i<=epsilons_used;++i){
-		equationsfile<<"eps"<<i<<" >= 0\neps"<<i<<" <= 1"<<endl;
+		if(equations.find(i)!=equations.end()){
+			equationsfile<<"eps"<<i<<" >= 0\neps"<<i<<" <= 1"<<endl;
+		}
 	}
 	
 
@@ -370,9 +374,9 @@ int main(){
 	// }
 	generate_equations();
 	ofstream equationsfile;
-	equationsfile.open("files/equations.lp");
 	for(int loop_counter=0;loop_counter<100;++loop_counter){	
 		cout<<"Iteration"<<loop_counter+1<<"->"<<endl;
+		equationsfile.open("files/equations.lp");
 		print_equations(equationsfile);
 		equationsfile.close();
 		// Jugaad for calling cplex from within the code
@@ -399,7 +403,6 @@ int main(){
 		command = "mv files/EquationsOutput files/EquationsOutput" + to_string(loop_counter);
 		system(command.c_str());
 		// temporarily for just one iteration
-		break;
 	}
 	return 0;
 }

### 2017-06-08 implementation/main.cpp
Debugged assignment case C's

@@ -45,6 +45,9 @@ struct cond{
 				}
 			}
 			else{
+				// cout<<"change is : src"<<src<<" dest:"<<dest1<<" toChange:"<<toChange<<" changed to:";
+				// change->print();
+				// cout<<endl;
 				buffer.clear();
 				buffer.str(string());
 				buffer<<"eps"<<equation_count<<"-f_"<<src<<"_0+f_"<<dest1<<"_0";
@@ -60,20 +63,23 @@ struct cond{
 					buffer.str(string());
 					if((i+1)!=toChange){
 						buffer<<"f_"<<dest1<<"_"<<i+1;
-						if(change->expression[i]>0.0){
-							buffer<<"+"<<change->expression[i]<<"f_"<<dest1<<"_"<<toChange;
+						if(change->expression[i+1]>0.0){
+							buffer<<"+"<<change->expression[i+1]<<"f_"<<dest1<<"_"<<toChange;
 						}
-						else if(change->expression[i]<0.0){
-							buffer<<change->expression[i]<<"f_"<<dest1<<"_"<<toChange;
+						else if(change->expression[i+1]<0.0){
+							buffer<<change->expression[i+1]<<"f_"<<dest1<<"_"<<toChange;
 						}
 					}
 					else{
-						if(change->expression[i]!=0.0){
-							buffer<<change->expression[i]<<"f_"<<dest1<<"_"<<toChange;
+						if(change->expression[i+1]!=0.0){
+							buffer<<change->expression[i+1]<<"f_"<<dest1<<"_"<<toChange;
 						}
+						// change->print();
+						// cout<<""<<endl;
 					}
 					buffer<<"-f_"<<src<<"_"<<i+1;
 					c[i] = buffer.str();
+					// cout<<c[i]<<endl;
 				}
 			}
 		}
@@ -387,13 +393,20 @@ int main(){
 		bool state = process_equations_output();
 		if(state==false){
 			cout<<"No solution possible"<<endl;
+			string command;
+			command = "mv files/EquationsOutput files/EquationsOutput" + to_string(loop_counter);
+			system(command.c_str());
 			break;
 		}
 		else{
 			//Some equation was deleted, some epsilon was 1
-			// cout<<equations.begin()->first<<endl;
+			// cout<<equations.begin()->first<<" "<<epsilons_used<<endl;
 			if(equations.begin()->first>epsilons_used){
 				cout<<"Solution found"<<endl;
+				string command;
[…]

### 2017-06-09 paper/prelims.tex
merging

--- 2017-05-29 paper/prelims.tex
@@
 $(\Omega,\mathcal{F},\probm)$, where $\Omega$ is a non-empty set (so called
-\emph{sample space}), $\mathcal{F}$ is a \emph{sigma-algebra} of measurable 
-sets over $\Omega$,
+\emph{sample space}), $\mathcal{F}$ is a \emph{sigma-algebra} over $\Omega$,
 i.e. a collection of subsets of $\Omega$ that contains the empty set
@@
 A \emph{probabilistic control flow graph (pCFG)} is a tuple
-$\pCFG=(\locs,\pvars,\locinit,\vecinit,\transitions,\updates,\probdist,\guards)$,
+$\pCFG=(\locs,\pvars,\locinit,\vecinit,\transitions,\probdist,\guards)$,
 where
 \begin{compactitem}
-\item $\locs$ is a finite set of \emph{locations} partitioned into four 
+\item $\locs$ is a finite set of \emph{locations} partitioned into three 
 pairwise
-disjoint subsets  $\locs_N$, $\locs_P$, $\locs_D$, and $\locs_A$ of 
-non-deterministic, 
-probabilistic, deterministic, and assignment locations;
+disjoint subsets  $\locs_N$, $\locs_P$, and $\locs_D$ of non-deterministic, 
+probabilistic, and deterministic locations;
 \item $\pvars=\{x_1,\dots,x_{|\pvars|}\}$ is a finite set of \emph{program 
@@
 % POKRACOVAT ODSUD
-\item $\transitions\subseteq\locs\times\locs$ is a transition relation'
-%whose members are tuples of
-%the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
-%program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target 
-%variable index}, and $u$ 
-%is an 
-%\emph{update element}, which can 
-%be one of the following mathematical objects: %\begin{itemize}
-%%\item 
-%(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
-%%\item 
-%(b)~a distribution $d\in \mathcal{D}$; or
-%%%\item 
-%(c)~a set $R\subseteq \Rset$.
-%\end{itemize}
-\item $\updates$ is a function assigning to each transition outgoing from an 
-assignment location a tuple $(i,\up)$, where $1\leq i \leq |\pvars|$ is a 
-\emph{target 
-	variable index} and $\up$ 
+\item $\transitions$ is a transition relation, whose members are tuples of
+the form $(\ell,i,\up,\ell')$, where $\ell$ and $\ell'$ are source and target
+program locations, respectively, $1\leq i \leq |\pvars|$ is a \emph{target variable index}, and $u$ 
 is an 
@@
 (b)~a distribution $d\in \mathcal{D}$; or
-%\item 
+%%\item 
 (c)~a set $R\subseteq \Rset$.
@@
 probability distributions, where each $\prob_{\ell}$ is a discrete probability
-distribution on the set of all transitions outgoing from~$\ell$;
+distribution on the set of all transitions outgoing from~$\ell$.
 \item $\guards$ is a function assigning a propositionally linear predicate
@@
 	line that have positive probability under $d$.} satisfies 
-$\support(d)\subseteq\exsat{\varphi_d}$. Finally, we assume that each 
-assignment location has at most (and thus exactly) one outgoing transition.
-%Finally, we assume that for each 
-%transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
-%deterministic. This is just a technical assumption yielding no loss of 
-%generality, and it somewhat simplifies notation.
+$\support(d)\subseteq\exsat{\varphi_d}$. Finally, we assume that for each 
+transition $(\loc,j,u,\loc')$ such that $u$ is a set the location $\loc$ is 
+deterministic. This is just a technical assumption yielding no loss of 
+generality, and it somewhat simplifies notation.
 %And we assume that each coordinate of $\mathcal{D}$ represents an integrable
@@
 $(\ell,\vec{x})$ if $\ell$ is the source location of $\tau$ and in addition,
-${\vec{x}}\models G(\tau)$ provided that $\ell$ is deterministic. 
-%A 
-%configuration 
-%$(\loc,\vec{x})$ is non-deterministic/probabilistic/deterministic/assignment 
-%one if 
-%$\loc$ is 
-%non-deterministic/probabilistic/deterministic/assignment one, respectively.
+${\vec{x}}\models G(\tau)$ provided that $\ell$ is deterministic. A 
+configuration 
+$(\loc,\vec{x})$ is non-deterministic/probabilistic/deterministic if $\loc$ is 
+non-deterministic/probabilistic/deterministic, respectively.
 
 \smallskip\noindent{\em Executions and reachable configurations.}
-We say that a configuration $(\loc',\vec{x}')$ is a \emph{successor} of a 
-configuration $(\loc,\vec{x})$ if 
-%A \emph{finite path} (or
-%\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
-%configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
-%%$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
-%for each
-%$0 \leq i < k$ 
-there is a transition  $\tau=(\loc,\loc')$ enabled in
-$(\loc,\vec{x})$ 
-%such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
-%satisfies one of the following:
-and $\vec{x}'$ satisfies the following:
-\begin{compactitem}
-\item if $\loc$ is not an assignment location, then $\vec{x}'=\vec{x}$;
-\item if $\loc$ is an assignment location with $\updates(\tau)=(j,\up)$, then 
-$\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ satisfies one of the 
-following 
-depending on the type of $\up$:
-\begin{compactitem}
-\item if $\up$ is an affine function, then $a=\up(\vec{x})$;
-\item if $\up$ is an integrable\footnote{A distribution on some numerical 
-domain 
-is integrable if its expected value exists and is finite. In particular, each 
-Dirac distribution is integrable.} distribution $d$, then 
-$a\in \support(d)$;
-\item if $\up$ is a set, then $a$ is some element of $\up$.
-\end{compactitem}
-\end{compactitem}
 A \emph{finite path} (or
-\emph{execution fragment}) of length $k$ in $\pCFG$ is a finite sequence of
+\emph{execution fragment}) in $\pCFG$ is a finite sequence of
 configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
@@
 for each
-$0 \leq i < k$ the configuration $(\loc_{i+1},\vec{x}_{i+1})$ is a successor of 
-$(\loc_i,\vec{x}_i)$.
+$0 \leq i < k$ there is a transition  $(\ell_i,j,\up,\ell_{i+1})$ enabled in
+$(\ell_i,\vec{x}_i)$ such that $\vec{x}_{i+1}=\vec{x}_i(j\leftarrow a)$ where $a$ 
+satisfies one of the following:
+\begin{compactitem}
+\item $\up$ is a function $f\colon\Rset^{|X|}\rightarrow\Rset$ and
+$a=f(\vec{x}_{i})$;
+\item $\up$ is an integrable\footnote{A distribution on some numerical domain 
+is integrable if its expected value exists and is finite. In particular, each 
+Dirac distribution is integrable.} distribution $d$ and 
+$a\in \support(d)$; or
+\item $\up$ is a set and $a\in \up$.
+\end{compactitem}
 A \emph{run} (or \emph{execution}) in
@@
 if there is a finite path starting in $(\locinit,\vecinit)$ that ends in
-$(\loc,\vec{x})$. We denote by $\Conf_\pCFG,\Fpath_\pCFG$ and $\Run_\pCFG$ the 
-sets of all configurations, finite paths and runs in $\pCFG$, respectively, 
-dropping the index $\pCFG$ when known from the context.
+$(\loc,\vec{x})$.
 
@@
 probability measure over the set of all its runs. Before this can be
-done, non-determinism in $\pCFG$ needs to be resolved. This is achieved using 
-the 
+done, non-determinism in $\pCFG$ needs to be resolved. This is done using the 
 standard notion of a \emph{scheduler}.
-
-We call a configuration $(\loc,\vec{x})$ non-deterministic if $\loc$ is a 
-non-deterministic location or if $\loc$ is an assignment location with a 
-non-deterministic assignment (i.e., the only transition $\tau$ outgoing from 
-$\loc$ has $\updates(\tau)=(j,\up)$ where $\up$ is a set).
 
@@
 \label{def:schedulers}
-A scheduler in a pCFG $\pCFG$ is a function $\sigma$
+A scheduler in an pCFG $\pCFG$ is a tuple $\sigma=(\sigmat,\sigmaa)$,
 %(the subscripts stand for 'transition' and 'assignment')
-%\begin{compactitem}
- assigning 
+where
+\begin{compactitem}
+\item $\sigmat$ (here '$t$' stands for 'transition') is a function assigning 
 to every finite path that ends in a non-deterministic configuration 
-$(\loc,\vec{x})$ a probability distribution on successor configurations of 
-$(\loc,\vec{x})$.
-
-%\end{compactitem}
+$(\loc,\vec{x})$ a probability distribution on transitions outgoing from 
+$\loc$; 
+and
+\item $\sigmaa$ (here '$a$' stands for 'assignment') is a function which takes 
+as an argument a finite path ending in a deterministic configuration in which 
+some transition 
+$(\loc,j,u,\loc')$ with $u$ being a set is enabled, and for such a path it 
+returns a probability distribution on $u$.
+\end{compactitem}
 %function
@@
 \end{definition}
-
-\smallskip\noindent{\em Measurable schedulers.}
-Note that schedulers can be viewed as partial functions from the set $\Fpath$ 
-to the set of probability distributions over the set $\Conf$.
-Since we deal with programs operating over real-valued variables, both $\Fpath$ 
-and $\Conf$ can be uncountable sets. Hence, we impose an 
-additional \emph{measurability} condition on schedulers, so as to ensure that 
-the semantics of probabilistic non-deterministic programs is defined in a 
-mathematically 
-sound way. First we need to clarify what are measurable sets of configurations 
-and histories. We define a sigma-algebra $\mathcal{F}_{\Conf}$ of measurable 
-sets of configurations to be the sigma-algebra 
-over $\Conf$ 
-generated\footnote{In general, it is known that for each set $\Omega$ and each 
-collection of its subsets $F\subseteq 2^{\Omega}$ there exists at least one 
-sigma-algebra $\mathcal{F}$ s.t. $F\subseteq \mathcal{F}$ and the intersection 
-of all such sigma-algebras is again a sigma algebra --  so called sigma-algebra 
-generated by $F$~\cite{Billingsley:book}.} by 
-all sets of the form $\{\loc\}\times B$, where $\loc$ is a location of $\pCFG$ 
-and  
-$B$ is a Borel-measurable subset 
-of $\Rset^{|\pvars|}$. Next, the set of finite paths $\Fpath$ can be viewed as 
-a subset of $\Conf\cup \Conf\times\Conf \cup \Conf\times\Conf\times\Conf \cup 
-\dots$. Hence, we define the sigma-algebra $\mathcal{H}$ of measurable sets of 
-finite paths 
-to be the sigma algebra generated by all sets of the form $\confset_1\times 
-\confset_2 \times 
-\cdots \times \confset_k \subseteq \Fpath$ such that $k\in \Nset$ and 
-$\confset_i \in 
-\mathcal{F}_{\Conf}$ for 
-all $1\leq i \leq k$. Now we can define the measurability of schedulers. Recall 
-that for each finite path $\genpath$ ending in a non-deterministic 
-configuration 
-we have that $\sigma(\genpath)$ is a probability distribution on $\Conf$, hence 
-for each measurable set of configurations $\confset$ we denote by 
-$\sigma(\genpath)(\confset)$ the probability that the random draw from 
-distribution $\sigma(\genpath)$ selects an element of $\confset$. We say that a 
-scheduler $\sigma$ is measurable if for each $\confset \in \mathcal{F}_{\Conf}$ 
-and each $p\in[0,1]$ the set $\{\pi \in \Fpath \mid 
-\sigma(\genpath)(\confset)\leq p\}$ belongs to $\mathcal{H}$, i.e., it is a 
-measurable set of paths.
-
-While the definition of a measurable scheduler might seem somewhat technical, 
-it is natural from measure-theoretic point of view and analogous definitions 
-inevitably emerge in works dealing with systems that exhibit both probabilistic 
-and non-deterministic behaviour over a continuous state 
-space~\cite{NSK:CTMDP-delayed,NK:CTMDP-bisimulation}. In particular, if all the 
-variables in a program range over a discrete subset (such as the integers), 
-then each scheduler in the associated pCFG is measurable.
-%Formally, let 
-%$\Borel(\Rset^{|\pvars|})$ be the standard sigma-algebra of Borel-measurable 
-%subsets of $\Rset^{|\pvars|}$. Now let $\genPathSet$ be any set of finite 
-%paths 
-%in 
-%$\pCFG$ and denote by $\proj_j (A)$ the set of all variable valuations that 
-%appear on some path in $A$ in step $j$, i.e. $\proj_j(A) = 
-%\{\vec{y}\in\Rset^{|\pvars|} \mid \exists  
-%(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k) \in A \text{ s.t. } k \geq j \} $
-%A \emph{measurable cylinder} of length 
-%$k$ is a 
-%set of paths $\cylinder_n$ in $\pCFG$ satisfying the following condition:
 
@@
 The translation from probabilistic programs to the corresponding pCFG is 
-standard~\cite{CFNH16:prob-termination}, and the details are presented 
-in~\cite{CNZ16:prob-invariants-arxiv}. We point out that the 
+standard~\cite{CFNH16:prob-termination}. We point out that the 
 construction produces pCFGs with a property that only transitions outgoing from 

### 2017-06-09 paper/invariants-martingales.tex

@@
 
-In this section we recall known methods and constructs for solving the 
-qualitative termination and reachability questions for \APP{s}, namely  
-linear invariants and ranking supermartingales. 
-We also demonstrate that these methods are not sufficient to address the 
-quantitative variants of these questions (i.e., probabilistic termination).
-In order to discuss the necessary concepts, we recall the basics of 
-martingales, which is relevant for both this and subsequent sections. 
+In this section we recall basic results on (one-dimensional) ranking 
+supermartingales in probabilistic programs.
+
+%In this section we recall known methods and constructs for solving the 
+%qualitative termination and reachability questions for \APP{s}, namely  
+%linear invariants and ranking supermartingales. 
+%We also demonstrate that these methods are not sufficient to address the 
+%quantitative variants of these questions (i.e., probabilistic termination).
+%In order to discuss the necessary concepts, we recall the basics of 
+%martingales, which is relevant for both this and subsequent sections. 
 %Since the definitions of this section are from the literature we relegate
@@
 %\vspace{-1em}
-\subsection{Pure Invariants}
+\subsection{Invariants}
 %\vspace{-0.5em}
@@
 whenever $\loc$ is 
-entered. To avoid confusion with stochastic invariants, 
-that we introduce later, we call these standard invariants \emph{pure invariants}.
+entered.
 
 \smallskip
-\begin{definition}[Linear Predicate Map (LPM) and Pure Invariant] We define the 
+\begin{definition}[Linear Predicate Maps (LPMs) and Invariants] We define the 
 following:
@@
 \item
-A \emph{pure linear invariant} (or just a pure invariant) for an \APP{} $P$ is 
+A \emph{linear invariant} (or just an invariant) for an \APP{} $P$ is 
 a linear predicate map $\inv$ for $P$ with
@@
 spaces of the form $(\Omega,\natfilt,\probm^\sigma)$, where $\Omega$ is a 
-set of runs in some $\pCFG$ and $\mathcal{F}$ is (the smallest) sigma-algebra 
+set of runs in some $\pCFG$ and $\natfilt$ is (the smallest) sigma-algebra 
 such that all the functions $\cfg{\sigma}{i}$, where $i\in \Nset_0$ and 

### 2017-06-09 paper/prelims.tex

--- 2017-06-01 paper/prelims.tex
@@
 terminating if under each scheduler $\sigma$ it holds that 
-$\E^{\sigma}(\{\ttime\}) \infty$. In 
+$\E^{\sigma}(\{\ttime\})<  \infty$. In 
 positive termination question for $P$ we aim to prove that $P$ 
@@
 $\sigma$ it holds that
-$E^{\sigma}(\treach{C})<\infty$.
+$\E^{\sigma}(\treach{C})<\infty$.
 \end{compactenum}

### 2017-06-09 paper/lexicographic.tex

@@
 B & \text{if $\omega \in D$ and $k<i$ }\\
-\vecseq{X}{k}{j}(\omega) & \text{ if $k\geq i$ and $\tilde{\stime}(\omega)>i$ 
+\vecseq{X}{k}{j}(\omega) & \text{if $\omega \in D$, $k\geq i$ and 
+$\tilde{\stime}(\omega)>i$ 
 }\\
-\vecseq{X}{\tilde{\stime}(\omega)}{j}(\omega) & \text{ if $k\geq i$ and 
+\vecseq{X}{\tilde{\stime}(\omega)}{j}(\omega) & \text{if $\omega\in D$, $k\geq 
+i$ and 
 $\tilde{\stime}(\omega)\leq i$ }.
@@
 number 
-such that $\vec{X}_j(\omega)\geq K$ for all $j$ and $\omega$ (such a number 
-exists from the definition of LexRSM). Second, denote by 
+such that $\vec{X}_j(\omega)\geq (K,\dots,K)$ for all $j$ and $\omega$ (such a 
+number 
+exists from the definition of LexRSM). Second, for any $k\in \Nset_0$ denote by 
+$\noofdec_k$ the random variable such that $\noofdec_k(\omega) = |\{i'\in 
+\Nset\mid i \leq i' \leq k \text{ and } \levelrank{\omega}{i'}=j\} |$.
+We claim 
+that for each $k\geq i$ 
+it holds 
+\begin{equation}
+\label{eq:lexrsm-soundness-main}
+\E[\tilde{X}_k]\leq B\cdot \probm(D) - \eps\cdot\sum_{\ell=0}^{k-i-1} \probm(D 
+\cap \tilde{\stime}\geq k \cap \noofdec_k 
+\geq 
+\ell).
+\end{equation}
+
+The proof of~\eqref{eq:lexrsm-soundness-main} goes by induction on $k$. For 
+$k=i$ the sum on the right-hand side is over an empty set of indexes, so the 
+inequality immediately follows from the definition of $\tilde{X}_k$. Now assume 
+that~\eqref{eq:lexrsm-soundness-main} holds for some $k\geq i$. 
 %\begin{enumerate}

### 2017-06-09 implementation/files/Parser.cpp
Invariant generation automation left

@@ -577,62 +577,69 @@ int find_variables(){
 	return variables.size();
 }
 
-void node::print(){
+void node::print(ostream& outputfile, string and_string, string or_string, string multiply_string){
 	if(type=="expr"){
+		bool something_printed = false;
 		if(expression[0]!=0.0){
-			cout<<expression[0];
+			outputfile<<expression[0];
+			something_printed = true;
 		}
 		for(int i = 1;i<=nVariables;i++){
 			if(expression[i]!=0.0){
 				if(expression[i]>0.0){
-					cout<<"+"<<expression[i]<<"x_"<<i;
+					if(something_printed==true){
+						outputfile<<"+"<<expression[i]<<multiply_string<<"x_"<<i;
+					}
+					else{
+						outputfile<<expression[i]<<multiply_string<<"x_"<<i;
+					}
+					something_printed = true;
 				}
 				else{
-					cout<<expression[i]<<"x_"<<i;
+					outputfile<<expression[i]<<multiply_string<<"x_"<<i;
+					something_printed = true;
 				}
 			}
 		}
 		return;
 	}
 	else if(type=="literal"){
-		children[0]->print();
-		cout<<constant<<"0";
+		children[0]->print(outputfile,and_string,or_string);
+		outputfile<<constant<<"0";
 		return;
 	}
 	else if(type=="affexpr"){
-		children[0]->print();
+		children[0]->print(outputfile,and_string,or_string);
 		for(int i=1;i<children.size();++i){
-			cout<<" and ";
-			children[i]->print();
+			outputfile<<" "<<and_string<<" ";
+			children[i]->print(outputfile,and_string,or_string);
 		}
 		return;
 	}
 	else if(type=="bexpr"){
-		cout<<"Add: "<<this<<"\t";
-		children[0]->print();
+		children[0]->print(outputfile,and_string,or_string);
 		for(int i=1;i<children.size();++i){
-			cout<<" or ";
-			children[i]->print();
+			outputfile<<" "<<or_string<<" ";
+			children[i]->print(outputfile,and_string,or_string);
 		}
-		cout<<endl<<"--------------------"<<endl;
 		return;
 	}
-	cout<<"Add: "<<this<<"\t";
-	cout<<"Type: "<<type<<"\t";
-	cout<<"Range: ["<<begin<<", "<<end<<")\t";
[…]

### 2017-06-09 implementation/main.cpp

@@ -152,7 +152,7 @@ node* and_node(node* one,node* two){
 	return node_and;
 }
 
-void generate_equations(){ //Would use the ofstream file to write the equations into it later
+void generate_equations(){ //Would use the ostream file to write the equations into it later
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
 		if(it->second->type=="det"){
 			//Invariant and guard imply the value decrease
@@ -242,7 +242,7 @@ void generate_equations(){ //Would use the ofstream file to write the equations 
 
 int last_used_lambda = 0;
 
-void print_equations(ofstream& equationsfile){
+void print_equations(ostream& equationsfile){
 	// <flag> to be changed
 	
 	equationsfile<<"maximize ";
@@ -329,6 +329,49 @@ bool process_equations_output(){
 	return to_return;
 }
 
+void print_fast(ostream& fastfile){
+	fastfile<<"model main{\n\n";
+	fastfile<<"\tvar x_1";
+	for(int i=2;i<=nVariables;++i){
+		fastfile<<", x_"<<i;
+	}
+	fastfile<<";\n\n";
+
+	fastfile<<"\tstates state_1";
+	for(int i=2;i<=last_used_label;++i){
+		fastfile<<", state_"<<i;
+	}
+	fastfile<<";\n\n";
+
+	for(int i=1;i<=last_used_label;++i){
+		CFG_location* state = label_map[i];
+		for(int j=0;j<state->edges.size();++j){
+			fastfile<<"\ttransition t_"<<i<<"_"<<j<<" := {\n";
+			fastfile<<"\t\tfrom\t:= state_"<<i<<";\n";
+			fastfile<<"\t\tto\t:= state_"<<state->edges[j].next->label<<";\n";
+			fastfile<<"\t\tguard\t:= ";
+			if(state->edges[j].guard==NULL){
+				fastfile<<"true";
+			}
+			else{
+				state->edges[j].guard->print(fastfile,"&&","||");
+			}
+			fastfile<<";\n";
+			fastfile<<"\t\taction\t:= ";
+			if(state->edges[j].change!=NULL){
+				fastfile<<"x_"<<state->edges[j].toChange<<"' = ";
+				state->edges[j].change->print(fastfile);
+			}
+			//Print action here
+			fastfile<<";\n\t};\n\n";
+		}
+	}
+
[…]

### 2017-06-09 implementation/files/invariant_script.sh
Progress

@@ -1,0 +1,4 @@
+./files/aspicV3.4 files/aspic.fast > files/InvariantOutput
+awk 'BEGIN {x=0} /^"*** Results :"/{x=1} x==1{print}' < files/InvariantOutput > InvariantOutput.tmp
+mv InvariantOutput.tmp files/InvariantOutput
+rm aspic.log

### 2017-06-09 implementation/files/script.sh

@@ -3,6 +3,5 @@
 /opt/ibm/ILOG/CPLEX_Studio_Community127/cplex/bin/x86-64_linux/cplex -c "read files/equations.lp" "optimize" "display solution variables *" > files/EquationsOutput
 # Awk script for trimming EquationsOutput
 awk 'BEGIN {x=0} /^CPLEX>/{x++} x==3{print}' < files/EquationsOutput > EquationsOutput.tmp
-cp EquationsOutput.tmp files/EquationsOutput
-rm EquationsOutput.tmp
+mv EquationsOutput.tmp files/EquationsOutput
 rm cplex.log

### 2017-06-09 implementation/main.cpp

@@ -1,5 +1,10 @@
 // Do I have to add a self loop on the end CFG location, because then the epsilon equation creates issues
 // end node will always be 2
+/*
+Comments:
+	Ask Petr	When invariant is NULL the program gives wrong result, because it ignores the implication in that case
+https://tapas.labri.fr/wp/wp-content/uploads/2017/02/FAST-manual.pdf
+*/
 #include <iostream>
 #include <algorithm>
 #include <cstdio>
@@ -231,7 +236,10 @@ void generate_equations(){ //Would use the ostream file to write the equations i
 	epsilons_used = equations_counter;
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
 		// Invariant implies the ranking function to be positive
-		int nEquations = it->second->invariant->children.size();
+		int nEquations = 0;
+		if(it->second->invariant!=NULL){
+			nEquations = it->second->invariant->children.size();
+		}
 		for(int i=0;i<nEquations;i++){
 			cond* condition = new cond(it->first);
 			equations[++equations_counter] = new equation(it->second->invariant->children[i],condition);
@@ -410,61 +418,68 @@ int main(){
 	}
 	skip_spaces(begin,endprog);
 	root=new node("stmt",begin,endprog,start,end);
+	// cout<<last_used_label<<endl;
 	ofstream fastfile;
 	fastfile.open("files/aspic.fast");
 	ostream* outfast = &fastfile;
 	print_fast(*outfast);
 	fastfile.close();
-	// cout<<"Input Code:"<<endl;
-	// cout<<program<<endl;
-	// cout<<"Parse Tree:"<<endl;
-	// root->print();
-	// cout<<"CFG:"<<endl;
-	// for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
-	// 	cout<<"------------------------"<<endl;
-	// 	cout<<"Node "<<it->first<<endl;
-	// 	it->second->print();
-	// 	// cout<<it->second->label<<endl;
-	// }
-	generate_equations();
-	ofstream equationsfile;
-	for(int loop_counter=0;loop_counter<100;++loop_counter){	
-		cout<<"Iteration"<<loop_counter+1<<"->"<<endl;
-		equationsfile.open("files/equations.lp");
-		ostream* equation_output_file = &equationsfile;
-		print_equations(*equation_output_file);
-		equationsfile.close();
[…]

### 2017-06-11 paper/lexicographic.tex

--- 2017-06-02 paper/lexicographic.tex
@@
 $\{\genfilt_i\}_{i=0}^{\infty}$.
-\item For each $i\in \Nset_0$ and each $1\leq j \leq n$ it holds 
-$\E[|\vecseq{X}{i}{j}|]<\infty$.
+%\item For each $i\in \Nset_0$ and each $1\leq j \leq n$ it holds 
+%$\E[|\vecseq{X}{i}{j}|]<\infty$.
 \item For each $i\in \Nset_0$ and each $\omega\in \Omega$ with 
@@
 \item For each $\omega \in \Omega$ and $i\in \Nset_0$ it holds 
-$\stime(\omega)>i 
-\Rightarrow \vec{X}_i (\omega)\geq \vec{0}$ (componentwise). Moreover, there 
-exists a constant $K<0$ such that $\stime(\omega)=i 
-\Rightarrow \vec{X}_i (\omega)\geq (K,\dots,K)$.
+$\vec{X}_i (\omega)\geq \vec{0}$ (componentwise).
 \end{compactenum}
@@
 $\eps$-LexRSM the value $\levelrank{\omega}{i}$ is well-defined for all 
-$\omega$. We denote by $\minlev(\omega)$ the smallest $0 \leq j \leq n$ such 
+$\omega$ and moreover, the random variable $\levelrank{}{i}$ is $\genfilt_i$-measurable. We denote by $\minlev(\omega)$ the smallest $0 \leq j \leq n$ such 
 that $j$ is a level of $\omega$ at infinitely many steps. Note that $\omega \in 
@@
 
-Now since $\genRunSet_{\infty}=M_1,\dots,M_n$, there must be $1\leq j \leq n$ 
-s.t. $\probm(M_j)>0$. For each $\omega\in M_j$ there is the smallest number
-$i_{\omega,j}\in\Nset_0$ such that for all $i\geq i_{\omega,j}$ it holds 
-$\levelrank{\omega}{i}\geq 
-j$. Denote by $S_{j,i}$ the set of all $\omega$'s in $M_j$ s.t. 
-$i_{\omega,j}=i$. 
-Since $M_j = M_{j,1} \cup M_{j,2} \cup M_{j,3} \cup \cdots$, there is $i\in 
+Now since $\genRunSet_{\infty}=M_1,\dots,M_n$, there must be $1\leq \fixn{j} \leq n$ 
+s.t. $\probm(M_{\fixn{j}})>0$. For each $\omega\in M_{\fixn{j}}$ there is the smallest number
+${i}_{\omega,\fixn{j}}\in\Nset_0$ such that for all $i\geq {i}_{\omega,{j}}$ it holds 
+$\levelrank{\omega}{i}\geq \fixn{j}$. Denote by $S_{\fixn{j},{i}}$ the set of all $\omega$'s in $M_{\fixn{j}}$ s.t. 
+$i_{\omega,\fixn{j}}={i}$. 
+Since $M_{\fixn{j}} = M_{\fixn{j},1} \cup M_{\fixn{j},2} \cup M_{\fixn{j},3} \cup \cdots$, there is $\fixn{i}\in 
 \Nset_0$ 
 s.t. 
-$\probm(M_{j,i})>0$. Continuing on the same note, for each $B\in \Nset$ we 
-denote by $M_{j,i}^{B}$ the set off all $\omega$'s in $M_{j,i}$ s.t. 
-$\vecseq{X}{i}{j}(\omega)\leq B$. Since $M_{j,i}= M_{j,i}^{1} \cup M_{j,i}^{2} 
-\cup M_{j,i}^{3} \cup \cdots  $, there is $B\in \Nset$ s.t. 
-$\probm(M_{j,i}^{B})>0$.
+$\probm(M_{\fixn{j},\fixn{i}})>0$. Continuing on the same note, for each $B\in \Nset$ we 
+denote by $M_{\fixn{j},\fixn{i}}^{B}$ the set off all $\omega$'s in $M_{\fixn{j},\fixn{i}}$ s.t. 
+$\vecseq{X}{\fixn{i}}{\fixn{j}}(\omega)\leq B$. Since $M_{\fixn{j},\fixn{i}}= M_{\fixn{j},\fixn{i}}^{1} \cup M_{\fixn{j},\fixn{i}}^{2} 
+\cup M_{\fixn{j},\fixn{i}}^{3} \cup \cdots  $, there is $\fixn{B}\in \Nset$ s.t. 
+$\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}})>0$. Denote $\fixn{p}=\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}})$.
+
+Let $D$ be the set of all $\omega\in \Omega$ such that 
+$\vecseq{X}{\fixn{i}}{\fixn{j}}(\omega)\leq \fixn{B}$. Note that $M_{\fixn{j},\fixn{i}}^{B} \subseteq D$ and $D\in 
+\genfilt_i$ (and thus also $D\in \genfilt_{i'}$ for all $i'\geq i$). 
+Define a stopping time $U$ w.r.t. filtration $\{\genfilt_i\}_{i=0}^{\infty}$ as follows: for all $\omega \in \Omega$ 
+we put 
+$U(\omega)=\inf\{k\in \Nset_0 \mid k\geq \fixn{i} \text{ and }
+\levelrank{\omega}{k} < \fixn{j} \}$.
+
+Define a 
+(one-dimensional) stochastic process $\{Y_k\}_{k=0}^{\infty}$ as 
+follows:
+$$
+Y_k(\omega) = \begin{dcases}
+0 & \text{if $\omega \not\in D$}\\
+\fixn{B} & \text{if $\omega \in D$ and $k<\fixn{i}$ }\\
+\vecseq{X}{k}{\fixn{j}}(\omega) & \text{if $\omega \in D$, $k\geq \fixn{i}$ and 
+$U(\omega)>k$ 
+}\\
+\vecseq{X}{U(\omega)}{\fixn{j}}(\omega) & \text{if $\omega\in D$, $k\geq 
+\fixn{i}$ and 
+$U(\omega)\leq k$ }.
+\end{dcases}
+$$
+
+We prove several properties of the process $\{Y_k\}_{k=0}^{\infty}$. 
+First, clearly for all $k\geq 0$, $Y_k(\omega)\geq 0$.
+Second, for each $k\geq \fixn{i}$, the variable $Y_k$ is $\genfilt_k$-measurable, as $D\in \genfilt_{\fixn{i}}$, $\{\vecseq{X}{i}{\fixn{j}}(\omega)\}_{i=0}^{\infty}$ is adapted to the filtration $\{\genfilt_i\}_{i=0}^{\infty}$ and $U$ is a stopping time w.r.t.
+this filtration.
+Finally, for any $k\in \Nset_0$ denote by
+$\noofdec_k$ the random variable such that $\noofdec_k(\omega) = |\{i'\in 
+\Nset\mid \fixn{i} \leq i' < k \text{ and } \levelrank{\omega}{i'}=\fixn{j}\} |$.
+We claim 
+that for each $k\geq \fixn{i}$ 
+it holds 
+\begin{equation}
+\label{eq:lexrsm-soundness-main}
+\E[Y_k]\leq \fixn{B}\cdot \probm(D) - \eps\cdot\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D 
+\cap \{U\geq k\} \cap \{\noofdec_k 
+= 
+\ell\}).
+\end{equation}
+
+The proof of~\eqref{eq:lexrsm-soundness-main} goes by induction on $k$. For 
+$k=\fixn{i}$ the sum on the right-hand side equals $0$, so the 
+inequality immediately follows from the definition of $Y_k$. Now assume 
+that~\eqref{eq:lexrsm-soundness-main} holds for some $k\geq \fixn{i}$. We have that 
+\begin{align}
+\label{eq:lexrsm-ind-1}
+\E[Y_{k+1}] &= \underbrace{\E[Y_{k+1}\cdot \indicator{\Omega\setminus D}]}_{=0=\E[Y_k\cdot\indicator{\Omega\setminus D}]} +  \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{U \leq k\}}]}_{=\E[Y_k\cdot\indicator{D\cap \{U\leq k\}}]} + \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{U > k\}}]}_{=\E[\vecseq{X}{k+1}{\fixn{j}}\cdot\indicator{D\cap \{U>k \} }]},
+\end{align}
+where the equality $\E[Y_{k+1}\cdot \indicator{D \cap \{U \leq k\}}] =\E[Y_k\cdot\indicator{D\cap \{U\leq k\}}] $ follows from the fact that $Y_{k+1}(\omega)=Y_k({\omega})=\vecseq{X}{U(\omega)}{\fixn{j}}(\omega)$ for $\omega\in \{U\leq k\}$, and similarly for the last term.
+We prove that
+\begin{equation}
+\label{eq:lexrsm-ind-2}
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \}}] \leq \E[Y_k\cdot \indicator{D\cap \{U > k \}} -\eps\cdot \indicator{D\cap \{U>k\} \cap \{\levelrank{}{k}= \fixn{j}\} }] .
+\end{equation}
+Indeed, it holds
+\begin{equation}
+\label{eq:lexrsm-ind-3}
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \}}] = \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }]  + \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j}\} }],
+\end{equation}
+since $\levelrank{\omega}{k}\geq \fixn{j}$ for all $\omega \in \{U>k\}$. Since the set $D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}$ is $\genfilt_k$-measurable, we get
+\begin{align}
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] &= \E[\E[\vecseq{X}{k+1}{\fixn{j}}\mid \genfilt_k]\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] \label{eq:lexrsm-ind-4}\\
+&\leq \E[(\vecseq{X}{k}{\fixn{j}} - \eps)\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-5}\\
+&=\E[(Y_k - \eps)\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-6},
+\end{align}
+where~\eqref{eq:lexrsm-ind-4} follows from the definition of conditional expectation~\eqref{eq:cond-exp},~\eqref{eq:lexrsm-ind-5} follows from the definition of $\{\levelrank{}{k}=\fixn{j}\}$, and~\eqref{eq:lexrsm-ind-6} holds since $Y_k(\omega)=\vecseq{X}{k}{\fixn{j}}(\omega)$ for $\omega$ with $U(\omega)> k$.
+Almost identical argument shows that
+\begin{equation}
+\label{eq:lexrsm-ind-7}
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j} \} }] \leq \E[Y_k\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j} \}}].
+\end{equation}
+Plugging~\eqref{eq:lexrsm-ind-6} and~\eqref{eq:lexrsm-ind-7} into~\eqref{eq:lexrsm-ind-3} yields~\eqref{eq:lexrsm-ind-2}. Now we can plug~\eqref{eq:lexrsm-ind-2} into~\eqref{eq:lexrsm-ind-1} to get
+\begin{align}
+\E[Y_{k+1}]&\leq \E[Y_k] - \eps\cdot \E[\indicator{D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \} }] = \E[Y_k] - \eps\cdot \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \} ) \nonumber\\
+&\leq \fixn{B}\cdot \probm(D) - \eps\cdot\left(\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D 
+\cap \{U\geq k\} \cap \{\noofdec_k = \ell\})\right) -\eps\cdot \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \}),
+\end{align}
+where the last inequality follows from induction hypothesis. Hence, using $D_{k,\ell}$ as a shorthand for $D 
+\cap \{U\geq k\} \cap \{\noofdec_k = \ell\}$, to prove~\eqref{eq:lexrsm-soundness-main} it remains to show that
+\begin{equation}
+\label{eq:lexrsm-ind-8}
+\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}) + \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \}) = \sum_{\ell=0}^{k+1-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}).
+\end{equation}
+The left-hand side of~\eqref{eq:lexrsm-ind-8} is equal to
+\begin{align}
+&\phantom{+}\;\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell} \cap \{\levelrank{}{k}=\fixn{j}\}) + \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} )\nonumber \\
+&+\sum_{\ell=0}^{k-\fixn{i}}\probm(\underbrace{D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j}\} \cap \{\noofdec_k = \ell\}}_{=D_{k,\ell} \cap \{\levelrank{}{k} = \fixn{j}\}})
+\nonumber \\
+&= \sum_{\ell=0}^{k-\fixn{i}}(\ell+1) \cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k} = \fixn{j}\}) +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} ) \nonumber\\
+&=\sum_{\ell=0}^{k-\fixn{i}} (\ell+1)\cdot \probm{(D_{k+1,\ell+1} \cap \{\levelrank{}{k} = \fixn{j}\})} +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} ) \label{eq:lexrsm-ind-9}\\
+&= \sum_{\ell=1}^{k+1-\fixn{i}} \ell\cdot \probm{(D_{k+1,\ell} \cap \{\levelrank{}{k} = \fixn{j}\})} +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} )
+   \nonumber \\
+&= (k+1-\fixn{i})\cdot \probm{(D_{k+1,k+1-\fixn{i}}\cap\{\levelrank{}{k} =\fixn{j} \} )} + \sum_{\ell=1}^{k-\fixn{i}} \ell \cdot \probm(D_{k+1,\ell})  \nonumber\\
+&= (k+1-\fixn{i})\cdot\probm(D_{k+1,k+1-\fixn{i}}) + \sum_{\ell=1}^{k-\fixn{i}}\ell\cdot\probm(D_{k+1,\ell}) = \text{ right-hand side of~\eqref{eq:lexrsm-ind-8}} \label{eq:lexrsm-ind-11}.
+\end{align}
+The individual steps in the above computation are justified as follows: in~\eqref{eq:lexrsm-ind-9} we use the facts that for all $\omega$'s whose level in step $k$ is $\fixn{j}$ it holds that $\noofdec_k(\omega)+1=\noofdec_{k+1}(\omega)$, and similarly, for $\omega$'s whose level in step $k$ is $>\fixn{j}$ it holds $\noofdec_k(\omega)=\noofdec_{k+1}(\omega)$.
+Moreover, for all $\omega\in D_{k,\ell}$ it holds that $\levelrank{k}{\omega}\geq \fixn{j} \Rightarrow U(\omega)\geq k+1$.
+Finally, in~\eqref{eq:lexrsm-ind-11} we use the fact that all $\omega\in D_{k+1,k+1-\fixn{i}}$ need to have level $\fixn{j}$ in step $k$, since otherwise such an $\omega$ would need to have level $\fixn{j}$ for at least $k+1-\fixn{i}$ times within steps $\{\fixn{i},\fixn{i}+1,\dots,k-1\}$, but there are $k-\fixn{i}$ such steps, a contradiction.
+This concludes the proof of~\eqref{eq:lexrsm-soundness-main}.
+
+Now according to~\eqref{eq:lexrsm-soundness-main} it holds  $\E[Y_k]\leq \fixn{B}\cdot \probm(D) - \eps\cdot\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D 
+\cap \{U\geq k\} \cap \{\noofdec_k 
+= 
+\ell\})$ for all $k\geq \fixn{i}$.
+Let $m=3\fixn{B}\cdot \probm(D)/(\eps\cdot \fixn{p})$.
+For each $\omega\in M_{\fixn{j},\fixn{i}}^{\fixn{B}} $ there exists $k_m(\omega)\geq \fixn{i}$ such that $\noofdec_{k_{m}(\omega)}\geq m$.
+Hence, $M_{\fixn{j},\fixn{i}}^{\fixn{B}} = \bigcup_{k=\fixn{i}}^{\infty} M_{\fixn{j},\fixn{i}}^{\fixn{B}} \cap \{k_{m}(\omega) \leq k\}$ and $\fixn{p}=\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}})=\lim_{k\rightarrow \infty}\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}} \cap \{k_{m} \leq k\})$.
+Thus, there exists $k_0$ such that $\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}} \cap \{\noofdec_{k_0}\geq m\}) \geq \fixn{p}/2$.
+Denote $\fixn{M}= M_{\fixn{j},\fixn{i}}^{\fixn{B}} \cap \{\noofdec_{k_0}\geq m\}$.
+Clearly $\fixn{M} \subseteq D\cap \{U\geq k_0 \} \cap \{\noofdec_{k_0}\geq m \}$.
+From \eqref{eq:lexrsm-soundness-main} it follows that
+\begin{align*}
+\E[Y_{k_0}]&\leq \fixn{B}\cdot \probm(D) - \eps\cdot\sum_{\ell=0}^{k_0-\fixn{i}} \ell\cdot\probm(D 
+\cap \{U\geq k_0\} \cap \{\noofdec_{k_0} 
+=
+\ell\}) \\
+&= \fixn{B}\cdot \probm(D) - \eps\cdot \sum_{\ell=1}^{k_0-\fixn{i}} \probm(D 
+\cap \{U\geq k_0\} \cap \{\noofdec_{k_0} 
+\geq 
+\ell\})\\
+& \leq \fixn{B}\cdot \probm(D) - \eps\cdot \sum_{\ell=1}^{m} \probm(D 
+\cap \{U\geq k_0\} \cap \{\noofdec_{k_0} 
+\geq 
+\ell\})\\
+&\leq \fixn{B}\cdot \probm(D) - \eps\cdot m \cdot \fixn{p}/2 < 0, 
+\end{align*}
+the last inequality following by substituting for $m$. But for each $k$ the random variable $Y_k$, so it must also have a non-negative expectation, a contradiction.
+%\begin{enumerate}
+%\
+%\end{itemize}
+%\end{enumerate}
 

### 2017-06-11 implementation/files/invariant_script.sh
formatting

@@ -1,4 +1,4 @@
 ./files/aspicV3.4 files/aspic.fast > files/InvariantOutput
-awk 'BEGIN {x=0} /^"*** Results :"/{x=1} x==1{print}' < files/InvariantOutput > InvariantOutput.tmp
+awk 'BEGIN {x=0} x==1{print} /^\*\*\* Results :/{x=1}' < files/InvariantOutput > InvariantOutput.tmp
 mv InvariantOutput.tmp files/InvariantOutput
 rm aspic.log

### 2017-06-11 implementation/main.cpp

@@ -2,8 +2,7 @@
 // end node will always be 2
 /*
 Comments:
-	Ask Petr	When invariant is NULL the program gives wrong result, because it ignores the implication in that case
-https://tapas.labri.fr/wp/wp-content/uploads/2017/02/FAST-manual.pdf
+	https://tapas.labri.fr/wp/wp-content/uploads/2017/02/FAST-manual.pdf
 */
 #include <iostream>
 #include <algorithm>
@@ -376,7 +375,7 @@ void print_fast(ostream& fastfile){
 	}
 
 	fastfile<<"}\n\nstrategy main_s{\n\n";
-	fastfile<<"\tRegion init := {state = state_1 && x_1 >= 0 && x_2 >= 1};\n\n";
+	fastfile<<"\tRegion init := {state = state_1};\n\n";
 	fastfile<<"}";
 }
 
@@ -427,6 +426,21 @@ int main(){
 	if(system("./files/invariant_script.sh")!=0){
 		cout<<"Something wrong with the script to find invariants";
 	}
+	
+	// Code for analysing files/InvariantOutput comes here
+	ifstream invariant_file("files/InvariantOutput");
+	for(int i=1;i<=last_used_label;){
+		string line;
+		getline(invariant_file,line);
+		size_t open = line.find('{');
+		if(open!=string::npos){
+			size_t close = line.find('}');
+			cout<<open<<" "<<close<<" "<<part(line,open+1,close)<<endl;
+			// <flag> Left here
+			i++;
+		}
+	}
+
 	/*
 	// Code to print the tree structure etc.
 	cout<<"Input Code:"<<endl;

### 2017-06-12 paper/prelims.tex
small changes in prelims

@@
 
-\smallskip\noindent{\em Affine expressions.}
+\smallskip\noindent{\em Expressions.}
+An \emph{expression} over the set of variables $\{x_1,\dots,x_n\}$ 
+is an expression in the standard programming-language sense, i.e. a formula 
+built in finite number of steps from constants, variables $x_1,\dots,x_n$, and 
+numerical operators from some fixed finite set. Each expression $E$ over 
+$\{x_1,\dots,x_n\}$ 
+determines a function which for each $m$-dimensional vector $\vec{x}$, where 
+$m\geq n$,  
+returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
+Slightly abusing our notation, we denote this function also by $E$ and the 
+value of this function on argument $\vec{x}$ by $E(\vec{x})$. 
+%An expression $E$ 
+%is \emph{measurable} if the function $E(\vec{x})$ is Borel 
+%measurable~\cite{xxx}.  In particular, expressions built using the standard 
+%arithmetic operators  are all measurable.
+
+\smallskip\noindent{\em Affine Expressions.}
 An \emph{affine expression} over the set of variables $\{x_1,\dots,x_n\}$ is an 
@@
 x_i$, where  $d,a_1,\dots,a_n$ are real-valued
-constants. Each affine expression $E$ over $\{x_1,\dots,x_n\}$ 
-determines a function which for each $m$-dimensional vector $\vec{x}$, where $m\geq n$,  
-returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
-Slightly abusing our notation, we denote this function also by $E$ and the 
-value of this function on argument $\vec{x}$ by $E(\vec{x})$. A function of the form $E(\vec{x})$ for some affine expression $E$ is called affine.
+constants.  A function of the form $E(\vec{x})$ for some affine expression $E$ 
+is called affine.
 %Following the terminology of~\cite{DBLP:conf/sas/KatoenMMM10} we fix the
@@
 \vspace{-1em}
-\subsection{Syntax of Affine Probabilistic Programs (\APP s)}\label{subsec:syntax}
+\subsection{Syntax of Probabilistic Programs}\label{subsec:syntax}
 \vspace{-0.5em}
 
+In this subsection we define the form of probabilistic programs that we 
+consider in our analysis. We consider two classes of probabilistic programs: 
+general probabilistic programs with arbitrary (measurable) expressions and 
+their subclass, affine probabilistic programs (\APP s) where are expressions 
+are restricted to be affine (see below for a precise definition). The reason 
+for this dual view is that our work also has two main points of focus: a 
+theoretical one, where we introduce new proof rules that can be used to prove 
+properties of general probabilistic programs; and an algorithmic one, where we 
+aim to prove properties of probabilistic programs automatically, using the 
+aforementioned proof rules. As already testified in the non-probabilistic 
+world, programs that contain only affine expressions allow for more efficient 
+automation of the analysis and at the same time, due to the presence of 
+non-determinism they can be used to form sound abstractions of programs with 
+non-linear arithmetic. Hence, we consider general programs when providing our 
+theoretical results and \APP s when presenting the automation of our techniques.
+
 \noindent{\em The Syntax.}
-We consider the standard syntax for affine probabilistic programs,
+We consider the standard syntax for probabilistic programs,
 which encompasses basic programming mechanisms such as assignment statement 
@@
 values of the variables are real numbers. 
-We allow only affine expressions in test statements and in the right-hand sides of assignments.
-We also assume that assume that each \APP{} $\program$ is preceded by an initialization preamble 
+We also assume that assume that each \APP{} $\program$ is preceded by an 
+initialization preamble 
 in 
@@
 each variable appearing in $\program$ is assigned some concrete number.
+the program variables are initialized. We do not a priori (\textbf{PETR: 
+elaborate one xpressions})
+
+\smallskip\noindent{\em Affine Probabilistic Programs (\APP s).}
+A probabilistic program is \emph{affine} if all the expressions that occur in 
+the program (i.e. in loop guards, conditionals and on right-hand sides of 
+assignments) are affine. We refer to the class of affine probabilistic programs 
+as \APP s.
+%We also assume that assume that each \APP{} $\program$ is preceded by an 
+%initialization preamble 
+%in 
+%which 
+%each variable appearing in $\program$ is assigned some concrete number.
 %the program variables are initialized.
-Due to space restrictions, details (such as grammar) are relegated to the Appendix.
+
+Due to space restrictions, details of syntax (such as grammar) are relegated to 
+the Appendix.
 For an example see Figure~\ref{fig:invariant-running}.
-We refer to this class of affine probabilistic programs as \APP s.
 

### 2017-06-12 implementation/files/Parser.cpp
Invariant automation

@@ -35,6 +35,17 @@ void skip_spaces(int &begin, int &end){ //skips the spaces in the beginning and 
 	}
 }
 
+void skip_spaces(string& line){
+	int begin = 0,end = line.length();
+	while(begin<line.size() and isspace(line[begin])){
+		begin++;
+	}
+	while(end>0 and isspace(line[end-1])){
+		end--;
+	}
+	line = part(line,begin,end);
+}
+
 node::node(string t){
 	bracket = NULL;
 	begin = -1;
@@ -53,6 +64,36 @@ node::node(string t, int b, int e, int s, int l, bool negate){
 		cerr<<"Error! begin= "<<b<<" end= "<<e<<endl;
 	}
 	process(s,l,negate);
+}
+
+node::node(string t, string l){
+	type = t;
+	if(t=="bexpr"){
+		children.resize(1);
+		children[0] = new node("affexpr");
+		stringstream line_stream;
+		line_stream.str(l);
+		string comma_sep;
+		while(getline(line_stream,comma_sep,',')){
+			skip_spaces(comma_sep);
+			node* formed_literal = new node("literal",comma_sep);
+			children[0]->children.pb(formed_literal);
+		}
+	}
+	else if(t=="literal"){
+		constant = "and";
+		proc_literal(false,false,l);
+		// <flag> Do something with the return value, ask Petr for the specific case
+	}
+	else if(t=="expr"){
+		constant = "expression";
+		proc_expr(l);
+	}
+	else{
+		cerr<<"Wrong usage"<<endl;
+		return;
+	}
+	
 }
 
 void vcopy(vector<node*> &sink,vector<node*> &tocopy){
@@ -111,7 +152,31 @@ node* negation(node* tonegate){
 	return NULL;
 }
 
+node* and_node(node* one,node* two){
+	if(one->type!="bexpr" or two->type!="bexpr"){
+		cerr<<"Taking and of wrong type of nodes"<<endl;
+		return NULL;
+	}
+	node* node_and = NULL;
+	node* temp_affexpr = NULL;
+	node_and = new node("bexpr");
+	node_and->constant = "or";
+	int a = one->children.size(), b = two->children.size();
+	for(int i = 0;i<a;i++){
+		for(int j = 0;j<b;j++){
+			temp_affexpr = new node("affexpr");
+			temp_affexpr->constant = "and";
+			temp_affexpr->children = one->children[i]->children;
[…]

### 2017-06-12 implementation/files/Parser.h

@@ -11,6 +11,7 @@ struct node{
 	//constructor
 	node(string t);
 	node(string t, int b, int e, int s = 0, int l = 0, bool negate = false);
+	node(string t, string line);
 private:
 //procs
 	void proc_stmt(int s,int l);
@@ -18,12 +19,15 @@ private:
 	void proc_affexpr();
 	void proc_pvar();
 	void analyse_expr(int begin,int end,bool negate);
+	void analyse_expr(string l,bool negate);
 	void proc_expr();
+	void proc_expr(string l);
 	void proc_constant();
-	void proc_literal(bool negate);
+	bool proc_literal(bool negate, bool strategic = true, string line = "");
 	void proc_bexpr();
 	void proc_ndbexpr();
 	void form_vector(int begin,int end,bool negate);
+	void form_vector(string l,bool negate);
 //main process
 	void process(int s, int l, bool negate);
 public:
@@ -58,6 +62,7 @@ public:
 void skip_spaces(int &begin, int &end);
 void vcopy(vector<node*> &sink,vector<node*> &tocopy);
 node* negation(node* tonegate);
+node* and_node(node* one,node* two);
 int find_variables();
 
 extern string program;

### 2017-06-12 implementation/main.cpp

@@ -2,6 +2,8 @@
 // end node will always be 2
 /*
 Comments:
+	Ask Petr what to do about the specific case <flag>
+	Considering start invariant and neglecting all other invariants, also the start invariant should be a polyhedra(Which I think would not pose a problem)
 	https://tapas.labri.fr/wp/wp-content/uploads/2017/02/FAST-manual.pdf
 */
 #include <iostream>
@@ -49,9 +51,6 @@ struct cond{
 				}
 			}
 			else{
-				// cout<<"change is : src"<<src<<" dest:"<<dest1<<" toChange:"<<toChange<<" changed to:";
-				// change->print();
-				// cout<<endl;
 				buffer.clear();
 				buffer.str(string());
 				buffer<<"eps"<<equation_count<<"-f_"<<src<<"_0+f_"<<dest1<<"_0";
@@ -78,12 +77,9 @@ struct cond{
 						if(change->expression[i+1]!=0.0){
 							buffer<<change->expression[i+1]<<"f_"<<dest1<<"_"<<toChange;
 						}
-						// change->print();
-						// cout<<""<<endl;
 					}
 					buffer<<"-f_"<<src<<"_"<<i+1;
 					c[i] = buffer.str();
-					// cout<<c[i]<<endl;
 				}
 			}
 		}
@@ -134,28 +130,6 @@ map<int,equation*> equations;
 int equations_counter = 0;
 int epsilons_used = 0;
 
-node* and_node(node* one,node* two){
-	if(one->type!="bexpr" or two->type!="bexpr"){
-		cerr<<"Taking and of wrong type of nodes"<<endl;
-		return NULL;
-	}
-	node* node_and = NULL;
-	node* temp_affexpr = NULL;
-	node_and = new node("bexpr");
-	node_and->constant = "or";
-	int a = one->children.size(), b = two->children.size();
-	for(int i = 0;i<a;i++){
-		for(int j = 0;j<b;j++){
-			temp_affexpr = new node("affexpr");
-			temp_affexpr->constant = "and";
-			temp_affexpr->children = one->children[i]->children;
-			temp_affexpr->children.insert(temp_affexpr->children.end(),two->children[j]->children.begin(),two->children[j]->children.end());
-			node_and->children.pb(temp_affexpr);
-		}
-	}
-	return node_and;
-}
-
 void generate_equations(){ //Would use the ostream file to write the equations into it later
[…]

### 2017-06-13 implementation/files/Parser.cpp
Minor additions

@@ -66,7 +66,7 @@ node::node(string t, int b, int e, int s, int l, bool negate){
 	process(s,l,negate);
 }
 
-node::node(string t, string l){
+node::node(string t, string l, bool& to_return){
 	type = t;
 	if(t=="bexpr"){
 		children.resize(1);
@@ -76,14 +76,32 @@ node::node(string t, string l){
 		string comma_sep;
 		while(getline(line_stream,comma_sep,',')){
 			skip_spaces(comma_sep);
-			node* formed_literal = new node("literal",comma_sep);
+			bool return_value = false;
+			node* formed_literal = new node("literal",comma_sep, return_value);
 			children[0]->children.pb(formed_literal);
+			if(return_value){
+				node* other_literal = new node("literal");
+				other_literal->constant = "<=";
+				other_literal->children.resize(1);
+				other_literal->children[0] = new node("expr");
+				other_literal->children[0]->expression.resize(nVariables+1);
+				for(int i=0;i<=nVariables;++i){
+					other_literal->children[0]->expression[i] = -1.0*formed_literal->children[0]->expression[i];
+				}
+				children[0]->children.pb(other_literal);
+				// formed_literal->print();
+				// cout<<endl;
+				// other_literal->print();
+				// cout<<endl;
+
+			}
 		}
 	}
 	else if(t=="literal"){
 		constant = "and";
-		proc_literal(false,false,l);
-		// <flag> Do something with the return value, ask Petr for the specific case
+		if(proc_literal(false,false,l)){
+			to_return = true;
+		}
 	}
 	else if(t=="expr"){
 		constant = "expression";
@@ -595,29 +613,30 @@ bool node::proc_literal(bool negate,bool strategic, string line){
 			}
 		}
 		children.resize(2);
-		children[0] = new node("expr",part(line,begin,sign));
+		bool temp;
+		children[0] = new node("expr",part(line,begin,sign),temp);
 		if(sign==-1){
 			cerr<<"Invalid literal";
 		}
 		else if(part(line,sign,sign+2)==">="){
 			constant = ">=";
-			children[1] = new node("expr",part(line,sign+2,end));
+			children[1] = new node("expr",part(line,sign+2,end),temp);
 		}
 		else if(part(line,sign,sign+2)=="<="){
 			constant = "<=";
-			children[1] = new node("expr",part(line,sign+2,end));
[…]

### 2017-06-13 implementation/files/Parser.h

@@ -11,7 +11,7 @@ struct node{
 	//constructor
 	node(string t);
 	node(string t, int b, int e, int s = 0, int l = 0, bool negate = false);
-	node(string t, string line);
+	node(string t, string line, bool& to_return);
 private:
 //procs
 	void proc_stmt(int s,int l);
@@ -69,5 +69,4 @@ extern string program;
 extern int nVariables;
 extern int last_used_label;
 extern map<int,CFG_location*> label_map;
-extern node* id1;
 extern node* root;

### 2017-06-13 implementation/main.cpp

@@ -2,7 +2,6 @@
 // end node will always be 2
 /*
 Comments:
-	Ask Petr what to do about the specific case <flag>
 	Considering start invariant and neglecting all other invariants, also the start invariant should be a polyhedra(Which I think would not pose a problem)
 	https://tapas.labri.fr/wp/wp-content/uploads/2017/02/FAST-manual.pdf
 */
@@ -132,6 +131,8 @@ int epsilons_used = 0;
 
 void generate_equations(){ //Would use the ostream file to write the equations into it later
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
+		it->second->invariant->print();
+		cout<<endl;
 		if(it->second->type=="det"){
 			//Invariant and guard imply the value decrease
 			if(it->second->edges.empty()){
@@ -142,9 +143,7 @@ void generate_equations(){ //Would use the ostream file to write the equations i
 				cond* condition = new cond(++equations_counter,it->second->edges[0].toChange,it->second->edges[0].change,it->first,it->second->edges[0].next->label);
 				// Guard would have been NULL here
 				if(it->second->invariant==NULL){
-					// <flag> Do something because true and false handling has changed
-					//This should never be the case as this would pose conditions that c==0 and d>0, which are not good
-					cerr<<"No invariant specified here"<<endl;
+					//It is the case when the invariant is false
 				}
 				else{
 					// Invariant implies the given condition
@@ -225,8 +224,6 @@ void generate_equations(){ //Would use the ostream file to write the equations i
 int last_used_lambda = 0;
 
 void print_equations(ostream& equationsfile){
-	// <flag> to be changed
-	
 	equationsfile<<"maximize ";
 	for(int i=1;i<=epsilons_used;++i){
 		if(equations.find(i)!=equations.end()){
@@ -406,18 +403,26 @@ int main(){
 			string invariant_string = part(line,open+1,close);
 			if(invariant_string=="true"){
 				if(label_map[i]->invariant==NULL){
-					label_map[i]->invariant = new node("true");
+					// label_map[i]->invariant = new node("true"); //Effectively but add 0<=0 instead
+					label_map[i]->invariant = new node("bexpr");
+					node* concerned_node = label_map[i]->invariant;
+					concerned_node->children.resize(1);
+					concerned_node->children[0] = new node("affexpr");
+					concerned_node->children[0]->children.resize(1);
[…]

### 2017-06-14 paper/lexicographic.tex
Streamlined the current vesrion of the paper a bit, added note on integrability.

@@
 supermartingale}, which generalizes the standard notion of ranking 
-supermartingales.
+supermartingales. However, to define any form of a supermartingale, we need the crucial notion of conditional expectation.
+
+\smallskip\noindent{\bf Conditional Expectation.} 
+%The notion of conditional expectation 
+%tightly connected to martingales. We present only a brief informal exposition 
+%here, for formal treatment see, e.g.~\cite[Chapter 9]{Williams:book}.
+Let $(\Omega,\mathcal{F},\probm)$ be a probability space, 
+$X\colon\Omega\rightarrow 
+\Rset$ an $\mathcal{F}$-measurable function, and $\mathcal{F}'\subseteq 
+\mathcal{F}$ sub-sigma-algebra of $\mathcal{F}$. The \emph{conditional 
+	expectation} of $X$ given $\mathcal{F}'$ is an $\mathcal{F}'$-measurable random 
+variable denoted by $\E[X| \mathcal{F}']$ which satisfies, for each set $A\in 
+\mathcal{F}'$, the following: 
+\begin{equation}
+\label{eq:cond-exp}
+\E[X\cdot 1_A] = \E[\E[X|\mathcal{F}]\cdot 1_A],
+\end{equation}
+where $1_A \colon \Omega\rightarrow \{0,1\}$ is an \emph{indicator function} of 
+$A$, i.e. function returning $1$ for 
+each $\omega\in A$ and $0$ for each $\omega\in \Omega\setminus A$. Note that 
+the left hand-side of~\eqref{eq:cond-exp} intuitively represents the expected 
+value 
+of $X(\omega)$ with domain restricted to $A$.
+
+In context of probabilistic programs we work with probability 
+spaces of the form $(\Omega,\natfilt,\probm^\sigma)$, where $\Omega$ is a 
+set of runs in some $\pCFG$ and $\natfilt$ is (the smallest) sigma-algebra 
+such that all the functions $\cfg{\sigma}{i}$, where $i\in \Nset_0$ and 
+$\sigma$ is a scheduler, are $\natfilt$-measurable. In such a setting we can 
+also consider sub-sigma-algebras $\natfilt_i$, $i\in \Nset_0$, of 
+$\natfilt$, where $\natfilt_i$ is the smallest sub-sigma-algebra of 
+$\natfilt$ such that all the functions $\cfg{\sigma}{j}$, $0\leq j \leq 
+i$, are $\natfilt_i$-measurable. Intuitively, each set $A$ belonging to such 
+an $\natfilt_i$ consists of runs whose first $i$ steps satisfy some 
+property, and the probability space $(\Omega,\natfilt_i,\probm^\sigma)$ 
+allows us to reason about probabilities of certain events happening in the 
+first 
+$i$ steps of program execution. 
+%(note that since $\natfilt_i \subseteq 
+%\natfilt$, we can view $\probm^\sigma$ also as a function of type 
+%$\natfilt_i\rightarrow [0,1]$). 
+Then, for each $A\in \natfilt_i$, the 
+value $\E[\E[X|\natfilt_i]\cdot 1_A]$ represents the expected value of 
+$X(\run)$ for the randomly generated run $\run$ provided that we restrict to 
+runs whose
+prefix of length $i$ satisfies the property given by $A$. 
+The sequence $\natfilt_0,\natfilt_1,\natfilt_2,\dots$ forms a 
+filtration of $\natfilt$, which we call a \emph{canonical filtration}.
 
@@
 %$\E[|\vecseq{X}{i}{j}|]<\infty$.
+\item For each $\omega \in \Omega$ and $i\in \Nset_0$ it holds 
+$\vec{X}_i (\omega)\geq \vec{0}$ (componentwise).
 \item For each $i\in \Nset_0$ and each $\omega\in \Omega$ with 
@@
 \end{compactitem}
-\item For each $\omega \in \Omega$ and $i\in \Nset_0$ it holds 
-$\vec{X}_i (\omega)\geq \vec{0}$ (componentwise).
 \end{compactenum}
 \end{definition}
+
+The 1-dimensional lexicographic $\eps$-ranking supermartingale is, to a large extent, equivalent to the notion of a ranking supermartingale as studied in~\cite{xxx}.
+There is one significant difference: in~\cite{xxx} there is an additional \emph{integrability} condition imposed on the one-dimensional process $\{X_i\}_{i=0}^{\infty}$, which requires that for each $i\geq 0$ it holds $\E[|X_i|]<\infty$ (or equivalently $\E[X_i]<\infty$, as the process is required to be non-negative).
+We do not impose this condition, and indeed, it is possible to define a 3-dimensional LexRSM $\{\vec{X}_{i}\}_{i=0}^{\infty}$ such that $\E[X[3]_i]=\infty$ for some $i$ \textbf{[PETR: EXAMPLE]}.
+However, closer inspection of proofs in~\cite{xxx} that link 1-dimensional ranking supermartingales to a.s.
+termination reveals that the integrability condition is never explicitly used there.
+Indeed, the only point in which integrability is needed is to ensure that the conditional expectations exist and are well-defined.
+However, the existence of conditional expectations is also guaranteed for random variables that are non-negative, even if their expected value if infinite, see Theorem~\ref{xxx}.
+This is exactly the case in both the original 1-dimensional ranking supermartingales and our generalization to LexRSMs.
+Below we show, that also in the LexRSM case we do not need integrability to prove a.s.
+termination, and hence we drop the condition altogether.
+This is particularly relevant when applying LexRSMs to programs with non-linear arithmetic, where proving integrability of LexRSM is more difficult than in the affine case, see Example~\ref{xxx}.
 

### 2017-06-14 paper/prelims.tex
Streamlined the current vesrion of the paper a bit, added note on integrability.

@@
 \vspace{-1em}
-\subsection{Basic Notions, Linear Predicates, Valuations}
+\subsection{Basic Notions}
 \vspace{-0.5em}
@@
 
-\smallskip\noindent{\em Expressions.}
-An \emph{expression} over the set of variables $\{x_1,\dots,x_n\}$ 
-is an expression in the standard programming-language sense, i.e. a formula 
-built in finite number of steps from constants, variables $x_1,\dots,x_n$, and 
-numerical operators from some fixed finite set. Each expression $E$ over 
-$\{x_1,\dots,x_n\}$ 
-determines a function which for each $m$-dimensional vector $\vec{x}$, where 
-$m\geq n$,  
-returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
-Slightly abusing our notation, we denote this function also by $E$ and the 
-value of this function on argument $\vec{x}$ by $E(\vec{x})$. 
-%An expression $E$ 
-%is \emph{measurable} if the function $E(\vec{x})$ is Borel 
-%measurable~\cite{xxx}.  In particular, expressions built using the standard 
-%arithmetic operators  are all measurable.
-
-\smallskip\noindent{\em Affine Expressions.}
-An \emph{affine expression} over the set of variables $\{x_1,\dots,x_n\}$ is an 
-expression of the form $d+\sum_{i=1}^{n}a_i
-x_i$, where  $d,a_1,\dots,a_n$ are real-valued
-constants.  A function of the form $E(\vec{x})$ for some affine expression $E$ 
-is called affine.
-%Following the terminology of~\cite{DBLP:conf/sas/KatoenMMM10} we fix the
-%following nomenclature:
-
-\smallskip\noindent{\em Linear constraint, assertion, predicates.}
-We use the following nomenclature:
-\begin{compactitem}
-\item {\em Linear Constraint.} A \emph{linear constraint} is a formula of the
-form $\psi$ or $\neg\psi$, where  $\psi$ is a non-strict inequality
-between affine expressions.
-\item {\em Linear Assertion.} A \emph{linear assertion} is a finite conjunction
-of linear constraints.
-\item {\em Propositionally Linear Predicate.}
-A  \emph{propositionally linear predicate} (PLP) is a finite disjunction of
-linear assertions.
-\end{compactitem}
-
-\noindent{\em Arity and satisfaction of PLP.}
-%The \emph{arity} of a PLP is the number of distinct variables appearing in it. 
-For a PLP $\varphi$ we denote by $\vars(\varphi)$ the set of all variables that 
-appear in $\varphi$.
+%\smallskip\noindent{\em Expressions.}
+%An \emph{expression} over the set of variables $\{x_1,\dots,x_n\}$ 
+%is an expression in the standard programming-language sense, i.e. a formula 
+%built in finite number of steps from constants, variables $x_1,\dots,x_n$, and 
+%numerical operators from some fixed finite set. Each expression $E$ over 
+%$\{x_1,\dots,x_n\}$ 
+%determines a function which for each $m$-dimensional vector $\vec{x}$, where 
+%$m\geq n$,  
+%returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
+%Slightly abusing our notation, we denote this function also by $E$ and the 
+%value of this function on argument $\vec{x}$ by $E(\vec{x})$. 
+%%An expression $E$ 
+%%is \emph{measurable} if the function $E(\vec{x})$ is Borel 
+%%measurable~\cite{xxx}.  In particular, expressions built using the standard 
+%%arithmetic operators  are all measurable.
+%
+%\smallskip\noindent{\em Affine Expressions.}
+%An \emph{affine expression} over the set of variables $\{x_1,\dots,x_n\}$ is an 
+%expression of the form $d+\sum_{i=1}^{n}a_i
+%x_i$, where  $d,a_1,\dots,a_n$ are real-valued
+%constants.  A function of the form $E(\vec{x})$ for some affine expression $E$ 
+%is called affine.
+%%Following the terminology of~\cite{DBLP:conf/sas/KatoenMMM10} we fix the
+%%following nomenclature
+%
+%\smallskip\noindent{\em Predicates}
+%A \emph{predicate} is a logical formula obtained by a finite number of applications of conjunction, disjunction and negation operations on \emph{atomic predicates} of the form $E(\vec{x})\leq E'(\vec{y})$, where $E$, $E'$ are expressions.
+%
+%\smallskip\noindent{\em Linear constraints, assertions, predicates.}
+%In the case of predicates involving only linear expression we use the following nomenclature:
+%\begin{compactitem}
+%\item {\em Linear Constraint.} A \emph{linear constraint} is a formula of the
+%form $\psi$ or $\neg\psi$, where  $\psi$ is a non-strict inequality
+%between affine expressions.
+%\item {\em Linear Assertion.} A \emph{linear assertion} is a finite conjunction
+%of linear constraints.
+%\item {\em Propositionally Linear Predicate.}
+%A  \emph{propositionally linear predicate} (PLP) is a finite disjunction of
+%linear assertions.
+%\end{compactitem}
+%
+%\noindent{\em Arity and satisfaction of PLP.}
+%%The \emph{arity} of a PLP is the number of distinct variables appearing in it. 
+%For a PLP $\varphi$ we denote by $\vars(\varphi)$ the set of all variables that 
+%appear in $\varphi$.
 As noted above, we stipulate that $\vars(\varphi)=\{x_1,\dots,x_{n(\varphi)}\}$ for some $n(\varphi)\in \Nset$.
 A vector $\vec{x}$ of dimension $m\geq n(\varphi)$ \emph{satisfies} $\varphi$, we write
-$\vec{x}\models\phi$, if 
- the arithmetic formula obtained by 
-substituting each occurrence of a variable $x_i$ in $\varphi$ by $\vec{x}[i]$ is 
-valid. We denote $\sat{\phi} = \{\vec{x}\in \Rset^m\mid m\geq n(\varphi) \wedge \vec{x} \models\phi\}$ 
-and $\exsat{\phi} = \sat{\phi}
-\cap \Rset^{n(\varphi)}$.
+%$\vec{x}\models\phi$, if 
+% the arithmetic formula obtained by 
+%substituting each occurrence of a variable $x_i$ in $\varphi$ by $\vec{x}[i]$ is 
+%valid. We denote $\sat{\phi} = \{\vec{x}\in \Rset^m\mid m\geq n(\varphi) \wedge \vec{x} \models\phi\}$ 
+%and $\exsat{\phi} = \sat{\phi}
+%\cap \Rset^{n(\varphi)}$.
 
@@
 
-\noindent{\em The Syntax.}
+\smallskip\noindent{\em Expressions.}
+An \emph{expression} over the set of variables $\{x_1,\dots,x_n\}$ 
+is an expression in the standard programming-language sense, i.e. a formula 
+built in finite number of steps from constants, variables $x_1,\dots,x_n$, and 
+numerical operators from some fixed finite set. Each expression $E$ over 
+$\{x_1,\dots,x_n\}$ 
+determines a function which for each $m$-dimensional vector $\vec{x}$, where 
+$m\geq n$,  
+returns a number resulting from substituting each $x_i$ in $E$ by $\vec{x}[i]$. 
+Slightly abusing our notation, we denote this function also by $E$ and the 
+value of this function on argument $\vec{x}$ by $E(\vec{x})$.
+We do not a priori fix a concrete set of operators that can be used to form expressions.
+However, in order to ensure that semantics of probabilistic programs with real-valued variables is defined correctly, we impose the following two conditions on the set of expressions used in each program:
+\begin{enumerate}
+\item
+For each expression $E$ over variables $\{x_1,\dots,x_n\}$ and each $n$-dimensional vector $\vec{x}$ the value $E(\vec{x})$ is well defined.\footnote{Our results can be easily extended to programs where encountering an expression of undefined value, such as division by zero, triggers an exception which terminates the program, but we abstract away from such details for the sake of clarity.}
+\item
+The function defined by each expression $E$ is Borel-measurable (for definition of Borel-measura\-bility, see, e.g.~\cite{Billingsley:book}).
+\end{enumerate}
+
+From measure theory it is known that these conditions hold in particular for programs where expressions are build using the standard arithmetic operators of addition, subtraction, multiplication, and division (provided that expressions evaluating to zero are not allowed as divisors).
+
+%An expression $E$ 
+%is \emph{measurable} if the function $E(\vec{x})$ is Borel 
+%measurable~\cite{xxx}.  In particular, expressions built using the standard 
+%arithmetic operators  are all measurable.
+
+\smallskip\noindent{\em Affine Expressions.}
+An \emph{affine expression} over the set of variables $\{x_1,\dots,x_n\}$ is an 
+expression of the form $d+\sum_{i=1}^{n}a_i
+x_i$, where  $d,a_1,\dots,a_n$ are real-valued
+constants.  A function of the form $E(\vec{x})$ for some affine expression $E$ 
+is called affine. As noted above
+%Following the terminology of~\cite{DBLP:conf/sas/KatoenMMM10} we fix the
+%following nomenclature
+
+\smallskip\noindent{\em Predicates}
+A \emph{predicate} is a logical formula obtained by a finite number of applications of conjunction, disjunction and negation operations on \emph{atomic predicates} of the form $E(\vec{x})\leq E'(\vec{y})$, where $E$, $E'$ are expressions.
+
+\smallskip\noindent{\em Linear constraints, assertions, predicates.}
+In the case of predicates involving only linear expression we use the following nomenclature:
+\begin{compactitem}
+	\item {\em Linear Constraint.} A \emph{linear constraint} is a formula of the
+	form $\psi$ or $\neg\psi$, where  $\psi$ is a non-strict inequality
+	between affine expressions.
+	\item {\em Linear Assertion.} A \emph{linear assertion} is a finite conjunction
+	of linear constraints.
+	\item {\em Propositionally Linear Predicate.}
+	A  \emph{propositionally linear predicate} (PLP) is a finite disjunction of
+	linear assertions.
+\end{compactitem}
+
+\textbf{[PETR: SAY SOMETHING ABOUT PREDICATE SATISFACTION.]}
+\textbf{[PETR: EXPLAIN BOREL-MEASURABILITY IN MORE DETAIL?]}
+
+%\noindent{\em Arity and satisfaction of PLP.}
+%%The \emph{arity} of a PLP is the number of distinct variables appearing in it. 
+%For a PLP $\varphi$ we denote by $\vars(\varphi)$ the set of all variables that 
+%appear in $\varphi$.
+As noted above, we stipulate that $\vars(\varphi)=\{x_1,\dots,x_{n(\varphi)}\}$ for some $n(\varphi)\in \Nset$.
+A vector $\vec{x}$ of dimension $m\geq n(\varphi)$ \emph{satisfies} $\varphi$, we write
+%$\vec{x}\models\phi$, if 
+%the arithmetic formula obtained by 
+%substituting each occurrence of a variable $x_i$ in $\varphi$ by $\vec{x}[i]$ is 
+%valid. We denote $\sat{\phi} = \{\vec{x}\in \Rset^m\mid m\geq n(\varphi) \wedge \vec{x} \models\phi\}$ 
+%and $\exsat{\phi} = \sat{\phi}
+%\cap \Rset^{n(\varphi)}$.
+
+\noindent{\em The Syntax of Probabilistic Programs {(\PP s)}.}
 We consider the standard syntax for probabilistic programs,
 which encompasses basic programming mechanisms such as assignment statement 
-(indicated by `:='), while-loop, if-branch. We also consider basic probabilistic mechanisms 
+(indicated by `:='), while-loop, if-branch.
+Expressions appear on right-hand sides of assignments, and predicates act as loop guards and conditions in if-then-else statements.
+We also consider basic probabilistic mechanisms
 such as probabilistic branch (indicated by `prob') and random sampling (e.g. 
@@
 values of the variables are real numbers. 
-We also assume that assume that each \APP{} $\program$ is preceded by an 
+We also assume that assume that each \PP{} $\program$ is preceded by an 
 initialization preamble 
@@
 each variable appearing in $\program$ is assigned some concrete number.
-the program variables are initialized. We do not a priori (\textbf{PETR: 
-elaborate one xpressions})
+the program variables are initialized.
 
@@
 \vspace{-1em}
-\subsection{Semantics of Affine Probabilistic Programs}\label{subsec:semantics}
+\subsection{Semantics of Probabilistic Programs}\label{subsec:semantics}
 \vspace{-0.5em}
 
-We now formally define the semantics of \APP's.
+We now formally define the semantics of \PP's.
 In order to do this, we first recall some fundamental concepts from probability
@@
 % POKRACOVAT ODSUD
-\item $\transitions\subseteq\locs\times\locs$ is a transition relation'
+\item $\transitions\subseteq\locs\times\locs$ is a transition relation;
 %whose members are tuples of
@@
 %\item 
-(a)~an affine function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
+(a)~a Borel-measurable function $u\colon \Rset^{|\pvars|}\rightarrow \Rset$;
 %\item 
@@
 \begin{compactitem}
-\item if $\up$ is an affine function, then $a=\up(\vec{x})$;
+\item if $\up$ is a Borel-measurable function, then $a=\up(\vec{x})$;
 \item if $\up$ is an integrable\footnote{A distribution on some numerical 
@@
 \smallskip\noindent{\em Termination, reachability, and termination time.}
-In the following, consider an \APP{} $P$ and its associated pCFG $\pCFG_P$. 
+In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. 
 We say that a run $\run$ of $\pCFG_P$ \emph{reaches} a set of configurations 

### 2017-06-17 paper/lexicographic.tex
corrected some stuff about conditional expectation

@@
 \Rset$ an $\mathcal{F}$-measurable function, and $\mathcal{F}'\subseteq 
-\mathcal{F}$ sub-sigma-algebra of $\mathcal{F}$. The \emph{conditional 
+\mathcal{F}$ sub-sigma-algebra of $\mathcal{F}$. A \emph{conditional 
 	expectation} of $X$ given $\mathcal{F}'$ is an $\mathcal{F}'$-measurable random 
@@
 of $X(\omega)$ with domain restricted to $A$.
+
+Note that any $\mathcal{F}'$-measurable random variable satisfying~\eqref{eq:cond-exp} can be called a conditional expectation.
+The definition does not guarantee that the conditional expectation is uniquely defined or that it exists at all.
+However, from probability theory we have the following:
+
+\begin{proposition}
+\label{prop:conditional-exp-existence}
+Let $(\Omega,\mathcal{F},\probm)$ be a probability space, 
+$X\colon\Omega\rightarrow 
+\Rset$ an $\mathcal{F}$-measurable function, and $\mathcal{F}'\subseteq 
+\mathcal{F}$ sub-sigma-algebra of $\mathcal{F}$. Assume that one of the following conditions hold:
+\begin{itemize}
+\item $\E[|X|]<\infty$; or
+\item $X$ is non-negative and for all $\omega\in\Omega$ it holds $X(\omega)\neq \infty$.
+\end{itemize}
+Then there exists a conditional expectation of $X$ given $\mathcal{F}'$ and it is almost-surely unique, i.e.
+for each two $\mathcal{F}'$-measurable functions $f$, $g$ that satisfy~\eqref{eq:cond-exp} it holds $\probm(\{\omega\mid f(\omega)\neq g(\omega)\})=0$.
+\end{proposition}
+\begin{proof}
+The proof for the case when $\E[|X|]$ is standard and appears in many textbooks on probability theory (e.g.~\cite{Billingsley:book,Ash:book,Rosenthal:book}).
+The proof for the second case is essentially the same: the condition that $X$ is non-negative and not admitting infinite value suffices for satisfying the assumptions of Radon-Nikodym Theorem, the main theoretical tool used in the proof.
+For the sake of completeness we present the proof in \AppendixMaterial.
+\end{proof}
+
+Since the constraint~\eqref{eq:cond-exp} defining conditional expectation is phrased in terms of expected values, the almost-sure uniqueness cannot be strengthened to uniqueness, as re-defining a random variable on a set of probability zero does not change its expectation.
+In the following, when we say that a conditional expectation of a random variable $X$ satisfies some inequality (e.g.
+$\E[X\mid\mathcal{F}]\geq 0$) on set $L\subseteq \Omega$, we mean that for each $\mathcal{F}$-measurable function $\E[X\mid\mathcal{F}]$ satisfying~\eqref{eq:cond-exp} the inequality is satisfied for almost all $\omega \in L$.
 
@@
 
+\begin{definition}
+Let $(\Omega,\genfilt,\probm)$ be a probability space and $O\subseteq \Omega$ some set.
+ We say that sets $L_1,\dots,L_n$ form an almost-sure partition of $O$ if $L_1,\dots,L_n$ are pairwise disjoint and $\probm(\{\omega\in O\mid \omega \not \in L_1\cup\dots \cup L_n\})=0$
+\end{definition}
+
 \begin{definition}[Lexicographic Ranking Supermartingale]
@@
 %$\E[|\vecseq{X}{i}{j}|]<\infty$.
-\item For each $\omega \in \Omega$ and $i\in \Nset_0$ it holds 
-$\vec{X}_i (\omega)\geq \vec{0}$ (componentwise).
-\item For each $i\in \Nset_0$ and each $\omega\in \Omega$ with 
-$\stime(\omega)>i$ there exists $1\leq j 
-\leq n$ such that
+\item For each $\omega \in \Omega$, $i\in \Nset_0$ and $1\leq j \leq n$ it holds 
+$\vec{X}_i (\omega)[j]\geq {0}$ and $\vec{X}_i(\omega)[j]\neq \infty$.
+\item For each $i\in \Nset_0$ there exists an almost-sure partition of the set $\{\stime>i\}$ into $n$ subsets $L^i_1,\dots,L^i_n$, all of them $\genfilt_i$-measurable, such that for each $1\leq j \leq n$:
 \begin{compactitem}
 	\item $\E[\vecseq{X}{i+1}{j}\mid 
-	\genfilt_i](\omega)\leq\vecseq{X}{i}{j}(\omega) - 
-	\eps$; and
-	\item for all $1 \leq j' < j$ it holds $\E[\vecseq{X}{i+1}{j'}\mid 
-	\genfilt_i](\omega)\leq\vecseq{X}{i}{j'}(\omega)$.
+	\genfilt_i]\leq\vecseq{X}{i}{j} - 
+	\eps$ on $L_j$; and
+	\item for all $1 \leq j' < j$ we have $\E[\vecseq{X}{i+1}{j'}\mid 
+	\genfilt_i]=\vecseq{X}{i}{j'}$ on $L_{j}$.
 \end{compactitem}
@@
 For $\omega\in \Omega$ we define the level of $\omega$ at step $i$ to be the 
-number $\levelrank{\omega}{i}$ equal to either $0$, if $\stime(\omega)\leq i$, 
-or otherwise to the smallest $1\leq j \leq n$ such that 
-$\E[\vecseq{X}{i+1}{j}\mid 
-\genfilt_i](\omega)\leq\vecseq{X}{i}{j}(\omega)-\eps$. By the definition of 
+number $\levelrank{\omega}{i}$ equal to either $0$, if $\stime(\omega)\leq i$ or $\omega$ does not belong to $L^i_1 \cup \dots\cup L^i_n$, 
+or otherwise to the $1\leq j \leq n$ such that $\omega\in L^i_j$. By the definition of 
 $\eps$-LexRSM the value $\levelrank{\omega}{i}$ is well-defined for all 
@@
 that $j$ is a level of $\omega$ at infinitely many steps. Note that $\omega \in 
-\genRunSet_{\infty}$ if and only if $\minlev(\omega)\neq 0$. We denote by $M_i$ 
+\genRunSet_{\infty}$ implies $\minlev(\omega)\neq 0$, and $\probm(\genRunSet_{\infty})=\probm(\{\minlev \neq 0 \})$, as for each $i$ almost all $\omega$'s for which $\stime(\omega)>i$ belong to $L^i_1 \cup \dots\cup L^i_n$.
+We denote by $M_i$
 the set of all $\omega$'s with $\minlev(\omega)=i$.

### 2017-06-18 paper/lexicographic.tex
started section on application of LexRSMs to prob programs

@@
 In the following, when we say that a conditional expectation of a random variable $X$ satisfies some inequality (e.g.
-$\E[X\mid\mathcal{F}]\geq 0$) on set $L\subseteq \Omega$, we mean that for each $\mathcal{F}$-measurable function $\E[X\mid\mathcal{F}]$ satisfying~\eqref{eq:cond-exp} the inequality is satisfied for almost all $\omega \in L$.
+$\E[X\mid\mathcal{F}]\geq 0$) on set $L\subseteq \Omega$, we mean that for each $\mathcal{F}$-measurable function $\E[X\mid\mathcal{F}]$ satisfying~\eqref{eq:cond-exp} the .
 

### 2017-06-18 paper/prelims.tex
corrected some stuff about conditional expectation
started section on application of LexRSMs to prob programs

@@
 
-\smallskip\noindent{\em Variables and valuations.}
+\smallskip\noindent{\em Variables.}
 Throughout the paper we fix a countable set of variables $\vars$. 
@@
 constants.  A function of the form $E(\vec{x})$ for some affine expression $E$ 
-is called affine. As noted above
+is called affine. As noted above, each affine function is Borel-measurable.
 %Following the terminology of~\cite{DBLP:conf/sas/KatoenMMM10} we fix the
@@
 
-\smallskip\noindent{\em Predicates}
-A \emph{predicate} is a logical formula obtained by a finite number of applications of conjunction, disjunction and negation operations on \emph{atomic predicates} of the form $E(\vec{x})\leq E'(\vec{y})$, where $E$, $E'$ are expressions.
+\smallskip\noindent{\em Predicates.}
+A \emph{predicate} is a logical formula obtained by a finite number of applications of conjunction, disjunction and negation operations on \emph{atomic predicates} of the form $E\leq E'$, where $E$, $E'$ are expressions.
 
@@
 \end{compactitem}
+
+Following the usual probabilistic terminology, we say that \emph{almost all} $\omega$ belonging to some set $O\subseteq \Omega$ satisfy some property $\Psi$ if it holds that $\probm{(\{\omega\in O\mid \omega \text{ does not satisfy }\Psi\})}=0$.
 

### 2017-06-20 implementation/files/Parser.cpp
Handling of variable names made a bit easy

@@ -25,6 +25,8 @@ string program;
 int nVariables;
 int last_used_label = 0;
 map<int,CFG_location*> label_map;
+map<string,int> variableId;
+map<int,string> variable;
 
 void skip_spaces(int &begin, int &end){ //skips the spaces in the beginning and the end
 	while(begin<program.size() and isspace(program[begin])){
@@ -394,7 +396,7 @@ void node::proc_assgn(int s,int l){
 				constant = part(program,temp,semicolon); //Gives the distribution
 				children[1] = new node("expr",semicolon+1,end-1); //Though it is just a constant but still storing it in an expression because of consistency
 				label_map[s]->type = "det"; //<flag> Considering that it is just the same as assigning the variable the constant value
-				CFG_edge temporary_edge(label_map[l],stoi(part(program,children[0]->begin+2,children[0]->end)),children[1]);
+				CFG_edge temporary_edge(label_map[l],variableId[part(program,children[0]->begin,children[0]->end)],children[1]);
 				label_map[s]->edges.pb(temporary_edge);
 			}
 		}
@@ -402,7 +404,7 @@ void node::proc_assgn(int s,int l){
 			constant = "simple assignment";
 			children[1] = new node("expr",pos+2,end); //rexpr is nothing but an expr, a expression
 			label_map[s]->type = "det";
-			CFG_edge temporary_edge(label_map[l],stoi(part(program,children[0]->begin+2,children[0]->end)),children[1]);
+			CFG_edge temporary_edge(label_map[l],variableId[part(program,children[0]->begin,children[0]->end)],children[1]);
 			label_map[s]->edges.pb(temporary_edge);
 		}
 	}
@@ -446,15 +448,7 @@ void node::analyse_expr(int begin,int end,bool negate){
 		}
 	}
 	if(star==-1){
-		if(part(program,begin,begin+2)=="x_"){
-			if(negate){
-				expression[stoi(part(program,begin+2,end))] += -1.0;
-			}
-			else{
-				expression[stoi(part(program,begin+2,end))] += 1.0;
-			}
-		}
-		else{
+		if(isdigit(program[begin])){
 			if(negate){
 				expression[0] -= stod(part(program,begin,end));
 			}
@@ -462,32 +456,42 @@ void node::analyse_expr(int begin,int end,bool negate){
[…]

### 2017-06-20 implementation/files/Parser.h

@@ -63,10 +63,12 @@ void skip_spaces(int &begin, int &end);
 void vcopy(vector<node*> &sink,vector<node*> &tocopy);
 node* negation(node* tonegate);
 node* and_node(node* one,node* two);
-int find_variables();
+int find_variables(int& begin, int& end);
 
 extern string program;
 extern int nVariables;
 extern int last_used_label;
 extern map<int,CFG_location*> label_map;
 extern node* root;
+extern map<string,int> variableId;
+extern map<int,string> variable;

### 2017-06-20 implementation/main.cpp

@@ -131,6 +131,9 @@ int epsilons_used = 0;
 
 void generate_equations(){ //Would use the ostream file to write the equations into it later
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
+		if(it->second->invariant==NULL){
+			continue;
+		}
 		it->second->invariant->print();
 		cout<<endl;
 		if(it->second->type=="det"){
@@ -320,9 +323,9 @@ string start_invariant(){
 
 void print_fast(ostream& fastfile){
 	fastfile<<"model main{\n\n";
-	fastfile<<"\tvar x_1";
+	fastfile<<"\tvar "<<variable[1];
 	for(int i=2;i<=nVariables;++i){
-		fastfile<<", x_"<<i;
+		fastfile<<", "<<variable[i];
 	}
 	fastfile<<";\n\n";
 
@@ -348,10 +351,9 @@ void print_fast(ostream& fastfile){
 			fastfile<<";\n";
 			fastfile<<"\t\taction\t:= ";
 			if(state->edges[j].change!=NULL){
-				fastfile<<"x_"<<state->edges[j].toChange<<"' = ";
+				fastfile<<variable[state->edges[j].toChange]<<"' = ";
 				state->edges[j].change->print(fastfile);
 			}
-			//Print action here
 			fastfile<<";\n\t};\n\n";
 		}
 	}
@@ -371,7 +373,10 @@ int main(){
 		input[i]=r;
 	input[i]=0;
 	program=input;
-	nVariables = find_variables(); //To find the number of different variables of the type x_i in the program
+	int begin = 0;
+	int endprog = program.length();
+	skip_spaces(begin,endprog);
+	nVariables = find_variables(begin,endprog); //To find the number of different variables of the type x_i in the program
 	start = ++last_used_label;
 	end = ++last_used_label;
 	label_map[start] = new CFG_location("det",start);
@@ -379,9 +384,6 @@ int main(){
 	// Code for adding self loop
 	// CFG_edge last_edge(label_map[end],-1,NULL);
 	// label_map[end]->edges.pb(last_edge);
-	int begin = 0;
-	int endprog = program.length();
-	skip_spaces(begin,endprog);
 	root=new node("stmt",begin,endprog,start,end);
 	ofstream fastfile;
 	fastfile.open("files/aspic.fast");

### 2017-06-20 paper/compositional.tex
added section on compositional termination proving

\section{Compositionality of Ranking Supermartingales Revisited}
\label{sec:compositional}

Compositionality in the context of termination proving means providing the 
proof of termination step-by-step, handling one loop at a time, rather than 
attempting to construct the proof (in our case, a LexRSM) at once, by solving a 
single large system of constraints. In non-probabilistic setting, the 
relationship between lexicographic ranking functions and compositional 
termination proving is well established, as well as the importance of 
compositional proofs for efficient implementation~\cite{xxx}. In the context of 
probabilistic programs, the work~\cite{HolgerPOPL} attempted to provide a 
compositional notion of almost-sure termination proof based on the 
\emph{probabilistic variant rule (V-rule),} which we explain in a more detail 
below. 
However, as already noted in~\cite{HolgerPOPL}, the probabilistic V-rule is not 
sound in general. In order to rectify this issue, in~\cite{HolgerPOPL} they 
impose an additional constraint of \emph{uniform integrability} on ranking 
supermartingales that prove termination of individual loops, 
under which the probabilistic V-rule is shown to be sound. Apart from uniform 
integrability being somewhat restrictive in itself, in~\cite{HolgerPOPL} it is 
argued that proving uniform integrability is beyond the capability of 
state-of-the-art automated theorem provers. As a substitute for 
these,~\cite{HolgerPOPL} introduces a type system that can be used to 
automatically prove 
uniform integrability of ranking supermartingales for a restricted class of 
programs. In particular, the method cannot handle programs in which termination 
is controlled by variable that can be modified by a non-constant value, e.g. 
variable $x$ which appears in an assignment of the form $x:=x-y$. In this 
section we show that using our insights in lexicographic supermartingales we 
can obtain a subtly different notion of a probabilistic V-rule which is sound 
without any additional assumptions, and which can be used to prove termination 
of programs that the previous method cannot handle.

### 2017-06-20 implementation/files/Parser.cpp
Strict inequalities would be inputted into aspic

@@ -54,12 +54,14 @@ node::node(string t){
 	end = -1;
 	type = t;
 	constant = "";
+	strict = false;
 }
 
 node::node(string t, int b, int e, int s, int l, bool negate){
 	bracket = NULL;
 	type = t;
 	begin = b;
+	strict = false;
 	end = e;
 	skip_spaces(begin,end);
 	if(begin>end){
@@ -69,6 +71,7 @@ node::node(string t, int b, int e, int s, int l, bool negate){
 }
 
 node::node(string t, string l, bool& to_return){
+	strict = false;
 	type = t;
 	if(t=="bexpr"){
 		children.resize(1);
@@ -155,6 +158,7 @@ node* negation(node* tonegate){
 				ch->children[0] = new node("expr");
 				ch->children[0]->constant = "expression";
 				ch->children[0]->expression.resize(nVariables+1);
+				ch->strict = !(tonegate->children[i]->children[j]->strict);
 				for(int k=0;k<=nVariables;++k){
 					ch->children[0]->expression[k] = -1.0*tonegate->children[i]->children[j]->children[0]->expression[k];
 				}
@@ -583,7 +587,7 @@ bool node::proc_literal(bool negate,bool strategic, string line){
 		}
 		int sign = -1;
 		for(int i = begin;i<end-1;++i){
-			if(part(program,i,i+2)=="<=" or part(program,i,i+2)==">="){
+			if(program[i]=='<' or program[i]=='>'){
 				sign = i;
 				break;
 			}
@@ -592,20 +596,44 @@ bool node::proc_literal(bool negate,bool strategic, string line){
 			cerr<<"Invalid literal between "<<begin<<" "<<end<<endl;
 		}
 		else{
-			if(!negate){
-				constant=part(program,sign,sign+2);
-			}
-			else{
-				if(part(program,sign,sign+2)==">="){
-					constant = "<=";
+			if(program[sign+1]=='='){
+				if(!negate){
+					constant=part(program,sign,sign+2);
 				}
 				else{
-					constant = ">=";
+					if(part(program,sign,sign+2)==">="){
+						constant = "<=";
+					}
+					else{
+						constant = ">=";
+					}
 				}
+				children.resize(2);
+				children[0] = new node("expr",begin,sign);
+				children[1] = new node("expr",sign+2,end);
 			}
-			children.resize(2);
-			children[0] = new node("expr",begin,sign);
[…]

### 2017-06-20 implementation/files/Parser.h

@@ -4,6 +4,7 @@ using namespace std;
 struct node{
 	string type; //The type of entity that this node corresponds to
 	string constant; //The constant factor (subtype) of this node
+	bool strict;
 	int begin,end; //begin and end point in program
 	vector<node *> children; //children in the parse tree
 	node *bracket; //bexpr written in bracket next to this node (for stmt only)
@@ -31,7 +32,7 @@ private:
 //main process
 	void process(int s, int l, bool negate);
 public:
-	void print(ostream& outputfile = cout,string and_string = "and", string or_string = "or", string multiply_string = "");
+	void print(ostream& outputfile = cout,string and_string = "and", string or_string = "or", string multiply_string = "", bool bruteforce = false);
 };
 
 struct CFG_location;

### 2017-06-20 implementation/main.cpp

@@ -346,13 +346,13 @@ void print_fast(ostream& fastfile){
 				fastfile<<"true";
 			}
 			else{
-				state->edges[j].guard->print(fastfile,"&&","||");
+				state->edges[j].guard->print(fastfile,"&&","||","",true);
 			}
 			fastfile<<";\n";
 			fastfile<<"\t\taction\t:= ";
 			if(state->edges[j].change!=NULL){
 				fastfile<<variable[state->edges[j].toChange]<<"' = ";
-				state->edges[j].change->print(fastfile);
+				state->edges[j].change->print(fastfile,"&&","||","",true);
 			}
 			fastfile<<";\n\t};\n\n";
 		}
@@ -437,7 +437,7 @@ int main(){
 	// cout<<"Input Code:"<<endl;
 	// cout<<program<<endl;
 	// cout<<"Parse Tree:"<<endl;
-	// root->print();
+	root->print(cout,"&&","||","*",false);
 	// cout<<"CFG:"<<endl;
 	// for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
 	// 	cout<<"------------------------"<<endl;

### 2017-06-20 implementation/examples/aaron12.c
Temporary shift

@@ -9,7 +9,7 @@ int aaron12() {
 
 		} else {
 			x=x-z;
-			y=y+(z*z);
+			y=y+(z*z); // Breakage point
 			z=z-1;
 
 		}

### 2017-06-20 implementation/main.cpp

@@ -437,14 +437,14 @@ int main(){
 	// cout<<"Input Code:"<<endl;
 	// cout<<program<<endl;
 	// cout<<"Parse Tree:"<<endl;
-	root->print(cout,"&&","||","*",false);
-	// cout<<"CFG:"<<endl;
-	// for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
-	// 	cout<<"------------------------"<<endl;
-	// 	cout<<"Node "<<it->first<<endl;
-	// 	it->second->print();
-	// 	// cout<<it->second->label<<endl;
-	// }
+	// root->print(cout,"&&","||","*",false);
+	cout<<"CFG:"<<endl;
+	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
+		cout<<"------------------------"<<endl;
+		cout<<"Node "<<it->first<<endl;
+		it->second->print();
+		// cout<<it->second->label<<endl;
+	}
 	
 	generate_equations();
 	ofstream equationsfile;

### 2017-06-20 paper/compositional.tex
started to write sec ib compositionality

@@
 of programs that the previous method cannot handle.
+
+Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
+\programbody \textbf{ od}$. We denote by $\loops(\program)$ the set of all 
+loops in $\program$ there are on the top level of $\programbody$, i.e. they are 
+nested loops of $\program$ nested directly "below" the outer loop. We define a 
+depth-1 slice of $\program$ to be a \PP{} $\slice(\program)$ obtained from 
+$\program$ by removing all loops in $\loops(\program)$. A formal definition of 
+both functions is given in \AppendixMaterial, we illustrate them in the 
+following example.
+
+\textbf{[PETR: INSERT EXAMPLE]}

### 2017-06-21 paper/lexicographic-programs.tex
Added file with application of LexRSMs to programs

\section{Applying Lexicographic Supermartingales to Probabilistic Programs}

We now discuss how to leverage the mathematical results of the previous section to provide a sound proof rule for almost-sure termination of probabilistic programs. Hence, for the rest of this section we fix a \PP{} $\program$ and the associated pCFG $\pCFG_\program=(\locs,\pvars,\locinit,\vecinit,\transitions,\updates,\probdist,\guards)$.

We aim to define a function assigning a non-negative vector to each configuration (so called measurable map) such that in each point of computation, the expected value of the function after performing one more computational step is smaller (in lexicographic ordering) than the current one. This property can be formalized using the standard  notion of \emph{pre-expectation}~\cite{xxx}.

\begin{definition}[Measurable Maps and Linear Expression Maps]
A 1-dimensional \emph{measurable map} for a \PP{} $\program$ is a function $\lem$ 
assigning to each program location $\loc$ of $\pCFG_P$ a Borel-measurable function $\lem(\loc)$  of program variables, i.e. each $\lem(\loc)$  is a function of type $\Rset^{|\pvars|}\rightarrow \Rset$. As a special case, if all the functions $\lem(\loc)$ are affine, then we call $\lem$ a 1-dimensional \emph{linear expression map (LEM)}. 
Am $n$-dimensional measurable/linear expression map is a vector $\vec{\lem}=(\lem_1,\dots,\lem_n)$ of 1-dimensional measurable/linear expression maps. 
\end{definition}

Each 1-dimensional measurable map $\lem$ and location $\loc$ determines a function $\lem(\loc)$ 
which takes as an argument an $|\pvars|$-dimensional vector. We use $\lem(\loc,\vec{x})$ as a shorthand 
notation for $\lem(\loc)(\vec{x})$.

\begin{definition}[Pre-Expectation]
	Let $\lem$ 
	be 1-dimensional measurable map for a \PP{} $\program$.
	%Let $\locs$ 
	%be the set of locations of $\pCFG_P$ and $\pvars$ its set of program 
	%variables. 
	The 
	pre-expectation of $\lem$ is a function $\preexp{\lem}\colon \locs\times 
	\Rset^{|\pvars|} \rightarrow \Rset\cup\{\infty\}$ defined as follows:
	\begin{compactitem} %\itemsep1pt \parskip0pt \parsep0pt
		\item 
		if $\loc$ is a probabilistic branching location, then
		$$\preexp{\lem}(\loc,\vec{x}):=\sum_{(\loc,\loc')\in\transitions} 
		Pr_{\loc}\left(\loc,\loc'\right)\cdot
		\lem(\loc',\vec{x});$$
		\item 
		if $\loc$ is a non-deterministic branching location, then
		$$\preexp{\lem}(\loc,\vec{x}):=
		\max_{(\loc,\loc')\in\transitions}\lem(\loc',\vec{x});$$

		%\item 
		%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\min_{(\loc,id,\ell')\in\transitions}\eta\left(\loc',\mathbf{x}\right)$
		% if $\loc$ is an angelic location;
		\item 
		if $\loc$ is a deterministic location, then for each $\vec{x}$ the value 
		$\preexp{\lem}(\loc,\vec{x})$ is determined as follows: there is exactly one 
		transition
		$\tau=(\loc,\loc')$ such that $\vec{x}\models G(\tau)$. We put $$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x})$$.
		\item 
		If $\loc$ is an assignment location, then there is exactly one transition $\tau=(\loc,\loc')$ outgoing from $\loc$.
		We distinguish 
		three cases, depending on $\updates(\tau)=(j,\up)$ (recall that $\up$ is an update element):
		\begin{compactitem}
			\item If $\up\colon \Rset^{|\pvars|}\rightarrow \Rset$ is a Borel-measurable function, then 
			$$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x}(j\leftarrow \up(\vec{x}))).$$
			\item If $\up$ is a distribution $d$, then $$ \preexp{\lem}(\loc,\vec{x}):= 
			\lem(\loc',\vec{x}(j\leftarrow \E[d])),$$ where $\E[d]$ is the expected value of the 
			distribution $d$.
			\item 
			If $\up$ is a set, then $$ \preexp{\lem}(\loc,\vec{x}):= \sup_{a\in\up}
			\lem(\loc',\vec{x}(j \leftarrow a)).$$
		\end{compactitem}
		%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\eta(\loc',\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right))$
		% if $\loc$ is a deterministic location, $(\loc,f,\loc')\in\transitions$ and 
		%$\mathbf{x}\in G(\loc,f,\loc')$, where 
		%$\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right)$ is the expected value of 
		%$f(\vec{x},\cdot)$. %on $\rvars$ if $\vec{x}$ is treated as a constant vector.
	\end{compactitem}
	%\end{definition}
\end{definition}

Intuitively, a pre-expectation of $\lem$ is a 
function which for each configuration $(\loc,\vec{x})$ returns the maximal
expected value of $\lem$ after one step is made from this configuration, where 
the maximum is taken over all possible non-deterministic choices.
Note that the only way in which pre-expectation can attain an infinite value is due to non-deterministic assignment. However, if the program in question has \emph{bounded non-determinism}, in the sense that all non-deterministic assignments can only choose the assignment value from a bounded set\footnote{A set $u \subseteq \mathcal{R}$ is bounded if it is contained in some interval of a finite length}, then the pre-expectation of any measurable map is a real-valued function.

As in termination analysis of non-probabilistic programs, our LexRSMs are typically supported by \emph{invariants}, i.e. overapproximations of the set of reachable configuration.

\begin{definition}[Invariant Map and Linear Invariant Map]
An \emph{invariant map} for a \PP{} $\program$ is a function $\inv$ assigning to each location of $\pCFG_{\program}$ a Borel-measurable set $\inv({\loc})\subseteq \Rset^{|\pvars|}$ of variable valuations, so called invariant of $\loc$, such that for each configuration $(\loc,\vec{x})$ reachable from the initial configuration it holds $\vec{x}\in \inv(\loc)$. Additionally, if each set $\inv(\loc)$ is of the form $\{\vec{x}\mid\vec{x}\models \Psi^\ell \}$ for some propositionally linear predicate $\Psi^\ell$, then we call $\inv$ a \emph{linear invariant map} (LIM).
\end{definition}

Slightly abusing the notation, we view each LIM equivalently as a function assigning linear predicates (whose satisfaction sets overapproximate the set of reachable valuations) to program locations.

We now have all the ingredients needed to define the notion of LexRSM maps for probabilistic programs.

\begin{definition}[Lexicographic Ranking Supermartingale Map]
Let $\eps>0$. An $n$-dimensional \emph{lexicographic $\eps$-ranking supermartingale map} ($\eps$-LexRSM map) for a program $\program$ supported by an invariant map $\inv$ is an $n$-dimensional measurable map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for $\program$ such that for each configuration $(\loc,\vec{x})$ where $\loc\neq \locterm$ and $\vec{x}\in \inv(\loc)$ the following conditions are satisfied:
 \begin{compactitem}
 	%\item For all $(\loc,\vec{x})\in \locs\times\Rset^{|\pvars|}$ it holds
 	%\item For all  it holds 
 	\item
 	for all $1\leq j \leq n$, $\lem_j(\loc,\vec{x})\geq 0$; and
 	\item 
 	there exists $1\leq j \leq$ n such that
 	\begin{compactitem}
 	\item
 	$\preexp{\lem_j}(\loc,\vec{x}) \leq \lem_j(\loc,\vec{x})-\eps$, and
 	\item
 	for all $1\leq j'<j$ it holds 
 	$\preexp{\lem_{j'}}(\loc,\vec{x}) \leq \lem_{j'}(\loc,\vec{x})$.
 	\end{compactitem}
 \end{compactitem}
If additionally $\lem$ is a linear expression map, then we call it a linear $\eps$-LexRSM map ($\eps$-LinLexRSM).
\end{definition}

\textbf{[PETR: DEFINE LOCTERM]}

### 2017-06-21 paper/lexicographic-programs.tex
expanded section on applying LexRSM to programs

@@
 \begin{definition}[Measurable Maps and Linear Expression Maps]
-A 1-dimensional \emph{measurable map} for a \PP{} $\program$ is a function $\lem$ 
+A 1-dimensional \emph{measurable map} for a \PP{} $\program$ is a  
+real-valued function $\lem$ 
 assigning to each program location $\loc$ of $\pCFG_P$ a Borel-measurable function $\lem(\loc)$  of program variables, i.e.
@@
 the maximum is taken over all possible non-deterministic choices.
-Note that the only way in which pre-expectation can attain an infinite value is due to non-deterministic assignment.
-However, if the program in question has \emph{bounded non-determinism}, in the sense that all non-deterministic assignments can only choose the assignment value from a bounded set\footnote{A set $u \subseteq \mathcal{R}$ is bounded if it is contained in some interval of a finite length}, then the pre-expectation of any measurable map is a real-valued function.
+%Note that the only way in which pre-expectation can attain an infinite value 
+%is 
+%due to non-deterministic assignment. However, if the program in question has 
+%\emph{bounded non-determinism}, in the sense that all non-deterministic 
+%assignments can only choose the assignment value from a bounded set\footnote{A 
+%set $u \subseteq \mathcal{R}$ is bounded if it is contained in some interval 
+%of 
+%a finite length}, then the pre-expectation of any real-valued measurable map 
+%is 
+%a real-valued function.
 
@@
 
-\textbf{[PETR: DEFINE LOCTERM]}
+\textbf{[PETR: DEFINE LOCATION TERMINATION]}
+
+The main result is the soundness of $\eps$-LexRSM maps for proving a.s. 
+termination.
+
+\begin{theorem}
+\label{thm:lexrsm-programs}
+Let $\program$ be a probabilistic program. Assume that there exists an $\eps>0$ 
+and an $n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for 
+$\program$ supported 
+by some 
+invariant map $\inv$. 
+Then $\program$ terminates almost surely.
+\end{theorem}
+\begin{proof}
+Let $\sigma$ be any measurable scheduler in $\program$.
+We define an $n$-dimensional stochastic process 
+$\{\vec{X}_{i}\}_{i=0}^{\infty} $ on the probability space 
+$(\OmegaRun,\natfilt,\probm^{\sigma})$ such 
+that for each 
+$i\geq 0$ and $1\leq j 
+\leq n$ and each run $\run$ we put $\vecseq{X}{i}{j}(\run) = 
+\lem_j(\cfg{\sigma}{i}(\run))$. We claim that $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
+is an $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
+the process is real-valued, componentwise non-negative, and adapted to the 
+canonical filtration of $\natfilt$. It remains to prove that condition (3) in 
+Definition~\ref{def:lexrsm} is satisfied. To this end, for each $i\geq 0$ we 
+define an almost-sure partition of the set $\{\run\in \OmegaRun\mid 
+\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting 
+\end{proof}

### 2017-06-21 paper/lexicographic.tex
expanded section on applying LexRSM to programs

@@
 \item $\E[|X|]<\infty$; or
-\item $X$ is non-negative and for all $\omega\in\Omega$ it holds $X(\omega)\neq \infty$.
+\item $X$ is real-valued and non-negative.
 \end{itemize}
@@
 \begin{definition}[Lexicographic Ranking Supermartingale]
+\label{def:lexrsm}
 Let $(\Omega,\genfilt,\probm)$ be a probability space, 
@@
 $\eps\geq 0$. 
-An $n$-dimensional stochastic process $\{\vec{X}_{i}\}_{i=0}^{\infty}$ is a 
+An $n$-dimensional real-valued stochastic process 
+$\{\vec{X}_{i}\}_{i=0}^{\infty}$ is a 
 \emph{lexicographic $\eps$-ranking supermartingale for $\stime$ 
@@
 \item For each $\omega \in \Omega$, $i\in \Nset_0$ and $1\leq j \leq n$ it holds 
-$\vec{X}_i (\omega)[j]\geq {0}$ and $\vec{X}_i(\omega)[j]\neq \infty$.
+$\vec{X}_i (\omega)[j]\geq {0}$. % and $\vec{X}_i(\omega)[j]\neq \infty$.
 \item For each $i\in \Nset_0$ there exists an almost-sure partition of the set $\{\stime>i\}$ into $n$ subsets $L^i_1,\dots,L^i_n$, all of them $\genfilt_i$-measurable, such that for each $1\leq j \leq n$:

### 2017-06-21 implementation/files/Parser.cpp
temp

@@ -27,6 +27,7 @@ int last_used_label = 0;
 map<int,CFG_location*> label_map;
 map<string,int> variableId;
 map<int,string> variable;
+map<string,int> stateId;
 
 void skip_spaces(int &begin, int &end){ //skips the spaces in the beginning and the end
 	while(begin<program.size() and isspace(program[begin])){
@@ -913,7 +914,7 @@ void CFG_edge::print(){
 		cout<<"Error! Error! Error!"<<endl;
 	}
 	if(change!=NULL){
-		cout<<"Change: x_"<<toChange<<" changed to ";
+		cout<<"Change: "<<variable[toChange]<<" changed to ";
 		change->print();
 		cout<<endl;
 	}

### 2017-06-21 implementation/files/Parser.h

@@ -61,6 +61,7 @@ public:
 };
 
 void skip_spaces(int &begin, int &end);
+void skip_spaces(string&);
 void vcopy(vector<node*> &sink,vector<node*> &tocopy);
 node* negation(node* tonegate);
 node* and_node(node* one,node* two);
@@ -73,3 +74,4 @@ extern map<int,CFG_location*> label_map;
 extern node* root;
 extern map<string,int> variableId;
 extern map<int,string> variable;
+extern map<string,int> stateId;

### 2017-06-21 implementation/files/invariant_script2.sh

@@ -1,0 +1,2 @@
+./files/aspicV3.4 -print inline files/aspic.fst
+rm aspic.log

### 2017-06-21 implementation/main.cpp

@@ -227,6 +227,7 @@ void generate_equations(){ //Would use the ostream file to write the equations i
 int last_used_lambda = 0;
 
 void print_equations(ostream& equationsfile){
+	last_used_lambda = 0;
 	equationsfile<<"maximize ";
 	for(int i=1;i<=epsilons_used;++i){
 		if(equations.find(i)!=equations.end()){
@@ -438,13 +439,13 @@ int main(){
 	// cout<<program<<endl;
 	// cout<<"Parse Tree:"<<endl;
 	// root->print(cout,"&&","||","*",false);
-	cout<<"CFG:"<<endl;
-	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
-		cout<<"------------------------"<<endl;
-		cout<<"Node "<<it->first<<endl;
-		it->second->print();
-		// cout<<it->second->label<<endl;
-	}
+	// cout<<"CFG:"<<endl;
+	// for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
+	// 	cout<<"------------------------"<<endl;
+	// 	cout<<"Node "<<it->first<<endl;
+	// 	it->second->print();
+	// 	// cout<<it->second->label<<endl;
+	// }
 	
 	generate_equations();
 	ofstream equationsfile;

### 2017-06-21 implementation/main2.cpp

@@ -1,0 +1,556 @@
+// Do I have to add a self loop on the end CFG location, because then the epsilon equation creates issues
+// end node will always be 2
+/*
+Comments:
+	Considering start invariant and neglecting all other invariants, also the start invariant should be a polyhedra(Which I think would not pose a problem)
+	https://tapas.labri.fr/wp/wp-content/uploads/2017/02/FAST-manual.pdf
+*/
+#include <iostream>
+#include <algorithm>
+#include <cstdio>
+#include <vector>
+#include <cstring>
+#include <sstream>
+#include <cctype>
+#include <set>
+#include <map>
+#include <fstream>
+#include <deque>
+#include <iomanip>
+#include <fstream>
+#include <queue>
+#include "files/Parser.h"
+using namespace std;
+#define MAXL 100000 //Maximum length of the program
+#define part(x,a,b) (x.substr((a),((b)-(a))))
+#define pb push_back
+#define A(i,j) (front->affexpr->children[i]->children[0]->expression[j+1])
+#define b(i) (-1.0*front->affexpr->children[i]->children[0]->expression[0])
+
+stringstream buffer;
+
+struct cond{
+	bool strict;
+	vector<string> c;
+	string negative_d; //negative component of d
+	cond(int equation_count,int toChange,node* change,int src,int dest1,int dest2 = -1,double probability = -1.0){
+		strict = false;
+		c.resize(nVariables);
+		if(dest2==-1){
+			if(change==NULL){
+				buffer.clear();
+				buffer.str(string());
+				buffer<<"eps"<<equation_count<<"-f_"<<src<<"_0+f_"<<dest1<<"_0";
+				negative_d = buffer.str();
+				for(int i=0;i<nVariables;i++){
+					buffer.clear();
+					buffer.str(string());
+					buffer<<"f_"<<dest1<<"_"<<i+1<<"-f_"<<src<<"_"<<i+1;
+					c[i] = buffer.str();
+				}
+			}
+			else{
+				buffer.clear();
+				buffer.str(string());
+				buffer<<"eps"<<equation_count<<"-f_"<<src<<"_0+f_"<<dest1<<"_0";
+				if(change->expression[0]<0.0){
+					buffer<<change->expression[0]<<"f_"<<dest1<<"_"<<toChange;
+				}
+				else if(change->expression[0]>0.0){
+					buffer<<"+"<<change->expression[0]<<"f_"<<dest1<<"_"<<toChange;
+				}
+				negative_d = buffer.str();
[…]

### 2017-06-22 implementation/files/Parser.cpp
reverting some changes

@@ -27,7 +27,6 @@ int last_used_label = 0;
 map<int,CFG_location*> label_map;
 map<string,int> variableId;
 map<int,string> variable;
-map<string,int> stateId;
 
 void skip_spaces(int &begin, int &end){ //skips the spaces in the beginning and the end
 	while(begin<program.size() and isspace(program[begin])){
@@ -914,7 +913,7 @@ void CFG_edge::print(){
 		cout<<"Error! Error! Error!"<<endl;
 	}
 	if(change!=NULL){
-		cout<<"Change: "<<variable[toChange]<<" changed to ";
+		cout<<"Change: x_"<<toChange<<" changed to ";
 		change->print();
 		cout<<endl;
 	}

### 2017-06-22 implementation/files/Parser.h

@@ -61,7 +61,6 @@ public:
 };
 
 void skip_spaces(int &begin, int &end);
-void skip_spaces(string&);
 void vcopy(vector<node*> &sink,vector<node*> &tocopy);
 node* negation(node* tonegate);
 node* and_node(node* one,node* two);
@@ -74,4 +73,3 @@ extern map<int,CFG_location*> label_map;
 extern node* root;
 extern map<string,int> variableId;
 extern map<int,string> variable;
-extern map<string,int> stateId;

### 2017-06-22 implementation/files/invariant_script2.sh

@@ -1,2 +1,0 @@
-./files/aspicV3.4 -print inline files/aspic.fst
-rm aspic.log

### 2017-06-22 implementation/main.cpp

@@ -227,7 +227,6 @@ void generate_equations(){ //Would use the ostream file to write the equations i
 int last_used_lambda = 0;
 
 void print_equations(ostream& equationsfile){
-	last_used_lambda = 0;
 	equationsfile<<"maximize ";
 	for(int i=1;i<=epsilons_used;++i){
 		if(equations.find(i)!=equations.end()){
@@ -439,13 +438,13 @@ int main(){
 	// cout<<program<<endl;
 	// cout<<"Parse Tree:"<<endl;
 	// root->print(cout,"&&","||","*",false);
-	// cout<<"CFG:"<<endl;
-	// for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
-	// 	cout<<"------------------------"<<endl;
-	// 	cout<<"Node "<<it->first<<endl;
-	// 	it->second->print();
-	// 	// cout<<it->second->label<<endl;
-	// }
+	cout<<"CFG:"<<endl;
+	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
+		cout<<"------------------------"<<endl;
+		cout<<"Node "<<it->first<<endl;
+		it->second->print();
+		// cout<<it->second->label<<endl;
+	}
 	
 	generate_equations();
 	ofstream equationsfile;

### 2017-06-22 implementation/main2.cpp

@@ -1,556 +1,0 @@
-// Do I have to add a self loop on the end CFG location, because then the epsilon equation creates issues
-// end node will always be 2
-/*
-Comments:
-	Considering start invariant and neglecting all other invariants, also the start invariant should be a polyhedra(Which I think would not pose a problem)
-	https://tapas.labri.fr/wp/wp-content/uploads/2017/02/FAST-manual.pdf
-*/
-#include <iostream>
-#include <algorithm>
-#include <cstdio>
-#include <vector>
-#include <cstring>
-#include <sstream>
-#include <cctype>
-#include <set>
-#include <map>
-#include <fstream>
-#include <deque>
-#include <iomanip>
-#include <fstream>
-#include <queue>
-#include "files/Parser.h"
-using namespace std;
-#define MAXL 100000 //Maximum length of the program
-#define part(x,a,b) (x.substr((a),((b)-(a))))
-#define pb push_back
-#define A(i,j) (front->affexpr->children[i]->children[0]->expression[j+1])
-#define b(i) (-1.0*front->affexpr->children[i]->children[0]->expression[0])
-
-stringstream buffer;
-
-struct cond{
-	bool strict;
-	vector<string> c;
-	string negative_d; //negative component of d
-	cond(int equation_count,int toChange,node* change,int src,int dest1,int dest2 = -1,double probability = -1.0){
-		strict = false;
-		c.resize(nVariables);
-		if(dest2==-1){
-			if(change==NULL){
-				buffer.clear();
-				buffer.str(string());
-				buffer<<"eps"<<equation_count<<"-f_"<<src<<"_0+f_"<<dest1<<"_0";
-				negative_d = buffer.str();
-				for(int i=0;i<nVariables;i++){
-					buffer.clear();
-					buffer.str(string());
-					buffer<<"f_"<<dest1<<"_"<<i+1<<"-f_"<<src<<"_"<<i+1;
-					c[i] = buffer.str();
-				}
-			}
-			else{
-				buffer.clear();
-				buffer.str(string());
-				buffer<<"eps"<<equation_count<<"-f_"<<src<<"_0+f_"<<dest1<<"_0";
-				if(change->expression[0]<0.0){
-					buffer<<change->expression[0]<<"f_"<<dest1<<"_"<<toChange;
-				}
-				else if(change->expression[0]>0.0){
-					buffer<<"+"<<change->expression[0]<<"f_"<<dest1<<"_"<<toChange;
-				}
-				negative_d = buffer.str();
[…]

### 2017-06-22 paper/lexicographic-programs.tex

@@
 define an almost-sure partition of the set $\{\run\in \OmegaRun\mid 
-\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting 
+\ttime(\run) >i\}$ into sets $L^{i}_1,\dots,L^{i}_n$ by putting $L^i_j$ to be 
+the set of all runs $\run$ such that $\ttime(\run)>i$ and for $\run$ the index 
+$j$ is the smallest one such that $\preexp{\lem_j}(\cfg{\sigma}{i}(\run)) \leq 
+\lem_j(\cfg{\sigma}{i}(\run))-\eps$ (which entails 
+$\preexp{\lem_{j'}}(\cfg{\sigma}{i}(\run)) =
+\lem_{j'}(\cfg{\sigma}{i}(\run))-\eps$ for all $1\leq j'< j$). Due to 
+definition of an $\eps$-LexRSM map such a $j$ exists for all 
+$\run\in\{\ttime>i\}$ and hence we indeed have a partition. It remains to prove 
+that irrespective of the initial choice of $\sigma$ it holds, for each $1\leq j 
+\leq n$, that $\E^\sigma[\vecseq{X}{i+1}{j}\mid 
+\natfilt_i]\leq\preexp{\lem_{j}}(\cfg{\sigma}{}) $. This can be achieved by a 
+standard~\cite{xxx}, through somewhat technical computation, which we defer to 
+\AppendixMaterial.
+
+Since  $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
+is an $\eps$-LexRSM for $\ttime$, from Theorem~\ref{thm:lexrsm-programs} it 
+follows that $\probm^{\sigma}\ttime <\infty) =1$, irrespective of scheduler 
+$\sigma$.
 \end{proof}

### 2017-06-22 paper/prelims.tex

@@
 a function such that for every $a\in \Rset \cup \{ \infty\}$ the set
-$\{\omega\in \Omega\mid R(\omega)\leq a\}$ belongs to $\mathcal{F}$.
+$\{\omega\in \Omega\mid R(\omega)\leq a\}$ belongs to $\mathcal{F}$. If 
+$R(\omega)\in \Rset$ for all $\omega\in \Omega$, we say that $R$ is 
+\emph{real-valued.}
 We denote by $\expv[R]$ the \emph{expected value} of a random variable $X$~(see \cite[Chapter 5]{Billingsley:book}
@@
 probabilistic 
-space $(\Omega,\natfilt,\probm^{\sigma})$ in which $\Omega$ is a set of all 
+space $(\OmegaRun,\natfilt,\probm^{\sigma})$ in which $\OmegaRun$ is a set of 
+all 
 runs in $\pCFG$, and a stochastic process 
@@
 such 
-that for each run $\run\in \Omega$ we have that $\cfg{\sigma}{i}(\run)$ is the 
+that for each run $\run\in \OmegaRun$ we have that $\cfg{\sigma}{i}(\run)$ is 
+the 
 $i$-th configuration on $\run$ (i.e., $\cfg{\sigma}{i}$ is a random vector 
@@
 set of configurations $\confset\in\mathcal{F}_{\Conf}$ it holds 
-$\{\omega\mid\cfg{\sigma}{i}(\omega)\in \confset\}\in \natfilt$). Equivalently, 
+$\{\run\mid\cfg{\sigma}{i}(\run)\in \confset\}\in \natfilt$). Equivalently, 
 $\natfilt$ can be defined as a sigma algebra generated by all set of runs of 
@@
 We denote by $\E^\sigma$ the expectation operator in probability space 
-$(\Omega,\natfilt,\probm^{\sigma})$.
+$(\OmegaRun,\natfilt,\probm^{\sigma})$.
 

### 2017-06-22 paper/compositional.tex
preparing to adjust the definition of pCFG

@@
 loops in $\program$ there are on the top level of $\programbody$, i.e. they are 
-nested loops of $\program$ nested directly "below" the outer loop. We define a 
-depth-1 slice of $\program$ to be a \PP{} $\slice(\program)$ obtained from 
-$\program$ by removing all loops in $\loops(\program)$. A formal definition of 
+nested loops of $\program$ nested directly "below" the outer loop. We define an 
+un-nested slice of $\program$ to be a \PP{} $\slice(\program)$ obtained from 
+$\program$ by removing all loops in $\loops(\program)$ and replacing them with 
+skip statements. We also define the set 
+$\alloops(\program)$ of all loops in $\program$, i.e. 
+$\alloops(\program):=\loops(\program)\cup \{\loops(\program')\mid \program'\in 
+\loops(\program)\}$. A 
+formal definition of 
 both functions is given in \AppendixMaterial, we illustrate them in the 
@@
 \textbf{[PETR: INSERT EXAMPLE]}
+
+\textbf{[PETR: INTRODUCE SETS OF INITIAL PROGRAM VALUATIONS RATHER THAN A 
+CONCRETE VALUATION]}
+
+Each measurable map for $\lem$ can be, in a natural way, restricted to a 
+measurable map $\lem_{\program'}$ for $\program'\in \loops(\program)$, 
+since each pCFG $\pCFG_{\program'}$ is a sub-pCFG of $\pCFG_{\program}$. 
+Similarly, a measurable map $\lem$ for $\program$ can be restricted to a 
+measurable map $\lem_{\slice}$ for $\slice(\program)$, since 
+$\pCFG_{\slice(\program)}$ is produced from $\pCFG_{\program}$ essentially by 
+removing certain locations (provided that we identify the locations forming 
+heads of nested loops with locations preceding skip statements), and 
+$\lem_{\slice}$ is defined to be equal to $\lem$ in all the locations that 
+remained.

### 2017-06-22 paper/lexicographic-programs.tex
changed definition of pCFG to allow for multiple initial valuations

@@
 
-We now discuss how to leverage the mathematical results of the previous section to provide a sound proof rule for almost-sure termination of probabilistic programs.
-Hence, for the rest of this section we fix a \PP{} $\program$ and the associated pCFG $\pCFG_\program=(\locs,\pvars,\locinit,\vecinit,\transitions,\updates,\probdist,\guards)$.
+We now discuss how to leverage the mathematical results of the previous section 
+to provide a sound proof rule for almost-sure termination of probabilistic 
+programs. Hence, for the rest of this section we fix a \PP{} $\program$ and the 
+associated pCFG 
+$\pCFG_\program=(\locs,\pvars,\locinit,\vecinitset,\transitions,\updates,\probdist,\guards)$.
 
@@
 \begin{proof}
-Let $\sigma$ be any measurable scheduler in $\program$.
+Let $\sigma$ be any measurable scheduler and $\vecinit\in\vecinitset$ any 
+initial variable valuation in $\program$.
 We define an $n$-dimensional stochastic process 
 $\{\vec{X}_{i}\}_{i=0}^{\infty} $ on the probability space 
-$(\OmegaRun,\natfilt,\probm^{\sigma})$ such 
+$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ such 
 that for each 
@@
 $\run\in\{\ttime>i\}$ and hence we indeed have a partition. It remains to prove 
-that irrespective of the initial choice of $\sigma$ it holds, for each $1\leq j 
-\leq n$, that $\E^\sigma[\vecseq{X}{i+1}{j}\mid 
+that irrespective of the initial choice of $\sigma$ and $\vecinit$ it holds, 
+for each $1\leq j 
+\leq n$, that $\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
 \natfilt_i]\leq\preexp{\lem_{j}}(\cfg{\sigma}{}) $. This can be achieved by a 
@@
 is an $\eps$-LexRSM for $\ttime$, from Theorem~\ref{thm:lexrsm-programs} it 
-follows that $\probm^{\sigma}\ttime <\infty) =1$, irrespective of scheduler 
-$\sigma$.
+follows that $\probm^{\sigma}_{\vecinit}\ttime <\infty) =1$, irrespective of 
+$\sigma$ and $\vecinit$.
 \end{proof}

### 2017-06-23 implementation/c_wrapper.cpp
Benchmarking started

@@ -1,0 +1,61 @@
+// Do I have to add a self loop on the end CFG location, because then the epsilon equation creates issues
+// end node will always be 2
+/*
+Comments:
+	Considering start invariant and neglecting all other invariants, also the start invariant should be a polyhedra(Which I think would not pose a problem)
+	https://tapas.labri.fr/wp/wp-content/uploads/2017/02/FAST-manual.pdf
+*/
+#include <iostream>
+#include <algorithm>
+#include <cstdio>
+#include <vector>
+#include <cstring>
+#include <sstream>
+#include <cctype>
+#include <set>
+#include <map>
+#include <fstream>
+#include <deque>
+#include <iomanip>
+#include <fstream>
+#include <queue>
+#include "files/Parser.h"
+using namespace std;
+#define MAXL 100000 //Maximum length of the program
+#define part(x,a,b) (x.substr((a),((b)-(a))))
+#define pb push_back
+
+
+int main(){
+	int start,end;
+	char input[MAXL];
+	// Setting precision to the printing of the double variables in program
+	// cout<<fixed<<setprecision(10);
+	int r,i;
+	for(i=0;(r=getchar())!=EOF;i++)
+		input[i]=r;
+	input[i]=0;
+	program=input;
+	int begin = 0;
+	int endprog = program.length();
+	skip_spaces(begin,endprog);
+	nVariables = find_variables(begin,endprog); //To find the number of different variables of the type x_i in the program
+	start = ++last_used_label;
+	end = ++last_used_label;
+	label_map[start] = new CFG_location("det",start);
+	label_map[end]	 = new CFG_location("det",end);
+	// Code for adding self loop
+	// CFG_edge last_edge(label_map[end],-1,NULL);
+	// label_map[end]->edges.pb(last_edge);
+	root=new node("stmt",begin,endprog,start,end);
+	cout<<"int main(){\n\n";
+	cout<<"int "<<variable[1];
+	for(int i=2;i<=nVariables;i++){
+		cout<<","<<variable[i];
+	}
+	// Ask Petr <flag> whether all the variables are integers or double because otherwise double causes problems for c2fsm
+	cout<<";\n";
+	root->print(cout,"&&","||","*",true);
+	cout<<"\n}\n";
+	return 0;
+}

### 2017-06-23 implementation/files/Parser.cpp

@@ -868,25 +868,63 @@ void node::print(ostream& outputfile, string and_string, string or_string, strin
 		}
 		return;
 	}
-	// <flag>
-	// outputfile<<"Add: "<<this<<"\t";
-	// outputfile<<"Type: "<<type<<"\t";
-	// outputfile<<"Range: ["<<begin<<", "<<end<<")\t";
-	// outputfile<<"Const: |"<<constant<<"|"<<endl;
-	// outputfile<<"Children: ";
-	// for(int i = 0;i<children.size();++i){
-	// 	outputfile<<children[i]<<"\t";
-	// }
-	if(bracket!=NULL){
-		// outputfile<<endl<<"Bracket: "<<bracket<<endl;
-		bracket->print(outputfile,and_string,or_string,multiply_string,bruteforce);
-		outputfile<<endl;
+	else if(type=="ndbexpr"){
+		if(constant=="single bexpr"){
+			children[0]->print(outputfile,and_string,or_string,multiply_string,bruteforce);
+		}
+		else{
+			// <flag>
+		}
 	}
-	for(int i=0;i<children.size();++i){
-		children[i]->print(outputfile,and_string,or_string,multiply_string,bruteforce);
+	else if(type=="stmt"){
+		if(constant=="skip"){
+			// Do nothing
+		}
+		else if(constant=="several statements"){
+			children[0]->print(outputfile,and_string,or_string,multiply_string,bruteforce);
+			outputfile<<endl;
+			children[1]->print(outputfile,and_string,or_string,multiply_string,bruteforce);
+		}
+		else if(constant=="while"){
+			outputfile<<"while(";
+			children[0]->print(outputfile,and_string,or_string,multiply_string,bruteforce);
+			outputfile<<"){\n\n";
+			children[1]->print(outputfile,and_string,or_string,multiply_string,bruteforce);
+			outputfile<<"\n}";
+		}
+		else if(constant=="if"){
+			outputfile<<"if(";
+			children[0]->print(outputfile,and_string,or_string,multiply_string,bruteforce);
+			outputfile<<"){\n\n";
+			children[1]->print(outputfile,and_string,or_string,multiply_string,bruteforce);
+			outputfile<<"\n}\nelse{\n\n";
+			children[2]->print(outputfile,and_string,or_string,multiply_string,bruteforce);
+			outputfile<<"\n}";
+		}
+		else if(constant=="single assgn"){
[…]

### 2017-06-23 implementation/files/Parser.h

@@ -17,15 +17,15 @@ private:
 //procs
 	void proc_stmt(int s,int l);
 	void proc_assgn(int s,int l);
-	void proc_affexpr();
+	void proc_affexpr(); //
 	void proc_pvar();
 	void analyse_expr(int begin,int end,bool negate);
 	void analyse_expr(string l,bool negate);
-	void proc_expr();
-	void proc_expr(string l);
+	void proc_expr(); //
+	void proc_expr(string l); //
 	void proc_constant();
 	bool proc_literal(bool negate, bool strategic = true, string line = "");
-	void proc_bexpr();
+	void proc_bexpr(); //
 	void proc_ndbexpr();
 	void form_vector(int begin,int end,bool negate);
 	void form_vector(string l,bool negate);

### 2017-06-23 implementation/files/invariant_script.sh

@@ -1,4 +1,4 @@
-./files/aspicV3.4 files/aspic.fast > files/InvariantOutput
+./files/aspicV3.4 -delay 10 files/aspic.fast > files/InvariantOutput
 awk 'BEGIN {x=0} x==1{print} /^\*\*\* Results :/{x=1}' < files/InvariantOutput > InvariantOutput.tmp
 mv InvariantOutput.tmp files/InvariantOutput
 rm aspic.log

### 2017-06-23 implementation/main.cpp

@@ -321,6 +321,8 @@ string start_invariant(){
 	}
 }
 
+
+
 void print_fast(ostream& fastfile){
 	fastfile<<"model main{\n\n";
 	fastfile<<"\tvar "<<variable[1];
@@ -346,7 +348,14 @@ void print_fast(ostream& fastfile){
 				fastfile<<"true";
 			}
 			else{
-				state->edges[j].guard->print(fastfile,"&&","||","",true);
+				node* temp;
+				if(label_map[i]->invariant!=NULL){
+					temp = and_node(state->edges[j].guard,label_map[i]->invariant);
+				}
+				else{
+					temp = state->edges[j].guard;
+				}
+				temp->print(fastfile,"&&","||","",true);
 			}
 			fastfile<<";\n";
 			fastfile<<"\t\taction\t:= ";
@@ -394,7 +403,7 @@ int main(){
 		cout<<"Something wrong with the script to find invariants";
 	}
 	
-	// Code for analysing files/InvariantOutput comes here
+	// // Code for analysing files/InvariantOutput comes here
 	ifstream invariant_file("files/InvariantOutput");
 	for(int i=1;i<=last_used_label;){
 		string line;
@@ -405,7 +414,6 @@ int main(){
 			string invariant_string = part(line,open+1,close);
 			if(invariant_string=="true"){
 				if(label_map[i]->invariant==NULL){
-					// label_map[i]->invariant = new node("true"); //Effectively but add 0<=0 instead
 					label_map[i]->invariant = new node("bexpr");
 					node* concerned_node = label_map[i]->invariant;
 					concerned_node->children.resize(1);
@@ -430,21 +438,19 @@ int main(){
 			}
 			i++;
 		}
-	}
-
-	
+	}	
 	// // Code to print the tree structure etc.
 	// cout<<"Input Code:"<<endl;
 	// cout<<program<<endl;
 	// cout<<"Parse Tree:"<<endl;
 	// root->print(cout,"&&","||","*",false);
-	cout<<"CFG:"<<endl;
-	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
-		cout<<"------------------------"<<endl;
-		cout<<"Node "<<it->first<<endl;
-		it->second->print();
-		// cout<<it->second->label<<endl;
-	}
+	// cout<<"CFG:"<<endl;
[…]

### 2017-06-23 paper/compositional.tex

@@
 $\alloops(\program):=\loops(\program)\cup \{\loops(\program')\mid \program'\in 
-\loops(\program)\}$. A 
-formal definition of 
+\loops(\program)\}$. The above definitions do not specify the initialization 
+preamble for the sub-programs $\program'\in\alloops(\program)$: we can deem 
+this pre-amble empty, i.e. consider these sub-programs with any initial 
+valuation, or, provided that some invariant map $\inv$ for $\program$ is 
+given, we can set the set of possible initial valuations of a sub-program 
+$\program'$ to be the set $\inv(\loc^{\lin}_{\program'})$, where 
+$\inv(\loc^{\lin}_{\program'})$ is the location of $\program$ corresponding to 
+the head of the loop $\program'$.
+A formal definition of 
 both functions is given in \AppendixMaterial, we illustrate them in the 
@@
 \textbf{[PETR: INSERT EXAMPLE]}
-
-\textbf{[PETR: INTRODUCE SETS OF INITIAL PROGRAM VALUATIONS RATHER THAN A 
-CONCRETE VALUATION]}
 
@@
 $\lem_{\slice}$ is defined to be equal to $\lem$ in all the locations that 
-remained.
+remained. Similarly,

### 2017-06-23 paper/prelims.tex
preparing to adjust the definition of pCFG

@@
 values of the variables are real numbers. 
-We also assume that assume that each \PP{} $\program$ is preceded by an 
-initialization preamble 
-in 
-which 
-each variable appearing in $\program$ is assigned some concrete number.
-the program variables are initialized.
+We also assume that assume that each \PP{} $\program$ is preceded by a preamble 
+specifying possible initial values of program variables: the preamble consists 
+of a single predicate characterizing possible initial valuations.
 
@@
 Due to space restrictions, details of syntax (such as grammar) are relegated to 
-the Appendix.
+\AppendixMaterial.
 For an example see Figure~\ref{fig:invariant-running}.
@@
 A \emph{probabilistic control flow graph (pCFG)} is a tuple
-$\pCFG=(\locs,\pvars,\locinit,\vecinit,\transitions,\updates,\probdist,\guards)$,
+$\pCFG=(\locs,\pvars,\locinit,\vecinitset,\transitions,\updates,\probdist,\guards)$,
 where
@@
 variables} (note that $\pvars \subseteq \vars$) ;
-\item $\locinit$ is an initial location and $\vecinit$ is an initial 
-\emph{assignment vector};
+\item $\locinit$ is an initial location and $\vecinitset$ is a set of initial 
+\emph{assignment vectors};
 % POKRACOVAT ODSUD
@@
 configurations $(\ell_0,\vec{x}_0)\cdots(\ell_k,\vec{x}_k)$ such that 
+$\loc_0=\locinit$, $\vec{x}_0\in\vecinitset$, and 
 %$(\loc_0,\vec{x}_0)=(\locinit,\vecinit)$ and 
@@
 configuration
-$(\locinit,\vecinit)$
+$(\locinit,\vecinit)$ (where, $\vecinit\in \vecinitset$)
 if there is a finite path starting in $(\locinit,\vecinit)$ that ends in
@@
 \smallskip\noindent{\em Stochastic process.}
-A pCFG $\pCFG$ together with a scheduler $\sigma$ define a stochastic 
+A pCFG $\pCFG$ together with a scheduler $\sigma$ and initial valuation 
+$\vecinit\in\vecinitset$ define a stochastic 
 process which produces a random run 
@@
 The above intuitive explanation can be formalized by showing that
-each pCFG $\pCFG$ together with a scheduler $\sigma$ uniquely determines a 
+each pCFG $\pCFG$ together with a scheduler $\sigma$ and initial valuation 
+$\vecinit\in \vecinitset$ uniquely determine a 
 certain 
 probabilistic 
-space $(\OmegaRun,\natfilt,\probm^{\sigma})$ in which $\OmegaRun$ is a set of 
+space $(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$ in which $\OmegaRun$ 
+is a set of 
 all 
@@
 measure 
-$\probm^{\sigma}$ is such that for each $i$, the distribution of 
+$\probm^{\sigma}_{\vecinit}$ is such that for each $i$, the distribution of 
 $\cfg{\sigma}{i}$ reflects the aforementioned way in which runs are randomly 
 generated. The formal construction 
-of $\natfilt$ and $\probm^{\sigma}$ proceeds via the standard \emph{cylinder 
+of $\natfilt$ and $\probm^{\sigma}_{\vecinit}$ proceeds via the standard 
+\emph{cylinder 
 construction}~\cite[Theorem 2.7.2]{Ash:book} and is
@@
 technical, hence we omit it. 
-We denote by $\E^\sigma$ the expectation operator in probability space 
-$(\OmegaRun,\natfilt,\probm^{\sigma})$.
+We denote by $\E^\sigma_{\vecinit}$ the expectation operator in probability 
+space 
+$(\OmegaRun,\natfilt,\probm^{\sigma}_{\vecinit})$.
 
@@
 
-\smallskip\noindent{\em Termination, reachability, and termination time.}
+\smallskip\noindent{\em Termination and termination time.}
 In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. 
-We say that a run $\run$ of $\pCFG_P$ \emph{reaches} a set of configurations 
-$\confset$ if it  contains a configuration from $\confset$. A run 
+We say that a run 
 \emph{terminates} if it reaches a 
@@
 executing $P$. 
-To each set of configurations $\confset$ we can assign a random variable 
-$\treach{\confset}$ 
-such that for each run $\run$ the value $\treach{\confset}(\run)$ represents 
+We define a random variable $\ttime$
+such that for each run $\run$ the value $\ttime(\run)$ represents 
 the first 
 point in time when 
-the current configuration on $\run$ is in $\confset$. If a run $\run$ does 
+the current location is $\loc_P^{\lout}$. If a run $\run$ does 
 \emph{not} 
-reach a set $\confset$, then 
-$\treach{\confset}(\run)=\infty$. We call $\treach{\confset}$ the 
-\emph{reachability time} of $\confset$. In particular, if $\confset$ is the set 
-of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
-terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
-called a \emph{termination time}, as it returns the number of steps after which 
-$\run$ terminates. Since termination time is an important concept on its own, 
-we use a special notation $\ttime$ for it.
+terminate, then 
+$\ttime(\run)=\infty$. We call $\ttime$ the 
+\emph{termination time} of $\program$. 
+%In particular, if $\confset$ is the set 
+%of all configurations $(\loc,\vec{x})$ such that $\loc=\loc^\lout_P$ (the 
+%terminal location of $\pCFG_P$), then $\treach{\confset}$ is 
+%called a \emph{termination time}, as it returns the number of steps after 
+%which 
+%$\run$ terminates. Since termination time is an important concept on its own, 
+%we use a special notation $\ttime$ for it.
 %\begin{example}
@@
 almost-surely (a.s.) 
-terminating if under each scheduler $\sigma$ it holds that 
-$\probm^{\sigma}(\{\run\mid \run \text{ terminates}\}) = 1$, or equivalently, 
+terminating if under each scheduler $\sigma$ and for each initial valuation 
+$\vecinit\in\vecinitset$ it holds that 
+$\probm^{\sigma}_{\vecinit}(\{\run\mid \run \text{ terminates}\}) = 1$, or 
+equivalently, 
 if for each $\sigma$ it holds $\probm^{\sigma}(\ttime<\infty)=1$. In 
@@
 positively
-terminating if under each scheduler $\sigma$ it holds that 
-$\E^{\sigma}(\{\ttime\})<  \infty$. In 
+terminating if under each scheduler $\sigma$ and for each initial valuation 
+$\vecinit\in\vecinitset$ it holds that 
+$\E^{\sigma}_{\vecinit}(\{\ttime\})<  \infty$. In 
 positive termination question for $P$ we aim to prove that $P$ 
@@
 
-We also define corresponding questions for the more general reachability 
-concept.
-
-\begin{compactenum}
-\item \emph{Almost-sure reachability:} For a set $C$ of configurations of a
-probabilistic program $P$, prove (if possible) that under each scheduler 
-$\sigma$ it holds that
-$\probm^{\sigma}(\treach{C}<\infty)=1$.
-\item \emph{Positive reachability:} For a set $C$ of configurations of 
-a probabilistic program $P$, prove (if possible) that under each scheduler 
-$\sigma$ it holds that
-$\E^{\sigma}(\treach{C})<\infty$.
-\end{compactenum}
-
-Since termination is a special case of reachability, each solution to the 
-almost-sure or probabilistic reachability questions provides solution for the 
-corresponding termination questions.
+%We also define corresponding questions for the more general reachability 
+%concept. 
+% 
+%\begin{compactenum}
+%\item \emph{Almost-sure reachability:} For a set $C$ of configurations of a
+%probabilistic program $P$, prove (if possible) that under each scheduler 
+%$\sigma$ it holds that
+%$\probm^{\sigma}(\treach{C}<\infty)=1$.
+%\item \emph{Positive reachability:} For a set $C$ of configurations of 
+%a probabilistic program $P$, prove (if possible) that under each scheduler 
+%$\sigma$ it holds that
+%$\E^{\sigma}(\treach{C})<\infty$.
+%\end{compactenum}
+%
+%Since termination is a special case of reachability, each solution to the 
+%almost-sure or probabilistic reachability questions provides solution for the 
+%corresponding termination questions.
 %\begin{example}\label{ex:illustrate}

### 2017-06-23 implementation/files/Parser.cpp
minor changes

@@ -951,7 +951,7 @@ void CFG_edge::print(){
 		cout<<"Error! Error! Error!"<<endl;
 	}
 	if(change!=NULL){
-		cout<<"Change: x_"<<toChange<<" changed to ";
+		cout<<"Change: "<<variable[toChange]<<" = ";
 		change->print();
 		cout<<endl;
 	}

### 2017-06-23 implementation/files/invariant_script.sh

@@ -1,4 +1,4 @@
-./files/aspicV3.4 -delay 10 files/aspic.fast > files/InvariantOutput
+./files/aspicV3.4 -delay 20 files/aspic.fast > files/InvariantOutput
 awk 'BEGIN {x=0} x==1{print} /^\*\*\* Results :/{x=1}' < files/InvariantOutput > InvariantOutput.tmp
 mv InvariantOutput.tmp files/InvariantOutput
 rm aspic.log

### 2017-06-23 implementation/main.cpp

@@ -132,6 +132,7 @@ int epsilons_used = 0;
 void generate_equations(){ //Would use the ostream file to write the equations into it later
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
 		if(it->second->invariant==NULL){
+			cout<<endl;
 			continue;
 		}
 		it->second->invariant->print();
@@ -227,6 +228,7 @@ void generate_equations(){ //Would use the ostream file to write the equations i
 int last_used_lambda = 0;
 
 void print_equations(ostream& equationsfile){
+	last_used_lambda = 0;
 	equationsfile<<"maximize ";
 	for(int i=1;i<=epsilons_used;++i){
 		if(equations.find(i)!=equations.end()){
@@ -444,13 +446,13 @@ int main(){
 	// cout<<program<<endl;
 	// cout<<"Parse Tree:"<<endl;
 	// root->print(cout,"&&","||","*",false);
-	// cout<<"CFG:"<<endl;
-	// for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
-	// 	cout<<"------------------------"<<endl;
-	// 	cout<<"Node "<<it->first<<endl;
-	// 	it->second->print();
-	// 	// cout<<it->second->label<<endl;
-	// }
+	cout<<"CFG:"<<endl;
+	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
+		cout<<"------------------------"<<endl;
+		cout<<"Node "<<it->first<<endl;
+		it->second->print();
+		// cout<<it->second->label<<endl;
+	}
 	
 	generate_equations();
 	ofstream equationsfile;

### 2017-06-23 implementation/main.cpp
Adding lambda from -inf to inf condition

@@ -288,6 +288,9 @@ void print_equations(ostream& equationsfile){
 			equationsfile<<"-inf<= f_"<<it->first<<"_"<<j<<" <= +inf"<<endl;
 		}
 	}
+	for(int i=0;i<=last_used_lambda;++i){
+		equationsfile<<"-inf<= l"<<i<<" <= +inf"<<endl;
+	}
 	equationsfile<<"end"<<endl;
 }
 

### 2017-06-23 implementation/main.cpp
necessary change

@@ -273,7 +273,7 @@ void print_equations(ostream& equationsfile){
 			}
 		}
 		if(front->condition->strict){
-			equationsfile<<" < 0"<<endl;
+			equationsfile<<" <= 0"<<endl;
 		}
 		else{
 			equationsfile<<" <= 0"<<endl;
@@ -287,9 +287,6 @@ void print_equations(ostream& equationsfile){
 		for(int j=0;j<=nVariables;j++){
 			equationsfile<<"-inf<= f_"<<it->first<<"_"<<j<<" <= +inf"<<endl;
 		}
-	}
-	for(int i=0;i<=last_used_lambda;++i){
-		equationsfile<<"-inf<= l"<<i<<" <= +inf"<<endl;
 	}
 	equationsfile<<"end"<<endl;
 }
@@ -449,13 +446,13 @@ int main(){
 	// cout<<program<<endl;
 	// cout<<"Parse Tree:"<<endl;
 	// root->print(cout,"&&","||","*",false);
-	cout<<"CFG:"<<endl;
-	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
-		cout<<"------------------------"<<endl;
-		cout<<"Node "<<it->first<<endl;
-		it->second->print();
-		// cout<<it->second->label<<endl;
-	}
+	// cout<<"CFG:"<<endl;
+	// for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
+	// 	cout<<"------------------------"<<endl;
+	// 	cout<<"Node "<<it->first<<endl;
+	// 	it->second->print();
+	// 	// cout<<it->second->label<<endl;
+	// }
 	
 	generate_equations();
 	ofstream equationsfile;

### 2017-06-23 paper/compositional.tex

--- 2017-06-20 paper/compositional.tex
@@
 Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
-\programbody \textbf{ od}$. We denote by $\loops(\program)$ the set of all 
-loops in $\program$ there are on the top level of $\programbody$, i.e. they are 
-nested loops of $\program$ nested directly "below" the outer loop. We define a 
-depth-1 slice of $\program$ to be a \PP{} $\slice(\program)$ obtained from 
-$\program$ by removing all loops in $\loops(\program)$. A formal definition of 
+\programbody \textbf{ od}$, and let $\pCFG_{\program}$ be the associated pCFG, whose set of locations we denote by $\locs$.
+We denote by $\loops(\program)$ the set of all locations of $\pCFG_{\program}$ that belong to a sub-pCFG of $\pCFG_{\program}$ corresponding to some nested loop of $\program$.
+We also define $\slice(\program)$ to be the set $\locs\setminus\loops(\program)$ of locations that do not belong to any nested sub-loop.
+A formal definition of
 both functions is given in \AppendixMaterial, we illustrate them in the 
 following example.
+%We denote by $\loops(\program)$ the set of all 
+%loops in $\program$ there are on the top level of $\programbody$, i.e. they are 
+%nested loops of $\program$ nested directly "below" the outer loop. We define an 
+%un-nested slice of $\program$ to be a \PP{} $\slice(\program)$ obtained from 
+%$\program$ by removing all loops in $\loops(\program)$ and replacing them with 
+%skip statements. We also define the set 
+%$\alloops(\program)$ of all loops in $\program$, i.e. 
+%$\alloops(\program):=\loops(\program)\cup \{\loops(\program')\mid \program'\in 
+%\loops(\program)\}$. The above definitions do not specify the initialization 
+%preamble for the sub-programs $\program'\in\alloops(\program)$: we can deem 
+%this pre-amble empty, i.e. consider these sub-programs with any initial 
+%valuation, or, provided that some invariant map $\inv$ for $\program$ is 
+%given, we can set the set of possible initial valuations of a sub-program 
+%$\program'$ to be the set $\inv(\loc^{\lin}_{\program'})$, where 
+%$\inv(\loc^{\lin}_{\program'})$ is the location of $\program$ corresponding to 
+%the head of the loop $\program'$.
+%A formal definition of 
+%both functions is given in \AppendixMaterial, we illustrate them in the 
+%following example.
 
 \textbf{[PETR: INSERT EXAMPLE]}
+
+We say that a 1-dimensional measurable map $\lem$ for $\pCFG_{\program}$ is $\eps$-ranking in location $\loc$, if for each $\vec{x}\in\Rset^{|\pvars|}$ such that $(\loc,\vec{x}) $ is reachable from some initial configuration it holds $\preexp{\lem}(\loc,\vec{x})\leq \lem(\loc,\vec{x})-\eps.$ The map $\lem$ is \emph{unaffected} in location $\loc$ if it is $0$-ranking in $\loc$, and non-negative in $\loc$ if $\lem(\loc,\vec{x})\geq 0$ for all $\vec{x}$ such that $(\loc,\vec{x})$ is reachable from an initial configuration.
+A measurable map $\lem$ is propositionally linear, if each function $\lem(\loc)$ is of the form $\indicator{G_1}\cdot E_1 + \cdots \indicator{G_k}\cdot E_k$, where each $\indicator{G_i}$ is an indicator function of some polyhedron and each $E_i$ is a linear expression.
+
+We recall the notion of compositional ranking supermartingale as introduced in~\cite{HolgerPOPL}.
+We call it a PV supermartingale, as it is based on so called probabilistic variant rule.
+Due to differences in syntax and semantics, the definition is syntactically slightly different from~\cite{HolgerPOPL}, but the essence is the same.
+
+\begin{definition}[PV-supermartingale]
+A 1-dimensional propositionally linear map $\lem$ is a PV supermartingale (PVSM) for a program $\program$ if there exists $\eps>0$ such that $\lem$ is $\eps$-ranking an non-negative in each location $\loc\in\slice{(\program)}$ and unaffected in each $\loc\in\loops(\program)$.
+\end{definition}
+
+In~\cite{HolgerPOPL} they show that even if all nested loops were already proved to terminate a.s.
+and there is a PVSM for the program, then the program itself might not terminate a.s.
+Then they investigate conditions under which existence of a PVSM together with a proof of a.s.
+termination of each nested sub-loop of $\program$ entails termination of the whole program $\program$.
+The condition were formulated in terms of a certain type system for \PP{}s, so we say that a PVSM satisfying the condition \emph{typechecks} correctly.
+In~\cite{HolgerPOPL} the following was proved:
+
+\begin{theorem}[\cite{HolgerPOPL}]
+\label{thm:holger-comp}
+Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
+\programbody \textbf{ od}$.
+Assume that each nested loop of $\program$ terminates almost surely from each reachable configuration, and that there exists a PVSM for $\program$ that typechecks correctly.
+Then $\program$ terminates almost surely.
+\end{theorem}
+
+We do not repeat the precise definition of the typesystem here, we just explain its motivation and give examples where it fails, i.e.
+examples where $\program$ terminates a.s.
+(and hence the same holds also for its nested loops) and a PV martingale exists, but no PVSM typechecks.
+
+The purpose of the typesystem is, in principle, to ensure that the stochastic process which, in each time step, returns the current value of the PVSM, is uniformly integrable inside the nested loops.
+Uniform integrability is a deep concept from probability and measure theory: a sequence $X_0,X_1,X_2,\dots$ of random variables is uniformly integrable if for each $\delta>0$ there exists an $K\in \Nset$ such that for all $n\geq 0$ it holds $\E[|X_n|\cdot\indicator{X_n \geq K}]\leq \delta$.
+As noted in~\cite{HolgerPOPL}, proofs of uniform integrability are hard to (semi)-automatize, and hence they devise the typechecking as a sound (but not complete) method of proving uniform integrability.
+Take for example \textbf{[PETR: ADD EXAMPLE]}.
+However, as we show below, for none of these programs there exists a PVSM that typechecks.
+
+Indeed, taking a closer look at typesystem in~\cite{HolgerPOPL}, there are several reasons for typechecking of PVSM to fail. The major ones are:
+\begin{compactenum}
+\item
+A PVSM $\lem$ for \PP{} $\program$ will not typecheck if $\program$ has a nested loop in which the value of $\lem$ can change unboundedly in a single step (see Example~\ref{xxx}). 
+\item
+A PVSM $\lem$ for  \PP{} $\program$ will not typecheck if $\program$ has a nested loop which itself has a nested loop in which some variable appearing in some expression in $\lem$ is modified, see Example~\ref{xxx}.\footnote{Both these statements regarding typechecking failure are somewhat simplified, even in these two cases the PVSM might sometimes typecheck correctly, in case where the nested loops are followed by assignments which completely overwrite the effect of these loops, e.g.
+if \textbf{[PETR: POINTER TO AN EXAMPLE]}.
+However, the statements intuitively summarize the major reasons for typechecking failure.
+}
+\end{compactenum}
+
+To overcome the rather strict
+
+%Each measurable map for $\lem$ can be, in a natural way, restricted to a 
+%measurable map $\lem_{\program'}$ for $\program'\in \loops(\program)$, 
+%since each pCFG $\pCFG_{\program'}$ is a sub-pCFG of $\pCFG_{\program}$. 
+%Similarly, a measurable map $\lem$ for $\program$ can be restricted to a 
+%measurable map $\lem_{\slice}$ for $\slice(\program)$, since 
+%$\pCFG_{\slice(\program)}$ is produced from $\pCFG_{\program}$ essentially by 
+%removing certain locations (provided that we identify the locations forming 
+%heads of nested loops with locations preceding skip statements), and 
+%$\lem_{\slice}$ is defined to be equal to $\lem$ in all the locations that 
+%remained. Similarly,

### 2017-06-25 paper/compositional.tex
expanded the section on compositionality

@@
 
-\begin{definition}[PV-supermartingale]
+\begin{definition}[PV-supermartingale {\cite[Definition 7.1.]{HolgerPOPL}}]
 A 1-dimensional propositionally linear map $\lem$ is a PV supermartingale (PVSM) for a program $\program$ if there exists $\eps>0$ such that $\lem$ is $\eps$-ranking an non-negative in each location $\loc\in\slice{(\program)}$ and unaffected in each $\loc\in\loops(\program)$.
 \end{definition}
+
+As a matter of fact, the condition that $\lem$ should be non-negative in locations of $\slice{(\program)}$ is not explicitly mentioned in \cite{HolgerPOPL}.
+However, it is implicitly used in some of the proofs and one can easily construct an example where, if the non-negativity in $\slice{(\program)}$ is not required, the Theorem~\ref{thm:holger-comp} below, which also comes from~\cite{HolgerPOPL}, does not hold.
+Hence, we state the condition explicitly.
 
@@
 
-To overcome the rather strict
+To overcome the rather strict typechecking, we use the results on LexRSMs to define a new notion of compositional ranking supermartingales, which we call \emph{nob-negative compositional (NC)} supermartingales.
+For the sake of generality, we allow NC martingales to be general measurable maps, not necessarily propositionally linear.
+
+\begin{definition}
+\label{def:nonneg-comp}
+A 1-dimensional measurable map $\lem$ is an NC supermartingale (NCSM) for a program $\program$ if there exists $\eps>0$ such that $\lem$ is:
+\begin{compactenum}
+\item  non-negative in each location of $\pCFG_{\program}$;
+\item 
+ $\eps$-ranking in each location $\loc\in\slice{(\program)}$; and
+\item 
+  unaffected in each $\loc\in\loops(\program)$.
+\end{compactenum}
+A (propositionally) linear NCSM (LinNCSM) is a NCSM which is also a (propositionally) linear expression map.
+\end{definition}
+
+We can prove that NCSMs are a sound method for proving a.s. termination in a compositional way, without any additional assumptions.
+
+\begin{theorem}
+Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
+\programbody \textbf{ od}$.
+Assume that each nested loop of $\program$ terminates almost surely from each reachable configuration, and that there exists a NCSM for $\program$.
+Then $\program$ terminates almost surely.
+\end{theorem}
+\begin{proof}
+\textbf{[PETR: TBD]}
+\end{proof}
+
+Hence, NCSMs effectively trade the uniform integrability condition for non-negativity over the whole program.
+We believe that the latter condition is substantially easier to impose and check automatically, especially in the case of affine probabilistic programs and (propositionally) linear NCSMs: there synthesizing NCSM entails synthesizing sufficient program invariants (for which there is a good automated tool support~\cite{xxx}) and encoding the non-negativity condition into a collection of linear constraints (as for general LexRSMs in Section~\ref{xxx}).
+
+%% I SHOULD MAYBE EXPLAIN THAT THE NEED TO PERFORM A STATIC ANALYSIS OF THE WHOLE PROGRAM IS 
+%% NOT LIMITING, AS SOMETHING SIMILAR NEEDS TO BE DONE FOR PVSM. BUT MAYBE LEAVE THIS FOR REBUTTAL. :)
 

### 2017-06-25 paper/compositional.tex
finished draft of compositionality section (no examples yet)

@@
 proof of termination step-by-step, handling one loop at a time, rather than 
-attempting to construct the proof (in our case, a LexRSM) at once, by solving a 
-single large system of constraints. In non-probabilistic setting, the 
+attempting to construct the proof (in our case, a LexRSM) at once. In non-probabilistic setting, the 
 relationship between lexicographic ranking functions and compositional 
@@
 below.
-However, as already noted in~\cite{HolgerPOPL}, the probabilistic V-rule is not 
+The reason for this attempt was to provide a method of proving a.s.
+termination for programs where 1-dimensional ranking supermartingales are not sufficient.
+Since our LexRSMs were conceived with a similar motivation, we devote this section to a thorough comparison of our approach with the one of~\cite{HolgerPOPL}.
+
+There are two main features of LexRSMs compared with the previous approach, which we call a PVSM (probabilistci V-rule supermartingale) approach.
+First, the PVSM approach is only applicable to programs in which termination of each loop can be witnessed by a suitable 1-dimensional ranking supermartingale.
+But there are already single-loop programs for which no 1-dimensional ranking supermartingale exists but a LexRSM does, see the following example.
+
+\textbf{[PETR: ADD EXAMPLE]}
+
+The need for multi-dimensional ranking arises within loops that might go through several ``computation phases'' before terminating, and loops containing non-deterministic choices.
+
+Second, as already noted in~\cite{HolgerPOPL}, the probabilistic V-rule is not 
 sound in general. In order to rectify this issue, in~\cite{HolgerPOPL} they 
@@
 programs. In particular, the method cannot handle programs in which termination 
-is controlled by variable that can be modified by a non-constant value, e.g. 
+is controlled by a variable that can be modified by a non-constant value in some sub-program, e.g. 
 variable $x$ which appears in an assignment of the form $x:=x-y$. In this 
-section we show that using our insights in lexicographic supermartingales we 
+section we show that using our insights into LexRSMs we 
 can obtain a subtly different notion of a probabilistic V-rule which is sound 
-without any additional assumptions, and which can be used to prove termination 
+without any additional assumptions, and which can be used to compositionally prove termination 
 of programs that the previous method cannot handle.
@@
 
-We say that a 1-dimensional measurable map $\lem$ for $\pCFG_{\program}$ is $\eps$-ranking in location $\loc$, if for each $\vec{x}\in\Rset^{|\pvars|}$ such that $(\loc,\vec{x}) $ is reachable from some initial configuration it holds $\preexp{\lem}(\loc,\vec{x})\leq \lem(\loc,\vec{x})-\eps.$ The map $\lem$ is \emph{unaffected} in location $\loc$ if it is $0$-ranking in $\loc$, and non-negative in $\loc$ if $\lem(\loc,\vec{x})\geq 0$ for all $\vec{x}$ such that $(\loc,\vec{x})$ is reachable from an initial configuration.
+Given an invariant map $\inv$, we say that a 1-dimensional measurable map $\lem$ for $\pCFG_{\program}$ is $\eps$-$\inv$-ranking in location $\loc$, if for each $\vec{x}\in\inv(\loc)$  it holds $\preexp{\lem}(\loc,\vec{x})\leq \lem(\loc,\vec{x})-\eps.$ The map $\lem$ is $\inv$-\emph{unaffected} in location $\loc$ if it is $0$-ranking in $\loc$, and $\inv$-non-negative in $\loc$ if $\lem(\loc,\vec{x})\geq 0$ for all $\vec{x} \in \inv(\loc)$.
 A measurable map $\lem$ is propositionally linear, if each function $\lem(\loc)$ is of the form $\indicator{G_1}\cdot E_1 + \cdots \indicator{G_k}\cdot E_k$, where each $\indicator{G_i}$ is an indicator function of some polyhedron and each $E_i$ is a linear expression.
@@
 \begin{definition}[PV-supermartingale {\cite[Definition 7.1.]{HolgerPOPL}}]
-A 1-dimensional propositionally linear map $\lem$ is a PV supermartingale (PVSM) for a program $\program$ if there exists $\eps>0$ such that $\lem$ is $\eps$-ranking an non-negative in each location $\loc\in\slice{(\program)}$ and unaffected in each $\loc\in\loops(\program)$.
+A 1-dimensional propositionally linear map $\lem$ is a PV supermartingale (PVSM) for a program $\program$ supported by an invariant map $\inv$ if there exists $\eps>0$ such that $\lem$ is $\eps$-$\inv$-ranking an $\inv$-non-negative in each location $\loc\in\slice{(\program)}$ and $\inv$-unaffected in each $\loc\in\loops(\program)$.
 \end{definition}
@@
 
+Thus, the typechecking algorithm may rule out programs where the termination-controlling variable represents e.g.
+a length of an array, which can be doubled/halved in some sub-program do to (de)allocation, merging, or splitting.
 To overcome the rather strict typechecking, we use the results on LexRSMs to define a new notion of compositional ranking supermartingales, which we call \emph{nob-negative compositional (NC)} supermartingales.
@@
 \label{def:nonneg-comp}
-A 1-dimensional measurable map $\lem$ is an NC supermartingale (NCSM) for a program $\program$ if there exists $\eps>0$ such that $\lem$ is:
+A 1-dimensional measurable map $\lem$ is an NC supermartingale (NCSM) for a program $\program$ supported by an invariant map $\inv$ if there exists $\eps>0$ such that $\lem$ is:
 \begin{compactenum}
-\item  non-negative in each location of $\pCFG_{\program}$;
+\item  $\inv$-non-negative in each location of $\pCFG_{\program}$;
 \item 
- $\eps$-ranking in each location $\loc\in\slice{(\program)}$; and
+ $\eps$-$\inv$-ranking in each location $\loc\in\slice{(\program)}$; and
 \item 
-  unaffected in each $\loc\in\loops(\program)$.
+  $\inv$-unaffected in each $\loc\in\loops(\program)$.
 \end{compactenum}
@@
 \begin{theorem}
+\label{thm:nonneg-comp}
 Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
 \programbody \textbf{ od}$.
-Assume that each nested loop of $\program$ terminates almost surely from each reachable configuration, and that there exists a NCSM for $\program$.
+Assume that each nested loop of $\program$ terminates almost surely from each reachable configuration, and that there exists a NCSM for $\program$ supported by some invariant map.
 Then $\program$ terminates almost surely.
@@
 Hence, NCSMs effectively trade the uniform integrability condition for non-negativity over the whole program.
-We believe that the latter condition is substantially easier to impose and check automatically, especially in the case of affine probabilistic programs and (propositionally) linear NCSMs: there synthesizing NCSM entails synthesizing sufficient program invariants (for which there is a good automated tool support~\cite{xxx}) and encoding the non-negativity condition into a collection of linear constraints (as for general LexRSMs in Section~\ref{xxx}).
+We believe that the latter condition is substantially easier to impose and check automatically, especially in the case of affine probabilistic programs and (propositionally) linear NCSMs: for \APP{}s, synthesizing NCSM entails synthesizing sufficient program invariants (for which there is a good automated tool support~\cite{xxx}) that overapproximate set of reachable configurations and encoding the rankingn, unaffection, and non-negativity conditions into a collection of linear constraints (as for general LinLexRSMs in Section~\ref{xxx}).
+Examples~\ref{xxx} show instances where attempts to prove of a.s.
+termination via PVSMs fail while proofs via LinNCSMs work.
+
+%While we defined NCSMs as one-dimensional measurable maps to make them analogous to
+
+Using LinNCSMs we can devise the following simple algorithm for compositional proving of almost-sure termination of \APP{}s:
+
+\begin{algorithm}
+\SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
+\DontPrintSemicolon
+
+\Input{An \APP{} $\program$ together with an invariant map $\inv$.}
+$d\leftarrow \text{ depth of loop nesting in $\program$}$\;
+\For{$i \leftarrow d$ \KwTo $0$}{
+\ForEach{sub-loop $\program'$ of $\program$ nested $i$ levels below the main loop}{
+\label{algoline:one}
+$\linsystem \leftarrow$ system of lin. constraints encoding the existence of LinNCSM for $\program'$ supported by $\inv$\;
+\label{algoline:two}\If{$\linsystem$ not solvable}{check existence of a PVSM for $\program'$~\cite{HolgerPOPL}
+\;
+\lIf{PVSM does not exist}{\Return{``cannot prove a.s. termination of $\program$''}}
+}
+}
+}
+\Return{``$\program$ terminates a.s.'' }
+\caption{Compositional Termination Proving}
+\label{algo:compositional}
+\end{algorithm}
+
+The soundness of the algorithm follows from Theorems~\ref{thm:nonneg-comp} and \ref{thm:holger-comp}.
+Note that we use the PVSM-based algorithm of~\cite{HolgerPOPL} as a back-up sub-procedure for the case when LinNCSM-based proof fails.
+Hence, Algorithm~\ref{algo:compositional} can compositionally prove a.s.
+termination of strictly larger class of programs than the PVSM-based algorithm alone.
+
+To summarize, the novelty of NCSMs is the following:
+\begin{compactenum}
+\item
+NCSMs allow compositional, fully automated proofs of a.s. termination without the need for reasoning about uniform integrability.
+\item LinNCSMs are capable of proving a.s. termination of programs for which no uniformly integrable PVSMs exist (and hence the method of~\cite{HolgerPOPL} cannot be used at all on such programs).
+\item LinNCSMs are capable of proving a.s. termination of programs for which the method of~\cite{HolgerPOPL} cannot be applied in an automated way, do to failure of the typechecking procedure.
+\end{compactenum}
+
+\subsection{Multidimensional Compositional Ranking}
+
+As noted previously, a fundamental advantage of LexRSMs over PVSMs is that LexRSMs can handle loops that go through several computational phases and as a result of this do not admit one-dimensional ranking supermartingales.
+Above, we defined NCSMs as one-dimensional objects, to make them analogous to PVSMs for better comparison.
+However, we can also define a multi-dimensional version of NCSMs.
+We say that, given an invariant map $\inv$, an $n$-dimensional measurable map is $\inv$-ranking in a location $\loc$ if for each $\vec{x}\in \inv(\loc)$ there exists $1\leq j \leq n$ such that $\preexp{\lem_j}(\loc,\vec{x}) \leq \lem_j(\loc,\vec{x})-\eps$ and for each $j'<j$ the map $\lem_{j'}$ is $\inv$-unaffected in $\loc$.
+
+\begin{definition}
+An $n$-dimensional measurable map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ is an NC supermartingale (NCSM) for a program $\program$ supported by an invariant map $\inv$ if there exists $\eps>0$ such that:
+\begin{compactenum}
+	\item  for each $1\leq i \leq n$, $\lem_i$ is non-negative in each location of $\pCFG_{\program}$;
+	\item 
+	$\vec{\lem}$ is $\eps$-ranking in each location $\loc\in\slice{(\program)}$; and
+	\item 
+	for each $1\leq i \leq n$, $\lem_i$ is
+	unaffected in each $\loc\in\loops(\program)$.
+\end{compactenum}
+An $n$-dimensional (propositionally) linear NCSM (LinNCSM) is an $n$-dimensional NCSM which is also a (propositionally) linear expression map.
+\end{definition}
+
+The following theorem can be proved in the same way as Theorem~\ref{thm:nonneg-comp}.
+
+\begin{theorem}
+Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
+\programbody \textbf{ od}$.
+Assume that each nested loop of $\program$ terminates almost surely from each reachable configuration, and that there exists a NCSM for $\program$ supported by some invariant map.
+Then $\program$ terminates almost surely.
+\end{theorem}
+
+For \APP{}s, we can generalize Algorithm~\ref{algo:compositional} by changing line~\ref{algoline:one} to ``check existence of a multi-dimensional LinNCSM for $\program'$'' and line~\ref{algoline:two} to ``\textbf{if} a multi-dimensional LinNCSM does not exist.'' The check of existence of a multi-dimensional LinNCSM for $\program'$ can be done by algorithm presented in Section~\ref{xxx}, modified so as to only pursue ranking for transitions outgoing from locations belonging to $\slice(\program')$.
+(I.e., only these transitions have $\eps_{\tau}$ included in the objective function, and algorithm terminates once all such transitions are ranked.) This further expands the usage of the algorithm.
+
+There are several advantages of using LexRSMs in a compositonal way.
+First, using Algorithm~\ref{algo:compositional} (modified for multi-dimensional LinNCSMs) we opt for solving a larger number of smaller linear systems rather than a smaller number of larger ones, which may be sometimes advantageous for performance reasons.
+Second, using compositional reasoning, we can leverage LexRSMs in a situation where some of the sub-loops were already proved to be a.s.
+terminating by some other argument, such as using an interactive proof assistant that leverages human intuition.
+This makes the LexRSM method applicable to programs whose parts are difficult to treat via automated static analysis.
 

### 2017-06-25 implementation/examples/cousot9.c
added some examples

@@ -1,7 +1,6 @@
 int cousot9(){
 
   int i,j,N;
-
   i=N;
   while(i>0) {
 

### 2017-06-25 implementation/files/invariant_script.sh

@@ -1,4 +1,4 @@
-./files/aspicV3.4 -delay 20 files/aspic.fast > files/InvariantOutput
+./files/aspicV3.4 -delay 30 files/aspic.fast > files/InvariantOutput
 awk 'BEGIN {x=0} x==1{print} /^\*\*\* Results :/{x=1}' < files/InvariantOutput > InvariantOutput.tmp
 mv InvariantOutput.tmp files/InvariantOutput
 rm aspic.log

### 2017-06-26 implementation/examples/counterex1a.c
Added some more examples

@@ -15,4 +15,3 @@ int counterex1a()
   }
   return 0;
 }
-

### 2017-06-26 implementation/examples/speedFails2.c

@@ -1,8 +1,4 @@
-
-int speedFails2(){
-  int i,x,n;
-  
-  for(i=x;i!=n;++i);
-  
-  return 0;
-}
+var i,x,n;
+[x>n]
+i:=x;
+while i>=n+1 or i+1<=n do i:=i+1 od

### 2017-06-26 implementation/files/Parser.cpp

@@ -269,14 +269,18 @@ void node::proc_stmt(int s,int l){
 		children.resize(2);
 		children[0] = new node("bexpr",begin+5,firstdo);
 		int mid = ++last_used_label;
+		int s_dup = ++last_used_label;
 		label_map[mid] = new CFG_location("det",mid);
+		label_map[s_dup] = new CFG_location("det",s_dup);
 		node* temporary_node;
 		temporary_node = negation(children[0]);
 		CFG_edge temporary_edge1(label_map[mid],-1,NULL,children[0]);
 		CFG_edge temporary_edge2(label_map[l],-1,NULL,temporary_node);
 		label_map[s]->edges.pb(temporary_edge1);
+		label_map[s_dup]->edges.pb(temporary_edge1);
 		label_map[s]->edges.pb(temporary_edge2);
-		children[1] = new node("stmt",firstdo+2,lastod,mid,s);
+		label_map[s_dup]->edges.pb(temporary_edge2);
+		children[1] = new node("stmt",firstdo+2,lastod,mid,s_dup);
 		return;
 	}
 	//check if it is an if

### 2017-06-26 implementation/files/invariant_script.sh

@@ -1,4 +1,4 @@
-./files/aspicV3.4 -delay 30 files/aspic.fast > files/InvariantOutput
+./files/aspicV3.4 files/aspic.fast > files/InvariantOutput
 awk 'BEGIN {x=0} x==1{print} /^\*\*\* Results :/{x=1}' < files/InvariantOutput > InvariantOutput.tmp
 mv InvariantOutput.tmp files/InvariantOutput
 rm aspic.log

### 2017-06-28 implementation/files/Parser.cpp
proceeding in expressions with uniform distributions

@@ -27,6 +27,7 @@ int last_used_label = 0;
 map<int,CFG_location*> label_map;
 map<string,int> variableId;
 map<int,string> variable;
+int dummy_states_needed = 0;
 
 void skip_spaces(int &begin, int &end){ //skips the spaces in the beginning and the end
 	while(begin<program.size() and isspace(program[begin])){
@@ -55,6 +56,7 @@ node::node(string t){
 	type = t;
 	constant = "";
 	strict = false;
+	delta = 0.0;
 }
 
 node::node(string t, int b, int e, int s, int l, bool negate){
@@ -63,6 +65,7 @@ node::node(string t, int b, int e, int s, int l, bool negate){
 	begin = b;
 	strict = false;
 	end = e;
+	delta = 0.0;
 	skip_spaces(begin,end);
 	if(begin>end){
 		cerr<<"Error! begin= "<<b<<" end= "<<e<<endl;
@@ -73,6 +76,7 @@ node::node(string t, int b, int e, int s, int l, bool negate){
 node::node(string t, string l, bool& to_return){
 	strict = false;
 	type = t;
+	delta = 0.0;
 	if(t=="bexpr"){
 		children.resize(1);
 		children[0] = new node("affexpr");
@@ -374,47 +378,47 @@ void node::proc_assgn(int s,int l){
 		while(isspace(program[samplepos])){
 			samplepos++;
 		}
-		if(part(program,samplepos,samplepos+6)=="sample"){
-			int semicolon = -1;
-			constant = "assignment from distribution";
-			for(int i = samplepos+6;i<end;i++){
-				if(program[i]==';'){
-					semicolon = i;
-					break;
-				}
-			}
-			if(semicolon == -1){
-				cerr<<"Mean value not found in the from distribution assignment between ["<<begin<<","<<end<<")"<<endl;
-			}
-			else{
-				int temp = samplepos+6;
-				while(temp<semicolon){
-					if(program[temp]=='('){
-						break;
-					}
-					temp++;
-				}
-				if(program[temp]!='('){
-					cerr<<"Wrong assignment from distribution between ["<<begin<<","<<end<<")"<<endl;
-				}
-				temp++;
-				while(isspace(program[temp]) and temp<semicolon){
-					temp++;
-				}
-				constant = part(program,temp,semicolon); //Gives the distribution
[…]

### 2017-06-28 implementation/files/Parser.h

@@ -9,6 +9,7 @@ struct node{
 	vector<node *> children; //children in the parse tree
 	node *bracket; //bexpr written in bracket next to this node (for stmt only)
 	vector<double> expression; //for expressions only
+	double delta;
 	//constructor
 	node(string t);
 	node(string t, int b, int e, int s = 0, int l = 0, bool negate = false);
@@ -73,3 +74,4 @@ extern map<int,CFG_location*> label_map;
 extern node* root;
 extern map<string,int> variableId;
 extern map<int,string> variable;
+extern int dummy_states_needed; //For adding the if * condition in fast file

### 2017-06-28 implementation/main.cpp

@@ -323,7 +323,7 @@ string start_invariant(){
 	}
 }
 
-
+int last_dummy_state_used = 0;
 
 void print_fast(ostream& fastfile){
 	fastfile<<"model main{\n\n";
@@ -337,35 +337,66 @@ void print_fast(ostream& fastfile){
 	for(int i=2;i<=last_used_label;++i){
 		fastfile<<", state_"<<i;
 	}
+	for(int i=1;i<=dummy_states_needed;++i){
+		fastfile<<",dummy_"<<i;
+	}
 	fastfile<<";\n\n";
 
 	for(int i=1;i<=last_used_label;++i){
 		CFG_location* state = label_map[i];
 		for(int j=0;j<state->edges.size();++j){
-			fastfile<<"\ttransition t_"<<i<<"_"<<j<<" := {\n";
-			fastfile<<"\t\tfrom\t:= state_"<<i<<";\n";
-			fastfile<<"\t\tto\t:= state_"<<state->edges[j].next->label<<";\n";
-			fastfile<<"\t\tguard\t:= ";
-			if(state->edges[j].guard==NULL){
-				fastfile<<"true";
-			}
-			else{
-				node* temp;
-				if(label_map[i]->invariant!=NULL){
-					temp = and_node(state->edges[j].guard,label_map[i]->invariant);
+			if(state->edges[j].change!=NULL){
+				if(state->edges[j].change->delta!=0){
+					//Add the four transitions and the job would be done
 				}
 				else{
-					temp = state->edges[j].guard;
+					fastfile<<"\ttransition t_"<<i<<"_"<<j<<" := {\n";
+					fastfile<<"\t\tfrom\t:= state_"<<i<<";\n";
+					fastfile<<"\t\tto\t:= state_"<<state->edges[j].next->label<<";\n";
+					fastfile<<"\t\tguard\t:= ";
+					if(state->edges[j].guard==NULL){
+						fastfile<<"true";
+					}
+					else{
+						node* temp;
+						if(label_map[i]->invariant!=NULL){
+							temp = and_node(state->edges[j].guard,label_map[i]->invariant);
+						}
+						else{
+							temp = state->edges[j].guard;
+						}
+						temp->print(fastfile,"&&","||","",true);
+					}
+					fastfile<<";\n";
+					fastfile<<"\t\taction\t:= ";
+					if(state->edges[j].change!=NULL){
[…]

### 2017-06-28 implementation/main.cpp
Sample assignments functionality added

@@ -348,6 +348,58 @@ void print_fast(ostream& fastfile){
 			if(state->edges[j].change!=NULL){
 				if(state->edges[j].change->delta!=0){
 					//Add the four transitions and the job would be done
+					//diverge 1
+					++last_dummy_state_used;
+					fastfile<<"\ttransition t_"<<i<<"_"<<j<<"_diverge1"<<" := {\n";
+					fastfile<<"\t\tfrom\t:= state_"<<i<<";\n";
+					fastfile<<"\t\tto\t:= dummy_"<<last_dummy_state_used<<";\n";
+					if(state->edges[j].guard!=NULL){
+						cerr<<"guard in an assignment"<<endl;
+					}
+					fastfile<<"\t\tguard\t:= true;\n";
+					fastfile<<"\t\taction\t:= ;\n\t};\n\n";
+					//converge 1
+					fastfile<<"\ttransition t_"<<i<<"_"<<j<<"_converge1"<<" := {\n";
+					fastfile<<"\t\tfrom\t:= dummy_"<<last_dummy_state_used<<";\n";
+					fastfile<<"\t\tto\t:= state_"<<state->edges[j].next->label<<";\n";
+					if(state->edges[j].guard!=NULL){
+						cerr<<"guard in an assignment"<<endl;
+					}
+					fastfile<<"\t\tguard\t:= true;\n";
+					fastfile<<"\t\taction\t:= ";
+					if(state->edges[j].change!=NULL){
+						fastfile<<variable[state->edges[j].toChange]<<"' = ";
+						state->edges[j].change->print(fastfile,"&&","||","",true);
+						fastfile<<"+"<<state->edges[j].change->delta;
+					}
+					fastfile<<";\n\t};\n\n";
+					
+
+					//diverge 2
+					++last_dummy_state_used;
+					fastfile<<"\ttransition t_"<<i<<"_"<<j<<"_diverge2"<<" := {\n";
+					fastfile<<"\t\tfrom\t:= state_"<<i<<";\n";
+					fastfile<<"\t\tto\t:= dummy_"<<last_dummy_state_used<<";\n";
+					if(state->edges[j].guard!=NULL){
+						cerr<<"guard in an assignment"<<endl;
+					}
+					fastfile<<"\t\tguard\t:= true;\n";
+					fastfile<<"\t\taction\t:= ;\n\t};\n\n";
+					//converge 2
[…]

### 2017-06-29 implementation/files/Parser.cpp
Large ifs and some examples added

@@ -78,32 +78,33 @@ node::node(string t, string l, bool& to_return){
 	type = t;
 	delta = 0.0;
 	if(t=="bexpr"){
-		children.resize(1);
-		children[0] = new node("affexpr");
 		stringstream line_stream;
 		line_stream.str(l);
-		string comma_sep;
-		while(getline(line_stream,comma_sep,',')){
-			skip_spaces(comma_sep);
-			bool return_value = false;
-			node* formed_literal = new node("literal",comma_sep, return_value);
-			children[0]->children.pb(formed_literal);
-			if(return_value){
-				node* other_literal = new node("literal");
-				other_literal->constant = "<=";
-				other_literal->children.resize(1);
-				other_literal->children[0] = new node("expr");
-				other_literal->children[0]->expression.resize(nVariables+1);
-				for(int i=0;i<=nVariables;++i){
-					other_literal->children[0]->expression[i] = -1.0*formed_literal->children[0]->expression[i];
-				}
-				children[0]->children.pb(other_literal);
-				// formed_literal->print();
-				// cout<<endl;
-				// other_literal->print();
-				// cout<<endl;
+		string semicolon_sep;
+		while(getline(line_stream,semicolon_sep,';')){	
+			node* affexpr_temp = new node("affexpr");
+			stringstream line_stream1;
+			line_stream1.str(semicolon_sep);
+			string comma_sep;
+			while(getline(line_stream1,comma_sep,',')){
+				skip_spaces(comma_sep);
+				bool return_value = false;
+				node* formed_literal = new node("literal",comma_sep, return_value);
+				affexpr_temp->children.pb(formed_literal);
+				if(return_value){
+					node* other_literal = new node("literal");
+					other_literal->constant = "<=";
+					other_literal->children.resize(1);
+					other_literal->children[0] = new node("expr");
+					other_literal->children[0]->expression.resize(nVariables+1);
+					for(int i=0;i<=nVariables;++i){
+						other_literal->children[0]->expression[i] = -1.0*formed_literal->children[0]->expression[i];
+					}
+					affexpr_temp->children.pb(other_literal);
 
+				}
 			}
+			children.pb(affexpr_temp);
 		}
 	}
 	else if(t=="literal"){
@@ -472,7 +473,7 @@ void node::analyse_expr(int begin,int end,bool negate){
 				skip_spaces(upper);
 				double lower_bound = stod(lower);
[…]

### 2017-06-29 implementation/large_ifs.cpp

@@ -1,0 +1,78 @@
+#include <iostream>
+#include <algorithm>
+#include <cstdio>
+#include <vector>
+#include <cstring>
+#include <sstream>
+#include <cctype>
+#include <set>
+#include <map>
+#include <fstream>
+#include <deque>
+#include <iomanip>
+#include <fstream>
+#include <queue>
+#include "files/Parser.h"
+
+using namespace std;
+
+#define MAXL 100000 //Maximum length of the program
+#define part(x,a,b) (x.substr((a),((b)-(a))))
+#define pb push_back
+
+int main(){
+	ofstream programfile;
+	programfile.open("example.prob");
+	programfile<<"var x,y";
+	cout<<"Input a value of n(>=0):\n";
+	int n;
+	cin>>n;
+	nVariables = n;
+	for(int i=1;i<=n;++i){
+		variable[i] = "y"+to_string(i);
+		variableId["y"+to_string(i)] = i;
+	}
+	for(int i=1;i<=n;i++){
+		programfile<<",y"<<i;
+	}
+	programfile<<";\nwhile x>=0 and y>=0 do\n\nif * then x:=x-1; y:=2*y else y:=y-1 fi";
+	if(n==0){
+		programfile<<"\n\n";
+	}
+	else{
+		programfile<<";\n\n";
+		bool temp;
+		node* condition = new node("bexpr","y1<=0;y1>=0",temp);
+		for(int i=2;i<=n;++i){
+			node* temp_condition = new node("bexpr","y" + to_string(i) + "<=0;y" + to_string(i) + ">=0",temp);
+			condition = and_node(condition,temp_condition);
+		}
+		for(int i=0;i<condition->children.size();++i){
+			if(i!=0){
+				programfile<<"else\n";
+			}
+			programfile<<"if ";
+			condition->children[i]->print(programfile,"and","or","*");
+			programfile<<" then skip";
+			programfile<<endl;
+		}
+		programfile<<"else skip\n";
+		for(int i=0;i<condition->children.size();++i){
+			programfile<<"fi";
+			programfile<<endl;
+		}
+		// for(int i=0;i<condition->children.size();++i){
+		// 	if(i!=0){
+		// 		programfile<<";\n";
+		// 	}
+		// 	programfile<<"if ";
+		// 	condition->children[i]->print(programfile,"and","or","*");
+		// 	programfile<<" then skip else skip fi";
+		// }
+	}
+
+	programfile<<"\n\nod"<<endl;
[…]

### 2017-06-29 implementation/main.cpp

@@ -132,11 +132,11 @@ int epsilons_used = 0;
 void generate_equations(){ //Would use the ostream file to write the equations into it later
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
 		if(it->second->invariant==NULL){
-			cout<<endl;
+			// cout<<endl;
 			continue;
 		}
-		it->second->invariant->print();
-		cout<<endl;
+		// it->second->invariant->print();
+		// cout<<endl;
 		if(it->second->type=="det"){
 			//Invariant and guard imply the value decrease
 			if(it->second->edges.empty()){
@@ -539,8 +539,9 @@ int main(){
 	
 	generate_equations();
 	ofstream equationsfile;
-	for(int loop_counter=0;loop_counter<100;++loop_counter){	
-		cout<<"Iteration"<<loop_counter+1<<"->"<<endl;
+	bool solved = false;
+	int loop_counter=0;
+	for(;loop_counter<100;++loop_counter){	
 		equationsfile.open("files/equations.lp");
 		ostream* equation_output_file = &equationsfile;
 		print_equations(*equation_output_file);
@@ -552,7 +553,6 @@ int main(){
 		//Processing the EquationsOutput file
 		bool state = process_equations_output();
 		if(state==false){
-			cout<<"No solution possible"<<endl;
 			string command;
 			command = "mv files/EquationsOutput files/EquationsOutput" + to_string(loop_counter);
 			system(command.c_str());
@@ -560,22 +560,27 @@ int main(){
 		}
 		else{
 			//Some equation was deleted, some epsilon was 1
-			// cout<<equations.begin()->first<<" "<<epsilons_used<<endl;
 			if(equations.begin()->first>epsilons_used){
-				cout<<"Solution found"<<endl;
+				solved = true;
 				string command;
 				command = "mv files/EquationsOutput files/EquationsOutput" + to_string(loop_counter);
 				system(command.c_str());
 				break;
 			}
 			else{
-				cout<<"Going into another iteration"<<endl;
 			}
 		}
 		string command;
 		command = "mv files/EquationsOutput files/EquationsOutput" + to_string(loop_counter);
 		system(command.c_str());
 		// temporarily for just one iteration
+	}
+	cout<<"The number of states in pCFG generated is "<<last_used_label<<endl;
+	if(solved){
[…]

### 2017-06-29 paper/compositional.tex

@@
 
-The purpose of the typesystem is, in principle, to ensure that the stochastic process which, in each time step, returns the current value of the PVSM, is uniformly integrable inside the nested loops.
-Uniform integrability is a deep concept from probability and measure theory: a sequence $X_0,X_1,X_2,\dots$ of random variables is uniformly integrable if for each $\delta>0$ there exists an $K\in \Nset$ such that for all $n\geq 0$ it holds $\E[|X_n|\cdot\indicator{X_n \geq K}]\leq \delta$.
-As noted in~\cite{HolgerPOPL}, proofs of uniform integrability are hard to (semi)-automatize, and hence they devise the typechecking as a sound (but not complete) method of proving uniform integrability.
-Take for example \textbf{[PETR: ADD EXAMPLE]}.
-However, as we show below, for none of these programs there exists a PVSM that typechecks.
+The purpose of the typesystem is, in principle, to ensure that the stochastic 
+process which, in each time step, returns the current value of the PVSM, is 
+uniformly integrable inside the nested loops. Uniform integrability is a deep 
+concept from probability and measure theory: a sequence $X_0,X_1,X_2,\dots$ of 
+random variables is uniformly integrable if for each $\delta>0$ there exists an 
+$K\in \Nset$ such that for all $n\geq 0$ it holds $\E[|X_n|\cdot\indicator{X_n 
+\geq K}]\leq \delta$. As noted in~\cite{HolgerPOPL}, proofs of uniform 
+integrability are hard to (semi)-automatize, and hence they devise the 
+typechecking as a sound (but not complete) method of proving uniform 
+integrability. This is illustrated in the following example.
+
+\begin{example}
+Consider the two \APP{}s in Figure~\ref{fig:uniint}, that differ only in one 
+coefficient in the 
+assignment on line 5. For the inner loop there exists (in both cases) a 
+1-dimensional linear ranking supermartingale whose value in each location is 
+equal to $c+d_{\loc}$, where $d_{\loc}$ is a location-specific constant. Since 
+the expected change of $c$ in each loop step is $-0.5$, this is indeed a LRSM. 
+Also, in both cases, a LEM of the form $x+d_{\loc}$, again for some suitable 
+location-specific constants $d_{\loc}$, is a PVSM for the outer loop, as $x$ 
+decreases and is non-negative within the outer loop and its expected change is 
+non-negative within the inner loop (more precisely, the inner-loop expected 
+change of $x$ zero in the left program and $-\frac{x}{4} - \frac{3}{4}$ in the 
+right 
+program). 
+However, the variable $x$ is uniformly integrable within the inner loop of the 
+right program while for the left program this does not hold. To see this, 
+imagine the inner loop as a stand-alone program and let $X_n$ be the value of 
+variable $x$ after $n$ steps of this stand-alone program (i.e., when the loop 
+terminates $x$ no longer changes). Solving a simple linear recurrence shows 
+that in the right program $\E[X_n] \rightarrow 0$ as $n\rightarrow \infty$, 
+which in particular shows uniform integrability of $X_0,X_1,X_2,\dots$. On the 
+other hand, in the left program for each $K>0$ we have $\probm(X_K \geq 
+2^K\cdot{x_0})\geq \frac{1}{2^K}$, where $x_0$ is the value of $x$ upon 
+entering the loop. Hence $\E[|X_K|cdot\indicator{X_K 
+\geq K}]\geq 1$ for each $K$ sufficiently large, which is incompatible with 
+uniform integrability. The example shows that proving uniform integrability 
+requires intricate reasoning about quantitative behaviour of the program. 
+Moreover, as shown below, none of the two programs have a PVSM that typechecks.
+
+% compute the distribution of 
+%$x$ in this stand-alone program after $n$ steps. In both cases, $x$ has value 
+%$0$ with probability 
+%$1-\frac{1}{2^n}$, as this is the probability of setting $x$ to zero in the 
+%first $n$ steps, after which the loop terminates. With the remaining 
+%probability $\frac{1}{2^n}$ the value of $x$ is $2^n \cdot x_0 - 2^{n+1}+2$ in 
+%the left program and $(\frac{3}{2})^n \cdot x_{0} -3\cdot((\frac{3}{2})^n +1)$ 
+%in the right program, where $x_0$ is the value of $x$ upon entering the inner 
+%loop (both equations are obtained by solving a simple linear recurrence). 
+%Hence, for the right program, the expectation of $x$ after $n$ steps of the 
+%inner loop (viewed as a stand-alone program) tends to zero as $n\rightarrow 
+%\infty$, which in particular shows uniform integrability. However, in left 
+%inner loop, $\E[|X_n|\cdot\indicator{X_n 
+%	\geq }]\geq 1$ 
+\end{example}
+%However, as we show below, for none of these programs there exists a PVSM that 
+%typechecks.
+
+\lstset{language=affprob}
+\lstset{tabsize=2}
+\newsavebox{\uniinto}
+\begin{lrbox}{\uniinto}
+\begin{lstlisting}[mathescape]
+while $x\geq 0$ do
+	$c:=1$;
+	while $x\geq 1$ and $c\geq 1$ do
+		if prob(0.5) then $x:=0$
+			else $x:=2(x-1)$ fi;
+		if prob(0.5) then $c:=0$ 
+			else skip fi
+	od;
+	$x:=x-1$
+od
+\end{lstlisting}
+\end{lrbox}
+\newsavebox{\uniintt}
+\begin{lrbox}{\uniintt}
+	\begin{lstlisting}[mathescape]
+while $x\geq 0$ do
+	$c:=1$;
+	while $x\geq 1$ and $c\geq 1$ do
+		if prob(0.5) then $x:=0$
+			else $x:=\frac{3}{2}(x-1)$ fi;
+		if prob(0.5) then $c:=0$ 
+			else skip fi
+	od;
+	$x:=x-1$
+od
+	\end{lstlisting}
+\end{lrbox}
+\begin{figure}[t]
+	%%\centering
+	\usebox{\uniinto}
+	\hspace{1cm}
+	\usebox{\uniintt}
+\label{fig:uniint}
+\caption{Examples of programs with (right) and without (left) uniformly 
+integrable PVSMs.}
+\end{figure}
+
+\lstset{language=affprob}
+\lstset{tabsize=2}
+\newsavebox{\unintthree}
+\begin{lrbox}{\unintthree}
+	\begin{lstlisting}[mathescape]
+	while $x\geq 0$ do
+	while $y\geq 0$ do
+	z:=x;
+	while $z\geq 0$ do
+	z:=z-1;
+	x:=x-1
+	od;
+	y:=y-1
+	od;
+	$x:=x-1$
+	od
+	\end{lstlisting}
+\end{lrbox}
+\begin{figure}[t]
+	%%\centering
+	\usebox{\unintthree}
+	%	\hspace{1cm}
+	%	\usebox{\uniintt}
+	\label{fig:uniint2}
+	\caption{Program where the outer loop does not have a PVSM that 
+		typechecks.}.
+\end{figure}
 
@@
 \item
-A PVSM $\lem$ for \PP{} $\program$ will not typecheck if $\program$ has a nested loop in which the value of $\lem$ can change unboundedly in a single step (see Example~\ref{xxx}). 
+A PVSM $\lem$ for \PP{} $\program$ will not typecheck if $\program$ has a 
+nested loop in which the value of $\lem$ can change unboundedly in a single 
+step (see Figure~\ref{fig:uniint}). 
 \item

### 2017-06-29 paper/compositional.tex
added examples

@@
 
-\textbf{[PETR: INSERT EXAMPLE]}
+%\textbf{[PETR: INSERT EXAMPLE]}
+\begin{example}
+Consider the program $\program$ in Figure~\ref{fig:uniint4} and its associated 
+pCFG. Then $\loops(\program)=\{\loc_2,\loc_3\}$ and 
+$\slice(\program)=\{\loc_0,\loc_1,\loc_4,\loc^{\lout}\}$.
+\end{example}
 
@@
 the expected change of $c$ in each loop step is $-0.5$, this is indeed a LRSM. 
-Also, in both cases, a LEM of the form $x+d_{\loc}$, again for some suitable 
-location-specific constants $d_{\loc}$, is a PVSM for the outer loop, as $x$ 
+Also, in both cases, a LEM of the form $x+d_{\loc}'$, again for some suitable 
+location-specific constants $d_{\loc}'$, is a PVSM for the outer loop, as $x$ 
 decreases and is non-negative within the outer loop and its expected change is 
@@
 	\usebox{\uniintt}
-\label{fig:uniint}
 \caption{Examples of programs with (right) and without (left) uniformly 
 integrable PVSMs.}
+\label{fig:uniint}
 \end{figure}
@@
 	while $y\geq 0$ do
-	z:=x;
+		$z:=x$;
 		while $z\geq 0$ do
-	z:=z-1;
-	x:=x-1
+			$z:=z-1$;
+			$x:=x-1$
 		od;
-	y:=y-1
+		$y:=y-1$
+	od;
+	$x:=x+$sample$(\mathit{Uniform[-3,1]})$
+od
+	\end{lstlisting}
+\end{lrbox}
+\newsavebox{\unintfour}
+\begin{lrbox}{\unintfour}
+	\begin{lstlisting}[mathescape]
+while $x\geq 0 $ do
+	$y:=x$;
+	while $y\geq 1$ do
+		$y:=y+$sample$(\mathit{Uniform[-3,1]})$
 	od;
@@
 	%%\centering
+	\begin{subfigure}{0.45\textwidth}
+		\usebox{\unintfour}
+		\vspace{0.2cm}
+		\begin{tikzpicture}[x=1.8cm,y=1.4cm]
+		\node[det] (init) at (0,0) {$\loc_0$};
+		\node[det] (fin) at (0,-1) {$\loc^{\lout}$};
+		\node[det] (as1) at (1,0) {$\loc_1$};
+		\node[det] (inner) at (2,0) {$\loc_2$};
+		\node[det] (as2) at (2,-1) {$\loc_3$};
+		\node[det] (as3) at (3,0) {$\loc_4$};
+%		
+		\draw[tran] (init) to node[font=\scriptsize,draw, fill=white, 
+		rectangle,pos=0.5] {$x<0$} (fin);
+		\draw[tran] (init) to node[font=\scriptsize,draw, fill=white, 
+		rectangle,pos=0.5] {$x\geq 0$} (as1);
+		\draw[tran] (as1) -- node[auto,font=\scriptsize] {$y:=x$} (inner);
+		\draw[tran] (inner) to node[font=\scriptsize,draw, fill=white, 
+		rectangle,pos=0.5] {$y<1$} (as3);
+		\draw[tran] (inner) to node[font=\scriptsize,draw, fill=white, 
+		rectangle,pos=0.5] {$y\geq 1$} (as2);
+		\draw[tran,rounded corners] (as2) -- (1.25,-1) -- 
+		node[font=\scriptsize, 
+		label={[xshift=-0.4cm,yshift=-0.2cm,font=\scriptsize] 
+		{$y:=\dots$}}] 
+		{} 
+		(inner);
+		\draw[tran, rounded corners] (as3) -- (3,0.5) -- node[font=\scriptsize, 
+		label={[font=\scriptsize, yshift=-0.1cm] {$x:=x-1$}}] {} (0,0.5) -- 
+		(init);
+		\end{tikzpicture}
+		%	\hspace{1cm}
+		%	\usebox{\uniintt}
+		\caption{Example for slicing illustration: program and its pCFG.}
+		\label{fig:uniint4}
+	\end{subfigure}
+	\hspace{0.5cm}
+	\begin{subfigure}{0.45\textwidth}
 	\usebox{\unintthree}
@@
 	%	\usebox{\uniintt}
+	\caption{Program where the outer loop does not have a PVSM that 
+		typechecks.}
 	\label{fig:uniint2}
-	\caption{Program where the outer loop does not have a PVSM that 
-		typechecks.}.
+	\end{subfigure}
+\caption{Program illustrations.}
+\label{fig:comp-ex}
 \end{figure}
@@
 \item
-A PVSM $\lem$ for  \PP{} $\program$ will not typecheck if $\program$ has a nested loop which itself has a nested loop in which some variable appearing in some expression in $\lem$ is modified, see Example~\ref{xxx}.\footnote{Both these statements regarding typechecking failure are somewhat simplified, even in these two cases the PVSM might sometimes typecheck correctly, in case where the nested loops are followed by assignments which completely overwrite the effect of these loops, e.g.
-if \textbf{[PETR: POINTER TO AN EXAMPLE]}.
-However, the statements intuitively summarize the major reasons for typechecking failure.
-}
+A PVSM $\lem$ for  \PP{} $\program$ will not typecheck if $\program$ has a 
+nested loop which itself has a nested loop in which some variable appearing in 
+some expression in $\lem$ is modified, see 
+Figure~\ref{fig:uniint2}.\footnote{Both these statements regarding typechecking 
+failure are somewhat simplified, even in these two cases the PVSM might 
+sometimes typecheck correctly, in case where the nested loops are followed by 
+assignments which completely overwrite the effect of these loops, e.g. if 
+the program in Figure~\ref{fig:uniint2} contained an assignment $x:=0$. 
+However, the statements intuitively 
+summarize the major reasons for typechecking failure. }
 \end{compactenum}
@@
 
-Hence, NCSMs effectively trade the uniform integrability condition for non-negativity over the whole program.
-We believe that the latter condition is substantially easier to impose and check automatically, especially in the case of affine probabilistic programs and (propositionally) linear NCSMs: for \APP{}s, synthesizing NCSM entails synthesizing sufficient program invariants (for which there is a good automated tool support~\cite{xxx}) that overapproximate set of reachable configurations and encoding the rankingn, unaffection, and non-negativity conditions into a collection of linear constraints (as for general LinLexRSMs in Section~\ref{xxx}).
-Examples~\ref{xxx} show instances where attempts to prove of a.s.
-termination via PVSMs fail while proofs via LinNCSMs work.
+Hence, NCSMs effectively trade the uniform integrability condition for 
+non-negativity over the whole program. We believe that the latter condition is 
+substantially easier to impose and check automatically, especially in the case 
+of affine probabilistic programs and (propositionally) linear NCSMs: for 
+\APP{}s, synthesizing NCSM entails synthesizing sufficient program invariants 
+(for which there is a good automated tool support~\cite{xxx}) that 
+overapproximate set of reachable configurations and encoding the rankingn, 
+unaffection, and non-negativity conditions into a collection of linear 
+constraints (as for general LinLexRSMs in Section~\ref{xxx}). 
+Figures~\ref{fig:uniint} and~\ref{fig:uniint2} show instances where attempts to 
+to prove of a.s. termination via PVSMs fail while proofs via LinNCSMs work.
 
+\begin{example}
+For all programs in Figures~\ref{fig:uniint} and~\ref{fig:uniint2} there are 
+LinNCSMs for all the loops in the program, which shows that the programs 
+terminate 
+a.s. In Figure~\ref{fig:uniint}, for both programs the inner loops have 
+LinNCSMs of the form $c+d_{\loc}$, for $d_{\loc}$ a location-specific constant, 
+while the outer loops have LinNCSMs of the form $x+d_{\loc}'$. In 
+Figure~\ref{fig:uniint2} the program similarly has LinNCSMs defined, proceeding 
+from the innermost loop and neglecting the location-specific constants, by 
+variables $z$, $y$, $x$.
+\end{example}
 %While we defined NCSMs as one-dimensional measurable maps to make them analogous to
@@
 
-The soundness of the algorithm follows from Theorems~\ref{thm:nonneg-comp} and \ref{thm:holger-comp}.
-Note that we use the PVSM-based algorithm of~\cite{HolgerPOPL} as a back-up sub-procedure for the case when LinNCSM-based proof fails.
-Hence, Algorithm~\ref{algo:compositional} can compositionally prove a.s.
-termination of strictly larger class of programs than the PVSM-based algorithm alone.
+The soundness of the algorithm follows from Theorems~\ref{thm:nonneg-comp} and 
+\ref{thm:holger-comp}. Note that we use the PVSM-based algorithm 
+of~\cite{HolgerPOPL} as a back-up sub-procedure for the case when LinNCSM-based 
+proof fails. Hence, Algorithm~\ref{algo:compositional} can compositionally 
+prove a.s. termination of strictly larger class of programs than the PVSM-based 
+algorithm alone.
 

### 2017-06-30 implementation/large_ifs.cpp
small memory optimisation

@@ -74,5 +74,5 @@ int main(){
 	programfile<<"\n\nod"<<endl;
 	programfile.close();
 
-	system("./a.out < example.prob > large_output");
+	// system("./a.out < example.prob > large_output");
 }

### 2017-06-30 implementation/main.cpp

@@ -19,9 +19,10 @@ Comments:
 #include <iomanip>
 #include <fstream>
 #include <queue>
+#include <ctime>
 #include "files/Parser.h"
 using namespace std;
-#define MAXL 100000 //Maximum length of the program
+#define MAXL 300000 //Maximum length of the program
 #define part(x,a,b) (x.substr((a),((b)-(a))))
 #define pb push_back
 #define A(i,j) (front->affexpr->children[i]->children[0]->expression[j+1])
@@ -459,14 +460,20 @@ void print_fast(ostream& fastfile){
 
 int main(){
 	int start,end;
-	char input[MAXL];
+	char* input = new char[MAXL];
 	// Setting precision to the printing of the double variables in program
 	// cout<<fixed<<setprecision(10);
 	int r,i;
-	for(i=0;(r=getchar())!=EOF;i++)
+	for(i=0;(r=getchar())!=EOF;i++){
+		if(i>MAXL){
+			cerr<<"Program longer than the maximum allowed limit of "<<MAXL<<endl;
+			exit(0);
+		}
 		input[i]=r;
+	}
 	input[i]=0;
 	program=input;
+	delete[] input;
 	int begin = 0;
 	int endprog = program.length();
 	skip_spaces(begin,endprog);

### 2017-06-30 implementation/files/Parser.h
Output format changed

@@ -62,6 +62,7 @@ public:
 };
 
 void skip_spaces(int &begin, int &end);
+void skip_spaces(string& line);
 void vcopy(vector<node*> &sink,vector<node*> &tocopy);
 node* negation(node* tonegate);
 node* and_node(node* one,node* two);

### 2017-06-30 implementation/main.cpp

@@ -22,7 +22,7 @@ Comments:
 #include <ctime>
 #include "files/Parser.h"
 using namespace std;
-#define MAXL 300000 //Maximum length of the program
+#define MAXL 700000 //Maximum length of the program
 #define part(x,a,b) (x.substr((a),((b)-(a))))
 #define pb push_back
 #define A(i,j) (front->affexpr->children[i]->children[0]->expression[j+1])
@@ -459,6 +459,7 @@ void print_fast(ostream& fastfile){
 }
 
 int main(){
+	double start_time = time(0);
 	int start,end;
 	char* input = new char[MAXL];
 	// Setting precision to the printing of the double variables in program
@@ -486,28 +487,52 @@ int main(){
 	// CFG_edge last_edge(label_map[end],-1,NULL);
 	// label_map[end]->edges.pb(last_edge);
 	root=new node("stmt",begin,endprog,start,end);
+	double parsing_over = time(0);
+	cout<<"Time:"<<(double(parsing_over-start_time))<<endl;
+	cout<<"Parsing over and CFG constructed"<<endl;
+	cout<<"The CFG constructed has "<<last_used_label<<" states"<<endl;
 	ofstream fastfile;
 	fastfile.open("files/aspic.fast");
 	ostream* outfast = &fastfile;
 	print_fast(*outfast);
 	fastfile.close();
+	double printed_fast = time(0);
+	cout<<endl<<"Time:"<<(double(printed_fast-start_time))<<endl;
+	cout<<"Fast file printed."<<endl;
 	if(system("./files/invariant_script.sh")!=0){
 		cout<<"Something wrong with the script to find invariants";
 	}
-	
+	double aspic_done = time(0);
+	cout<<endl<<"Time:"<<(double(aspic_done-start_time))<<endl;
+	cout<<"Invariants generated using aspic"<<endl;
 	// // Code for analysing files/InvariantOutput comes here
 	ifstream invariant_file("files/InvariantOutput");
 	for(int i=1;i<=last_used_label;){
 		string line;
 		getline(invariant_file,line);
+		skip_spaces(line);
 		size_t open = line.find('{');
 		if(open!=string::npos){
+			int first_space = -1;
+			for(int i=0;i<line.length();++i){
+				if(isspace(line[i])){
+					first_space = i;
+					break;
+				}
+			}
+			if(first_space==-1){
+				cerr<<"Error! No space in the Invariant output concerned line"<<endl;
+			}
+			if(part(line,0,6)!="state_"){
+				continue;
[…]

### 2017-06-30 paper/lexicographic-programs.tex
added example with infinite expected termination time

@@
 \end{proof}
+
+We conclude this section by showing that (Lin)LexRSMs can, unlike 1-dimensional 
+RSMs, prove a.s. 
+termination of programs whose expected termination time is infinite.
+\lstset{language=affprob}
+\lstset{tabsize=2,escapechar=\&}
+\newsavebox{\infas}
+\begin{lrbox}{\infas}
+	\begin{lstlisting}[mathescape]
+$x:=1$;$c:=1$ 
+while $c\geq 1$ do				
+	if prob(0.5) then			
+		$x:=2\cdot x$			
+	else						
+		$c:=0$					
+	fi							
+od								
+while $x\geq 0$ do $x:=x-1$ od	
+	\end{lstlisting}
+\end{lrbox}
+\newsavebox{\infast}
+\begin{lrbox}{\infast}
+	\begin{lstlisting}[mathescape]
+
+$(6c+2,x)$
+$(6c+1,x)$
+$(6c+3,x)$
+
+$(3,x)$
+
+$(0,x)$	
+	\end{lstlisting}
+\end{lrbox}
+\newsavebox{\infastinv}
+\begin{lrbox}{\infastinv}
+	\begin{lstlisting}[mathescape]
+
+$[c\geq 0 \wedge x\geq 1]$
+$[c\geq 1 \wedge x\geq 1]$
+$[c\geq 1 \wedge x\geq 1]$
+
+$[c\geq 1 \wedge x \geq 1]$
+
+$[x\geq 0]$	
+	\end{lstlisting}
+\end{lrbox}
+\begin{figure}[t]
+	\centering
+	\usebox{\infas}
+	\hspace{0.1cm}
+	\usebox{\infast}
+	\hspace{0.1cm}
+	\usebox{\infastinv}
+\caption{An a.s. terminating program with infinite expected termination time. A 
+2-dimensional $1$-LinLexRSM map for the program is given on the right, along 
+with the 
+supporting invariants in square brackets. The invariants and a LinLexRSM on 
+each 
+line belong to the program location in which the program is \emph{before} 
+executing the command on that line. The function is indeed a $1$-LinLexRSM, 
+since in the probabilistic branching location $\loc$ we have 
+$\preexp{\lem}(\loc,(x,c))=3c+3$ and $3c+3\leq 6c+1-1$ for all $c\geq 1$.} 
+\label{fig:inftime}
+\end{figure}
+
+\begin{example}
+Consider the program in Figure~\ref{fig:inftime}. It is easy to see that it 
+terminates a.s., but the expected termination time is infinite: to see this, 
+note that that the expected value of variable $x$ upon reaching the second loop 
+is $\frac{1}{2}\cdot 1 + \frac{1}{4}\cdot 2 + \frac{1}{8}\cdot 4 + \cdots = 
+\frac{1}{2}+\frac{1}{2}+\frac{1}{2}+\cdots=\infty$ and that the time needed to 
+get out of the second loop is equal to the value of $x$ upon entering the loop. 
+However, a.s. termination of the program is proved by a 2-dimensional LinLexRSM 
+$\lem$ pictured in the figure.
+\end{example}

### 2017-06-30 implementation/files/Parser.cpp
Adding comments in the code

@@ -1,6 +1,4 @@
 //Compile with C++11
-//The variable names should be of the form x_i where i is a number and the indexing starts from 1 i.e. the variables in the program are of the form x_1, x_2 ... x_n.
-//Assuming * is for multiplication rather than .
 #include <iostream>
 #include <algorithm>
 #include <cstdio>
@@ -21,6 +19,7 @@ using namespace std;
 #define part(x,a,b) (x.substr((a),((b)-(a))))
 #define pb push_back
 
+//variable declarations
 string program;
 int nVariables;
 int last_used_label = 0;
@@ -28,7 +27,14 @@ map<int,CFG_location*> label_map;
 map<string,int> variableId;
 map<int,string> variable;
 int dummy_states_needed = 0;
+node* root;
 
+/*
+
+	begin and end are the pointers to the concerned string in program
+	So, for parsing purposes, skip_spaces changes the pointers accordingly so that the extra spaces at the start and end are out of consideration
+
+*/
 void skip_spaces(int &begin, int &end){ //skips the spaces in the beginning and the end
 	while(begin<program.size() and isspace(program[begin])){
 		begin++;
@@ -38,6 +44,11 @@ void skip_spaces(int &begin, int &end){ //skips the spaces in the beginning and 
 	}
 }
 
+/*
+
+	line is the concerned string and may contain extra spaces at the begin and end, skip_spaces deletes the extra spaces from start and end of line
+
+*/
 void skip_spaces(string& line){
 	int begin = 0,end = line.length();
 	while(begin<line.size() and isspace(line[begin])){
@@ -49,6 +60,12 @@ void skip_spaces(string& line){
 	line = part(line,begin,end);
 }
 
+
+/*
+
+	Constructor when only the type of node is specified
+
+*/
 node::node(string t){
 	bracket = NULL;
 	begin = -1;
@@ -59,6 +76,16 @@ node::node(string t){
 	delta = 0.0;
 }
 
+/*
+
+	A general constructor when the
+		b: start pointer of the program
+		e: end pointer of the program
+		s: the id of start node in pCFG
+		l: rhe id of final node in pCFG
+		negate: a special bool passed for nodes of type "expr" when the expression was of the form -(string1) and now the concerned string becomes string1 and negate becomes 1
+
+*/
 node::node(string t, int b, int e, int s, int l, bool negate){
 	bracket = NULL;
 	type = t;
[…]

### 2017-06-30 implementation/files/Parser.h

@@ -1,21 +1,29 @@
 #include <string>
 using namespace std;
 
+/*
+
+	A class used to parse the program easily
+	This helps in the recursive implementation of parsing
+
+*/
 struct node{
 	string type; //The type of entity that this node corresponds to
 	string constant; //The constant factor (subtype) of this node
-	bool strict;
+	bool strict; // For the literal nodes, this defines whether the inequality is strict or not
 	int begin,end; //begin and end point in program
-	vector<node *> children; //children in the parse tree
-	node *bracket; //bexpr written in bracket next to this node (for stmt only)
-	vector<double> expression; //for expressions only
-	double delta;
-	//constructor
+	vector<node *> children; //children in the parse tree, the order of children is custom wrt the type of node
+	node *bracket; //bexpr written in bracket next to this node (for stmt only), it helps to store the invariant if any was present in the initial program description
+	vector<double> expression; //for expressions only, it stores the coefficient vector of the corresponfing expression
+	double delta; // For the expressions in which a probabilistic assignment is there, this is the padding which has to be applied on the either side of rhs value
+	
+	//constructors
 	node(string t);
 	node(string t, int b, int e, int s = 0, int l = 0, bool negate = false);
 	node(string t, string line, bool& to_return);
 private:
-//procs
+	
+	//procs
 	void proc_stmt(int s,int l);
 	void proc_assgn(int s,int l);
 	void proc_affexpr(); //
@@ -30,14 +38,25 @@ private:
 	void proc_ndbexpr();
 	void form_vector(int begin,int end,bool negate);
 	void form_vector(string l,bool negate);
-//main process
+	
+	//main process
 	void process(int s, int l, bool negate);
 public:
 	void print(ostream& outputfile = cout,string and_string = "and", string or_string = "or", string multiply_string = "", bool bruteforce = false);
 };
 
+/*
+
+	A node in the CFG
+
+*/
 struct CFG_location;
 
+/*
+
+	An edge (directed) in the CFG
+
+*/
 struct CFG_edge{
 public:
 	CFG_location* next;
@@ -68,11 +87,12 @@ node* negation(node* tonegate);
 node* and_node(node* one,node* two);
 int find_variables(int& begin, int& end);
 
-extern string program;
-extern int nVariables;
-extern int last_used_label;
-extern map<int,CFG_location*> label_map;
-extern node* root;
-extern map<string,int> variableId;
-extern map<int,string> variable;
+//extern because these variables are needed in main file also
+extern string program; // The program which is to be parsed
[…]

### 2017-06-30 implementation/main.cpp

@@ -462,8 +462,6 @@ int main(){
 	double start_time = time(0);
 	int start,end;
 	char* input = new char[MAXL];
-	// Setting precision to the printing of the double variables in program
-	// cout<<fixed<<setprecision(10);
 	int r,i;
 	for(i=0;(r=getchar())!=EOF;i++){
 		if(i>MAXL){
@@ -473,10 +471,15 @@ int main(){
 		input[i]=r;
 	}
 	input[i]=0;
+/*
+
+	program is the string used for generating the pCFG and stores the program as a whole
+
+*/
 	program=input;
 	delete[] input;
 	int begin = 0;
-	int endprog = program.length();
+	int endprog = program.length();	
 	skip_spaces(begin,endprog);
 	nVariables = find_variables(begin,endprog); //To find the number of different variables of the type x_i in the program
 	start = ++last_used_label;

### 2017-06-30 implementation/files/Parser.cpp
Further commenting and documentation

@@ -1,4 +1,3 @@
-//Compile with C++11
 #include <iostream>
 #include <algorithm>
 #include <cstdio>
@@ -26,7 +25,7 @@ int last_used_label = 0;
 map<int,CFG_location*> label_map;
 map<string,int> variableId;
 map<int,string> variable;
-int dummy_states_needed = 0;
+int dummy_states_needed = 0; // The states needed for the if * transformation of non deterministic assignments
 node* root;
 
 /*

### 2017-06-30 implementation/files/Parser.h

@@ -71,12 +71,12 @@ public:
 
 struct CFG_location{
 public:
-	int label;
-	string type; // det ndet and prob
-	vector<CFG_edge> edges;
 	CFG_location(string type,int label);
-	vector<int> ranking_function;
-	node* invariant; // A bexpr node
+	int label;	// The label of this node, or the reverse mapping of what is stored in the "label_map" map
+	string type; // det ndet or prob
+	vector<CFG_edge> edges; // The set of edges that emerge from this node
+	vector<int> ranking_function; // The coefficient matrix for the affine ranking function for this node
+	node* invariant; // A bexpr node, the invariant at this particular node
 	void print();
 };
 

### 2017-06-30 implementation/main.cpp

@@ -130,33 +130,31 @@ map<int,equation*> equations;
 int equations_counter = 0;
 int epsilons_used = 0;
 
-void generate_equations(){ //Would use the ostream file to write the equations into it later
+/*
+
+	This function fills in the map "equations" with all the equations initially with the relevant conditions and changes
+
+*/
+void generate_equations(){
+	// Iterate through all the states and their edges to generate the relevant decrement conditions
 	for(map<int,CFG_location*>::iterator it = label_map.begin();it!=label_map.end();++it){
 		if(it->second->invariant==NULL){
-			// cout<<endl;
+			// The invariant was "false", so just leave all the transitions of this state
 			continue;
 		}
-		// it->second->invariant->print();
-		// cout<<endl;
 		if(it->second->type=="det"){
 			//Invariant and guard imply the value decrease
 			if(it->second->edges.empty()){
-				//Code for the last node
+				//Code for the last node. No condition
 			}
 			else if(it->second->edges.size()==1){
 				//First make a condition
 				cond* condition = new cond(++equations_counter,it->second->edges[0].toChange,it->second->edges[0].change,it->first,it->second->edges[0].next->label);
-				// Guard would have been NULL here
-				if(it->second->invariant==NULL){
-					//It is the case when the invariant is false
-				}
-				else{
-					// Invariant implies the given condition
-					int nEquations = it->second->invariant->children.size();
-					for(int i=0;i<nEquations;i++){
-						equations[equations_counter] = new equation(it->second->invariant->children[i],condition);
-						// equations.push(new equation(it->second->invariant->children[i],condition));
-					}
+				// Guard would have been true here
+				// Invariant implies the given condition
+				int nEquations = it->second->invariant->children.size();
+				for(int i=0;i<nEquations;i++){
+					equations[equations_counter] = new equation(it->second->invariant->children[i],condition);
 				}
 			}
 			else{
@@ -228,6 +226,11 @@ void generate_equations(){ //Would use the ostream file to write the equations i
 
 int last_used_lambda = 0;
 
+/*
+
+	Print the equations.lp file, it just prints the constraints which are still there in the map "equations"
+
+*/
 void print_equations(ostream& equationsfile){
 	last_used_lambda = 0;
[…]

### 2017-06-30 paper/lexicographic-programs.tex
removed pre-expectation in favor of ranking transitions re-arranged the section according to this change

@@
 
-We aim to define a function assigning a non-negative vector to each configuration (so called measurable map) such that in each point of computation, the expected value of the function after performing one more computational step is smaller (in lexicographic ordering) than the current one.
-This property can be formalized using the standard  notion of \emph{pre-expectation}~\cite{xxx}.
+We aim to define a function assigning a non-negative vector to each 
+configuration (so called measurable map) such that in each point of 
+computation, the expected value of the function after performing one more 
+computational step is smaller (in lexicographic ordering) than the current one. 
+We formalize this property below.
+%This property can be formalized using the standard  notion of 
+%\emph{pre-expectation}~\cite{xxx}.
 
@@
 
-\begin{definition}[Pre-Expectation]
-	Let $\lem$ 
-	be 1-dimensional measurable map for a \PP{} $\program$.
-	%Let $\locs$ 
-	%be the set of locations of $\pCFG_P$ and $\pvars$ its set of program 
-	%variables. 
-	The 
-	pre-expectation of $\lem$ is a function $\preexp{\lem}\colon \locs\times 
-	\Rset^{|\pvars|} \rightarrow \Rset\cup\{\infty\}$ defined as follows:
+We now formalize the notion of a transition in a pCFG being ranked by a 
+measurable map. We first define this notion for transitions that do not go out 
+of a probabilistic branching location, as these require a special treatment.
+
+\begin{definition}
+\label{def:rank1}
+Let $\lem$ be a measurable map, $(\loc,\vec{x})$ be a configuration such that 
+$\loc\not\in \locsPB$ and let 
+$\tau=(\loc,\loc')$ be a 
+transition outgoing from $\loc$. For an $\eps\geq 0$ we say that $\tau$ is 
+\emph{$\eps$-ranked by $\lem$ from $(\loc,\vec{x})$} if the following 
+conditions are 
+satisfied, depending on the type of $\loc$:
 \begin{compactitem} %\itemsep1pt \parskip0pt \parsep0pt
+%		\item 
+%		if $\loc$ is a probabilistic branching location, then
+%		$$\preexp{\lem}(\loc,\vec{x}):=\sum_{(\loc,\loc')\in\transitions} 
+%		Pr_{\loc}\left(\loc,\loc'\right)\cdot
+%		\lem(\loc',\vec{x});$$
 		\item 
-		if $\loc$ is a probabilistic branching location, then
-		$$\preexp{\lem}(\loc,\vec{x}):=\sum_{(\loc,\loc')\in\transitions} 
-		Pr_{\loc}\left(\loc,\loc'\right)\cdot
-		\lem(\loc',\vec{x});$$
-		\item 
-		if $\loc$ is a non-deterministic branching location, then
-		$$\preexp{\lem}(\loc,\vec{x}):=
-		\max_{(\loc,\loc')\in\transitions}\lem(\loc',\vec{x});$$
+		if $\loc$ is a deterministic or non-deterministic branching location, 
+		then
+		$$\lem(\loc',\vec{x}) \leq 
+		\lem(\loc,\vec{x})-\eps;$$
 
 		%\item
-		%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\min_{(\loc,id,\ell')\in\transitions}\eta\left(\loc',\mathbf{x}\right)$
+
+%%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\min_{(\loc,id,\ell')\in\transitions}\eta\left(\loc',\mathbf{x}\right)$
 		% if $\loc$ is an angelic location;
+%		\item 
+%		if $\loc$ is a deterministic location, then there is 
+%exactly one 
+%		transition
+%		$\tau=(\loc,\loc')$ such that $\vec{x}\models G(\tau)$. We put 
+%$$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x})$$.
 		\item 
-		if $\loc$ is a deterministic location, then for each $\vec{x}$ the value 
-		$\preexp{\lem}(\loc,\vec{x})$ is determined as follows: there is exactly one 
-		transition
-		$\tau=(\loc,\loc')$ such that $\vec{x}\models G(\tau)$. We put $$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x})$$.
-		\item 
-		If $\loc$ is an assignment location, then there is exactly one transition $\tau=(\loc,\loc')$ outgoing from $\loc$.
-		We distinguish 
-		three cases, depending on $\updates(\tau)=(j,\up)$ (recall that $\up$ is an update element):
+		If $\loc$ is an assignment location, then we distinguish 
+		three cases, depending on $\updates(\tau)=(j,\up)$ (recall that $\up$ 
+is an update element):
 		\begin{compactitem}
-			\item If $\up\colon \Rset^{|\pvars|}\rightarrow \Rset$ is a Borel-measurable function, then 
-			$$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x}(j\leftarrow \up(\vec{x}))).$$
-			\item If $\up$ is a distribution $d$, then $$ \preexp{\lem}(\loc,\vec{x}):= 
-			\lem(\loc',\vec{x}(j\leftarrow \E[d])),$$ where $\E[d]$ is the expected value of the 
+			\item If $\up\colon \Rset^{|\pvars|}\rightarrow \Rset$ is a 
+Borel-measurable function, then we require
+			$$\lem(\loc',\vec{x}(j\leftarrow 
+\up(\vec{x}))) \leq \lem(\loc,\vec{x})-\eps$$
+			\item If $\up$ is a distribution $d$, then we require $$ 
+			\lem(\loc',\vec{x}(j\leftarrow \E[d])) \leq 
+			\lem(\loc,\vec{x})-\eps,$$ 
+			where $\E[d]$ is the 
+expected value of the 
 			distribution $d$.
 			\item 
-			If $\up$ is a set, then $$ \preexp{\lem}(\loc,\vec{x}):= \sup_{a\in\up}
-			\lem(\loc',\vec{x}(j \leftarrow a)).$$
+			If $\up$ is a set, then we require $$\sup_{a\in\up}
+			\lem(\loc',\vec{x}(j \leftarrow a)) \leq \lem(\loc,\vec{x})-\eps.$$
 		\end{compactitem}
-		%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\eta(\loc',\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right))$
-		% if $\loc$ is a deterministic location, $(\loc,f,\loc')\in\transitions$ and 
+
+%%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\eta(\loc',\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right))$
+		% if $\loc$ is a deterministic location, 
+%$(\loc,f,\loc')\in\transitions$ and 
 		%$\mathbf{x}\in G(\loc,f,\loc')$, where 
-		%$\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right)$ is the expected value of 
-		%$f(\vec{x},\cdot)$. %on $\rvars$ if $\vec{x}$ is treated as a constant vector.
+		%$\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right)$ is the expected 
+%value of 
+		%$f(\vec{x},\cdot)$. %on $\rvars$ if $\vec{x}$ is treated as a constant 
+%%vector.
 	\end{compactitem}
-	%\end{definition}
 \end{definition}
 
-Intuitively, a pre-expectation of $\lem$ is a 
-function which for each configuration $(\loc,\vec{x})$ returns the maximal
-expected value of $\lem$ after one step is made from this configuration, where 
-the maximum is taken over all possible non-deterministic choices.
+Since ranking supermartingales are required to decrease on average, for 
+individual transitions outgoing from $\locsPB$ it does not make sense to say 
+that they are ranked or not. Instead, for each $\loc\in\locsPB$ we consider all 
+outgoing transitions together.
+
+\begin{definition}
+\label{def:rank2}
+Let $\lem$ be a measurable map, and let $(\loc,\vec{x})$ be a configuration with
+$\loc\in \locsPB$. For an $\eps\geq 0$ we say that $\loc$ is $\eps$-ranked by 
+$\lem$ from $(\loc,\vec{x})$ if 
+$$\sum_{(\loc,\loc')\in\transitions} 
+		Pr_{\loc}\left(\loc,\loc'\right)\cdot
+		\lem(\loc',\vec{x}\leq \lem(\loc,\vec{x})-\eps.$$
+\end{definition}
+
+To capture the specific of $\locsPB$, we introduce the 
+notion of \emph{generalized transition}.
+
+\begin{definition}
+A generalized transition of a pCFG $\pCFG$ is either a transition of $\pCFG$ 
+outgoing 
+from location not in $\locsPB$ or a location $\loc\in\locsPB$.
+\end{definition}
+
+Intuitively, we represent the set of transitions outgoing from $\loc\in\locsPB$ 
+by the source location $\loc$.  For generalized transitions 
+$\tilde{\tau}=\loc\in\locsPB$ we say that $\tilde{\tau}$ is outgoing from 
+$\loc$.
+
+Definitions~\ref{def:rank1} and~\ref{def:rank2} 
+define when is a generalized transition $\eps$-ranked by $\lem$ from 
+configuration $(\loc,\vec{x})$. We say that a 
+generalized transition is \emph{unaffected} by $\lem$ from $(\loc,\vec{x})$ if 
+it is $0$-ranked by $\lem$ from $(\loc,\vec{x})$.
+
+%\begin{definition}[Pre-Expectation]
+%	Let $\lem$ 
+%	be 1-dimensional measurable map for a \PP{} $\program$.
+%	%Let $\locs$ 
+%	%be the set of locations of $\pCFG_P$ and $\pvars$ its set of program 
+%	%variables. 
+%	The 
+%	pre-expectation of $\lem$ is a function $\preexp{\lem}\colon \locs\times 
+%	\Rset^{|\pvars|} \rightarrow \Rset\cup\{\infty\}$ defined as follows:
+%	\begin{compactitem} %\itemsep1pt \parskip0pt \parsep0pt
+%		\item 
+%		if $\loc$ is a probabilistic branching location, then
+%		$$\preexp{\lem}(\loc,\vec{x}):=\sum_{(\loc,\loc')\in\transitions} 
+%		Pr_{\loc}\left(\loc,\loc'\right)\cdot
+%		\lem(\loc',\vec{x});$$
+%		\item 
+%		if $\loc$ is a non-deterministic branching location, then
+%		$$\preexp{\lem}(\loc,\vec{x}):=
+%		\max_{(\loc,\loc')\in\transitions}\lem(\loc',\vec{x});$$
+%		
+%		%\item 
+%		
+%%%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\min_{(\loc,id,\ell')\in\transitions}\eta\left(\loc',\mathbf{x}\right)$
+%		% if $\loc$ is an angelic location;
+%		\item 
+%		if $\loc$ is a deterministic location, then for each $\vec{x}$ the 
+%value 
+%		$\preexp{\lem}(\loc,\vec{x})$ is determined as follows: there is 
+%exactly one 
+%		transition
+%		$\tau=(\loc,\loc')$ such that $\vec{x}\models G(\tau)$. We put 
+%$$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x})$$.
+%		\item 
+%		If $\loc$ is an assignment location, then there is exactly one 
+%transition $\tau=(\loc,\loc')$ outgoing from $\loc$.
+%		We distinguish 
+%		three cases, depending on $\updates(\tau)=(j,\up)$ (recall that $\up$ 
+%is an update element):
+%		\begin{compactitem}
+%			\item If $\up\colon \Rset^{|\pvars|}\rightarrow \Rset$ is a 
+%Borel-measurable function, then 
+%			$$\preexp{\lem}(\loc,\vec{x}):= \lem(\loc',\vec{x}(j\leftarrow 
+%\up(\vec{x}))).$$
+%			\item If $\up$ is a distribution $d$, then $$ 
+%\preexp{\lem}(\loc,\vec{x}):= 
+%			\lem(\loc',\vec{x}(j\leftarrow \E[d])),$$ where $\E[d]$ is the 
+%expected value of the 
+%			distribution $d$.
+%			\item 
+%			If $\up$ is a set, then $$ \preexp{\lem}(\loc,\vec{x}):= 
+%\sup_{a\in\up}
+%			\lem(\loc',\vec{x}(j \leftarrow a)).$$
+%		\end{compactitem}
+%		
+%%%$\mathrm{pre}_\eta(\loc,\mathbf{x}):=\eta(\loc',\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right))$
+%		% if $\loc$ is a deterministic location, 
+%%$(\loc,f,\loc')\in\transitions$ and 
+%		%$\mathbf{x}\in G(\loc,f,\loc')$, where 
+%		%$\expv_{\rvars}\left(f(\vec{x},\mathbf{r})\right)$ is the expected 
+%%value of 
+%		%$f(\vec{x},\cdot)$. %on $\rvars$ if $\vec{x}$ is treated as a constant 
+%%%vector.
+%	\end{compactitem}
+%	%\end{definition}
+%\end{definition}
+
+%Intuitively, a pre-expectation of $\lem$ is a 
+%function which for each configuration $(\loc,\vec{x})$ returns the maximal
+%expected value of $\lem$ after one step is made from this configuration, where 
+%the maximum is taken over all possible non-deterministic choices.
 %Note that the only way in which pre-expectation can attain an infinite value 
@@
  	\item 
- 	there exists $1\leq j \leq$ n such that
+ 	for each generalized transition $\tilde{\tau}$ outgoing from $\loc$ there 
+ 	exists $1\leq j 
+ 	\leq$ n such that
  	\begin{compactitem}
  	\item
- 	$\preexp{\lem_j}(\loc,\vec{x}) \leq \lem_j(\loc,\vec{x})-\eps$, and
+ 	$\tilde{\tau}$ is $\eps$-ranked by $\lem_j$ from $(\loc,\vec{x})$
  	\item
- 	for all $1\leq j'<j$ it holds 
- 	$\preexp{\lem_{j'}}(\loc,\vec{x}) \leq \lem_{j'}(\loc,\vec{x})$.
+ 	for all $1\leq j'<j$ we have that $\tilde{\tau}$ is unaffected by 
+ 	$\lem_{j'}$ from $(\loc,\vec{x})$.
  	\end{compactitem}
@@
 the set of all runs $\run$ such that $\ttime(\run)>i$ and for $\run$ the index 
-$j$ is the smallest one such that $\preexp{\lem_j}(\cfg{\sigma}{i}(\run)) \leq 
-\lem_j(\cfg{\sigma}{i}(\run))-\eps$ (which entails 
-$\preexp{\lem_{j'}}(\cfg{\sigma}{i}(\run)) =
-\lem_{j'}(\cfg{\sigma}{i}(\run))-\eps$ for all $1\leq j'< j$). Due to 
+$j$ is the smallest one such that the $(i+1)$-th transition on $\run$ is ranked 
+by $\lem_j$ from $\cfg{\sigma}{i}(\run)$. Due to 
 definition of an $\eps$-LexRSM map such a $j$ exists for all 
@@
 for each $1\leq j 
-\leq n$, that $\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
-\natfilt_i]\leq\preexp{\lem_{j}}(\cfg{\sigma}{}) $. This can be achieved by a 
-standard~\cite{xxx}, through somewhat technical computation, which we defer to 
-\AppendixMaterial.
+\leq n$ and $j'<j$, that $\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
+\natfilt_i]\leq X_i[j]-\eps $ on $L_j^i$ and 
+$\E^\sigma_{\vecinit}[\vecseq{X}{i+1}{j}\mid 
+\natfilt_i]\leq X_i[j] $ on $L_{j'}^i$. This follows easily from 
+the definition of $L^{i}_1,\dots,L^{i}_n$ and from the definition of a 
+transition being $\eps$-ranked by $\lem_j$.
 

### 2017-07-01 paper/lexicographic-programs.tex
started writing the section on algorithm

@@
 We now have all the ingredients needed to define the notion of LexRSM maps for probabilistic programs.
+For notational convenience, we extend the function $\guards$ (which assigns guards to deterministic transitions) to the set of all generalized transition: for a generalized transition $\tau'$ which is not a standard transition outgoing from deterministic location, we put $\guards(\tau')=0\leq 0 \equiv \mathit{true}$.
 
@@
  	\item 
- 	for each generalized transition $\tilde{\tau}$ outgoing from $\loc$ there 
+ 	for each generalized transition $\tilde{\tau}$ outgoing from $\loc$ such that $\vec{x}\models\guards(\tilde\tau)$ there 
  	exists $1\leq j 

### 2017-07-01 paper/synthesis.tex
started writing the section on algorithm

\section{Algorithmic Aspects}

In this section we describe a polynomial-time algorithm for synthesizing linear $\eps$-LexRSM maps in affine probabilistic programs supported by a given linear invariant map $\inv$. The algorithm, based on iterative solving of linear constraints, is an adaptation of a similar algorithm for finding lexicographic ranking functions in non-probabilistic programs~\cite{xxx}. Hence, we provide only a high-level description, focusing on new aspects of our adaptation.

The main idea is to iteratively synthesize 1-dimensional linear expression map that $1$-rank a subset of generalized transitions. These maps form the individual components of the sought-after $1$-LinLexRSM map. In each iteration, we start with a set $U$ of the yet-unranked generalized transitions. We seek a 1-dimensional LEM which ranks the maximal number of elements if $U$, and is unaffected by the remaining elements of $U$ (here by ranking a generalized transition $\tilde\tau$ we mean ranking it from each configuration $(\loc,\vec{x})$, where $\loc$ is the source location of $\tilde\tau$ and $\vec{x}\in\inv(\loc)$). If no 1-dimensional LEM that would rank at least one element in $U$ exists, then there is no LinLexRSM map for the program. Otherwise, we remove the newly ranked elements from $U$ and continue into the next iteration, until $U$ becomes empty. The process is summarized in Algorithm~\ref{xxx}.

Hence, the main computational task of the algorithm is to check, for a given set of generalized transition $U$, whether there exists a 1-dimensional LEM $\lem$ such that:
\begin{compactenum}
\item for each $\tilde{\tau}\in U$ and  each configuration $(\loc,\vec{x})$ where $\loc$ is the source of $\tilde{\tau}$ and $\vec{x}\in \inv(\loc)$ we have that $\lem(\loc,\vec{x})\geq 0$ and that $\tilde{\tau}$ is unaffected by $\lem$; and
\item there is $\tilde{\tau}\in U$ that is $1$-ranked by $\lem$, from each configuration $(\loc,\vec{x})$ where $\loc$ is the source of $\tilde{\tau}$ and $\vec{x}\in \inv(\loc)\cap\{\vec{x}'\mid \vec{x}'\models \guards(\tilde{\tau})\}$; we then say that $\lem$ ranks $\tilde{\tau}$ w.r.t. $\inv$.
\end{compactenum}
  Moreover, if such an LEM $\lem$ exists, the algorithm has to find one that maximizes the number of gen. transitions in $U$ ranked by it. Both these tasks can be accomplished by the standard  method of linear constraints based on the use of Farkas's lemma, which was widely use for synthesis of termination proofs in both probabilistic and non-probabilistic programs~\cite{xxx}. That is, the algorithm first constructs, for each location $\loc$ a \emph{template} for $\lem$, i.e. an expression of the form $a_1^{\loc}x_1 + \cdots + a_{|\pvars|}^{\loc}x_{|\pvars|} = b^{\loc} $, where $x_1,\dots,x_{|\pvars|}$ are program variables and $a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc} $ are yet unknown coefficients. That is, supplying concrete values for all the unknown coefficients yields an LEM. Using the construction provided e.g. in~\cite{xxx} we construct in polynomial time, for each generalized transition $\tilde{\tau}$, a system of linear constraints over set of variables $\{ a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc}\mid \loc\in\locs \}\cup \{\eps_{\tilde\tau}\}$

### 2017-07-01 paper/synthesis.tex
added the description of the algorithm

@@
 That is, supplying concrete values for all the unknown coefficients yields an LEM.
-Using the construction provided e.g.
-in~\cite{xxx} we construct in polynomial time, for each generalized transition $\tilde{\tau}$, a system of linear constraints over set of variables $\{ a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc}\mid \loc\in\locs \}\cup \{\eps_{\tilde\tau}\}$
+Now the conditions (1) and (2) above can be expressed using linear constraints on the coefficients.
+More precisely, using the construction provided e.g.
+in~\cite{xxx} (which includes a use of the Farkas's lemma) we construct in polynomial time, for each generalized transition $\tilde{\tau}$, a system of linear constraints $\linsystem_{\tilde{\tau}}$ over set of variables $\{ a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc}\mid \loc\in\locs \}\cup \{\eps_{\tilde\tau}\}\cup F$, where $F$ is the set of fresh variables (not appearing in any template) and $\eps_{\tilde\tau} $ is constrained to be non-negative.
+Each solution of the system $\linsystem_{\tilde{\tau}}$ yields a LEM which satisfies the constraint (1) for $\tilde{\tau}$.
+Moreover, each solution of $\linsystem_{\tilde{\tau}}$ yields a LEM which $\eps_{\tilde{\tau}}$-ranks $\tilde{\tau}$.
+To find a LEM which satisfies both constraints (1) and (2) as well as maximizes the number of 1-ranked elements of $U$ it is sufficient to construct $\linsystem_{\tilde{\tau}}$ for each $\tilde{\tau}\in U$ and solve the following linear program $\lp_{U}$:
+  \begin{align*}
+\text{maximize }  &\sum_{\tilde{\tau} \in U} \eps_{\tilde\tau} \text{ subject to constraints}\\[0.3cm]
+&\linsystem_{\tilde{\tau}}\,;\quad\quad \quad\quad\tilde{\tau}\in U\\
+&0 \leq \eps_{\tilde{\tau}} \leq 1\,; \quad \tilde{\tau}\in U
+  \end{align*}
+
+Each system $\linsystem_{\tilde{\tau}}$ is constructed in such a way that if it admits a solution with some $\eps_{\tilde{\tau}}$ positive, then decreasing the value of $\eps_{\tilde{\tau}}$ in that solution to any non-negative value still yield a valid solution (this corresponds to the fact that if some transition is $\eps$-ranked by $\lem$, than it is $\eps'$-ranked by $\lem$ for each $0\leq \eps'\leq \eps$).
+Morever, each solution where $\eps_{\tilde{\tau}}$ is positive can be rescaled into another solution in which $\eps_{\tilde{\tau}}$ is at least $1$.
+It follows that of $\lp_{U}$ has at least one feasible solution, then it has an optimal solution in which each $\eps_{\tilde{\tau}}$ is either $0$ or $1$.
+If the system does not have a solution or all the $\eps_{\tilde{\tau}}$ are equal to zero then there is no LEM satisfying (1) and (2).
+Otherwise, the optimal solution of $\lp_{U}$ yields a LEM $\lem$ which satisfies both (1) and (2) and maximizes the number of 1-ranked elements of $U$.
+
+This polynomial-time linear-programming step is used as a sub-procedure in Algorithm~\ref{algo:linlexrsm} for LinLexRSM synthesis.
+
+\begin{algorithm}
+\SetKwInOut{Input}{input}\SetKwInOut{Output}{output}
+\DontPrintSemicolon
+
+\Input{An \APP{} $\program$ together with an invariant map $\inv$.}
+\Output{A multi-dimensional LinLexRSM if it exists, otherwise ``No LinLexRSM''}
+$U \leftarrow \text{ all generalized transitions of } \pCFG_{\program}$\;
+$d\leftarrow 0$\;
+\While{$U$ is non-empty}{
+	$d \leftarrow d+1$\;
+	construct and solve $\lp_{U}$\;
+	\If{$\lp_U$ does not have feasible solution or optimal value is $0$}{\Return{No LinLexRSM}}
+	\Else{$\sol\leftarrow$ optimal solution of $\lp_U$\;
+		$\lem_d \leftarrow $ the LEM $\lem$ induced by $\sol$\;
+		$U\leftarrow U \setminus \{\tilde{\tau}\mid \eps_{\tilde{\tau}}=1 \text{ in }\sol \}$}
+}
+\Return{$(\lem_1,\dots,\lem_d)$}
+\caption{Synthesis of LinLexRSMs for \APP{} s}
+\label{algo:linlexrsm}
+\end{algorithm}
+
+Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} follow by arguments identical to those presented in~\cite{xxx}. 
+\begin{theorem}
+	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ together with linear invariant map $\inv$.
+If teh algorithm returns a $d$-dimensional LEM $\vec{\lem}=(\lem_1,\dots,\lem_d)$, then $\vec{\lem}$ is a $1$-LinLexRSM map for $\program$ supported by $\inv$.
+Conversely, if the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and $\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  supported by $\inv$.
+\end{theorem}
+\begin{proof}[Proof (sketch)]
+For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the $i$-th iteration of the while loop.
+Then each $\tilde{\tau}\in U_{i}\setminus U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$.
+Since $U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 1-ranked by some component of $\vec{\lem}$.
+Since
+\end{proof}
+
+\

### 2017-07-02 paper/synthesis.tex
added small comment
changed some def

@@
 \begin{compactenum}
-\item for each $\tilde{\tau}\in U$ and  each configuration $(\loc,\vec{x})$ where $\loc$ is the source of $\tilde{\tau}$ and $\vec{x}\in \inv(\loc)$ we have that $\lem(\loc,\vec{x})\geq 0$ and that $\tilde{\tau}$ is unaffected by $\lem$; and
+\item for each location $\loc\in\locs$ and all $\vec{x}\in \inv(\loc)$ it holds $\lem(\loc,\vec{x})\geq 0$;
+\item for each $\tilde{\tau}\in U$ and  each configuration $(\loc,\vec{x})$ where $\loc$ is the source of $\tilde{\tau}$ and $\vec{x}\in \inv(\loc)\cap\{\vec{x}'\mid \vec{x}'\models \guards(\tilde{\tau})\}$ we have that $\tilde{\tau}$ is unaffected by $\lem$ from $(\loc,\vec{x})$; and
 \item there is $\tilde{\tau}\in U$ that is $1$-ranked by $\lem$, from each configuration $(\loc,\vec{x})$ where $\loc$ is the source of $\tilde{\tau}$ and $\vec{x}\in \inv(\loc)\cap\{\vec{x}'\mid \vec{x}'\models \guards(\tilde{\tau})\}$; we then say that $\lem$ ranks $\tilde{\tau}$ w.r.t.
@@
 in~\cite{xxx} (which includes a use of the Farkas's lemma) we construct in polynomial time, for each generalized transition $\tilde{\tau}$, a system of linear constraints $\linsystem_{\tilde{\tau}}$ over set of variables $\{ a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc}\mid \loc\in\locs \}\cup \{\eps_{\tilde\tau}\}\cup F$, where $F$ is the set of fresh variables (not appearing in any template) and $\eps_{\tilde\tau} $ is constrained to be non-negative.
-Each solution of the system $\linsystem_{\tilde{\tau}}$ yields a LEM which satisfies the constraint (1) for $\tilde{\tau}$.
+Each solution of the system $\linsystem_{\tilde{\tau}}$ yields a LEM which satisfies the constraints (1) and (2) for $\tilde{\tau}$.
 Moreover, each solution of $\linsystem_{\tilde{\tau}}$ yields a LEM which $\eps_{\tilde{\tau}}$-ranks $\tilde{\tau}$.
-To find a LEM which satisfies both constraints (1) and (2) as well as maximizes the number of 1-ranked elements of $U$ it is sufficient to construct $\linsystem_{\tilde{\tau}}$ for each $\tilde{\tau}\in U$ and solve the following linear program $\lp_{U}$:
+To find a LEM which satisfies all constraints (1)--(3) as well as maximizes the number of 1-ranked elements of $U$ it is sufficient to construct $\linsystem_{\tilde{\tau}}$ for each $\tilde{\tau}\in U$ and solve the following linear program $\lp_{U}$:
   \begin{align*}
@@
 It follows that of $\lp_{U}$ has at least one feasible solution, then it has an optimal solution in which each $\eps_{\tilde{\tau}}$ is either $0$ or $1$.
-If the system does not have a solution or all the $\eps_{\tilde{\tau}}$ are equal to zero then there is no LEM satisfying (1) and (2).
-Otherwise, the optimal solution of $\lp_{U}$ yields a LEM $\lem$ which satisfies both (1) and (2) and maximizes the number of 1-ranked elements of $U$.
+If the system does not have a solution or all the $\eps_{\tilde{\tau}}$ are equal to zero then there is no LEM satisfying (1)--(3).
+Otherwise, the optimal solution of $\lp_{U}$ yields a LEM $\lem$ which satisfies (1)--(3) and maximizes the number of 1-ranked elements of $U$.
 
@@
 \Return{$(\lem_1,\dots,\lem_d)$}
-\caption{Synthesis of LinLexRSMs for \APP{} s}
+\caption{Synthesis of LinLexRSMs for \APP{}s}
 \label{algo:linlexrsm}
@@
 Since $U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 1-ranked by some component of $\vec{\lem}$.
-Since
+Since \textbf{[PETR: REWRITE (1) TO REQUIRE NON-NEG EVERYWHERE]}
 \end{proof}

### 2017-07-02 paper/synthesis.tex
finished the algorithmic section

@@
 Since $U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 1-ranked by some component of $\vec{\lem}$.
-Since \textbf{[PETR: REWRITE (1) TO REQUIRE NON-NEG EVERYWHERE]}
+Non-negativity of each $\lem_i$ is ensured directly by $\lp_{U_i}$.
+Hence, it remains to show that each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$.
+But this follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the fact that each gen.
+transition in $U_j$ is unaffected by $\lem_j$.
+
+Proving completeness is more intricate; as pointed out in~\cite{xxx}, one needs to show that the greedy strategy of selecting LEM that ranks maximal number of remaining transitions does not cut off some possible LinLexRSMs.
+In~\cite{xxx} the completeness of the greedy strategy is proved by using several geometric arguments that exploit the fact that the underlying programs are affine.
+The same geometric properties hold for our \APP{}s (all ranking conditions in Definitions~\ref{def:rank1} and~\ref{def:rank2} are linear in program variables), the result is easily transferable.
 \end{proof}

### 2017-07-02 paper/extensions.tex
minor changes

\section{Bounds on Expected Termination Time}

As shown in Example~\ref{ex:infinite-time}, already LinLexRSM maps are capable of proving almost-sure termination of programs whose expected termination time is infinite. However, it is often desirable to obtain bounds on expected runtime of a program. In this section, we present a LexRSM-based proof rule for obtaining bounds on expected runtime, and we show how to automatize the usage of this proof rule to obtain polynomial bounds on expected runtime in a subclass of \PP{}s.

\begin{theorem}
\label{thm:runtime-bound}
\end{theorem}

### 2017-07-03 paper/conclu.tex
Added abs, intro, related, conclusion

\section{Related Work}
In this section we discuss the related works.

\smallskip\noindent{\em Probabilistic programs and termination.}
In early works the termination for concurrent probabilistic programs was studied 
as fairness~\cite{SPH84}, which ignored precise probabilities.
For countable state space a sound and complete characterization of almost-sure termination 
was presented in~\cite{HS85}, but nondeterminism was absent.
A sound and complete method for proving termination of finite-state programs
was given in~\cite{EGK12}.
For probablistic programs with countable state space and without 
nondeterminism, the {\em Lyapunov ranking functions} provide a sound and 
complete method to prove positive termination~\cite{BG05,Foster53}.
For probabilistic programs with nondeterminism, but restricted to discrete probabilistic
choices, the termination problem was studied in~\cite{MM04,MM05}.
The RSM-based (ranking supermatingale-based) approach extending ranking functions was first 
presented in~\cite{SriramCAV} for probabilistic programs without non-determinism,
but with real-valued variables, and its extension for probabilistic programs
with non-determinism has been studied in~\cite{HolgerPOPL,CFNH16:prob-termination,CAV-POLY,POPL,ARXIV}.
While all these results deeply clarify the role of RSMs for probabilistic programs, 
the notion of lexicographic RSMs to obtain a practical approach for termination 
analysis for probabilistic programs has not been studied before, which we consider in 
this work.

\smallskip\noindent{\em Other approaches.}
Besides RSMs, other approaches has also been considered for probabilistic programs.
A sound approach~\cite{DBLP:conf/sas/Monniaux01} for almost-sure termination 
is to explore the exponential decrease of probabilities upon bounded-termination 
through abstract interpretation~\cite{DBLP:conf/popl/CousotC77}.
MENTION PROOF RULES AND NOT AN AUTOMATED APPROACH.

\smallskip\noindent{\em Non-probabilistic programs.}
Termination analysis of non-probabilistic programs has also been extensively 
studied~\cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91,BMS05b,CSZ13,LJB01}.
Ranking functions are at the heart of the termination analysis, and lexicographic 
ranking function has emerged as one of the most efficient and practical approaches
for termination analysis~\cite{}.
In this work we extend such lexicographich ranking functions to probabilistic programs,
and present lexicographic RSMs for almost-sure termination analysis of probabilistic programs
with non-determinism.

\section{Conclusion and Future Work}
In this work we considered lexicographic RSMs for termination analysis of probabilistic
programs with nondeterminism.
We showed it presents a sound approach for almost-sure termination, that is compositional,
algorithmically efficient, and leads to approach that can handle realistic programs.
There are several interesting directions of future work.
Lexicographic ranking functions has been considered in several works to 
provide different practical methods for analysis of non-probabilistic programs.
First, while our work presents the foundations of lexicographic RSMs for probabilistic 
programs, extending other practical methods of lexicographic ranking functions to 
lexicographic RSMs is an interesting direction of future work.
Second, while our algorithmic approaches focus on the linear case, it would be interesting
to consider other non-linear, and polynomial functions in the future.

### 2017-07-03 paper/introduction.tex
Added abs, intro, related, conclusion

\begin{abstract}
Probabilistic programs extend classical imperative programs with 
real-valued random variables.
The most basic liveness property for such programs is the termination 
property.
The qualitative (aka almost-sure) termination problem asks given a probabilistic program
whether the program terminates with probability~1.
While ranking function provide a sound and complete method for non-probabilistic
programs, the extension of them to probabilistic programs is achieved
via ranking supermartingales (RSMs). 
While deep theoretical results have been establishes about RSMs, 
their application to probabilistic programs with nondeterminism has been limited
to academic examples. 
For non-probabilistic programs lexicographic ranking functions provide a compositional
and practical approach for termination analysis, which has been missing for
probabilistic programs. 
In this work we introduce lexicographic RSMs and show that they present a sound
method for almost-sure termination .
We show that lexicographic RSMs enjoy the desired property of compositionality,
and for programs with linear arithmetic they can be synthesized efficiently (in 
polynomial time).
We also show with additional restrictions even asymptotic bounds on expected
termination time can be obtained through lexicographi RSMs.
Finally, we present experimental results on abstraction of real-world programs
to demonstrate the effectiveness of our approach.
\end{abstract}

\section{Introduction}\label{sec:introduction}

\noindent{\em Probabilistic programs with nondeterminism.} 
Randomness play a fundamental role in many areas across science, and in 
computer science in particular.
In applications such as stochastic network protocols~\cite{BaierBook,prism},
randomized algorithms~\cite{RandBook,RandBook2}, machine learning~\cite{},
NATURE PAPER, the probabilistic behavior must be considered to faithfully model
the underlying dynamic system.
The extension of classical imperative programs with \emph{random value generators}, 
that produce random values according to some desired probability distribution, 
naturally gives probabilistic programs.
Along with probability, nondeterminism also plays a crucial role.
In particular in program analysis, for effective analysis of large programs,
all variables cannot be considered, and abstraction ignores some variables,
and the worst-case analysis is represented by adversarial nondeterminism.
Hence probabilistic programs with nondeterminism has become an active and 
important research area in program analysis.

\smallskip\noindent{\em Termination problem.} 
In static analysis of programs the most basic, as well most important, 
liveness property is the {\em termination} problem.
While for non-probabilistic programs the termination question asks whether
an input program {\em always} terminates, for probabilistic programs 
the termination questions accounts for the probabilistic behaviors. 
The most basic and fundamental extensions of the termination problem 
for probabilistic programs are:

\begin{compactenum}
\item \emph{Almost-sure termination.} 
The \emph{almost-sure termination} problem asks whether the program terminates with probability~1.

\item \emph{Positive termination.} 
The \emph{positive termination} problem asks whether the expected termination time is finite.
A related quantitative generalization of the positive termination question is to obtain 
asymptotic bounds on the expected termination time.

\end{compactenum}
While the positive termination implies almost-sure termination, the converse is not true (see~\cite[]{}).

\smallskip\noindent{\em Ranking functions and ranking supermartingales (RSMs).}
The key technique that applies for liveness analysis of non-probabilistic programs is 
the notion of {\em ranking functions}, which provides a sound and complete 
method for termination of non-probabilistic programs~\cite{rwfloyd1967programs}.
There exist a wide variety of different approaches for construction 
of ranking functions for non-probabilistic programs~\cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91}.
The generalization of ranking function to probabilistic programs is achieved through the
{\em ranking supermartingales (RSMs)}~\cite{}.
The ranking supermartingales provide a powerful and automated approach for termination 
analysis of probabilistic programs~\cite{SRIRAM,HOLGER,HONGFEI}, and algorithmic 
approaches for special cases such as linear and polynomial RSMs has
also been considered~\cite{POPL,CAV,POPL}.

\smallskip\noindent{\em .} 
While a impressive set of theoretical results related to RSMs has been
established, for probabilistic programs with nondeterminism the current approaches  
are only applicable to academic examples of variants of random walks. 
The key reason can be understood as follows: even for non-probabilistic programs while 
ranking functions are sound and complete, they do not necessarily provide a practical 
approach. 
In contrast, as a practical and scalable approach for non-probabilistic programs the 
notion of lexicographic ranking functions has been widely studied~\cite{}. 
Algorithmic approaches for lexicographic ranking functions allow the termination analysis 
to be applicable to real-world non-probabilistic programs (after abstraction).
However both the theoretical foundations as well as practical approaches related to 
such lexicographic ranking functions are completely missing for probabilistic programs,
which we address in this work.

\smallskip\noindent{\em Our contributions.} 
In this work our main contributions ranges from theoretical foundations of lexicographic
RSMs, to algorithmic approaches for them, to experimental results
showing their applicability for programs.
We describe our main contributions below:
\begin{enumerate}

\item {\em Theoretical foundations.} 
First, we introduce the notion of lexicographic RSMs, and show that
such supermartingales ensures almost-sure termination (Theorem~\ref{} in Section~\ref{}). 
Our first result is a purely mathematical result that introduces a new concept, and 
proves almost-sure termination, that is independent of any probabilistic program.
Based on the mathematical result we show that probabilistic programs with nondeterminism
the existence of a lexicographic ranking functions with respect to an invariant ensures
almost-sure termination (Theorem~\ref{} in Section~\ref{}).

\item {\em Compositionality.} 
Second we study the compositional properties of lexicographic RSMs.
A key limitation of the previous approaches related to RSMs is the lack 
of compositionality, and we show (in Section~\ref{}) how lexicographic RSMs 
present a compositional approach for almost-sure termination of probabilistic programs.

\item {\em Algorithm.} We then consider algorithms for synthesis of lexicographic RSMs, 
and for efficient algorithms we consider nondeterministic probabilistic 
programs that are {\em affine} (i.e., the arithmetic operations are linear)
[VARIABLE RANGE BOUNDED?]
We present a polynomial-time algorithm for synthesis of lexicographic RSMs 
for affine programs (see Theorem~\ref{}).

\item {\em Asymptotic bounds.} 
In general, the existence of lexicographic RSMs does not imply positive
termination. 
In other words, we present an example (Example~\ref{}) where lexicographic RSM exists ensuring
almost-sure termination, but the expected termination time is infinite.
We then present a natural restriction (i.e., a subclass of programs) such the lexicographic 
RSMs not only imply positive termination, but even asymptotic bounds on the expected termination 
time can be derived from them (see Theorem~\ref{}).

\item {\em Experimental results.} 
We present experimental results of our approach on realistic programs to show the applicability 
of our approach. 
To demonstrate the effectiveness of our approach we consider the benchmarks of non-probabilistic
programs from~\cite{} which are obtained as abstraction of real-world programs,where 
lexicographic ranking functions were applied for termination analysis. 
We extend these benchmarks with probabilistic statements and apply lexicographic RSMs to these
programs. 
Our experimental results show that our approach can handle these programs very efficiently.

\end{enumerate}

### 2017-07-03 paper/introduction.tex
OK

@@
-\begin{abstract}
-Probabilistic programs extend classical imperative programs with 
-real-valued random variables.
-The most basic liveness property for such programs is the termination 
-property.
-The qualitative (aka almost-sure) termination problem asks given a probabilistic program
-whether the program terminates with probability~1.
-While ranking function provide a sound and complete method for non-probabilistic
-programs, the extension of them to probabilistic programs is achieved
-via ranking supermartingales (RSMs). 
-While deep theoretical results have been establishes about RSMs, 
-their application to probabilistic programs with nondeterminism has been limited
-to academic examples. 
-For non-probabilistic programs lexicographic ranking functions provide a compositional
-and practical approach for termination analysis, which has been missing for
-probabilistic programs. 
-In this work we introduce lexicographic RSMs and show that they present a sound
-method for almost-sure termination .
-We show that lexicographic RSMs enjoy the desired property of compositionality,
-and for programs with linear arithmetic they can be synthesized efficiently (in 
-polynomial time).
-We also show with additional restrictions even asymptotic bounds on expected
-termination time can be obtained through lexicographi RSMs.
-Finally, we present experimental results on abstraction of real-world programs
-to demonstrate the effectiveness of our approach.
-\end{abstract}
-
 \section{Introduction}\label{sec:introduction}

### 2017-07-03 paper/introduction.tex
OK

@@
 that produce random values according to some desired probability distribution, 
-naturally gives probabilistic programs.
+naturally gives rise probabilistic programs.
 Along with probability, nondeterminism also plays a crucial role.
@@
 an input program {\em always} terminates, for probabilistic programs 
-the termination questions accounts for the probabilistic behaviors. 
+the termination questions must account for the probabilistic behaviors. 
 The most basic and fundamental extensions of the termination problem 
@@
 method for termination of non-probabilistic programs~\cite{rwfloyd1967programs}.
-There exist a wide variety of different approaches for construction 
+There exist a wide variety of approaches for construction 
 of ranking functions for non-probabilistic programs~\cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91}.
-The generalization of ranking function to probabilistic programs is achieved through the
+The generalization of ranking functions to probabilistic programs is achieved through the
 {\em ranking supermartingales (RSMs)}~\cite{}.
@@
 
-\smallskip\noindent{\em .} 
+\smallskip\noindent{\em Practical limitations of existing approaches.} 
 While a impressive set of theoretical results related to RSMs has been
-established, for probabilistic programs with nondeterminism the current approaches  
+established~\cite{}, for probabilistic programs with nondeterminism the current approaches  
 are only applicable to academic examples of variants of random walks. 
@@
 \smallskip\noindent{\em Our contributions.} 
-In this work our main contributions ranges from theoretical foundations of lexicographic
+In this work our main contributions range from theoretical foundations of lexicographic
 RSMs, to algorithmic approaches for them, to experimental results
@@
 Based on the mathematical result we show that probabilistic programs with nondeterminism
-the existence of a lexicographic ranking functions with respect to an invariant ensures
+the existence of a lexicographic RSM with respect to an invariant ensures
 almost-sure termination (Theorem~\ref{} in Section~\ref{}).
@@
 termination. 
-In other words, we present an example (Example~\ref{}) where lexicographic RSM exists ensuring
-almost-sure termination, but the expected termination time is infinite.
-We then present a natural restriction (i.e., a subclass of programs) such the lexicographic 
+In other words, we present an example (Example~\ref{}) where a lexicographic RSM exists ensuring
+almost-sure termination, yet the expected termination time is infinite.
+We then present a natural restriction (i.e., a subclass of programs) such that the lexicographic 
 RSMs not only imply positive termination, but even asymptotic bounds on the expected termination 

### 2017-07-03 paper/conclu.tex
expanded section on expected termination time

@@
 In early works the termination for concurrent probabilistic programs was studied 
-as fairness~\cite{SPH84}, which ignored precise probabilities.
+as fairness~1\cite{SPH84}, which ignored precise probabilities.
 For countable state space a sound and complete characterization of almost-sure termination 

### 2017-07-03 paper/extensions.tex
expanded section on expected termination time

@@
 
-As shown in Example~\ref{ex:infinite-time}, already LinLexRSM maps are capable of proving almost-sure termination of programs whose expected termination time is infinite.
-However, it is often desirable to obtain bounds on expected runtime of a program.
-In this section, we present a LexRSM-based proof rule for obtaining bounds on expected runtime, and we show how to automatize the usage of this proof rule to obtain polynomial bounds on expected runtime in a subclass of \PP{}s.
+As shown in Example~\ref{ex:infinite-time}, already LinLexRSM maps are capable 
+of proving almost-sure termination of programs whose expected termination time 
+is infinite. However, it is often desirable to obtain bounds on expected 
+runtime of a program. In this section, we present a LexRSM-based proof rule for 
+obtaining bounds on expected runtime, and we show how to automatize the usage 
+of this proof rule to obtain polynomial bounds on expected runtime in a 
+subclass of \PP{}s.
+
+As in the case of a.s. termination we start with general mathematical statement 
+about LexRSMs. We define a restricted class of LexRSMs with \emph{bounded 
+expected conditional increase} property. Recall from 
+definition~\ref{def:lexrsm} that 
+LexRSM for a stopping time $\stime$ is characterized by the possibility to 
+a.s. partition, for each $i\in \Nset_0$, the set $\{\omega\in \Omega\mid 
+\stime(\omega)>i \}$ into $d$ sets $L^i_1,\dots,L^i_d$ such that, intuitively, 
+on $L^i_j$ the conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
+is smaller than $\vec{X}_i[j]$, and for all $j'<j$, on $L^i_{j'}$ the 
+conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
+is no larger than $\vec{X}_i[j]$. This leaves the opportunity of conditional 
+expectation of $\vec{X}_{i+1}[j]$ being larger than $\vec{X}_i[j]$ on 
+$L^i_{j''}$ with $j''$. The conditional expected increase property bounds the 
+possibility of this increase.
+
+\begin{definition}
+Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be a 
+$d$-dimensional LexRSM for some stopping time $\stime$, defined w.r.t. some 
+filtration $\{\genfilt_i\}_{i=0}^{\infty}$. We say that 
+$\{\vec{X}_{i}\}_{i=0}^{\infty}$ has \emph{$c$-bounded expected conditional 
+increase (ECI),} for 
+some $c\geq 0$, if for each $i\in \Nset_0 $ and each $1\leq j \leq d $ it holds 
+$\E[\vecseq{X}{i+1}{j}\mid \genfilt_i]\leq \vecseq{X}{i}{j}+c$ on $L^i_{j''}$, 
+for all $j''>j$ 
+(here $L^i_1,\dots,L^i_d$ are as in Definition~\ref{def:lexrsm}).
+\end{definition}
+
+For LexRSMs with $c$-bounded ECI we have the following result.
 
@@
 \label{thm:runtime-bound}
+
 \end{theorem}

### 2017-07-03 paper/lexicographic.tex
added some examples
removed pre-expectation in favor of ranking transitions re-arranged the section according to this change
expanded section on expected termination time

@@
 \end{proposition}
+\textbf{[PETR: POLISH, ADD INTUITION]}
 \begin{proof}
@@
 
-Since the constraint~\eqref{eq:cond-exp} defining conditional expectation is phrased in terms of expected values, the almost-sure uniqueness cannot be strengthened to uniqueness, as re-defining a random variable on a set of probability zero does not change its expectation.
-In the following, when we say that a conditional expectation of a random variable $X$ satisfies some inequality (e.g.
-$\E[X\mid\mathcal{F}]\geq 0$) on set $L\subseteq \Omega$, we mean that for each $\mathcal{F}$-measurable function $\E[X\mid\mathcal{F}]$ satisfying~\eqref{eq:cond-exp} the .
+Since the constraint~\eqref{eq:cond-exp} defining conditional expectation is 
+phrased in terms of expected values, the almost-sure uniqueness cannot be 
+strengthened to uniqueness, as re-defining a random variable on a set of 
+probability zero does not change its expectation. In the following, when we say 
+that a conditional expectation of a random variable $X$ satisfies some 
+inequality (e.g. $\E[X\mid\mathcal{F}]\geq 0$) on set $L\subseteq \Omega$, we 
+mean that for each $\mathcal{F}$-measurable function $\E[X\mid\mathcal{F}]$ 
+satisfying~\eqref{eq:cond-exp} the inequality holds on some subset $L'\subseteq 
+L$ such that $\probm(L')=\probm(L)$.
 
@@
 	\genfilt_i]\leq\vecseq{X}{i}{j} - 
-	\eps$ on $L_j$; and
+	\eps$ on $L_j^i$; and
 	\item for all $1 \leq j' < j$ we have $\E[\vecseq{X}{i+1}{j'}\mid 
-	\genfilt_i]=\vecseq{X}{i}{j'}$ on $L_{j}$.
+	\genfilt_i]=\vecseq{X}{i}{j'}$ on $L_{j}^i$.
 \end{compactitem}

### 2017-07-04 implementation/examples/aaron12.c
Array examples added

@@ -1,0 +1,23 @@
+int aaron12() {
+	int x,y,z;
+
+	while (x>=y) {
+		if (nondet()) {
+			x=x+1;
+			y=y+x;
+
+
+		} else {
+			x=x-z;
+			y=y+(z*z); // Breakage point
+			z=z-1;
+
+		}
+
+
+	}
+	return 0;
+
+
+
+}

### 2017-07-04 paper/extensions.tex
small changes
small esits

@@
 obtaining bounds on expected runtime, and we show how to automatize the usage 
-of this proof rule to obtain polynomial bounds on expected runtime in a 
+of this proof rule to obtain bounds on expected runtime in a 
 subclass of \PP{}s.
@@
 filtration $\{\genfilt_i\}_{i=0}^{\infty}$. We say that 
-$\{\vec{X}_{i}\}_{i=0}^{\infty}$ has \emph{$c$-bounded expected conditional 
+$\{\vec{X}_{i}\}_{i=0}^{\infty}$ has \emph{$\vec{c}$-bounded expected 
+conditional 
 increase (ECI),} for 
-some $c\geq 0$, if for each $i\in \Nset_0 $ and each $1\leq j \leq d $ it holds 
-$\E[\vecseq{X}{i+1}{j}\mid \genfilt_i]\leq \vecseq{X}{i}{j}+c$ on $L^i_{j''}$, 
+some non-negative vector $\vec{c}\in\Rset^d$, if for each $i\in \Nset_0 $ and 
+each $1\leq 
+j \leq d $ it holds 
+$\E[\vecseq{X}{i+1}{j}\mid \genfilt_i]\leq \vecseq{X}{i}{j}+\vec{c}[j]$ on 
+$L^i_{j''}$, 
 for all $j''>j$ 
@@
 
-For LexRSMs with $c$-bounded ECI we have the following result.
+For LexRSMs with $\vec{c}$-bounded ECI we have the following result. For 
+simplicity, 
+we formulate the result for 1-LexRSMs, though it is easy to prove analogous 
+result for general $\eps$-LexRSMs, $\eps>0$, at the cost of obtaining less 
+readable formula.
 
@@
 \label{thm:runtime-bound}
-
+Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be a 
+$d$-dimensional LexRSM with $c$-bounded ECI for some stopping time $\stime$. 
+Then $\E[\stime]\leq  
+\sum_{j=1}^{d}\E[\vecseq{X}{0}{j}]\cdot(\vec{c}[j]+1)^{d-j}$.
 \end{theorem}
+\begin{proof}
+Denote $\noofdecrank_j(\omega)$ the number of steps $i$ in which $\omega\in 
+L_j^i$. Since by Theorem~\ref{thm:lexrsm-main} the existence of LexRSM entails 
+$\probm(\stime<\infty)=1$, the value $\noofdecrank_j(\omega)$ is a.s. finite. 
+We prove, by induction on $1\$
+\end{proof}

### 2017-07-04 paper/lexicographic.tex
small esits

@@
 \begin{theorem}
+\label{thm:lexrsm-main}
 Let $(\Omega,\genfilt,\probm)$ be a probability space, 

### 2017-07-04 paper/extensions.tex
expanded proof of Thm 2

@@
 $\probm(\stime<\infty)=1$, the value $\noofdecrank_j(\omega)$ is a.s. finite. 
-We prove, by induction on $1\$
+We prove that for each $1\leq j \leq d$ it holds $\E[\noofdecrank_j]\leq 
+\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\noofdecrank_{j'}]\right) + 
+\E[\vecseq{X}{0}{j}].$ Since $\stime(\omega)=\sum_{1\leq j \leq d} 
+\noofdecrank_j(\omega)$, for each $\omega\in \Omega$ (and hence, due to 
+linearity of expectation $\E[\stime]=\sum_{1\leq j \leq d} 
+E[\noofdecrank_j]$), the statement of the 
+Theorem follows by an easy induction.
+
+To prove the required inequality, let $\nodecrank{k}{j}(\omega)$ be the number 
+of steps $i$ \emph{within the first $k$} steps such that $\omega\in L_j^i$. We 
+prove, by induction on $k$, that for each $k$ it holds 
+$\E[\nodecrank{k}{j}(\omega)]\leq 
+\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\noofdecrank_{j'}] \right)+ 
+\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k}{j}]$. Once this is proved, the 
+desired inequality follows by taking $k$ to $\infty$, since 
+$\lim_{k\rightarrow \infty}\E[\nodecrank{k}{j}] = \E[\noofdecrank_j]$ and 
+$\lim_{k\rightarrow \infty}\E[\vecseq{X}{k}{j}] \geq 0$.
 \end{proof}

### 2017-07-04 implementation/examples/aaron3.c
More examples added

@@ -1,17 +1,0 @@
-int aaron3() {
-	int x,y,z,tx;
-	while (x>=y) {
-		if (nondet()) {
-			z=z-1;
-			tx=x;
-			x=nondet();
-			if (x>tx+z){ //Breakdown for probabilistic case
-				return 0;
-			}
-		}
-		else {
-			y=y+1;
-		}
-	}
-	return 0;
-}

### 2017-07-04 implementation/main.cpp

@@ -142,6 +142,8 @@ void generate_equations(){
 			// The invariant was "false", so just leave all the transitions of this state
 			continue;
 		}
+		it->second->invariant->print();
+		cout<<endl;
 		if(it->second->type=="det"){
 			//Invariant and guard imply the value decrease
 			if(it->second->edges.empty()){

### 2017-07-04 paper/extensions.tex
expanded the expected termination time section

@@
 prove, by induction on $k$, that for each $k$ it holds 
-$\E[\nodecrank{k}{j}(\omega)]\leq 
-\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\noofdecrank_{j'}] \right)+ 
+$\E[\nodecrank{k}{j}]\leq 
+\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k}{j'}] \right)+ 
 \E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k}{j}]$. Once this is proved, the 
@@
 $\lim_{k\rightarrow \infty}\E[\vecseq{X}{k}{j}] \geq 0$.
+
+The base case $k=0$ is simple as both sides of the inequality are zero. Assume 
+that the inequality holds for some $k\geq 0$. We have 
+$\E[\nodecrank{k+1}{j}]=\E[\nodecrank{k}{j}]+\probm{(L_j^{k})}$, so from 
+induction hypothesis we get 
+\begin{equation}
+\label{eq:time1}
+\E[\nodecrank{k+1}{j}]\leq 
+\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k}{j'}] \right)+ 
+\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k}{j}] + \probm{(L_j^{k})}.\end{equation} 
+Now denote 
+$L^k_{<j} = L^k_1 \cup \dots\cup L^k_{j-1}$ and $L^k_{>j}= 
+L^k_{j+1}\cup\dots\cup L^k_{d}$. We have  
+$\E[\vecseq{X}{k}{j}] = \E[\vecseq{X}{k}{j}\cdot 
+\indicator{ L_{<j}^k}] + \E[\vecseq{X}{k}{j}\cdot 
+\indicator{ L_j^k}] + \E[\vecseq{X}{k}{j}\cdot 
+	\indicator{L_{>j}^k}] \geq 
+\E[\vecseq{X}{k+1}{j}\cdot 
+	\indicator{ L_{<j}^k}] -\vec{c}[j]\cdot \probm(L_{<j}^k) + 
+	\E[\vecseq{X}{k+1}{j}\cdot 
+	\indicator{ L_{j}^k}] + \probm(L_{j}^k) + \E[\vecseq{X}{k+1}{j}\cdot 
+	\indicator{ L_{>j}^k}]= \E[\vecseq{X}{k+1}{j}] -\vec{c}[j]\cdot 
+	\probm(L_{<j}^k)+ \probm(L_{j}^k)$. Plugging this 
+	into~\ref{eq:time1} yields
+\begin{align*}
+\E[\nodecrank{k+1}{j}]&\leq 
+\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k}{j'}] \right)+ 
+\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k+1}{j}] + \vec{c}[j]\cdot 
+\probm(L_{<j}^k)\\
+&=\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k+1}{j'}] \right)+ 
+\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k+1}{j}].
+\end{align*}
+%
+%\end{align}
 \end{proof}
+
+To transfer this mathematical result to probabilistic programs, we want to 
+impose a restriction on LexRSM maps that ensures that all components of a 
+LexRSM map have, from each reachable configuration, an expected one-step 
+increase of at most $c$. Here $c$ can be a constant, but it can also be a value 
+that depends on the initial configurations: this is to handle cases where some 
+variables are periodically reset to a value related to the initial variable 
+values, such as variable $z$ in Figure~\ref{fig:uniint2}. To this end, let 
+$\program$ be a \PP{} with a pCFG $\pCFG_\program$ and let 
+$\vec{\lem}=(\lem_1,\dots,\lem_d)$ be a 
+$d$-dimensional $1$-LexRSM map for $\program$. Consider a $d$-dimensional 
+vector 
+$\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_d)$ whose each component is an 
+expression over variables of 
+$\program$. We say 
+that $\vec{\lem}$ has 
+$\vec{\bar{c}}$-bounded ECI w.r.t. invariant map $\inv$ if the following holds 
+for each initial configuration $(\locinit,\vecinit)$ with $\vecinit\in 
+\vecinitset$: for 
+each 
+configuration $(\loc,\vec{x})$ with $\vec{x}\in \inv(\loc)$ and generalized 
+transition $\tilde{\tau}$ 
+of 
+$\pCFG_\program$ outgoing from $(\loc,\vec{x})$ it holds that if $j$ is 
+the smallest index such that 
+$\tilde{\tau}$ is $1$-ranked by $\lem_j$ from $(\loc,\vec{x})$, then for all 
+$j'>j$ the gen. transition $\tilde{\tau}$ is $f$-ranked by $\lem_{j'}$ from 
+$(\loc,\vec{x})$, where $f=-c_{j'}(\vecinit)$. From 
+Theorem~\ref{thm:runtime-bound} we have the following:
+
+\begin{corollary}
+\label{col:runtime-progs}
+Let $\program$ be a probabilistic program. Assume that there exists a 
+$d$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_d)$ for 
+$\program$ supported 
+by some 
+invariant map $\inv$, scuh that $\vec{\lem}$ has $\vec{\bar{c}}$-bounded ECI 
+(w.r.t. $\inv$) 
+for some vector of expressions $\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_d)$. 
+Then under each scheduler $\sigma$ and for each initial valuation of program 
+variables $\vecinit\in\vecinitset$ it holds $\E^{\sigma}_{\vecinit}[\ttime]\leq 
+\sum_{j=1}^{d}\lem_j(\locinit,\vecinit)\cdot(\bar{c}_j(\vecinit))^{d-j}$.
+\end{corollary}
+
+\textbf{[PETR: FINISH THIS SECTION]}

### 2017-07-05 implementation/files/Parser.cpp
Output Format changed

@@ -1133,7 +1133,7 @@ CFG_edge::CFG_edge(CFG_location* next1,int toChange1,node* change1,node* guard1,
 	Used for debugging only in this program
 
 */
-void CFG_edge::print(){
+void CFG_edge::print(bool print_probability){
 	if(next!=NULL){
 		cout<<"Destination: "<<next->label<<endl;
 	}
@@ -1146,10 +1146,13 @@ void CFG_edge::print(){
 		cout<<endl;
 	}
 	if(guard!=NULL){
-		cout<<"Guard is: "<<endl;
+		cout<<"Guard is: ";
 		guard-> print();
+		cout<<endl;
 	}
-	cout<<"Probability to occur is"<<probability<<endl<<endl; 
+	if(print_probability){
+		cout<<"Probability to occur is "<<probability<<endl; 
+	}
 }
 
 /*
@@ -1168,22 +1171,50 @@ CFG_location::CFG_location(string type,int label){
 /*
 
 	Prints the details about the CFG_location
-	It was used only for debussing in this program
 
 */
 void CFG_location::print(){
 	cout<<"Type: "<<type<<endl;
+	cout<<"Invariant: ";
 	if(invariant!=NULL){
-		cout<<"Invariant: ";
-		cout<<invariant<<" : ";
 		invariant->print();
-		cout<<endl;
 	}
-
+	else{
+		cout<<"false";
+	}
+	cout<<endl;
+	cout<<"LexRSM: ";
+	for(int i=0;i<lexrsm.size();++i){
+		bool printed = false;
+		if(lexrsm[i][0]!=0.0){
+			cout<<lexrsm[i][0];
+			printed = true;
+		}
+		for(int j=1;j<=nVariables;++j){
+			if(lexrsm[i][j]>0){
+				if(printed==true){
+					cout<<"+";
+				}
+				cout<<lexrsm[i][j]<<"*"<<variable[j];
+				printed = true;
+			}
+			else if(lexrsm[i][j]<0){
+				cout<<lexrsm[i][j]<<"*"<<variable[j];
+				printed = true;
+			}
+		}
+		if(printed == false){
+			cout<<"0";
+		}
+		if(i!=lexrsm.size()-1){
+			cout<<" ; ";
+		}
+	}
+	cout<<endl;
 	int i=1;
 	for(vector<CFG_edge>::iterator it = edges.begin();it!=edges.end();++it){
[…]

### 2017-07-05 implementation/files/Parser.h

@@ -66,7 +66,7 @@ public:
 	double probability;
 	CFG_edge();
 	CFG_edge(CFG_location* next,int toChange,node* change,node* guard=NULL,double probability=1.0);
-	void print();
+	void print(bool print_probability = false);
 };
 
 struct CFG_location{
@@ -78,6 +78,7 @@ public:
 	vector<int> ranking_function; // The coefficient matrix for the affine ranking function for this node
 	node* invariant; // A bexpr node, the invariant at this particular node
 	void print();
+	vector<vector<double> > lexrsm;
 };
 
 void skip_spaces(int &begin, int &end);

### 2017-07-05 paper/extensions.tex
finished sec on expected runtime

@@
 
-\textbf{[PETR: FINISH THIS SECTION]}
+Now assume that we have synthesized a $1$-LexRSM map $\vec{\lem}$ and we want 
+to check if there exists $\bar{\vec{c}}$ such that $\vec{\lem}$ has 
+$\bar{\vec{c}}$-bounded ECI. In the linear setting (i.e. the program is an 
+\APP{}, masp $\lem$ and $\inv$ are linear, and we seek $\vec{\bar{c}}$ which is 
+a vector of affine expressions) we can encode the existence of $\vec{\bar{c}}$ 
+into a system of linear inequalities in the similar way as the existence of a 
+linear LexRSM maps was encoded in Section~\ref{sec:lex-programs}. That is, we 
+set up a linear template with unknown coefficients for each component of 
+$\vec{\bar{c}}$ and using Farkas's lemma we set up a system of linear 
+constraints, which includes the unknown coefficients as variables, encoding the 
+fact that $\vec{\lem}$ has $\bar{\vec{c}}$-bounded ECI. Details are provided 
+in~\AppendixMaterial. In this way, we can check in polynomial time if 
+Corollary~\ref{col:runtime-progs} can be applied to $\vec{\lem}$, and if yes, 
+we can synthesize the witness vector $\bar{\vec{c}}$. Since $\bar{\vec{c}}$ 
+consists of affine expressions, Corollary~\ref{col:runtime-progs} provides a 
+polynomial upper bound on expected runtime (polynomial in the ``size'' of 
+initial variable valuation).
+
+%\textbf{[PETR: FINISH THIS SECTION]}

### 2017-07-05 paper/extensions.tex
small edit

@@
 consists of affine expressions, Corollary~\ref{col:runtime-progs} provides a 
-polynomial upper bound on expected runtime (polynomial in the ``size'' of 
-initial variable valuation).
+polynomial (in the size of 
+initial variable valuation) upper bound on expected runtime.
 

### 2017-07-05 paper/conclu.tex
conclusions

--- 2017-07-03 paper/conclu.tex
@@
 Besides RSMs, other approaches has also been considered for probabilistic programs.
+Logical calculi for reasoning about properties of 
+probabilistic programs (including termination) were studied 
+in~\cite{Kozen:prob-semantics,FH:prdl,Kozen:probabilistic-PDL,Feldman:propositional-probdl}
+ and extended to programs with demonic non-determinism 
+ in~\cite{MM04,MM05,KKMO16:wp-expected-runtime,OKKM16:recursive-prob-wp-calculus}.
 A sound approach~\cite{DBLP:conf/sas/Monniaux01} for almost-sure termination 
-is to explore the exponential decrease of probabilities upon bounded-termination 
+is to explore the exponential decrease of probabilities upon 
+bounded-termination 
 through abstract interpretation~\cite{DBLP:conf/popl/CousotC77}.
-MENTION PROOF RULES AND NOT AN AUTOMATED APPROACH.
 

### 2017-07-05 paper/introduction.tex
conclusions

@@
 \noindent{\em Probabilistic programs with nondeterminism.} 
-Randomness play a fundamental role in many areas across science, and in 
+Randomness plays a fundamental role in many areas across science, and in 
 computer science in particular.
@@
 that produce random values according to some desired probability distribution, 
-naturally gives rise probabilistic programs.
+naturally gives rise to probabilistic programs.
 Along with probability, nondeterminism also plays a crucial role.
@@
 and the worst-case analysis is represented by adversarial nondeterminism.
-Hence probabilistic programs with nondeterminism has become an active and 
-important research area in program analysis.
+Hence, probabilistic programs with nondeterminism have become an active and 
+important research focus in program analysis.
 
@@
 The ranking supermartingales provide a powerful and automated approach for termination 
-analysis of probabilistic programs~\cite{SRIRAM,HOLGER,HONGFEI}, and algorithmic 
-approaches for special cases such as linear and polynomial RSMs has
+analysis of probabilistic programs~\cite{SriramCAV,HolgerPOPL,HONGFEI}, and 
+algorithmic 
+approaches for special cases such as linear and polynomial RSMs have
 also been considered~\cite{POPL,CAV,POPL}.
@@
 \smallskip\noindent{\em Practical limitations of existing approaches.} 
-While a impressive set of theoretical results related to RSMs has been
+While an impressive set of theoretical results related to RSMs has been
 established~\cite{}, for probabilistic programs with nondeterminism the current approaches  
@@
 RSMs, to algorithmic approaches for them, to experimental results
-showing their applicability for programs.
+showing their applicability to programs.
 We describe our main contributions below:
@@
 First, we introduce the notion of lexicographic RSMs, and show that
-such supermartingales ensures almost-sure termination (Theorem~\ref{} in Section~\ref{}). 
+such supermartingales ensure almost-sure termination (\ref{thm:lexrsm-main} in 
+Section~\ref{sec:lexicographic}). 
 Our first result is a purely mathematical result that introduces a new concept, and 
 proves almost-sure termination, that is independent of any probabilistic program.
-Based on the mathematical result we show that probabilistic programs with nondeterminism
+Based on the mathematical result we show that for probabilistic programs with 
+nondeterminism
 the existence of a lexicographic RSM with respect to an invariant ensures
-almost-sure termination (Theorem~\ref{} in Section~\ref{}).
+almost-sure termination (Theorem~\ref{thm:lexrsm-programs} in 
+Section~\ref{sec:lex-programs}).
 
@@
 Second we study the compositional properties of lexicographic RSMs.
-A key limitation of the previous approaches related to RSMs is the lack 
-of compositionality, and we show (in Section~\ref{}) how lexicographic RSMs 
-present a compositional approach for almost-sure termination of probabilistic programs.
+A key limitation of the previous approaches related to compositional 
+RSMs~\cite{HolgerPOPL} is that it imposes a technical \emph{uniform 
+integrability} conditions, which is hard to reason about automatically. We show 
+(in Section~\ref{sec:compositional}) how 
+lexicographic RSMs 
+present an easy-to-automatize compositional approach for almost-sure 
+termination of probabilistic programs.
 
@@
 and for efficient algorithms we consider nondeterministic probabilistic 
-programs that are {\em affine} (i.e., the arithmetic operations are linear)
-[VARIABLE RANGE BOUNDED?]
+programs that are {\em affine} (i.e., the arithmetic operations are linear).
 We present a polynomial-time algorithm for synthesis of lexicographic RSMs 
-for affine programs (see Theorem~\ref{}).
+for affine programs (Theorem~\ref{thm:algo}).
 
@@
 termination. 
-In other words, we present an example (Example~\ref{}) where a lexicographic RSM exists ensuring
+In other words, we present an example (Example~\ref{ex:infinite-time}) where a 
+lexicographic RSM exists ensuring
 almost-sure termination, yet the expected termination time is infinite.
@@
 RSMs not only imply positive termination, but even asymptotic bounds on the expected termination 
-time can be derived from them (see Theorem~\ref{}).
+time can be derived from them (Theorem~\ref{thm:runtime-bound} and 
+Corollary~\ref{col:runtime-progs}).
 
@@
 To demonstrate the effectiveness of our approach we consider the benchmarks of non-probabilistic
-programs from~\cite{} which are obtained as abstraction of real-world programs,where 
+programs from~\cite{ADFG10:lexicographic} which are obtained as abstraction of 
+real-world programs, 
+where 
 lexicographic ranking functions were applied for termination analysis. 

### 2017-07-05 paper/conclu.tex
edits of intro and added some references

--- 2017-07-03 paper/conclu.tex
@@
 
+\smallskip\noindent{\em Compositional a.s. termination proving.} A 
+compositional rule for proving almost-sure termination was studied 
+in~\cite{HolgerPOPL} under the uniform integrability assumption. 
+In~\cite{MM05}, a soundness of the probabilistic variant rule is proved for 
+programs with finitely many configurations.
+
 \smallskip\noindent{\em Other approaches.}
 Besides RSMs, other approaches has also been considered for probabilistic programs.
+Logical calculi for reasoning about properties of 
+probabilistic programs (including termination) were studied 
+in~\cite{Kozen:prob-semantics,FH:prdl,Kozen:probabilistic-PDL,Feldman:propositional-probdl}
+ and extended to programs with demonic non-determinism 
+ in~\cite{MM04,MM05,KKMO16:wp-expected-runtime,OKKM16:recursive-prob-wp-calculus,GKI14:prob-semantics,
+  DBLP:conf/sas/KatoenMMM10}. However, none of these approaches is readily 
+  automatizable.
 A sound approach~\cite{DBLP:conf/sas/Monniaux01} for almost-sure termination 
-is to explore the exponential decrease of probabilities upon bounded-termination 
-through abstract interpretation~\cite{DBLP:conf/popl/CousotC77}.
-MENTION PROOF RULES AND NOT AN AUTOMATED APPROACH.
+is to explore the exponential decrease of probabilities upon 
+bounded-termination 
+through abstract interpretation~\cite{DBLP:conf/popl/CousotC77}. A method for 
+a.s. termination of weakly finite programs (where number of reachable 
+configurations is finite from each initial configuration) based on 
+\emph{patterns} was presented in~\cite{EGK12}.
 
@@
 Termination analysis of non-probabilistic programs has also been extensively 
-studied~\cite{DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91,BMS05b,CSZ13,LJB01}.
+studied~\cite{PR04:transition-invariants, 
+CPR06:terminator,DBLP:conf/cav/BradleyMS05,DBLP:conf/tacas/ColonS01, 
+DBLP:conf/vmcai/PodelskiR04,DBLP:conf/pods/SohnG91,BMS05b,LJB01,KSTW10:compositional-transition-invariants,
+ CPR11:termination-cacm}.
 Ranking functions are at the heart of the termination analysis, and lexicographic 
 ranking function has emerged as one of the most efficient and practical approaches
-for termination analysis~\cite{}.
-In this work we extend such lexicographich ranking functions to probabilistic programs,
+for termination analysis~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal}, 
+being used e.g. in 
+the prominent \texttt{T2} temporal prover~\cite{BCIKP16:T2}.
+In this work we extend lexicographic ranking functions to probabilistic 
+programs,
 and present lexicographic RSMs for almost-sure termination analysis of probabilistic programs
-with non-determinism.
+with non-determinism. Theoretical complexity of synthesizing lexicographic 
+ranking functions in non-probabilistic programs was studied 
+in~\cite{BG13:integer-ranking,BG15:lexicographic-complexity}.
 
@@
 In this work we considered lexicographic RSMs for termination analysis of probabilistic
-programs with nondeterminism.
-We showed it presents a sound approach for almost-sure termination, that is compositional,
-algorithmically efficient, and leads to approach that can handle realistic programs.
+programs with non-determinism.
+We showed it presents a sound approach for almost-sure termination, that is
+algorithmically efficient, enables compositional reasoning about termination, 
+and leads to approach that can handle realistic programs.
 There are several interesting directions of future work.
@@
 First, while our work presents the foundations of lexicographic RSMs for probabilistic 
-programs, extending other practical methods of lexicographic ranking functions to 
+programs, extending other practical methods based on lexicographic ranking 
+functions to 
 lexicographic RSMs is an interesting direction of future work.

### 2017-07-05 paper/introduction.tex
edits of intro and added some references

@@
 ranking functions are sound and complete, they do not necessarily provide a practical 
-approach. 
+approach. This is because to prove termination, a witness in the form of a 
+ranking function has to be computed: to this automatically, ranking functions 
+of a restricted shape (such as linear ranking functions) have to be considered, 
+and 1-dimensional ranking functions of a restricted type can only prove 
+termination of a limited class of programs. 
 In contrast, as a practical and scalable approach for non-probabilistic programs the 
-notion of lexicographic ranking functions has been widely studied~\cite{}. 
-Algorithmic approaches for lexicographic ranking functions allow the termination analysis 
+notion of lexicographic ranking functions has been widely 
+studied~\cite{}. 
+Algorithmic approaches for linear lexicographic ranking functions allow the 
+termination analysis 
 to be applicable to real-world non-probabilistic programs (after abstraction).
@@
 almost-sure termination, yet the expected termination time is infinite.
-We then present a natural restriction (i.e., a subclass of programs) such that the lexicographic 
+We then present a natural restriction under which the lexicographic 
 RSMs not only imply positive termination, but even asymptotic bounds on the expected termination 

### 2017-07-05 paper/prelims.tex
added examples
edits of intro and added some references

@@
 \section{Preliminaries}\label{sec:prelim}
+
+We use standard notation in the field of probabilistic program 
+analysis~\cite{POPL17PAPER}.
 
@@
 vectors, e.g. $\vec{x}$, $\vec{y}$, etc., and we denote an $i$-th component of a
-vector $\vec{x}$ by $\vec{x}[i]$. For the purpose of matrix calculations we 
-assume that (non-transposed) vectors are row vectors. If $\vec{v},\vec{v}'$ are 
-$n$ and $m$ dimensional vectors, respectively, then $(\vec{v},\vec{v}')$ is an 
-$(n+m)$-dimensional vector obtained by ``concatenation'' of $\vec{v}$ and 
-$\vec{v}'$. We identify 1-dimensional vectors with numbers. For an 
+vector $\vec{x}$ by $\vec{x}[i]$. 
+%For the purpose of matrix calculations we 
+%assume that (non-transposed) vectors are row vectors. If $\vec{v},\vec{v}'$ 
+%are 
+%$n$ and $m$ dimensional vectors, respectively, then $(\vec{v},\vec{v}')$ is an 
+%$(n+m)$-dimensional vector obtained by ``concatenation'' of $\vec{v}$ and 
+%$\vec{v}'$. 
+We identify 1-dimensional vectors with numbers. For an 
 $n$-dimensional vector 
@@
 Throughout the paper we fix a countable set of variables $\vars$. 
-We consider some arbitrary but fixed linear order on the set of all variables. 
-Hence, given some set of variables $V$ we can enumerate its members in 
-ascending order (w.r.t. the fixed ordering) and write 
-$V=\{x_1,x_2,x_3,\dots\}$. 
+We consider some arbitrary but fixed linear order on the set of all variables, 
+hence we write 
+$\vars=\{x_1,x_2,x_3,\dots\}$. 
 %A \emph{valuation} is a partial function 
@@
 general probabilistic programs with arbitrary (measurable) expressions and 
-their subclass, affine probabilistic programs (\APP s) where are expressions 
+their subclass, affine probabilistic programs (\APP s) where all expressions 
 are restricted to be affine (see below for a precise definition). The reason 
@@
 However, in order to ensure that semantics of probabilistic programs with real-valued variables is defined correctly, we impose the following two conditions on the set of expressions used in each program:
-\begin{enumerate}
-\item
+%\begin{enumerate}
+(1)
 For each expression $E$ over variables $\{x_1,\dots,x_n\}$ and each $n$-dimensional vector $\vec{x}$ the value $E(\vec{x})$ is well defined.\footnote{Our results can be easily extended to programs where encountering an expression of undefined value, such as division by zero, triggers an exception which terminates the program, but we abstract away from such details for the sake of clarity.}
-\item
+(2)
 The function defined by each expression $E$ is Borel-measurable (for definition of Borel-measura\-bility, see, e.g.~\cite{Billingsley:book}).
-\end{enumerate}
+%\end{enumerate}
 
@@
 \smallskip\noindent{\em Linear constraints, assertions, predicates.}
-In the case of predicates involving only linear expression we use the following nomenclature:
+In the case of predicates involving only linear expression we use the following 
+standard nomenclature:
 \begin{compactitem}

### 2017-07-05 paper/introduction.tex
Citations added, small changes

@@
 In applications such as stochastic network protocols~\cite{BaierBook,prism},
-randomized algorithms~\cite{RandBook,RandBook2}, machine learning~\cite{},
-NATURE PAPER, the probabilistic behavior must be considered to faithfully model
-the underlying dynamic system.
+randomized algorithms~\cite{RandBook,RandBook2}, machine learning~\cite{LearningSurvey,G15},
+the probabilistic behavior must be considered to faithfully model the underlying dynamic system.
 The extension of classical imperative programs with \emph{random value generators}, 
@@
 \end{compactenum}
-While the positive termination implies almost-sure termination, the converse is not true (see~\cite[]{}).
+While the positive termination implies almost-sure termination, the converse is not true 
+(e.g., see Example~\ref{ex:infinite-time}).
 
@@
 The generalization of ranking functions to probabilistic programs is achieved through the
-{\em ranking supermartingales (RSMs)}~\cite{}.
+{\em ranking supermartingales (RSMs)}~\cite{SriramCAV,HolgerPOPL,CF17}.
 The ranking supermartingales provide a powerful and automated approach for termination 
-analysis of probabilistic programs~\cite{SriramCAV,HolgerPOPL,HONGFEI}, and 
-algorithmic 
-approaches for special cases such as linear and polynomial RSMs have
-also been considered~\cite{POPL,CAV,POPL}.
+analysis of probabilistic programs, and algorithmic approaches for special cases such as 
+linear and polynomial RSMs have also been considered~\cite{SriramCAV,CFNH16,CFG16,CNZ17}.
 
@@
 While an impressive set of theoretical results related to RSMs has been
-established~\cite{}, for probabilistic programs with nondeterminism the current approaches  
+established~\cite{SriramCAV,HolgerPOPL,CF17,CFNH16,CFG16,CNZ17}, 
+for probabilistic programs with nondeterminism the current approaches  
 are only applicable to academic examples of variants of random walks. 
@@
 notion of lexicographic ranking functions has been widely 
-studied~\cite{}. 
+studied~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal,BCIKP16:T2}. 
 Algorithmic approaches for linear lexicographic ranking functions allow the 
@@
 First, we introduce the notion of lexicographic RSMs, and show that
-such supermartingales ensure almost-sure termination (\ref{thm:lexrsm-main} in 
+such supermartingales ensure almost-sure termination (Theorem~\ref{thm:lexrsm-main} in 
 Section~\ref{sec:lexicographic}). 

### 2017-07-05 paper/app.tex
app

\section{Details of Program Syntax}
In this subsection we present the details of the syntax of (affine) 
probabilistic 
programs.

Recall that $\vars$ %and $\mathcal{R}$ 
is a collection of 
\emph{variables}.
%and \emph{random} variables, respectively. 
Moreover, let $\mathcal{D}$ be a set of \emph{probability distributions} on 
real numbers.
The abstract syntax of affine
probabilistic programs (\APP s)
is given by the grammar in Figure~\ref{fig:syntax}, where
the expressions $\langle \mathit{pvar}\rangle$ and $\langle
\mathit{dist}\rangle$  range over $\vars$ and $\mathcal{D}$, respectively.
We allow for non-deterministic assignments, expressed by a statement $x:=
\text{\textbf{ndet($\mathit{dom}$)}}$, where $\mathit{dom}$ is a \emph{domain
specifier} determining the set from which the value can be chosen: for general 
programs it can be any Borel-measureable set, for \APP{}s it has to be an 
interval (possibly of infinite length).  
The grammar is such that $\langle \mathit{expr} \rangle$ %and $\langle
%\mathit{rexpr} \rangle$ 
may evaluate to an arbitrary affine expression over the
program variables.
%, and the program and random variables, respectively (note that
%random variables can only be used in the RHS of an assignment). 
Next, $\langle
\mathit{bexpr}\rangle$ may evaluate to an arbitrary propositionally linear
predicate.

For general (not necessary affine) \PP{}s we set $\langle\mathit{expr}\rangle$ 
to be the set of all expressions permitted by the set of mathematical 
operations of the underlying language. Similarly, $\langle\mathit{bexpr} 
\rangle$ is the set of all predicates, as defined in Section~\ref{sec:prelim}.

The guard of each if-then-else statement is either $\star$, 
representing a (demonic) non-deterministic choice between the branches,
a keyword \textbf{prob}($p$), where $p\in [0,1]$ is a number given in decimal
representation (represents a probabilistic choice, where  the if-branch is
executed with probability $p$ and the then-branch with probability $1-p$), or
the guard is a propositionally linear predicate, in which case the statement
represents a standard deterministic conditional branching.

%We assume that each \APP{} $\program$ is preceded by an initialization 
%preamble 
%in 
%which 
%each variable appearing in $\program$ is assigned some concrete number. 
Regarding distributions, for each $d\in \mathcal{D}$ we assume the
existence of a program primitive denoted by '\textbf{sample($d$)}' implementing 
sampling from $d$. In practice, the distributions appearing in a program would 
be those for which sampling is 
provided by suitable libraries (such as 
uniform distribution over some interval, Bernoulli, geometric, etc.), 
but we 
abstract away from these implementation details. For the purpose of our 
analysis, it is sufficient that for each distribution $d$ appearing in the 
program the following characteristics: expected value $\expv[d]$ of $d$ and a 
set $SP_d$ containing \emph{support} of $d$  (the support of $d$ is the 
smallest 
closed set of real numbers whose complement has probability zero 
under $d$). \footnote{In 
particular, 
a support of a \emph{discrete} probability 
distribution $d$ is simply the at most countable set of all points on a real 
line that have positive probability under $d$. For continuous distributions, 
e.g. a normal distribution, uniform, etc., the support is typically either 
$\Rset$ or some closed real interval. } For \APP{}s, $SP_d$ is required to be 
an 
interval. 
\begin{figure}
\begin{align*}
\langle \mathit{stmt}\rangle &::= 
%\,\langle\mathit{pvar}\rangle
%\,\text{'$:=$'}\, \langle\mathit{rexpr} \rangle \mid 
%\langle\mathit{pvar}\rangle \,\text{'$:=$}\,
%\text{\textbf{ndet($\langle\mathit{dom}\rangle$)}'}  \\
\langle \mathit{assgn} \rangle \mid \text{'\textbf{skip}'} \mid 
\langle\mathit{stmt}\rangle \, \text{';'} \, \langle \mathit{stmt}\rangle \\
%
&\mid   \text{'\textbf{if}'} \,
\langle\mathit{ndbexpr}\rangle\,\text{'\textbf{then}'} \, \langle
\mathit{stmt}\rangle \, \text{'\textbf{else}'} \, \langle \mathit{stmt}\rangle
\,\text{'\textbf{fi}'}
\\
%
&\mid  \text{'\textbf{while}'}\, \langle\mathit{bexpr}\rangle \,
\text{'\textbf{do}'} \, \langle \mathit{stmt}\rangle \, \text{'\textbf{od}'}
\\
\langle \mathit{assgn} \rangle &::= 
\,\langle\mathit{pvar}\rangle
\,\text{'$:=$'}\, \langle\mathit{expr} \rangle \mid 
\langle\mathit{pvar}\rangle \,\text{'$:=$}\,
\text{\textbf{ndet($\langle\mathit{dom}\rangle$)}'}
\\
&\mid \langle\mathit{pvar}\rangle \,\text{'$:=$}\,
\text{\textbf{sample($\langle\mathit{dist}\rangle$)}'}
%
%
%
%
%&
%
%\\
%
%\\
\\
\vspace{0.5\baselineskip}
%\\
%\langle\mathit{var} \rangle &::= \langle\mathit{pvar} \rangle \mid 
%\langle\mathit{rvar}\rangle
%\\
%%\vspace{\baselineskip}
%\vspace{0.5\baselineskip}
%\\
\langle\mathit{expr} \rangle &::= \langle \mathit{constant} \rangle \mid
\langle\mathit{pvar}\rangle
%\\
\mid \langle \mathit{constant} \rangle \,\text{'$\cdot$'} \,
\langle\mathit{pvar}\rangle
\\
%
&\mid \langle\mathit{expr} \rangle\, \text{'$+$'} \,\langle\mathit{expr} \rangle
\mid \langle\mathit{expr} \rangle\, \text{'$-$'} \,\langle\mathit{expr} \rangle
%\\
%%
%\vspace{0.5\baselineskip}
%%\\
%\langle\mathit{rexpr} \rangle &::= \langle\mathit{expr} \rangle \mid 
%\langle\mathit{rvar}\rangle 
%%\mid
%% \langle\mathit{pvar}\rangle 
%\mid\langle\mathit{constant} \rangle \,\text{'$*$'} \,
%\langle\mathit{rvar}\rangle \\
%%
%&\mid
%%\langle\mathit{constant} \rangle \,\text{'$*$'} \, \langle\mathit{pvar}\rangle
%%\mid 
%\langle\mathit{rexpr} \rangle\, \text{'$+$'} \,\langle\mathit{rexpr} \rangle
%\mid \langle\mathit{rexpr} \rangle\, \text{'$-$'} \,\langle\mathit{rexpr}
%\rangle
%%\\
%%\vspace{0.5\baselineskip}
%%\\ \langle\mathit{bexpr} \rangle &::= \langle \mathit{predicate}\rangle \mid 
%%\neg \langle\mathit{predicate}\rangle
\\
%
%\vspace{0.5\baselineskip}
%\langle \mathit{dom} \rangle &::= \text{'\textbf{Int}'} \mid
%\text{'\textbf{Real}'} \mid
%\text{'\textbf{Int}$[\langle\mathit{constant}\rangle,\langle\mathit{constant}\rangle]$'}
% \\ 
%% 
% &\mid
%\text{'\textbf{Real}$[\langle\mathit{constant}\rangle,\langle\mathit{constant}\rangle]$'}
%\mid \langle \mathit{dom} \rangle \text{'\textbf{or}'}\langle \mathit{dom}
%\rangle
%\\
%
%\vspace{0.5\baselineskip}
%\\
\langle \mathit{bexpr}\rangle &::=  \langle \mathit{affexpr} \rangle \mid
\langle \mathit{affexpr} \rangle \, \text{'\textbf{or}'} \,
\langle\mathit{bexpr}\rangle
\vspace{0.5\baselineskip}
\\
%
%\vspace{\baselineskip}
%\\
\langle\mathit{affexpr} \rangle &::=  \langle\mathit{literal} \rangle\mid
\langle\mathit{literal} \rangle\, \text{'\textbf{and}'}
\,\langle\mathit{affexpr} \rangle
\\
%
\langle\mathit{literal} \rangle &::= \langle\mathit{expr} \rangle\,
\text{'$\leq$'} \,\langle\mathit{expr} \rangle \mid \langle\mathit{expr}
\rangle\, \text{'$\geq$'} \,\langle\mathit{expr} \rangle
\\
%
&\mid \neg \langle \mathit{literal} \rangle
\\
%
%\vspace{0.5\baselineskip}
%\\
\langle\mathit{ndbexpr} \rangle &::= {\star}\mid
\text{'\textbf{prob($p$)}'} \mid \langle\mathit{bexpr} \rangle
\end{align*}
\caption{Syntax of affine probabilistic programs (\APP 's).}
\label{fig:syntax}
\end{figure}

%The syntax is such that $\langle \mathit{expr} \rangle$ and $\langle
%\mathit{rexpr} \rangle$ may evaluate to an arbitrary affine expression over the
%program variables, and the program and random variables, respectively (note
%that random variables can only be used in the RHS of an assignment). Next,
%$\langle \mathit{bexpr}\rangle$ may evaluate to an arbitrary disjunction of
%polyhedral constraints (i.e. conjunctions of linear inequalities and equalities
%over the program variables).

%Additionally to the program body generated by the grammar, we assume that each
%program has a preamble in which both program and random variables are declared.
%We assume that every program variable is initialized to some fixed constant
%upon declaration, and that for each of the random variables its
%For random variables we need to suitably specify their distribution in the
%preamble. This amounts to either directly specifying the joint distribution of
%these variables, or stipulating the variables to be stochastically independent
%and specifying a probability distribution for each one of them. We point out
%that the only parameters of the program's random variables that we use in the
%process of deriving ranking supermartingales are their expected values (which
%we assume to be well-defined and finite) and, for deriving concentration
%inequalities, also bounds on their range. \PN{Is this true? Do we not need
%bounds on their range for as termination as well?} The exact distributions are
%only needed for the actual execution of the program.

%\begin{example}\label{ex:prog}
%We present an example of an affine probabilistic program shown in
%Figure~\ref{ex:prob}.
%The program variable is $x$, and there is a while loop, where given a
%probabilistic
%choice one of two statement blocks $Q_1$ or $Q_2$ is executed.
%The block $Q_1$ (resp. $Q_2$) is executed if the probabilistic choice is at
%least
%$0.6$ (resp. less than $0.4$).
%The statement block $Q_1$ (resp., $Q_2$) is an angelic (resp. demonic)
%conditional
%statement to either increment or decrement $x$.
%\lstset{language=affprob}
%\lstset{tabsize=3}
%\newsavebox{\affproblist}
%\begin{lrbox}{\affproblist}
%\begin{lstlisting}[mathescape]
%$x:=0$;
%while $x \geq 0$ do
%	if prob(0.6) then
%		if angel then
%			$x:=x+1$
%		else
%			$x:=x-1$
%		fi
%	else
%		if demon then
%			$x:=x+1$
%		else
%			$x:=x-1$
%		fi
%	fi
%od
%\end{lstlisting}
%\end{lrbox}
%\begin{figure}[t]
%%%\centering
%\usebox{\affproblist}
%\caption{An example of a probabilistic program}\label{ex:prob}
%\end{figure}
%\end{example}

\subsection{Details of Section~\ref{subsec:semantics}}

\begin{remark}[Use of random variables]
%\label{rem:randuse}
In the paper we sometimes work with random variables 
that are functions of the type $R\colon\Omega \rightarrow S$ for some finite 
set $S$. These can be captured by the definition given in 
Section~\ref{sec:prelim} by identifying the 
elements of $S$ with distinct real numbers.\footnote{This is equivalent to 
saying that a function $R\colon \Omega\rightarrow S$, with $S$ finite, is a 
random variable if for each $s\in S$ the set $\{\omega\in \Omega\mid 
R(\omega)=s\}$ belongs to $\mathcal{F}$.} The exact choice of numbers is 
irrelevant in such a case, as we are not interested in, e.g. computing expected 
values of such random variables, or similar operations. 
\end{remark}

\paragraph*{From Programs to pCFGs}
To every probabilistic program $P$ we can assign a pCFG $\pCFG_P$ whose 
locations correspond to the values of the
program counter of $P$ and whose transition relation captures the behaviour of
$P$. We illustrate the construction for \APP{}s, for general programs it is 
similar. To obtain $\pCFG_{P}$, we first rename 
the variables in $P$ to 
$x_1,\dots,x_n$, where $n$ is the number of distinct variables in the program. 
The
construction of $\pCFG_P$ can be described inductively.
For each program $P$ the pCFG $\pCFG_P$ contains two distinguished
locations, $\ell^{\lin}_{P}$ and $\ell^{\lout}_{P}$, the latter one being always
deterministic, that intuitively represent the state of the program counter
before and after executing $P$, respectively. In the following, we denote by 
$\id_1$ a function such that for each $\vec{x}$ we have 
$\id_{1}(\vec{x})=\vec{x}[1]$.
\begin{compactenum}
\item {\em Deterministic Assignments and Skips.}
For $P= {x_j}{:=}{E}$ where $x_j$ is a program variable and $E$ is an 
expression, or $P = \textbf{skip}$, the pCFG $\pCFG_P$ consists only of
%these two (deterministic) locations
locations $\ell^{\lin}_P$ and $\ell^{\lout}_P$ (first assignment location, 
second one deterministic) and a
transition $(\ell^{\lin}_{P},\ell^{\lout}_P)$. In the first case, 
$U(\ell^{\lin}_{P},\ell^{\lout}_P)=(j,E)$.
\item {\em Probabilistic and Non-Deterministic Assignemnts}
For $P= {x_j}{:=}{\textbf{sample($d$)}}$ where $x_j$ is a program variable and 
$d$ is a distribution, the pCFG $\pCFG_P$ consists locations $\ell^{\lin}_P$ 
and $\ell^{\lout}_P$ and a
transition $\tau=(\ell^{\lin}_{P},\ell^{\lout}_P)$ with $U(\tau)=(j,d)$. For 
$P= 
{x_j}{:=}{\textbf{ndet($\mathit{dom}$)}}$, the construction is similar, with 
the only transition being $\tau=(\ell^{\lin}_{P}\ell^{\lout}_P)$ and 
$U(\tau)=(j,D)$, where 
$D$ is 
the set specified by the domain specifier $\mathit{dom}$.

\item {\em Sequential Statements.}
For $P = Q_1;Q_2$ we take the pCFGs $\pCFG_{Q_1}$, $\pCFG_{Q_2}$ and
join them by identifying the location $\ell^{\lout}_{Q_1}$ with
$\ell^{\lin}_{Q_2}$, putting $\ell^{\lin}_{P}=\ell^{\lin}_{Q_1}$ and
$\ell^{\lout}_{P}=\ell^{\lout}_{Q_2}$.

\item {\em While Statements.}
For $P = \textbf{while $\phi$ do }Q \textbf{ od}$ we add a new deterministic
location $\ell^{\lin}_{P}$ which we identify with $\ell^{\lout}_{Q}$, a new
deterministic location $\ell^{\lout}_{P}$, and transitions
$\tau=(\ell^{\lin}_{P},\ell^{\lin}_{Q})$,
$\tau'=(\ell^{\lin}_{P},\ell^{\lout}_{P})$ such that $G(\tau)=\phi$ and
$G(\tau')=\neg\phi$.

\item {\em If Statements.}
Finally, for $P = \textbf{if $\mathit{ndb}$ then }Q_1 \textbf{ else } Q_2
\textbf{ fi}$ we add a new location $\ell^{\lin}_{P}$ (which is not an 
assignment location) together with two
transitions $\tau_1 = (\ell^{\lin}_{P},\ell^{\lin}_{Q_1})$, $\tau_2 =
(\ell^{\lin}_{P},\ell^{\lin}_{Q_2})$, and we identify the locations 
$\ell^{\lout}_{Q_1}$ and $\ell^{\lout}_{Q_1}$ with $\ell^{\lout}_{P}$. (If both
$Q_j$'s consist of a single statement, we also identify $\ell^\lin_{P}$ with 
$\ell^{\lin}_{Q_j}$'s.) In this
case the newly added location $\ell^\lin_{P}$ is non-deterministic branching if 
and only 
if
$ndb$ is the keyword '$\star$'. If
$\mathit{ndb}$ is of the form $\textbf{prob($p$)}$, the location $\ell^\lin_{P}$
is probabilistic branching with $\probdist_{\ell^\lin_{P}}(\tau_1)=p$ and
$\probdist_{\ell^\lin_{P}}(\tau_2)=1-p$. Otherwise (i.e. if $\mathit{ndb}$ is a
predicate), $\ell^\lin_{P}$ is a deterministic location
with $G(\tau_1)=\mathit{ndb}$ and $G(\tau_2)=\neg \mathit{ndb}$.
\end{compactenum}
Once the pCFG $\pCFG_P$ is constructed using the above rules, we put
$G(\tau)=\textit{true}$ for all transitions $\tau$ outgoing from deterministic
locations whose guard was not set in the process, and finally we add a self-loop
on the location $\ell^{\lout}_P$. This ensures that the assumptions in
Definition~\ref{def:stochgame} are satisfied.
Furthermore note that for pCFG obtained for a program $P$, since the only
branching is conditional branching, every location $\loc$ has at most two
successors $\loc_1,\loc_2$.

### 2017-07-05 paper/prelims.tex
app

@@
 consider in our analysis. We consider two classes of probabilistic programs: 
-general probabilistic programs with arbitrary (measurable) expressions and 
+general probabilistic programs (\PP{}s ) with arbitrary (measurable) 
+expressions and 
 their subclass, affine probabilistic programs (\APP s) where all expressions 
@@
 \smallskip\noindent{\em Predicates.}
-A \emph{predicate} is a logical formula obtained by a finite number of applications of conjunction, disjunction and negation operations on \emph{atomic predicates} of the form $E\leq E'$, where $E$, $E'$ are expressions.
+A \emph{predicate} is a logical formula obtained by a finite number of 
+applications of conjunction, disjunction and negation operations on 
+\emph{atomic predicates} of the form $E\leq E'$, where $E$, $E'$ are 
+expressions. We denote by $\vec{x}\models E$ the fact that $E$ is satisfied by 
+substituting values from of $\vec{x}$ for the corresponding variables in $E$.
 
@@
 
-\textbf{[PETR: SAY SOMETHING ABOUT PREDICATE SATISFACTION.]}
-\textbf{[PETR: EXPLAIN BOREL-MEASURABILITY IN MORE DETAIL?]}
+%\textbf{[PETR: SAY SOMETHING ABOUT PREDICATE SATISFACTION.]}
+%\textbf{[PETR: EXPLAIN BOREL-MEASURABILITY IN MORE DETAIL?]}
 
@@
 A probabilistic program is \emph{affine} if all the expressions that occur in 
-the program (i.e. in loop guards, conditionals and on right-hand sides of 
-assignments) are affine. We refer to the class of affine probabilistic programs 
+the program (i.e. in loop guards, conditionals, right-hand sides of 
+assignments) are affine and if the set of possible initial valuations is a 
+polyhedron. We refer to the class of affine probabilistic programs 
 as \APP s.
@@
 %%We now recall basic definitions of probability theory.
-The crucial notion is the one of a probability space. A probability space is a triple
+ A probability space is a triple
 $(\Omega,\mathcal{F},\probm)$, where $\Omega$ is a non-empty set (so called
@@
 %non-determinism
-The semantics can be defined as the semantics of an uncountable state-space
+We consider standard operational semantics of \PP{}s defined via an 
+uncountable state-space
 Markov decision process (MDP) (uncountable due to real-valued variables).
-We take an operational approach to define the semantics, and associate to each program 
-a certain stochastic process~\cite{SriramCAV,HolgerPOPL,Kozen:prob-semantics}.
+That is, we associate to each program 
+a certain stochastic process.
 To define this process, we first define so called 
-\emph{probabilistic control flow graphs}~\cite{CFG16:positivstellensatz-arxiv}.
+\emph{probabilistic control flow graphs}~\cite{HONGFEI-CAV16}.
 
@@
 %\item 
-(c)~a set $R\subseteq \Rset$.
+(c)~a set $R\subseteq \Rset$ (representing a non-deterministic update).
 %\end{itemize}
@@
 \vee \dots \vee G(\tau_k) \equiv \mathit{true}$ and $G(\tau_i) \wedge G(\tau_j)
-\equiv \mathit{false}$ for each $1\leq i < j \leq k$. Moreover, for each 
+\equiv \mathit{false}$ for each $1\leq i < j \leq k$. For each 
 distribution $d$ appearing in the 
 pCFG we assume the following features are known: expected value $\expv[d]$ of 
-$d$ and a 
-single-variable PLP $\varphi_d$ such that the \emph{support} of $d$ (i.e. the 
+$d$ and a set $SP_d$ containing the \emph{support} of $d$. Tthe support is the  
 smallest 
 closed set of real numbers whose complement has probability zero 
-under $d$)\footnote{In particular, a support of a \emph{discrete} probability 
+under $d$\footnote{In particular, a support of a \emph{discrete} probability 
 	distribution $d$ is simply the at most countable set of all points on a 
 	real 
-	line that have positive probability under $d$.} satisfies 
-$\support(d)\subseteq\exsat{\varphi_d}$. Finally, we assume that each 
-assignment location has at most (and thus exactly) one outgoing transition.
+	line that have positive probability under $d$.}) Finally, we assume that 
+	each 
+assignment location has at most (and thus exactly) one outgoing transition. The 
+translation from probabilistic programs to the corresponding pCFG is 
+standard~\cite{CFNH16:prob-termination}, and the details are presented 
+in~\AppendixMaterial.
 %Finally, we assume that for each 
@@
 \end{definition}
-
-%We can naturally transfer the linear ordering of variables in $\vars$ to 
-%variables in $X$. Hence, the set of variables $X$ can be uniquely written as a 
-%vector $\vec{x}=(x_1,\dots,x_{|X|})$.
-%Since $\pvars$ is a linearly ordered set of variables, there is a natural 
-%correspondence between valuations in $\vals(\pvars)$ and 
-%$|\pvars|$-dimensional 
-%vectors. To formalize this correspondence, to any valuation 
-%$\val\in\vals(X)$ we assign a vector 
-%$\vec{x}_{\val}=(\val(x_1),\dots,\val(x_{|X|}))$, while to an 
-%$|X|$-dimensional vector $\vec{x}$ we assign a valuation $\val_{\vec{x}}\in 
-%\vals(\pvars)$ such that $\val_{\vec{x}}(x_i) = \vec{x}[i]$ for all $1\leq i 
-%\leq |X|$.
-
-%Given a pCFG $\pCFG$ 
-%
-%For notational convenience we assume that the sets $\pvars$ and $\rvars$ are
-%endowed with some fixed linear ordering, which allows us to write $\pvars =
-%\{x_1,x_2,\dots,x_{|\pvars|}\}$ and $\rvars = \{r_1,r_2,\dots,r_{|\rvars|}\}$.
-%Every update function $f$ in a stochastic game can then be viewed as a tuple
-%$(f_1,\dots,f_{|\pvars|})$, where each $f_i$ is of type
-%$\Rset^{|\pvars\cup\rvars|}\rightarrow\Rset$. We denote by
-%$\mathbf{x}=(\vec{x}[i])_{i=1}^{|\pvars|}$ and
-%$\mathbf{r}=(\vec{r}[i])_{i=1}^{|\rvars|}$ the vectors of concrete valuations 
-%of
-%program and random variables, respectively.
-%In particular, we assume that each component of $\vec{r}$ lies within the range
-%of the corresponding random variable. %A valuation $\vec{r}$ is admissible if
-%%each its component lies within the range of the corresponding random variable.
-%We use the following succinct notation for special update functions: by $\id$ 
-%we
-%denote a function which does not change the program variables at all, i.e. for
-%every $1\leq i \leq |\pvars|$ we have $f_i(\vec{x},\vec{r})= \vec{x}[i]$. For a
-%function $g$ over the program and random variables we denote by 
-%$\assgn{x_j}{g}$
-%the update function $f$ such that $f_j(\vec{x},\vec{r})=g(\vec{x},\vec{r})$ and
-%$f_i(\vec{x},\vec{r})=\vec{x}[i]$ for all $i\neq j$.
-
-%We say that an SGS $\mathcal{G}$ is \emph{normalized} if all guards of all
-%transitions in $\mathcal{G}$ are in a disjunctive normal form.
-
-%\begin{example}\label{ex:sgs}
-%Figure~\ref{fig:Q4:SGS} shows an example of stochastic game structure.
-%Deterministic locations are represented by boxes, angelic locations by
-%triangles, demonic locations by diamonds,
-%and stochastic locations by circles. Transitions are labelled with update
-%functions, while guards and probabilities
-%of transitions outgoing from deterministic and stochastic locations,
-%respectively, are given in rounded rectangles
-%on these transitions. For the sake of succinctness we do not picture
-%tautological guards and identity update functions. Note that the SGS is
-%normalized.
-%We will describe in Example~\ref{ex:illustrate} how the stochastic game
-%structure shown corresponds to the program described in
-%Example~\ref{ex:prog}.
-%\end{example}
-
-%\paragraph*{Dynamics of Stochastic Games}
-
-The translation from probabilistic programs to the corresponding pCFG is 
-standard~\cite{CFNH16:prob-termination}, and the details are presented 
-in~\AppendixMaterial.
 
@@
 \smallskip\noindent{\em Non-determinism and Schedulers.}
-Due to the presence of non-determinism and probabilistic choices, a pCFG
-$\pCFG$ may represent a multitude of possible behaviours. The probabilistic
+%Due to the presence of non-determinism and probabilistic choices, a pCFG
+%$\pCFG$ may represent a multitude of possible behaviours. 
+The probabilistic
 behaviour of $\pCFG$ can be captured by constructing a suitable
@@
 the 
-standard notion of a \emph{scheduler}.
-
-There are two sources of non-determinism in our programs: one in branching and 
+standard notion of a \emph{scheduler}. Note that there are two sources of 
+non-determinism in our programs: one in branching and 
 one in assignments. We call a location $\loc$ non-deterministic if $\loc$ is a 
 non-deterministic branching location or if $\loc$ is an assignment location 
-with a 
-non-deterministic assignment (i.e., the only transition $\tau$ outgoing from 
-$\loc$ has $\updates(\tau)=(j,\up)$ where $\up$ is a set). A configuration 
+with the only transition $\tau$ outgoing from 
+$\loc$ having a non-deterministic assignment. A configuration 
 $(\loc,\vec{x})$ is non-deterministic if $\loc$ is non-deterministic.
@@
 $|\pvars|$ consisting of real-valued random variables). The sigma-algebra 
-$\natfilt$ is the smallest (w.r.t. inclusion) sigma algebra under which all 
+$\natfilt$ is the smallest (w.r.t. inclusion) sigma-algebra under which all 
 the functions $\cfg{\sigma}{i}$, for all $i\geq 0$, 
@@
 \vspace{-0.5em}
-We consider computational problems related to the basic liveness properties of 
-\APP{s}, namely \emph{termination} and its generalization, \emph{reachability}.
+Termination is the basic liveness property of \PP{}s.
 
 \smallskip\noindent{\em Termination and termination time.}
-In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. 
+In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. This 
+pCFG has a special location  $\loc_{P}^{\lout}$ corresponding to the value of 
+the program counter after 
+executing $P$.
 We say that a run 
 \emph{terminates} if it reaches a 
-configuration whose first component (i.e. a location of $\pCFG_P$) is the 
-location
-$\loc_{P}^{\lout}$ corresponding to the value of the program counter after 
-executing $P$. 
+configuration whose first component is
+$\loc_{P}^{\lout}$.
+%
 We define a random variable $\ttime$

### 2017-07-05 implementation/main.cpp
Time changed to float from int

@@ -20,6 +20,9 @@ Comments:
 #include <fstream>
 #include <queue>
 #include <ctime>
+#include <sys/time.h>
+
+
 #include "files/Parser.h"
 using namespace std;
 #define MAXL 700000 //Maximum length of the program
@@ -498,7 +501,8 @@ void print_fast(ostream& fastfile){
 }
 
 int main(){
-	double start_time = time(0);
+	struct timeval  tv1, tv2;
+	gettimeofday(&tv1, NULL);
 	int start,end;
 	char* input = new char[MAXL];
 	int r,i;
@@ -525,8 +529,8 @@ int main(){
 	label_map[end]	 = new CFG_location("det",end);
 	// Parsing program from begin to endprog(pointers) and the start and end nodesid's have also been passed
 	root=new node("stmt",begin,endprog,start,end);
-	double parsing_over = time(0);
-	cout<<"Time:"<<(double(parsing_over-start_time))<<endl;
+	gettimeofday(&tv2, NULL);
+	cout<<"Time:"<<(double) (tv2.tv_usec - tv1.tv_usec) / 1000000 +(double) (tv2.tv_sec - tv1.tv_sec)<<endl;
 	cout<<"Parsing over and CFG constructed"<<endl;
 	cout<<"The CFG constructed has "<<last_used_label<<" states"<<endl;
 	ofstream fastfile;
@@ -535,15 +539,17 @@ int main(){
 	// Printing the CFG in fast format into files/aspic.fast file
 	print_fast(*outfast);
 	fastfile.close();
-	double printed_fast = time(0);
-	cout<<endl<<"Time:"<<(double(printed_fast-start_time))<<endl;
+	gettimeofday(&tv2, NULL);
+	cout<<"Time:"<<(double) (tv2.tv_usec - tv1.tv_usec) / 1000000 +(double) (tv2.tv_sec - tv1.tv_sec)<<endl;
 	cout<<"Fast file printed."<<endl;
+	double aspic_start_time = (double) (tv2.tv_usec - tv1.tv_usec) / 1000000 +(double) (tv2.tv_sec - tv1.tv_sec);
 	// Calling the invariant_script.sh which inturn calls aspic to generate invariants into files/InvariantOutput	
 	if(system("./files/invariant_script.sh")!=0){
 		cerr<<"Something wrong with the script to find invariants";
 	}
-	double aspic_done = time(0);
-	cout<<endl<<"Time:"<<(double(aspic_done-start_time))<<endl;
[…]

### 2017-07-05 paper/app.tex

@@
 
-\subsection{Details of Section~\ref{subsec:semantics}}
+\section{Details of Program Semantics}
 
@@
 Furthermore note that for pCFG obtained for a program $P$, since the only
-branching is conditional branching, every location $\loc$ has at most two
+branching is if-then-else branching, every location $\loc$ has at most two
 successors $\loc_1,\loc_2$.

### 2017-07-05 paper/conclu.tex

@@
 but with real-valued variables, and its extension for probabilistic programs
-with non-determinism has been studied in~\cite{HolgerPOPL,CFNH16:prob-termination,CAV-POLY,POPL,ARXIV}.
+with non-determinism has been studied in~\cite{HolgerPOPL,CF17,CFNH16,CFG16,CNZ17}.
 While all these results deeply clarify the role of RSMs for probabilistic programs, 

### 2017-07-05 paper/synthesis.tex

@@
 \begin{theorem}
-	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ together with linear invariant map $\inv$.
-If teh algorithm returns a $d$-dimensional LEM $\vec{\lem}=(\lem_1,\dots,\lem_d)$, then $\vec{\lem}$ is a $1$-LinLexRSM map for $\program$ supported by $\inv$.
-Conversely, if the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and $\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  supported by $\inv$.
+\label{thm:algo}
+	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ 
+	together with linear invariant map $\inv$. If the algorithm returns a 
+	$d$-dimensional LEM $\vec{\lem}=(\lem_1,\dots,\lem_d)$, then $\vec{\lem}$ 
+	is a $1$-LinLexRSM map for $\program$ supported by $\inv$. Conversely, if 
+	the algorithm returns ``No LinLexRSM'', then for any $d'\in \Nset$ and 
+	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
+	supported by $\inv$. The algorithm runs in time polynomial in size of 
+	$\program$ and $\inv$. \textbf{[PETR: POLYHEDRAL GUARDS]}
 \end{theorem}

### 2017-07-05 paper/app.tex
re-shuffled the proof of the main theorem a bit

@@
+%\nouppercaseheads
+%\def\@secfont{\sffamily\bfseries}
 \section{Details of Program Syntax}
@@
 successors $\loc_1,\loc_2$.
+
+\def\xx{\ref{prop:conditional-exp-existence}}
+
+\section{Proof of Proposition~\xx}
+
+We first recall the general statement of the Radon-Nikodym theorem.
+Given two measurable spaces\footnote{A generalization of a probability space where the measure of $\Omega$ does not have to be 1, but any non-negative number or even infinity.} $(\Omega,\genfilt,\mu)$ and $(\Omega,\genfilt,\nu)$, we say that $\nu$ dominates $\mu$, written $\mu<<\nu$ if for all $A\in \genfilt$, $\nu(A)=0$ implies $\mu(A)=0$.
+Radon-Nikodym theorem states that if both $\mu$ and $\nu$ are sigma-finite (that is, $\Omega$ is a union of countably many sets of finite measure under $\nu$ and $\mu$), then $\mu<<\nu$ implies that there exists an almost-surely unique $\genfilt$-measurable function $f\colon \Omega\rightarrow [0,\infty)$ such that for each $A \in\genfilt$, the Lebesgue integral of the function $f\cdot\indicator{A}$ in measurable space  $(\Omega,\genfilt,\nu)$ is equal to $\mu(A)$.
+The function $f$ is called a Radon-Nikodym derivative of $\mu$ w.r.t.
+$\nu$, and we denote in by $\frac{d\mu}{d\nu}$.
+
+Now assume that $X$ is a non-negative real-valued random variable in some probability space $(\Omega,\genfilt,\probm)$ and $\genfilt'$ is a sub-sigma algebra of $\genfilt$.
+Note that $\probm$ is sigma-finite.
+Define a measure $\mu$ on $\genfilt'$ by putting $\mu(A)=\E[X\cdot\indicator{A}]$, for each $A\in \genfilt$ (here $\E$ is the expectation operator, i.e.
+the Lebesgue integral, in probability space $(\Omega,\genfilt',\probm')$, where $\probm'$ is a restriction of $\probm$ to $\genfilt'$).
+Then $\mu$ is sigma-finite: indeed, for any $n\in \Nset$ let $A_n = \{\omega\in\Omega\mid X(\omega)\leq n\}$.
+Then $\mu(A_n)\in [0,n]$, in particular it is finite, and since $X$ is real-valued, we have $\Omega=\bigcup_{n=1}^{\infty} A_n$.
+Hence, $\frac{d\mu}{d\probm'}$ exists and is almost-surely unique.
+It is now easy to check that $\frac{d\mu}{d\probm'}$ satisfies the condition defining the conditional expectation $\E[X\mid \genfilt']$: indeed, the condition is equivalent to $\E[X\mid \genfilt']$ being a derivative of $\mu$ w.r.t.
+$\probm'$.
+This concludes the proof.
+
+%%\section{Integrability of Variables in Non-Affine Programs}
+%%
+%%Consider the o
+
+\section{Computations for the proof of Theorem~\ref{THM:LEXRSM-MAIN}}
+
+Recall that we aim to prove equation~\eqref{eq:lexrsm-soundness-main}.
+
+For 
+$k=\fixn{i}$ the sum on the right-hand side equals $0$, so the 
+inequality immediately follows from the definition of $Y_k$. Now assume 
+that~\eqref{eq:lexrsm-soundness-main} holds for some $k\geq \fixn{i}$. We have that 
+\begin{align}
+\label{eq:lexrsm-ind-1}
+\E[Y_{k+1}] &= \underbrace{\E[Y_{k+1}\cdot \indicator{\Omega\setminus D}]}_{=0=\E[Y_k\cdot\indicator{\Omega\setminus D}]} +  \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{U \leq k\}}]}_{=\E[Y_k\cdot\indicator{D\cap \{U\leq k\}}]} + \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{U > k\}}]}_{=\E[\vecseq{X}{k+1}{\fixn{j}}\cdot\indicator{D\cap \{U>k \} }]},
+\end{align}
+where the equality $\E[Y_{k+1}\cdot \indicator{D \cap \{U \leq k\}}] =\E[Y_k\cdot\indicator{D\cap \{U\leq k\}}] $ follows from the fact that $Y_{k+1}(\omega)=Y_k({\omega})=\vecseq{X}{U(\omega)}{\fixn{j}}(\omega)$ for $\omega\in \{U\leq k\}$, and similarly for the last term.
+We prove that
+\begin{equation}
+\label{eq:lexrsm-ind-2}
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \}}] \leq \E[Y_k\cdot \indicator{D\cap \{U > k \}} -\eps\cdot \indicator{D\cap \{U>k\} \cap \{\levelrank{}{k}= \fixn{j}\} }] .
+\end{equation}
+Indeed, it holds
+\begin{equation}
+\label{eq:lexrsm-ind-3}
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \}}] = \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }]  + \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j}\} }],
+\end{equation}
+since $\levelrank{\omega}{k}\geq \fixn{j}$ for all $\omega \in \{U>k\}$. Since the set $D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}$ is $\genfilt_k$-measurable, we get
+\begin{align}
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] &= \E[\E[\vecseq{X}{k+1}{\fixn{j}}\mid \genfilt_k]\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] \label{eq:lexrsm-ind-4}\\
+&\leq \E[(\vecseq{X}{k}{\fixn{j}} - \eps)\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-5}\\
+&=\E[(Y_k - \eps)\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-6},
+\end{align}
+where~\eqref{eq:lexrsm-ind-4} follows from the definition of conditional expectation~\eqref{eq:cond-exp},~\eqref{eq:lexrsm-ind-5} follows from the definition of $\{\levelrank{}{k}=\fixn{j}\}$, and~\eqref{eq:lexrsm-ind-6} holds since $Y_k(\omega)=\vecseq{X}{k}{\fixn{j}}(\omega)$ for $\omega$ with $U(\omega)> k$.
+Almost identical argument shows that
+\begin{equation}
+\label{eq:lexrsm-ind-7}
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j} \} }] \leq \E[Y_k\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j} \}}].
+\end{equation}
+Plugging~\eqref{eq:lexrsm-ind-6} and~\eqref{eq:lexrsm-ind-7} into~\eqref{eq:lexrsm-ind-3} yields~\eqref{eq:lexrsm-ind-2}. Now we can plug~\eqref{eq:lexrsm-ind-2} into~\eqref{eq:lexrsm-ind-1} to get
+\begin{align}
+\E[Y_{k+1}]&\leq \E[Y_k] - \eps\cdot \E[\indicator{D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \} }] = \E[Y_k] - \eps\cdot \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \} ) \nonumber\\
+&\leq \fixn{B}\cdot \probm(D) - \eps\cdot\left(\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D 
+\cap \{U\geq k\} \cap \{\noofdec_k = \ell\})\right) -\eps\cdot \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \}),
+\end{align}
+where the last inequality follows from induction hypothesis. Hence, using $D_{k,\ell}$ as a shorthand for $D 
+\cap \{U\geq k\} \cap \{\noofdec_k = \ell\}$, to prove~\eqref{eq:lexrsm-soundness-main} it remains to show that
+\begin{equation}
+\label{eq:lexrsm-ind-8}
+\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}) + \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \}) = \sum_{\ell=0}^{k+1-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}).
+\end{equation}
+The left-hand side of~\eqref{eq:lexrsm-ind-8} is equal to
+\begin{align}
+&\phantom{+}\;\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell} \cap \{\levelrank{}{k}=\fixn{j}\}) + \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} )\nonumber \\
+&+\sum_{\ell=0}^{k-\fixn{i}}\probm(\underbrace{D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j}\} \cap \{\noofdec_k = \ell\}}_{=D_{k,\ell} \cap \{\levelrank{}{k} = \fixn{j}\}})
+\nonumber \\
+&= \sum_{\ell=0}^{k-\fixn{i}}(\ell+1) \cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k} = \fixn{j}\}) +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} ) \nonumber\\
+&=\sum_{\ell=0}^{k-\fixn{i}} (\ell+1)\cdot \probm{(D_{k+1,\ell+1} \cap \{\levelrank{}{k} = \fixn{j}\})} +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} ) \label{eq:lexrsm-ind-9}\\
+&= \sum_{\ell=1}^{k+1-\fixn{i}} \ell\cdot \probm{(D_{k+1,\ell} \cap \{\levelrank{}{k} = \fixn{j}\})} +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} )
+\nonumber \\
+&= (k+1-\fixn{i})\cdot \probm{(D_{k+1,k+1-\fixn{i}}\cap\{\levelrank{}{k} =\fixn{j} \} )} + \sum_{\ell=1}^{k-\fixn{i}} \ell \cdot \probm(D_{k+1,\ell})  \nonumber\\
+&= (k+1-\fixn{i})\cdot\probm(D_{k+1,k+1-\fixn{i}}) + \sum_{\ell=1}^{k-\fixn{i}}\ell\cdot\probm(D_{k+1,\ell}) = \text{ right-hand side of~\eqref{eq:lexrsm-ind-8}} \label{eq:lexrsm-ind-11}.
+\end{align}
+The individual steps in the above computation are justified as follows: in~\eqref{eq:lexrsm-ind-9} we use the facts that for all $\omega$'s whose level in step $k$ is $\fixn{j}$ it holds that $\noofdec_k(\omega)+1=\noofdec_{k+1}(\omega)$, and similarly, for $\omega$'s whose level in step $k$ is $>\fixn{j}$ it holds $\noofdec_k(\omega)=\noofdec_{k+1}(\omega)$.
+Moreover, for all $\omega\in D_{k,\ell}$ it holds that $\levelrank{k}{\omega}\geq \fixn{j} \Rightarrow U(\omega)\geq k+1$.
+Finally, in~\eqref{eq:lexrsm-ind-11} we use the fact that all $\omega\in D_{k+1,k+1-\fixn{i}}$ need to have level $\fixn{j}$ in step $k$, since otherwise such an $\omega$ would need to have level $\fixn{j}$ for at least $k+1-\fixn{i}$ times within steps $\{\fixn{i},\fixn{i}+1,\dots,k-1\}$, but there are $k-\fixn{i}$ such steps, a contradiction.
+This concludes the proof of~\eqref{eq:lexrsm-soundness-main}.

### 2017-07-05 paper/lexicographic.tex
re-shuffled the proof of the main theorem a bit

@@
 \label{prop:conditional-exp-existence}
+\label{PROP:CONDITIONAL-EXP-EXISTENCE}
 Let $(\Omega,\mathcal{F},\probm)$ be a probability space, 
@@
 \end{proposition}
-\textbf{[PETR: POLISH, ADD INTUITION]}
-\begin{proof}
+%\textbf{[PETR: POLISH, ADD INTUITION]}
+\begin{proof}[Proof (Key ideas)]
 The proof for the case when $\E[|X|]$ is standard and appears in many textbooks on probability theory (e.g.~\cite{Billingsley:book,Ash:book,Rosenthal:book}).
@@
 
-\begin{definition}
-Let $(\Omega,\genfilt,\probm)$ be a probability space and $O\subseteq \Omega$ some set.
- We say that sets $L_1,\dots,L_n$ form an almost-sure partition of $O$ if $L_1,\dots,L_n$ are pairwise disjoint and $\probm(\{\omega\in O\mid \omega \not \in L_1\cup\dots \cup L_n\})=0$
-\end{definition}
+%\begin{definition}
+%Let $(\Omega,\genfilt,\probm)$ be a probability space and $O\subseteq \Omega$ some set.
+% We say that sets $L_1,\dots,L_n$ form an almost-sure partition of $O$ if $L_1,\dots,L_n$ are pairwise disjoint and $\probm(\{\omega\in O\mid \omega \not \in L_1\cup\dots \cup L_n\})=0$
+%\end{definition}
 
@@
 $\vec{X}_i (\omega)[j]\geq {0}$. % and $\vec{X}_i(\omega)[j]\neq \infty$.
-\item For each $i\in \Nset_0$ there exists an almost-sure partition of the set $\{\stime>i\}$ into $n$ subsets $L^i_1,\dots,L^i_n$, all of them $\genfilt_i$-measurable, such that for each $1\leq j \leq n$:
+\item For each $i\in \Nset_0$ there exists a partition of the set $\{\stime>i\}$ into $n$ subsets $L^i_1,\dots,L^i_n$, all of them $\genfilt_i$-measurable, such that for each $1\leq j \leq n$:
 \begin{compactitem}
@@
 
-The 1-dimensional lexicographic $\eps$-ranking supermartingale is, to a large extent, equivalent to the notion of a ranking supermartingale as studied in~\cite{xxx}.
-There is one significant difference: in~\cite{xxx} there is an additional \emph{integrability} condition imposed on the one-dimensional process $\{X_i\}_{i=0}^{\infty}$, which requires that for each $i\geq 0$ it holds $\E[|X_i|]<\infty$ (or equivalently $\E[X_i]<\infty$, as the process is required to be non-negative).
-We do not impose this condition, and indeed, it is possible to define a 3-dimensional LexRSM $\{\vec{X}_{i}\}_{i=0}^{\infty}$ such that $\E[X[3]_i]=\infty$ for some $i$ \textbf{[PETR: EXAMPLE]}.
-However, closer inspection of proofs in~\cite{xxx} that link 1-dimensional ranking supermartingales to a.s.
-termination reveals that the integrability condition is never explicitly used there.
-Indeed, the only point in which integrability is needed is to ensure that the conditional expectations exist and are well-defined.
-However, the existence of conditional expectations is also guaranteed for random variables that are non-negative, even if their expected value if infinite, see Theorem~\ref{xxx}.
+The 1-dimensional lexicographic $\eps$-ranking supermartingale is, to a large extent, equivalent to the notion of a ranking supermartingale as studied in~\cite{HolgerPOPL,CFNH16:prob-termination}.
+There is one significant difference: in these works there is an additional \emph{integrability} condition imposed on the one-dimensional process $\{X_i\}_{i=0}^{\infty}$, which requires that for each $i\geq 0$ it holds $\E[|X_i|]<\infty$ (or equivalently $\E[X_i]<\infty$, as the process is required to be non-negative).
+We do not impose this condition, which simplifies possible application of LexRSMs to programs with non-linear arithmetic, where, as already shown in~\cite{HolgerPOPL}, integrability of program variables is not guaranteed.
+%We do not impose this condition, and indeed, it is possible to define a 3-dimensional LexRSM $\{\vec{X}_{i}\}_{i=0}^{\infty}$ such that $\E[X[3]_i]=\infty$ for some $i$ (example in \AppendixMaterial).
+The reason why integrability condiiton can be dropped is that it is only needed in the previous works to ensure that the conditional expectations exist and are well-defined.
+However, the existence of conditional expectations is also guaranteed for random variables that are real-valued non-negative, see Proposition~\ref{prop:conditional-exp-existence}.
 This is exactly the case in both the original 1-dimensional ranking supermartingales and our generalization to LexRSMs.
-Below we show, that also in the LexRSM case we do not need integrability to prove a.s.
+
+The following theorem states our main mathematical result on LexRSMs.
+%Below we show, that also in the LexRSM case we do not need integrability to prove a.s.
 termination, and hence we drop the condition altogether.
-This is particularly relevant when applying LexRSMs to programs with non-linear arithmetic, where proving integrability of LexRSM is more difficult than in the affine case, see Example~\ref{xxx}.
+This is particularly relevant when applying LexRSMs to programs with non-linear arithmetic, where integrability of program variables is not guaranteed.
 
@@
 \label{thm:lexrsm-main}
+\label{THM:LEXRSM-MAIN}
 Let $(\Omega,\genfilt,\probm)$ be a probability space, 
@@
 stopping 
-time w.r.t. that filtration. If there exists an $\eps$-LexRSM for $\stime$, 
+time w.r.t. that filtration. If there exists an $n$-dimensional $\eps$-LexRSM for $\stime$, 
 then $\probm(\stime<\infty)=1$.
@@
 For $\omega\in \Omega$ we define the level of $\omega$ at step $i$ to be the 
-number $\levelrank{\omega}{i}$ equal to either $0$, if $\stime(\omega)\leq i$ or $\omega$ does not belong to $L^i_1 \cup \dots\cup L^i_n$, 
+number $\levelrank{\omega}{i}$ equal to either $0$, if $\stime(\omega)\leq i$, 
 or otherwise to the $1\leq j \leq n$ such that $\omega\in L^i_j$. By the definition of 
@@
 that $j$ is a level of $\omega$ at infinitely many steps. Note that $\omega \in 
-\genRunSet_{\infty}$ implies $\minlev(\omega)\neq 0$, and $\probm(\genRunSet_{\infty})=\probm(\{\minlev \neq 0 \})$, as for each $i$ almost all $\omega$'s for which $\stime(\omega)>i$ belong to $L^i_1 \cup \dots\cup L^i_n$.
-We denote by $M_i$
+\genRunSet_{\infty}$ if and only if $\minlev(\omega)\neq 0$, so $\probm(\genRunSet_{\infty})=\probm(\{\minlev \neq 0 \})$. We denote by $M_i$ 
 the set of all $\omega$'s with $\minlev(\omega)=i$.
@@
 Now since $\genRunSet_{\infty}=M_1,\dots,M_n$, there must be $1\leq \fixn{j} \leq n$ 
-s.t. $\probm(M_{\fixn{j}})>0$. For each $\omega\in M_{\fixn{j}}$ there is the smallest number
-${i}_{\omega,\fixn{j}}\in\Nset_0$ such that for all $i\geq {i}_{\omega,{j}}$ it holds 
-$\levelrank{\omega}{i}\geq \fixn{j}$. Denote by $S_{\fixn{j},{i}}$ the set of all $\omega$'s in $M_{\fixn{j}}$ s.t. 
+s.t. $\probm(M_{\fixn{j}})>0$, i.e. with positive probability the smallest level appearing infinitely often is $\fixn{j}$. For each $\omega\in M_{\fixn{j}}$ there is the smallest number
+${i}_{\omega,\fixn{j}}\in\Nset_0$ such that for all $i\geq {i}_{\omega,\fixn{j}}$ it holds 
+$\levelrank{\omega}{i}\geq \fixn{j}$, i.e.
+after step ${i}_{\omega,\fixn{j}}$ the level of $\omega$ in all steps up to infinity is at least $\fixn{j}$.
+Denote by $S_{\fixn{j},{i}}$ the set of all $\omega$'s in $M_{\fixn{j}}$ s.t.
 $i_{\omega,\fixn{j}}={i}$. 
@@
 s.t. 
-$\probm(M_{\fixn{j},\fixn{i}})>0$. Continuing on the same note, for each $B\in \Nset$ we 
+$\probm(M_{\fixn{j},\fixn{i}})>0$.
+That is, there is a point in time such that with positive probability, after this point the level of $\omega$ is at least $\fixn{j}$, and it is equal to $\fixn{j}$ infinitely many times.
+Continuing on the same note, for each $B\in \Nset$ we
 denote by $M_{\fixn{j},\fixn{i}}^{B}$ the set off all $\omega$'s in $M_{\fixn{j},\fixn{i}}$ s.t. 
@@
 \cup M_{\fixn{j},\fixn{i}}^{3} \cup \cdots  $, there is $\fixn{B}\in \Nset$ s.t. 
-$\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}})>0$. Denote $\fixn{p}=\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}})$.
+$\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}})>0$. 
+%Denote $\fixn{p}=\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}})$.
+
+So there is a set of positive probability $M=M_{\fixn{j},\fixn{i}}^{\fixn{B}}$  such that for all $\omega$'s in the set: after step $\fixn{i}$ the level of $\omega$ is at least $\fixn{j}$ (which, intuitively, means that $\vec{X}[\fixn{j}]$ does not have a tendency to increase after this time step on $\omega$'s in $M$), the level of $\omega$ is infinitely often equal to $\fixn{j}$ (intuitively, $\vec{X}[\fixn{j}]$ has infinitely often the tendency to decrease by $\geq \eps$ for $\omega$'s in $M$), and at time $\fixn{i}$ the value of $\vec{X}[\fixn{j}]$ is bounded (by $B$) on $M$.
+This should, again intuitively, lead to a conclusion, that when ``restricted to $M$'', $\vec{X}[\fixn{j}]$ has tendency to decrease unboundedly over time, a contradiction with non-negativeness of $\vec{X}[\fixn{j}]$.
+However, proving this intuitive result is much more intricate: most importantly, it is not clear what ``restricted to $M$'' stands for.
+The stochastic process $\{\vec{X}\}_{i=0}^{\infty}$ as well as the LexRSM conditions are tied to the filtration $\{\genfilt_i\}_{i=0}^{\infty}$, but the set $M$ is not necessarily $\genfilt_i$ measurable for any concrete $i$, since whether $\omega$ belongs to $M$ depends on values of $\levelrank{\omega}{i}$ for \emph{infinitely many $i$}.
+Hence, we use a work-around.
 
 Let $D$ be the set of all $\omega\in \Omega$ such that 
-$\vecseq{X}{\fixn{i}}{\fixn{j}}(\omega)\leq \fixn{B}$. Note that $M_{\fixn{j},\fixn{i}}^{B} \subseteq D$ and $D\in 
-\genfilt_i$ (and thus also $D\in \genfilt_{i'}$ for all $i'\geq i$). 
-Define a stopping time $U$ w.r.t. filtration $\{\genfilt_i\}_{i=0}^{\infty}$ as follows: for all $\omega \in \Omega$ 
+$\vecseq{X}{\fixn{i}}{\fixn{j}}(\omega)\leq \fixn{B}$. Note that $M \subseteq D$ and $D\in 
+\genfilt_{\fixn{i}}$ (and thus also $D\in \genfilt_{i'}$ for all $i'\geq \fixn{i}$). 
+Define a stopping time $F$ w.r.t. filtration $\{\genfilt_i\}_{i=0}^{\infty}$ as follows: for all $\omega \in \Omega$ 
 we put 
-$U(\omega)=\inf\{k\in \Nset_0 \mid k\geq \fixn{i} \text{ and }
+$F(\omega)=\inf\{k\in \Nset_0 \mid k\geq \fixn{i} \text{ and }
 \levelrank{\omega}{k} < \fixn{j} \}$.
@@
 }\\
-\vecseq{X}{U(\omega)}{\fixn{j}}(\omega) & \text{if $\omega\in D$, $k\geq 
+\vecseq{X}{F(\omega)}{\fixn{j}}(\omega) & \text{if $\omega\in D$, $k\geq 
 \fixn{i}$ and 
-$U(\omega)\leq k$ }.
+$F(\omega)\leq k$ }.
 \end{dcases}
@@
 
-We prove several properties of the process $\{Y_k\}_{k=0}^{\infty}$. 
+Intuitively, the process $\{Y_k\}_{k=0}^{\infty}$ is an over-approximation of what we would like to call ``$\vec{X}[\fixn{j}]$ restricted to $M$.''
+We prove several properties of the process. 
 First, clearly for all $k\geq 0$, $Y_k(\omega)\geq 0$.
@@
 $\noofdec_k$ the random variable such that $\noofdec_k(\omega) = |\{i'\in 
-\Nset\mid \fixn{i} \leq i' < k \text{ and } \levelrank{\omega}{i'}=\fixn{j}\} |$.
-We claim 
+\Nset\mid \fixn{i} \leq i' < k \text{ and } \levelrank{\omega}{i'}=\fixn{j}\} |$, i.e. $\noofdec_k(\omega)$ counts the number of steps between $\fixn{i}$ and $k$ in which level is $\fixn{j}$.
+We prove
 that for each $k\geq \fixn{i}$ 
@@
 \E[Y_k]\leq \fixn{B}\cdot \probm(D) - \eps\cdot\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D 
-\cap \{U\geq k\} \cap \{\noofdec_k 
+\cap \{F\geq k\} \cap \{\noofdec_k 
 = 
@@
 
-The proof of~\eqref{eq:lexrsm-soundness-main} goes by induction on $k$. For 
-$k=\fixn{i}$ the sum on the right-hand side equals $0$, so the 
-inequality immediately follows from the definition of $Y_k$. Now assume 
-that~\eqref{eq:lexrsm-soundness-main} holds for some $k\geq \fixn{i}$. We have that 
-\begin{align}
-\label{eq:lexrsm-ind-1}
-\E[Y_{k+1}] &= \underbrace{\E[Y_{k+1}\cdot \indicator{\Omega\setminus D}]}_{=0=\E[Y_k\cdot\indicator{\Omega\setminus D}]} +  \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{U \leq k\}}]}_{=\E[Y_k\cdot\indicator{D\cap \{U\leq k\}}]} + \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{U > k\}}]}_{=\E[\vecseq{X}{k+1}{\fixn{j}}\cdot\indicator{D\cap \{U>k \} }]},
-\end{align}
-where the equality $\E[Y_{k+1}\cdot \indicator{D \cap \{U \leq k\}}] =\E[Y_k\cdot\indicator{D\cap \{U\leq k\}}] $ follows from the fact that $Y_{k+1}(\omega)=Y_k({\omega})=\vecseq{X}{U(\omega)}{\fixn{j}}(\omega)$ for $\omega\in \{U\leq k\}$, and similarly for the last term.
-We prove that
-\begin{equation}
-\label{eq:lexrsm-ind-2}
-\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \}}] \leq \E[Y_k\cdot \indicator{D\cap \{U > k \}} -\eps\cdot \indicator{D\cap \{U>k\} \cap \{\levelrank{}{k}= \fixn{j}\} }] .
-\end{equation}
-Indeed, it holds
-\begin{equation}
-\label{eq:lexrsm-ind-3}
-\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \}}] = \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }]  + \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j}\} }],
-\end{equation}
-since $\levelrank{\omega}{k}\geq \fixn{j}$ for all $\omega \in \{U>k\}$. Since the set $D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}$ is $\genfilt_k$-measurable, we get
-\begin{align}
-\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] &= \E[\E[\vecseq{X}{k+1}{\fixn{j}}\mid \genfilt_k]\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] \label{eq:lexrsm-ind-4}\\
-&\leq \E[(\vecseq{X}{k}{\fixn{j}} - \eps)\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-5}\\
-&=\E[(Y_k - \eps)\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-6},
-\end{align}
-where~\eqref{eq:lexrsm-ind-4} follows from the definition of conditional expectation~\eqref{eq:cond-exp},~\eqref{eq:lexrsm-ind-5} follows from the definition of $\{\levelrank{}{k}=\fixn{j}\}$, and~\eqref{eq:lexrsm-ind-6} holds since $Y_k(\omega)=\vecseq{X}{k}{\fixn{j}}(\omega)$ for $\omega$ with $U(\omega)> k$.
-Almost identical argument shows that
-\begin{equation}
-\label{eq:lexrsm-ind-7}
-\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j} \} }] \leq \E[Y_k\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j} \}}].
-\end{equation}
-Plugging~\eqref{eq:lexrsm-ind-6} and~\eqref{eq:lexrsm-ind-7} into~\eqref{eq:lexrsm-ind-3} yields~\eqref{eq:lexrsm-ind-2}. Now we can plug~\eqref{eq:lexrsm-ind-2} into~\eqref{eq:lexrsm-ind-1} to get
-\begin{align}
-\E[Y_{k+1}]&\leq \E[Y_k] - \eps\cdot \E[\indicator{D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \} }] = \E[Y_k] - \eps\cdot \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \} ) \nonumber\\
-&\leq \fixn{B}\cdot \probm(D) - \eps\cdot\left(\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D 
-\cap \{U\geq k\} \cap \{\noofdec_k = \ell\})\right) -\eps\cdot \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \}),
-\end{align}
-where the last inequality follows from induction hypothesis. Hence, using $D_{k,\ell}$ as a shorthand for $D 
-\cap \{U\geq k\} \cap \{\noofdec_k = \ell\}$, to prove~\eqref{eq:lexrsm-soundness-main} it remains to show that
-\begin{equation}
-\label{eq:lexrsm-ind-8}
-\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}) + \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \}) = \sum_{\ell=0}^{k+1-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}).
-\end{equation}
-The left-hand side of~\eqref{eq:lexrsm-ind-8} is equal to
-\begin{align}
-&\phantom{+}\;\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell} \cap \{\levelrank{}{k}=\fixn{j}\}) + \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} )\nonumber \\
-&+\sum_{\ell=0}^{k-\fixn{i}}\probm(\underbrace{D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j}\} \cap \{\noofdec_k = \ell\}}_{=D_{k,\ell} \cap \{\levelrank{}{k} = \fixn{j}\}})
-\nonumber \\
-&= \sum_{\ell=0}^{k-\fixn{i}}(\ell+1) \cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k} = \fixn{j}\}) +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} ) \nonumber\\
-&=\sum_{\ell=0}^{k-\fixn{i}} (\ell+1)\cdot \probm{(D_{k+1,\ell+1} \cap \{\levelrank{}{k} = \fixn{j}\})} +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} ) \label{eq:lexrsm-ind-9}\\
-&= \sum_{\ell=1}^{k+1-\fixn{i}} \ell\cdot \probm{(D_{k+1,\ell} \cap \{\levelrank{}{k} = \fixn{j}\})} +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} )
-   \nonumber \\
-&= (k+1-\fixn{i})\cdot \probm{(D_{k+1,k+1-\fixn{i}}\cap\{\levelrank{}{k} =\fixn{j} \} )} + \sum_{\ell=1}^{k-\fixn{i}} \ell \cdot \probm(D_{k+1,\ell})  \nonumber\\
-&= (k+1-\fixn{i})\cdot\probm(D_{k+1,k+1-\fixn{i}}) + \sum_{\ell=1}^{k-\fixn{i}}\ell\cdot\probm(D_{k+1,\ell}) = \text{ right-hand side of~\eqref{eq:lexrsm-ind-8}} \label{eq:lexrsm-ind-11}.
-\end{align}
-The individual steps in the above computation are justified as follows: in~\eqref{eq:lexrsm-ind-9} we use the facts that for all $\omega$'s whose level in step $k$ is $\fixn{j}$ it holds that $\noofdec_k(\omega)+1=\noofdec_{k+1}(\omega)$, and similarly, for $\omega$'s whose level in step $k$ is $>\fixn{j}$ it holds $\noofdec_k(\omega)=\noofdec_{k+1}(\omega)$.
-Moreover, for all $\omega\in D_{k,\ell}$ it holds that $\levelrank{k}{\omega}\geq \fixn{j} \Rightarrow U(\omega)\geq k+1$.
-Finally, in~\eqref{eq:lexrsm-ind-11} we use the fact that all $\omega\in D_{k+1,k+1-\fixn{i}}$ need to have level $\fixn{j}$ in step $k$, since otherwise such an $\omega$ would need to have level $\fixn{j}$ for at least $k+1-\fixn{i}$ times within steps $\{\fixn{i},\fixn{i}+1,\dots,k-1\}$, but there are $k-\fixn{i}$ such steps, a contradiction.
-This concludes the proof of~\eqref{eq:lexrsm-soundness-main}.
+The proof of~\eqref{eq:lexrsm-soundness-main} goes by induction on $k$. The computations being somewhat technical, we defer them to \AppendixMaterial.
 
@@
 \ell\})$ for all $k\geq \fixn{i}$.
-Let $m=3\fixn{B}\cdot \probm(D)/(\eps\cdot \fixn{p})$.
-For each $\omega\in M_{\fixn{j},\fixn{i}}^{\fixn{B}} $ there exists $k_m(\omega)\geq \fixn{i}$ such that $\noofdec_{k_{m}(\omega)}\geq m$.
-Hence, $M_{\fixn{j},\fixn{i}}^{\fixn{B}} = \bigcup_{k=\fixn{i}}^{\infty} M_{\fixn{j},\fixn{i}}^{\fixn{B}} \cap \{k_{m}(\omega) \leq k\}$ and $\fixn{p}=\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}})=\lim_{k\rightarrow \infty}\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}} \cap \{k_{m} \leq k\})$.
-Thus, there exists $k_0$ such that $\probm(M_{\fixn{j},\fixn{i}}^{\fixn{B}} \cap \{\noofdec_{k_0}\geq m\}) \geq \fixn{p}/2$.
-Denote $\fixn{M}= M_{\fixn{j},\fixn{i}}^{\fixn{B}} \cap \{\noofdec_{k_0}\geq m\}$.
-Clearly $\fixn{M} \subseteq D\cap \{U\geq k_0 \} \cap \{\noofdec_{k_0}\geq m \}$.
+Let $m=3\fixn{B}\cdot \probm(D)/(\eps\cdot \probm(M))$.
+For each $\omega\in M$ we see level $\fixn{j}$ infinitely often, so there exists step $k(\omega)\geq \fixn{i}$ such that $\noofdec_{k(\omega)}\geq m$, i.e.
+$\omega$ has level $\fixn{j} $ at least in $m$ steps between steps $\fixn{i}$ and $k(\omega)$.
+Clearly, $M = \bigcup_{\ell=\fixn{i}}^{\infty} (M \cap \{k(\omega) \leq \ell\})$ and hence $\probm(M)=\lim_{\ell\rightarrow \infty}\probm(M \cap \{k_{m} \leq \ell\})$.
+Thus, there exists $\ell_0 \geq m+\fixn{i}$ such that $\probm(M \cap \{\noofdec_{\ell_0}\geq m\}) \geq \probm(M)/2$.
+Clearly $M \cap \{\noofdec_{\ell_0}\geq m\} \subseteq D\cap \{F\geq \ell_0 \} \cap \{\noofdec_{\ell_0}\geq m \}$.
 From \eqref{eq:lexrsm-soundness-main} it follows that
 \begin{align*}
-\E[Y_{k_0}]&\leq \fixn{B}\cdot \probm(D) - \eps\cdot\sum_{\ell=0}^{k_0-\fixn{i}} \ell\cdot\probm(D 
-\cap \{U\geq k_0\} \cap \{\noofdec_{k_0} 
+\E[Y_{\ell_0}]&\leq \fixn{B}\cdot \probm(D) - \eps\cdot\sum_{\ell=0}^{\ell_0-\fixn{i}} \ell\cdot\probm(D 
+\cap \{U\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
 =
 \ell\}) \\
-&= \fixn{B}\cdot \probm(D) - \eps\cdot \sum_{\ell=1}^{k_0-\fixn{i}} \probm(D 
-\cap \{U\geq k_0\} \cap \{\noofdec_{k_0} 
+&= \fixn{B}\cdot \probm(D) - \eps\cdot \sum_{\ell=1}^{\ell_0-\fixn{i}} \probm(D 
+\cap \{U\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
 \geq 
@@
 & \leq \fixn{B}\cdot \probm(D) - \eps\cdot \sum_{\ell=1}^{m} \probm(D 
-\cap \{U\geq k_0\} \cap \{\noofdec_{k_0} 
+\cap \{U\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
 \geq 
-\ell\})\\
-&\leq \fixn{B}\cdot \probm(D) - \eps\cdot m \cdot \fixn{p}/2 < 0, 
-\end{align*}
-the last inequality following by substituting for $m$. But for each $k$ the random variable $Y_k$, so it must also have a non-negative expectation, a contradiction.
+\ell\}) \\&\leq \fixn{B}\cdot \probm(D) - \eps\cdot \sum_{\ell=1}^{m} \probm(D 
+\cap \{U\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
+\geq 
+m\})\\
+&\leq \fixn{B}\cdot \probm(D) - \eps\cdot m \cdot \probm(M)/2 < 0, 
+\end{align*}where the second line follows by standard re-arranging of terms, the third line follows from the fact that $m\leq \ell_0 + \fixn{i}$, the fourth line follows from $\{\noofdec_{\ell_0} 
+\geq 
+m \} \subseteq \noofdec_{\ell_0} 
+\geq 
+\ell$ for each $\ell \leq m$, the first inequality on the last line follows by  using $\probm(M)/2\leq \probm(M \cap \{\noofdec_{\ell_0}\geq m\}) \leq \probm(D\cap \{F\geq \ell_0 \} \cap \{\noofdec_{\ell_0}\geq m \})$, and
+the last inequality follows by expanding the definition of $m$. But for each $k$ the random variable $Y_k$ is non-negative, so it must also have a non-negative expectation, a contradiction.
 %\begin{enumerate}

### 2017-07-05 paper/synthesis.tex
re-shuffled the proof of the main theorem a bit

@@
 In this section we describe a polynomial-time algorithm for synthesizing linear $\eps$-LexRSM maps in affine probabilistic programs supported by a given linear invariant map $\inv$.
-The algorithm, based on iterative solving of linear constraints, is an adaptation of a similar algorithm for finding lexicographic ranking functions in non-probabilistic programs~\cite{xxx}.
+The algorithm, based on iterative solving of linear constraints, is an adaptation of a similar algorithm for finding lexicographic ranking functions in non-probabilistic programs~\cite{ADFG10:lexicographic}.
 Hence, we provide only a high-level description, focusing on new aspects of our adaptation.
@@
 Otherwise, we remove the newly ranked elements from $U$ and continue into the next iteration, until $U$ becomes empty.
-The process is summarized in Algorithm~\ref{xxx}.
+The process is summarized in Algorithm~\ref{algo:linlexrsm}.
 
@@
 transitions in $U$ ranked by it.
-Both these tasks can be accomplished by the standard  method of linear constraints based on the use of Farkas's lemma, which was widely use for synthesis of termination proofs in both probabilistic and non-probabilistic programs~\cite{xxx}.
+Both these tasks can be accomplished by the standard  method of linear constraints based on the use of Farkas's lemma, which was widely use for synthesis of termination proofs in both probabilistic and non-probabilistic programs~\cite{DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,SriramCAV,CFNH16:prob-termination}.
 That is, the algorithm first constructs, for each location $\loc$ a \emph{template} for $\lem$, i.e.
@@
 \end{theorem}
-\begin{proof}[Proof (sketch)]
+\begin{proof}[Proof (Key Ideas)]
 For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the $i$-th iteration of the while loop.
@@
 
-Proving completeness is more intricate; as pointed out in~\cite{xxx}, one needs to show that the greedy strategy of selecting LEM that ranks maximal number of remaining transitions does not cut off some possible LinLexRSMs.
-In~\cite{xxx} the completeness of the greedy strategy is proved by using several geometric arguments that exploit the fact that the underlying programs are affine.
-The same geometric properties hold for our \APP{}s (all ranking conditions in Definitions~\ref{def:rank1} and~\ref{def:rank2} are linear in program variables), the result is easily transferable.
+Proving completeness is more intricate; as pointed out in~\cite{ADFG10:lexicographic}, one needs to show that the greedy strategy of selecting LEM that ranks maximal number of remaining transitions does not cut off some possible LinLexRSMs.
+In~\cite{ADFG10:lexicographic} the completeness of the greedy strategy is proved by using several geometric arguments that exploit the fact that the underlying programs are affine.
+The same geometric properties hold for our generalization to \APP{}s (all ranking conditions in Definitions~\ref{def:rank1} and~\ref{def:rank2} are linear in program variables), so the result is easily transferable.
 \end{proof}
 
-\
+%\

### 2017-07-06 implementation/examples/realheapsort.c
Examples tested

@@ -1,50 +1,0 @@
-//realheapsort
-void realheapsort()
-{
-  float t[100], temp;
-  int N;
-  int k,j,m;
-  if(N > 2){
-  //construction
-    for (k=1;k<=N-1;k++)
-      {
-        j=k;
-        // t[pere j] > t[j]
-        while ((j>0) && (t[(j+1)/2-1]>t[j]))
-	  {
-	    //swap j avec pere j puis j<-pere j
-            temp = t[j];
-	    t[j] = t[(j+1)/2-1];
-	    t[(j+1)/2-1] = temp;
-	    j = (j+1)/2-1;
-	  }
-      }
-
-  //destruction
-  
-    for (k=0;k<=N-2;k++)
-      {
-        j = 0; m=0;
-        temp = t[N-k-1];
-        t[N-k-1]=t[0];
-        t[0]=temp;
-        while(2*j+1<=N-2-k)
-	  {
-	    if ((2*j+1==N-2-k) || (t[2*j+1]<t[2*j+2]))
-	      m=2*j+1;
-	    else
-	      m=2*j+2;
-            if (t[j]>t[m])
-	      {//swap j m
-	        temp=t[m];
-	        t[m]=t[j];
-                t[j]=temp;
-	        j=m;
-	      }
-            else j=N;
-	  
-	  }
-     
-      }
-  }
-}

### 2017-07-06 paper/exp_appendix.tex
Added experiments

\section{Experimental Results}\label{sec:app_ex}
% The table which is to be there in appendix
In Table~\ref{tab:exp1} in Section~\ref{} we presented results for twenty five benchmarks.
We now present additional result for other twenty seven benchmarks in Table~\ref{tab:exp3}.

\begin{center}
\begin{table}[]
  \centering
   \begin{tabular}{c|c|c|c|c|c}

   {Benchmark} & {Time (s)} & {Solution} & {Dimension} & {Prob. loops} & {Prob. Assignments
} \\\hline
{aaron2} & {0.07} & {yes} & {2} & {yes} & {yes
} \\\hline
{ax} & {0.09} & {yes} & {3} & {yes} & {yes
} \\\hline
{complex} & {0.08} & {yes} & {1} & {yes} & {yes
} \\\hline
{counterex1b} & {0.08} & {yes} & {3} & {yes} & {yes
} \\\hline
{cousot9} & {0.08} & {yes} & {3} & {yes} & {no
} \\\hline
{easy2} & {0.07} & {yes} & {2} & {yes} & {yes
} \\\hline
{loops} & {0.08} & {no} & {} & {no} & {no
} \\\hline
{nestedloop} & {0.09} & {yes} & {3} & {yes} & {no
} \\\hline
{perfect1} & {0.08} & {yes} & {3} & {yes} & {yes
} \\\hline
{random1d} & {0.08} & {yes} & {2} & {yes} & {yes
} \\\hline
{realbubble} & {0.18} & {yes} & {3} & {yes} & {yes
} \\\hline
{realheapsort} & {0.18} & {no} & {} & {no} & {no
} \\\hline
{realheapsort\_step1} & {0.09} & {no} & {} & {no} & {no
} \\\hline
{realselect} & {0.09} & {yes} & {3} & {yes} & {yes
} \\\hline
{rsd} & {0.08} & {yes} & {1} & {yes} & {yes
} \\\hline
{sipma91} & {0.11} & {yes} & {2} & {yes} & {yes
} \\\hline
{speedDis1} & {0.07} & {no} & {} & {no} & {no
} \\\hline
{speedFails1} & {0.07} & {yes} & {2} & {yes} & {yes
} \\\hline
{speedFails4} & {0.07} & {no} & {} & {no} & {no
} \\\hline
{speedNestedMultipleDep} & {0.08} & {yes} & {3} & {yes} & {yes
} \\\hline
{speedpldi3} & {0.08} & {yes} & {3} & {yes} & {yes
} \\\hline
{speedSimpleMultiple} & {0.08} & {no} & {} & {no} & {no
} \\\hline
{speedSingleSingle} & {0.09} & {yes} & {2} & {yes} & {yes
} \\\hline
{terminate} & {0.07} & {yes} & {2} & {yes} & {yes
} \\\hline
{wcet0} & {0.09} & {yes} & {2} & {yes} & {yes
} \\\hline
{wcet2} & {0.07} & {yes} & {2} & {yes} & {yes
} \\\hline
{wise} & {0.07} & {no} & {} & {no} & {no} \\\hline 
    \end{tabular}
    \caption{Caption}
    \label{tab:my_label}
\end{table}
\end{center}

% The table for constructed long examples

### 2017-07-06 paper/experiments.tex
Added experiments

\section{Experimental Results}
We have implemented our algorithm (in C++) and present two sets of experimental results.
For all the experimental results we use Aspic~\cite{} for invariant generation, 
and our algorithm requires linear-programming solution for which we use CPlex~\cite{}.
Our experimental results were obtained on [SPEC OF COMPUTER].

\smallskip\noindent{\em Abstraction of real-world programs.}
We consider benchmarks considered for real-world non-probabilistic programs 
in~\cite{ADFG10:lexicographic}.
Note that there are no abstraction tools available for probabilistic programs (to the
best of our knowledge). 
Hence we consider the abstract programs obtained from real-world benchmarks as 
considered in the benchmark suite of~\cite{ADFG10:lexicographic}.
Given these non-probabilistic programs we obtain probabilistic programs in two ways:
(a)~{\em probabilistic loops} where the existing while loops are made probabilistic 
by executing the statements with probability~1/2, and with remaining probabilities 
executing skip statements;
(b)~{\em probabilistic assignments} where the existing assignments are changed to 
probabilistic assignments with variables generated uniformly in range $[-1,1]$ 
(i.e., we consider a variable $r$ whose value is generated by probabilistic 
assignment uniformly in the range $[-1,1]$ and add the variable to the existing 
assignment).
We report our results on twentfy five benchmarks in Table~\ref{tab:exp1} (since we consider around 
fifty benchmark examples the results on the remaining ones are presented in Appendix, Section~\ref{sec:app_ex}).
We show the time taken by our approach, which is always less than $1/10$-th of second.
We show whether solution exists or not (i.e., whether linear lexicographic RSMs exist
or not), and if the solution exists we present the dimension of the lexicographic RSM
we obtain. In all cases the dimension is quite small (at most~3). 
We also show whether we enhance the program with probabilistic loops and probabilistic assignments.

\smallskip\noindent{\em Constructed large programs.} 
The programs obtained as abstractions of real-world programs in the benchmarks 
as mentioned above have between 10-100 lines of code. 
To test the how does our approach scale with larger codes we consider large
probabilistic programs generated as follows.
Given $n$ Boolean variables, we consider probabilistic while loops,
some nondeterministic conditional branches, and all possible $2^n$ if conditions
based on the Boolean variables. 
Hence given $n$ variables we have probabilistic programs of size $O(2^n)$, and 
we first ran an invariant generation tool, followed by our algorithm.
In all these examples lexicographic RSMs exist, and has dimension at most~3.
Even for programs with around 12K lines of codes the total time taken is around 
one hour, where the invariant generation (i.e., running Aspic) takes the maximum time, 
and our algorithm requires around two minutes.
The results are presented in Table~\ref{tab:exp2}.

% The main table which contains 25 entries

\begin{center}
\begin{table}[]
  \centering
   \begin{tabular}{c|c|c|c|c|c}

   {Benchmark} & {Time (s)} & {Solution} & {Dimension} & {Prob. loops} & {Prob. Assignments
} \\\hline
{alain} & {0.09} & {yes} & {2} & {yes} & {yes
} \\\hline
{catmouse} & {0.06} & {yes} & {2} & {yes} & {yes
} \\\hline
{counterex1a} & {0.08} & {no} & {} & {no} & {no
} \\\hline
{counterex1c} & {0.09} & {yes} & {3} & {yes} & {yes
} \\\hline
{easy1} & {0.07} & {yes} & {1} & {yes} & {yes
} \\\hline
{exmini} & {0.07} & {yes} & {2} & {yes} & {yes
} \\\hline
{insertsort} & {0.08} & {yes} & {3} & {yes} & {yes
} \\\hline
{ndecr} & {0.07} & {yes} & {2} & {yes} & {yes
} \\\hline
{perfect} & {0.09} & {yes} & {3} & {yes} & {yes
} \\\hline
{perffect2} & {0.1} & {yes} & {3} & {yes} & {no
} \\\hline
{real2} & {0.07} & {no} & {} & {no} & {no
} \\\hline
{realheapsort\_step2} & {0.07} & {yes} & {1} & {no} & {no
} \\\hline
{realshellsort} & {0.07} & {yes} & {1} & {no} & {no
} \\\hline
{serpent} & {0.08} & {yes} & {1} & {yes} & {yes
} \\\hline
{sipmabubble} & {0.08} & {yes} & {3} & {yes} & {yes
} \\\hline
{speedDis2} & {0.07} & {no} & {} & {no} & {no
} \\\hline
{speedFails2} & {0.06} & {yes} & {2} & {yes} & {no
} \\\hline
{speedNestedMultiple} & {0.08} & {yes} & {3} & {yes} & {yes
} \\\hline
{speedpldi2} & {0.07} & {yes} & {2} & {yes} & {yes
} \\\hline
{speedpldi4} & {0.07} & {yes} & {3} & {yes} & {yes
} \\\hline
{speedSimpleMultipleDep} & {0.07} & {no} & {} & {no} & {no
} \\\hline
{speedSingleSingle2} & {0.08} & {yes} & {2} & {yes} & {no
} \\\hline
{unperfect} & {0.08} & {yes} & {2} & {yes} & {no
} \\\hline
{wcet1} & {0.09} & {yes} & {2} & {yes} & {yes
} \\\hline
{while2} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline

\end{tabular}
\caption{Experimental results for benchmarks from~\cite{ADFG10:lexicographic} extended with probabilistic loops and/or probabilistic assignments.} \label{tab:exp1}
\end{table}
\end{center}

\begin{center}
\begin{table}[]
  \centering
   \begin{tabular}{c|c|c|c|c}

{Variables} & {LOC} & {Inv Time (s)} & {Our Time (s)} & {Total Time (s)} \\\hline
{2} & {20} & {0.050189} & {0.028366} & {0.078555} \\\hline
{3} & {32} & {0.060078} & {0.02359} & {0.083668} \\\hline
{4} & {56} & {0.074388} & {0.030035} & {0.104423} \\\hline
{5} & {104} & {0.131122} & {0.051545} & {0.182667} \\\hline
{6} & {200} & {0.354702} & {0.097943} & {0.452645} \\\hline
{7} & {392} & {1.30833} & {0.29947} & {1.6078} \\\hline
{8} & {776} & {7.55238} & {0.69566} & {8.24804} \\\hline
{9} & {1544} & {33.0638} & {2.4977} & {35.5615} \\\hline
{10} & {3080} & {164.088} & {8.763} & {172.851} \\\hline
{11} & {6152} & {817.914} & {35.37} & {853.284} \\\hline
{12} & {12296} & {4260.96} & {145.18} & {4406.14} \\\hline
\end{tabular}
\caption{Experimental results for synthetic examples.}\label{tab:exp2}
\end{table}
\end{center}

### 2017-07-06 paper/exp_appendix.tex
changes to experiments

@@
 % The table which is to be there in appendix
-In Table~\ref{tab:exp1} in Section~\ref{} we presented results for twenty five benchmarks.
+In Table~\ref{tab:exp1} in Section~\ref{sec:experiments} we presented results for twenty five benchmarks.
 We now present additional result for other twenty seven benchmarks in Table~\ref{tab:exp3}.
@@
     \end{tabular}
-    \caption{Caption}
-    \label{tab:my_label}
+    \caption{Additonal experimental results for benchmarks from~\cite{ADFG10:lexicographic} extended with probabilistic loops and/or probabilistic assignments.}
+    \label{tab:exp3}
 \end{table}

### 2017-07-06 paper/exp_appendix.tex
table cleaned

@@
 } \\\hline
-{aaron2} & {0.07} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{ax} & {0.09} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{complex} & {0.08} & {yes} & {1} & {yes} & {yes
-} \\\hline
-{counterex1b} & {0.08} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{cousot9} & {0.08} & {yes} & {3} & {yes} & {no
-} \\\hline
-{easy2} & {0.07} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{loops} & {0.08} & {no} & {} & {no} & {no
-} \\\hline
-{nestedloop} & {0.09} & {yes} & {3} & {yes} & {no
-} \\\hline
-{perfect1} & {0.08} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{random1d} & {0.08} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{realbubble} & {0.18} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{realheapsort} & {0.18} & {no} & {} & {no} & {no
-} \\\hline
-{realheapsort\_step1} & {0.09} & {no} & {} & {no} & {no
-} \\\hline
-{realselect} & {0.09} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{rsd} & {0.08} & {yes} & {1} & {yes} & {yes
-} \\\hline
-{sipma91} & {0.11} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{speedDis1} & {0.07} & {no} & {} & {no} & {no
-} \\\hline
-{speedFails1} & {0.07} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{speedFails4} & {0.07} & {no} & {} & {no} & {no
-} \\\hline
-{speedNestedMultipleDep} & {0.08} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{speedpldi3} & {0.08} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{speedSimpleMultiple} & {0.08} & {no} & {} & {no} & {no
-} \\\hline
-{speedSingleSingle} & {0.09} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{terminate} & {0.07} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{wcet0} & {0.09} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{wcet2} & {0.07} & {yes} & {2} & {yes} & {yes
-} \\\hline
+{aaron2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
+{ax} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
+{complex} & {0.08} & {yes} & {1} & {yes} & {yes} \\\hline
+{counterex1b} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
+{cousot9} & {0.08} & {yes} & {3} & {yes} & {no} \\\hline
+{easy2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
+{loops} & {0.08} & {no} & {} & {no} & {no} \\\hline
+{nestedloop} & {0.09} & {yes} & {3} & {yes} & {no} \\\hline
+{perfect1} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
+{random1d} & {0.08} & {yes} & {2} & {yes} & {yes} \\\hline
+{realbubble} & {0.18} & {yes} & {3} & {yes} & {yes} \\\hline
+{realheapsort} & {0.18} & {no} & {} & {no} & {no} \\\hline
+{realheapsort\_step1} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{realselect} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
+{rsd} & {0.08} & {yes} & {1} & {yes} & {yes} \\\hline
+{sipma91} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
+{speedDis1} & {0.07} & {no} & {} & {no} & {no} \\\hline
+{speedFails1} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
+{speedFails4} & {0.07} & {no} & {} & {no} & {no} \\\hline
+{speedNestedMultipleDep} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedpldi3} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedSimpleMultiple} & {0.08} & {no} & {} & {no} & {no} \\\hline
+{speedSingleSingle} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{terminate} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
+{wcet0} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{wcet2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
 {wise} & {0.07} & {no} & {} & {no} & {no} \\\hline 

### 2017-07-06 paper/experiments.tex
changes to experiments
table cleaned

@@
-\section{Experimental Results}
+\section{Experimental Results}\label{sec:experiments}
 We have implemented our algorithm (in C++) and present two sets of experimental results.
@@
 
-   {Benchmark} & {Time (s)} & {Solution} & {Dimension} & {Prob. loops} & {Prob. Assignments
-} \\\hline
-{alain} & {0.09} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{catmouse} & {0.06} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{counterex1a} & {0.08} & {no} & {} & {no} & {no
-} \\\hline
-{counterex1c} & {0.09} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{easy1} & {0.07} & {yes} & {1} & {yes} & {yes
-} \\\hline
-{exmini} & {0.07} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{insertsort} & {0.08} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{ndecr} & {0.07} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{perfect} & {0.09} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{perffect2} & {0.1} & {yes} & {3} & {yes} & {no
-} \\\hline
-{real2} & {0.07} & {no} & {} & {no} & {no
-} \\\hline
-{realheapsort\_step2} & {0.07} & {yes} & {1} & {no} & {no
-} \\\hline
-{realshellsort} & {0.07} & {yes} & {1} & {no} & {no
-} \\\hline
-{serpent} & {0.08} & {yes} & {1} & {yes} & {yes
-} \\\hline
-{sipmabubble} & {0.08} & {yes} & {3} & {yes} & {yes
-} \\\hline
+   {Benchmark} & {Time (s)} & {Solution} & {Dimension} & {Prob. loops} & {Prob. Assignments} \\\hline
+{alain} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{catmouse} & {0.06} & {yes} & {2} & {yes} & {yes} \\\hline
+{counterex1a} & {0.08} & {no} & {} & {no} & {no} \\\hline
+{counterex1c} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
+{easy1} & {0.07} & {yes} & {1} & {yes} & {yes} \\\hline
+{exmini} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
+{insertsort} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
+{ndecr} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
+{perfect} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
+{perffect2} & {0.1} & {yes} & {3} & {yes} & {no} \\\hline
+{real2} & {0.07} & {no} & {} & {no} & {no} \\\hline
+{realheapsort\_step2} & {0.07} & {yes} & {1} & {no} & {no} \\\hline
+{realshellsort} & {0.07} & {yes} & {1} & {no} & {no} \\\hline
+{serpent} & {0.08} & {yes} & {1} & {yes} & {yes} \\\hline
+{sipmabubble} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
 {speedDis2} & {0.07} & {no} & {} & {no} & {no
 } \\\hline
-{speedFails2} & {0.06} & {yes} & {2} & {yes} & {no
-} \\\hline
-{speedNestedMultiple} & {0.08} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{speedpldi2} & {0.07} & {yes} & {2} & {yes} & {yes
-} \\\hline
-{speedpldi4} & {0.07} & {yes} & {3} & {yes} & {yes
-} \\\hline
-{speedSimpleMultipleDep} & {0.07} & {no} & {} & {no} & {no
-} \\\hline
-{speedSingleSingle2} & {0.08} & {yes} & {2} & {yes} & {no
-} \\\hline
-{unperfect} & {0.08} & {yes} & {2} & {yes} & {no
-} \\\hline
-{wcet1} & {0.09} & {yes} & {2} & {yes} & {yes
-} \\\hline
+{speedFails2} & {0.06} & {yes} & {2} & {yes} & {no} \\\hline
+{speedNestedMultiple} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedpldi2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
+{speedpldi4} & {0.07} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedSimpleMultipleDep} & {0.07} & {no} & {} & {no} & {no} \\\hline
+{speedSingleSingle2} & {0.08} & {yes} & {2} & {yes} & {no} \\\hline
+{unperfect} & {0.08} & {yes} & {2} & {yes} & {no} \\\hline
+{wcet1} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
 {while2} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
@@
 
-{Variables} & {LOC} & {Inv Time (s)} & {Our Time (s)} & {Total Time (s)} \\\hline
-{2} & {20} & {0.050189} & {0.028366} & {0.078555} \\\hline
-{3} & {32} & {0.060078} & {0.02359} & {0.083668} \\\hline
-{4} & {56} & {0.074388} & {0.030035} & {0.104423} \\\hline
-{5} & {104} & {0.131122} & {0.051545} & {0.182667} \\\hline
-{6} & {200} & {0.354702} & {0.097943} & {0.452645} \\\hline
-{7} & {392} & {1.30833} & {0.29947} & {1.6078} \\\hline
-{8} & {776} & {7.55238} & {0.69566} & {8.24804} \\\hline
-{9} & {1544} & {33.0638} & {2.4977} & {35.5615} \\\hline
-{10} & {3080} & {164.088} & {8.763} & {172.851} \\\hline
-{11} & {6152} & {817.914} & {35.37} & {853.284} \\\hline
+{Variables} & {LOC} & {Inv Time (s)} & {Our Time (s)} & {Total Time (s)
+} \\\hline
+{0} & {8} & {0.09} & {0.02} & {0.11} \\\hline
+{1} & {14} & {0.05} & {0.02} & {0.07} \\\hline
+{2} & {20} & {0.06} & {0.03} & {0.08} \\\hline
+{3} & {32} & {0.07} & {0.03} & {0.09} \\\hline
+{4} & {56} & {0.08} & {0.04} & {0.11} \\\hline
+{5} & {104} & {0.14} & {0.06} & {0.19} \\\hline
+{6} & {200} & {0.36} & {0.1} & {0.46} \\\hline
+{7} & {392} & {1.31} & {0.3} & {1.61} \\\hline
+{8} & {776} & {7.56} & {0.7} & {8.25} \\\hline
+{9} & {1544} & {33.07} & {2.5} & {35.57} \\\hline
+{10} & {3080} & {164.09} & {8.77} & {172.86} \\\hline
+{11} & {6152} & {817.92} & {35.37} & {853.29} \\\hline
 {12} & {12296} & {4260.96} & {145.18} & {4406.14} \\\hline

### 2017-07-06 paper/synthesis.tex
small changes

@@
 \end{compactenum}
-  Moreover, if such an LEM $\lem$ exists, the algorithm has to find one that maximizes the number of gen.
-transitions in $U$ ranked by it.
-Both these tasks can be accomplished by the standard  method of linear constraints based on the use of Farkas's lemma, which was widely use for synthesis of termination proofs in both probabilistic and non-probabilistic programs~\cite{DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,SriramCAV,CFNH16:prob-termination}.
-That is, the algorithm first constructs, for each location $\loc$ a \emph{template} for $\lem$, i.e.
-an expression of the form $a_1^{\loc}x_1 + \cdots + a_{|\pvars|}^{\loc}x_{|\pvars|} = b^{\loc} $, where $x_1,\dots,x_{|\pvars|}$ are program variables and $a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc} $ are yet unknown coefficients.
-That is, supplying concrete values for all the unknown coefficients yields an LEM.
-Now the conditions (1) and (2) above can be expressed using linear constraints on the coefficients.
-More precisely, using the construction provided e.g.
-in~\cite{xxx} (which includes a use of the Farkas's lemma) we construct in polynomial time, for each generalized transition $\tilde{\tau}$, a system of linear constraints $\linsystem_{\tilde{\tau}}$ over set of variables $\{ a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc}\mid \loc\in\locs \}\cup \{\eps_{\tilde\tau}\}\cup F$, where $F$ is the set of fresh variables (not appearing in any template) and $\eps_{\tilde\tau} $ is constrained to be non-negative.
-Each solution of the system $\linsystem_{\tilde{\tau}}$ yields a LEM which satisfies the constraints (1) and (2) for $\tilde{\tau}$.
-Moreover, each solution of $\linsystem_{\tilde{\tau}}$ yields a LEM which $\eps_{\tilde{\tau}}$-ranks $\tilde{\tau}$.
-To find a LEM which satisfies all constraints (1)--(3) as well as maximizes the number of 1-ranked elements of $U$ it is sufficient to construct $\linsystem_{\tilde{\tau}}$ for each $\tilde{\tau}\in U$ and solve the following linear program $\lp_{U}$:
+  Moreover, if such an LEM $\lem$ exists, the algorithm has to find one that 
+  maximizes the number of gen. transitions in $U$ ranked by it. Both these 
+  tasks can be accomplished by the standard  method of linear constraints based 
+  on the use of Farkas's lemma, which was widely use for synthesis of 
+  termination proofs in both probabilistic and non-probabilistic 
+  programs~\cite{DBLP:conf/tacas/ColonS01,DBLP:conf/vmcai/PodelskiR04,SriramCAV,CFNH16:prob-termination}.
+   That is, the algorithm first constructs, for each location $\loc$ a 
+  \emph{template} for $\lem$, i.e. an expression of the form $a_1^{\loc}x_1 + 
+  \cdots + a_{|\pvars|}^{\loc}x_{|\pvars|} = b^{\loc} $, where 
+  $x_1,\dots,x_{|\pvars|}$ are program variables and 
+  $a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc} $ are yet unknown 
+  coefficients. That is, supplying concrete values for all the unknown 
+  coefficients yields an LEM. Now the conditions (1) and (2) above can be 
+  expressed using linear constraints on the coefficients. More precisely, using 
+  the construction provided e.g. in~\cite{SriramCAV,CFNH16:prob-termination} 
+  (which includes a use of the Farkas's lemma) we construct in polynomial time, 
+  for each generalized transition $\tilde{\tau}$, a system of linear 
+  constraints $\linsystem_{\tilde{\tau}}$ over set of variables $\{ 
+  a_1^{\loc},\dots,a_{|\pvars|}^{\loc},b^{\loc}\mid \loc\in\locs \}\cup 
+  \{\eps_{\tilde\tau}\}\cup F$, where $F$ is the set of fresh variables (not 
+  appearing in any template) and $\eps_{\tilde\tau} $ is constrained to be 
+  non-negative. Each solution of the system $\linsystem_{\tilde{\tau}}$ yields 
+  a LEM which satisfies the constraints (1) and (2) for $\tilde{\tau}$. 
+  Moreover, each solution of $\linsystem_{\tilde{\tau}}$ yields a LEM which 
+  $\eps_{\tilde{\tau}}$-ranks $\tilde{\tau}$. To find a LEM which satisfies all 
+  constraints (1)--(3) as well as maximizes the number of 1-ranked elements of 
+  $U$ it is sufficient to construct $\linsystem_{\tilde{\tau}}$ for each 
+  $\tilde{\tau}\in U$ and solve the following linear program $\lp_{U}$:
   \begin{align*}
@@
 
-Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} follow by arguments identical to those presented in~\cite{xxx}. 
+Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} 
+follow by arguments identical to those presented 
+in~\cite{ADFG10:lexicographic}. 
 \begin{theorem}

### 2017-07-06 paper/experiments.tex
Computer specs

@@
 We have implemented our algorithm (in C++) and present two sets of experimental results.
-For all the experimental results we use Aspic~\cite{} for invariant generation, 
-and our algorithm requires linear-programming solution for which we use CPlex~\cite{}.
-Our experimental results were obtained on [SPEC OF COMPUTER].
+For all the experimental results we use the tool Aspic~\cite{} for invariant generation, 
+and our algorithm requires linear-programming solver for which we use CPlex~\cite{}.
+All our experimental results were obtained on the following platform: 
+Ubuntu16.04, 7.7GB, Intel-Core i3-4130 CPU 3.40GHz QuadCore 64-bit.
+{\bf KRISH: TWO CITATIONS.}
 
 \smallskip\noindent{\em Abstraction of real-world programs.}
-We consider benchmarks considered for real-world non-probabilistic programs 
+We consider benchmarks from abstraction of real-world non-probabilistic programs 
 in~\cite{ADFG10:lexicographic}.
-Note that there are no abstraction tools available for probabilistic programs (to the
-best of our knowledge). 
+Note that there are no abstraction tools available for probabilistic programs 
+(to the best of our knowledge). 
 Hence we consider the abstract programs obtained from real-world benchmarks as 
@@
 (a)~{\em probabilistic loops} where the existing while loops are made probabilistic 
-by executing the statements with probability~1/2, and with remaining probabilities 
-executing skip statements;
+by executing the existing statements with probability~1/2, and with remaining 
+probability executing skip statements; and
 (b)~{\em probabilistic assignments} where the existing assignments are changed to 
 probabilistic assignments with variables generated uniformly in range $[-1,1]$ 
-(i.e., we consider a variable $r$ whose value is generated by probabilistic 
-assignment uniformly in the range $[-1,1]$ and add the variable to the existing 
-assignment).
-We report our results on twentfy five benchmarks in Table~\ref{tab:exp1} (since we consider around 
-fifty benchmark examples the results on the remaining ones are presented in Appendix, Section~\ref{sec:app_ex}).
-We show the time taken by our approach, which is always less than $1/10$-th of second.
-We show whether solution exists or not (i.e., whether linear lexicographic RSMs exist
+(i.e., we consider an additional variable $r$ whose value is generated by 
+probabilistic assignment uniformly in the range $[-1,1]$ and we add the variable 
+to the existing assignment).
+We report our results on twentfy five benchmarks in Table~\ref{tab:exp1} (we 
+consider around fifty benchmark examples and the results on the remaining ones 
+are presented in Appendix, Section~\ref{sec:app_ex}).
+The experimental results show that the time taken by our approach is always less 
+than $1/10$-th of second.
+In the table, along with the benchmark name, and time in seconds, we show 
+whether a solution exists or not (i.e., whether linear lexicographic RSMs exist
 or not), and if the solution exists we present the dimension of the lexicographic RSM
 we obtain. In all cases the dimension is quite small (at most~3).
-We also show whether we enhance the program with probabilistic loops and probabilistic assignments.
+The final two columns of the table represent whether the non-probabilistic 
+program is extended with probabilistic loops and/or probabilistic assignments.
 
-\smallskip\noindent{\em Constructed large programs.} 
+\smallskip\noindent{\em Synthetic examples of large programs.} 
 The programs obtained as abstractions of real-world programs in the benchmarks 
 as mentioned above have between 10-100 lines of code. 
-To test the how does our approach scale with larger codes we consider large
-probabilistic programs generated as follows.
+To test the how does our approach scale with larger codes we consider synthetic examples 
+of large probabilistic programs generated as follows.
 Given $n$ Boolean variables, we consider probabilistic while loops,
-some nondeterministic conditional branches, and all possible $2^n$ if conditions
+with some nondeterministic conditional branches, and all possible $2^n$ if conditions
 based on the Boolean variables. 
-Hence given $n$ variables we have probabilistic programs of size $O(2^n)$, and 
-we first ran an invariant generation tool, followed by our algorithm.
+Hence given $n$ variables we have probabilistic programs of size $O(2^n)$.
+For such programs we first run an invariant generation tool, followed by our algorithm.
 In all these examples lexicographic RSMs exist, and has dimension at most~3.
@@
 and our algorithm requires around two minutes.
-The results are presented in Table~\ref{tab:exp2}.
+The results are presented in Table~\ref{tab:exp2} where we present the number of 
+variables, then lines of code, followed by the time taken for invariant generation
+by Aspic, then the time taken by our algorithm, and finally the total time.
 

### 2017-07-06 paper/compositional.tex
re-structuring sec with compositionality

@@
 
+\subsection{One-Dimensional Compositional Proofs of Almost-Sure Termination}
+
 Compositionality in the context of termination proving means providing the 
 proof of termination step-by-step, handling one loop at a time, rather than 
-attempting to construct the proof (in our case, a LexRSM) at once. In non-probabilistic setting, the 
-relationship between lexicographic ranking functions and compositional 
-termination proving is well established, as well as the importance of 
-compositional proofs for efficient implementation~\cite{xxx}. In the context of 
+attempting to construct the proof (in our case, a LexRSM) at 
+once~\cite{KSTW10:compositional-transition-invariants}. 
+%In non-probabilistic setting, the 
+%relationship between lexicographic ranking functions and compositional 
+%termination proving is well established, as well as the importance of 
+%compositional proofs for efficient implementation. 
+In the context of 
 probabilistic programs, the work~\cite{HolgerPOPL} attempted to provide a 
@@
 \emph{probabilistic variant rule (V-rule),} which we explain in a more detail 
-below.
-The reason for this attempt was to provide a method of proving a.s.
-termination for programs where 1-dimensional ranking supermartingales are not sufficient.
-Since our LexRSMs were conceived with a similar motivation, we devote this section to a thorough comparison of our approach with the one of~\cite{HolgerPOPL}.
+below. However, for the method to work,~\cite{HolgerPOPL} imposes a 
+technical~\emph{uniform integrability} condition, whose checking is hard to 
+automatize. In this 
+section we show that using our insights into LexRSMs we 
+can obtain a different notion of a probabilistic V-rule which is sound 
+without any additional assumptions, and which can be used to compositionally 
+prove termination 
+of programs that the previous method cannot handle.
 
-There are two main features of LexRSMs compared with the previous approach, which we call a PVSM (probabilistci V-rule supermartingale) approach.
-First, the PVSM approach is only applicable to programs in which termination of each loop can be witnessed by a suitable 1-dimensional ranking supermartingale.
-But there are already single-loop programs for which no 1-dimensional ranking supermartingale exists but a LexRSM does, see the following example.
+%The reason for this attempt was to provide a method of 
+%proving a.s. termination for programs where 1-dimensional ranking 
+%supermartingales are not sufficient. Since our LexRSMs were conceived with a 
+%similar motivation, we devote this section to a thorough comparison of our 
+%approach with the one of~\cite{HolgerPOPL}.
 
-\textbf{[PETR: ADD EXAMPLE]}
-
-The need for multi-dimensional ranking arises within loops that might go through several ``computation phases'' before terminating, and loops containing non-deterministic choices.
-
-Second, as already noted in~\cite{HolgerPOPL}, the probabilistic V-rule is not 
-sound in general. In order to rectify this issue, in~\cite{HolgerPOPL} they 
-impose an additional constraint of \emph{uniform integrability} on ranking 
-supermartingales that prove termination of individual loops, 
-under which the probabilistic V-rule is shown to be sound. Apart from uniform 
-integrability being somewhat restrictive in itself, in~\cite{HolgerPOPL} it is 
-argued that proving uniform integrability is beyond the capability of 
-state-of-the-art automated theorem provers. As a substitute for 
-these,~\cite{HolgerPOPL} introduces a type system that can be used to 
-automatically prove 
-uniform integrability of ranking supermartingales for a restricted class of 
-programs. In particular, the method cannot handle programs in which termination 
-is controlled by a variable that can be modified by a non-constant value in some sub-program, e.g. 
-variable $x$ which appears in an assignment of the form $x:=x-y$. In this 
-section we show that using our insights into LexRSMs we 
-can obtain a subtly different notion of a probabilistic V-rule which is sound 
-without any additional assumptions, and which can be used to compositionally prove termination 
-of programs that the previous method cannot handle.
+%There are two main features of LexRSMs compared with the previous approach, 
+%which we call a PVSM (probabilistic V-rule supermartingale) approach. First, 
+%the PVSM approach is only applicable to programs in which termination of each 
+%loop can be witnessed by a suitable 1-dimensional ranking supermartingale. But 
+%there are already single-loop programs for which no 1-dimensional ranking 
+%supermartingale exists but a LexRSM does, see the following example.
+%
+%\textbf{[PETR: ADD EXAMPLE]}
+%
+%%The need for multi-dimensional ranking arises within loops that might go 
+%%through several ``computation phases'' before terminating, and loops 
+%%containing 
+%%non-deterministic choices.
+%
+%Second, as already noted in~\cite{HolgerPOPL}, the probabilistic V-rule is not 
+%sound in general. In order to rectify this issue, in~\cite{HolgerPOPL} they 
+%impose an additional constraint of \emph{uniform integrability} on ranking 
+%supermartingales that prove termination of individual loops, 
+%under which the probabilistic V-rule is shown to be sound. Apart from uniform 
+%integrability being somewhat restrictive in itself, in~\cite{HolgerPOPL} it is 
+%argued that proving uniform integrability is beyond the capability of 
+%state-of-the-art automated theorem provers. As a substitute for 
+%these,~\cite{HolgerPOPL} introduces a type system that can be used to 
+%automatically prove 
+%uniform integrability of ranking supermartingales for a restricted class of 
+%programs. In particular, the method cannot handle programs in which 
+%termination 
+%is controlled by a variable that can be modified by a non-constant value in 
+%some sub-program, e.g. 
+%variable $x$ which appears in an assignment of the form $x:=x-y$.
 
@@
 
-Given an invariant map $\inv$, we say that a 1-dimensional measurable map $\lem$ for $\pCFG_{\program}$ is $\eps$-$\inv$-ranking in location $\loc$, if for each $\vec{x}\in\inv(\loc)$  it holds $\preexp{\lem}(\loc,\vec{x})\leq \lem(\loc,\vec{x})-\eps.$ The map $\lem$ is $\inv$-\emph{unaffected} in location $\loc$ if it is $0$-ranking in $\loc$, and $\inv$-non-negative in $\loc$ if $\lem(\loc,\vec{x})\geq 0$ for all $\vec{x} \in \inv(\loc)$.
-A measurable map $\lem$ is propositionally linear, if each function $\lem(\loc)$ is of the form $\indicator{G_1}\cdot E_1 + \cdots \indicator{G_k}\cdot E_k$, where each $\indicator{G_i}$ is an indicator function of some polyhedron and each $E_i$ is a linear expression.
+Given an invariant map $\inv$, we say that a 1-dimensional measurable map 
+$\lem$ for $\pCFG_{\program}$ is $\eps$-$\inv$-ranking in location $\loc$, if 
+for each $\vec{x}\in\inv(\loc)$  each generalized transition~$\tilde{\tau}$ 
+outgoing from $\loc$ is $\eps$-ranked by $\lem$.
 
-We recall the notion of compositional ranking supermartingale as introduced in~\cite{HolgerPOPL}.
-We call it a PV supermartingale, as it is based on so called probabilistic variant rule.
-Due to differences in syntax and semantics, the definition is syntactically slightly different from~\cite{HolgerPOPL}, but the essence is the same.
+We recall the notion of compositional ranking supermartingale as introduced 
+in~\cite{HolgerPOPL}. We call it a PV supermartingale, as it is based on so 
+called probabilistic variant rule. Due to differences in syntax and semantics, 
+the definition is syntactically slightly different from~\cite{HolgerPOPL}, but 
+the essence is the same. A measurable map $\lem$ is propositionally linear, if 
+each function $\lem(\loc)$ is of the form $\indicator{G_1}\cdot E_1 + \cdots 
+\indicator{G_k}\cdot E_k$, where each $\indicator{G_i}$ is an indicator 
+function of some polyhedron and each $E_i$ is a linear expression.
 
@@
 
-In~\cite{HolgerPOPL} they show that even if all nested loops were already proved to terminate a.s.
-and there is a PVSM for the program, then the program itself might not terminate a.s.
-Then they investigate conditions under which existence of a PVSM together with a proof of a.s.
-termination of each nested sub-loop of $\program$ entails termination of the whole program $\program$.
-The condition were formulated in terms of a certain type system for \PP{}s, so we say that a PVSM satisfying the condition \emph{typechecks} correctly.
+In~\cite{HolgerPOPL} they show that even if all nested loops were already 
+proved to terminate a.s. and there is a PVSM for the program, then the program 
+itself might not terminate a.s. Then they impose a \emph{uniform integrability} 
+constraint on the PVSM under which a 
+PVSM together with a proof of a.s. termination of each nested sub-loop of 
+$\program$ entails termination of the whole program $\program$. 
+%they 
+%impose an additional constraint of \emph{uniform integrability} on ranking 
+%supermartingales that prove termination of individual loops, 
+%under which the probabilistic V-rule is shown to be sound. 
+Uniform integrability is a deep 
+concept from probability and measure theory: a sequence $X_0,X_1,X_2,\dots$ of 
+random variables is uniformly integrable if for each $\delta>0$ there exists an 
+$K\in \Nset$ such that for all $n\geq 0$ it holds $\E[|X_n|\cdot\indicator{X_n 
+	\geq K}]\leq \delta$.
+Apart from uniform 
+integrability being somewhat restrictive in itself, in~\cite{HolgerPOPL} it is 
+argued that proving uniform integrability is beyond the capability of 
+state-of-the-art automated theorem provers. As a substitute for 
+these,~\cite{HolgerPOPL} introduces a type system that can be used to 
+automatically prove 
+uniform integrability of ranking supermartingales for a restricted class of 
+programs.
+We do not repeat the precise definition of the typesystem here, we just say 
+that a PVSM satisfying the condition imposed by the type system 
+\emph{typechecks} correctly. 
 In~\cite{HolgerPOPL} the following was proved:
@@
 Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
-\programbody \textbf{ od}$.
-Assume that each nested loop of $\program$ terminates almost surely from each reachable configuration, and that there exists a PVSM for $\program$ that typechecks correctly.
-Then $\program$ terminates almost surely.
+\programbody \textbf{ od}$. Assume that each nested loop of $\program$ 
+terminates almost surely from each reachable configuration, and that there 
+exists a PVSM for $\program$ that typechecks correctly. Then $\program$ 
+terminates almost surely.
 \end{theorem}
 
-We do not repeat the precise definition of the typesystem here, we just explain its motivation and give examples where it fails, i.e.
-examples where $\program$ terminates a.s.
-(and hence the same holds also for its nested loops) and a PV martingale exists, but no PVSM typechecks.
+%, we just explain its motivation and give examples where it fails, i.e. 
+%examples where $\program$ terminates a.s. (and hence the same holds also for 
+%its nested loops) and a PV martingale exists, but no PVSM typechecks.
 
-The purpose of the typesystem is, in principle, to ensure that the stochastic 
-process which, in each time step, returns the current value of the PVSM, is 
-uniformly integrable inside the nested loops. Uniform integrability is a deep 
-concept from probability and measure theory: a sequence $X_0,X_1,X_2,\dots$ of 
-random variables is uniformly integrable if for each $\delta>0$ there exists an 
-$K\in \Nset$ such that for all $n\geq 0$ it holds $\E[|X_n|\cdot\indicator{X_n 
-\geq K}]\leq \delta$. As noted in~\cite{HolgerPOPL}, proofs of uniform 
-integrability are hard to (semi)-automatize, and hence they devise the 
-typechecking as a sound (but not complete) method of proving uniform 
-integrability. This is illustrated in the following example.
+%The purpose of the typesystem is, in principle, to ensure that the stochastic 
+%process which, in each time step, returns the current value of the PVSM, is 
+%uniformly integrable inside the nested loops.  As noted in~\cite{HolgerPOPL}, 
+%proofs of uniform 
+%integrability are hard to (semi)-automatize, and hence they devise the 
+%typechecking as a sound (but not complete) method of proving uniform 
+%integrability. This is illustrated in the following example.
+The intricacies of uniform integrability are shown in the following example.
 
 \begin{example}
+\label{ex:uniform}
+\label{EX:UNIFORM}
 Consider the two \APP{}s in Figure~\ref{fig:uniint}, that differ only in one 
@@
 However, the variable $x$ is uniformly integrable within the inner loop of the 
-right program while for the left program this does not hold. To see this, 
-imagine the inner loop as a stand-alone program and let $X_n$ be the value of 
-variable $x$ after $n$ steps of this stand-alone program (i.e., when the loop 
-terminates $x$ no longer changes). Solving a simple linear recurrence shows 
-that in the right program $\E[X_n] \rightarrow 0$ as $n\rightarrow \infty$, 
-which in particular shows uniform integrability of $X_0,X_1,X_2,\dots$. On the 
-other hand, in the left program for each $K>0$ we have $\probm(X_K \geq 
-2^K\cdot{x_0})\geq \frac{1}{2^K}$, where $x_0$ is the value of $x$ upon 
-entering the loop. Hence $\E[|X_K|cdot\indicator{X_K 
-\geq K}]\geq 1$ for each $K$ sufficiently large, which is incompatible with 
-uniform integrability. The example shows that proving uniform integrability 
+right program while for the left program this does not hold: we show this 
+in~\AppendixMaterial. The example shows that proving uniform integrability 
 requires intricate reasoning about quantitative behaviour of the program. 
@@
 Let $\program$ be a \PP{} of the form $\textbf{while } \Psi \textbf{ do } 
-\programbody \textbf{ od}$.
-Assume that each nested loop of $\program$ terminates almost surely from each reachable configuration, and that there exists a NCSM for $\program$ supported by some invariant map.
-Then $\program$ terminates almost surely.
+\programbody \textbf{ od}$. Assume that each nested loop of $\program$ 
+terminates almost surely from each reachable configuration, and that there 
+exists a NCSM for $\program$ supported by some invariant map. Then $\program$ 
+terminates almost surely from each initial configuration.
 \end{theorem}
-\begin{proof}
-\textbf{[PETR: TBD]}
+\begin{proof}[Proof (Key Idea)]
+Assume that there is set of non-terminating runs $A$ of positive probability.
+Let $\{X_i\}_{i=0}^{\infty}$ be a stochastic process returning the value of 
+NCSM $\lem$ in step $i$. We show that for each $i$ it holds $\E[X_i] 
+\leq \lem(\locinit,\vecinit) - \eps\cdot \E[\sharp_i]$, where $\sharp_i$ is the 
+random variable returning the number of times a location from 
+$\slice(\program)$ is visited. Since all sub-loops of $\program$ terminate 
+a.s., $\lim_{i\rightarrow \infty}\sharp_i(\run)=\infty$ for $\run\in A$. Since 
+$A$ has positive probability $\lim_{i\rightarrow 
+\infty}\E[\sharp_i(\run)=\infty]$, it holds $\lim_{i\rightarrow 
+\infty}\E[X_i] = -\infty$, a contradiction with non-negativity of $\lem$.
 \end{proof}
@@
 \APP{}s, synthesizing NCSM entails synthesizing sufficient program invariants 
-(for which there is a good automated tool support~\cite{xxx}) that 
-overapproximate set of reachable configurations and encoding the rankingn, 
+(for which there is a good automated tool support~\cite{FG10:aspic})  encoding 
+the ranking, 
 unaffection, and non-negativity conditions into a collection of linear 
-constraints (as for general LinLexRSMs in Section~\ref{xxx}). 
+constraints (as for general LinLexRSMs in Section~\ref{sec:algo}). 
 Figures~\ref{fig:uniint} and~\ref{fig:uniint2} show instances where attempts to 

### 2017-07-06 paper/experiments.tex
re-structuring sec with compositionality

@@
 We have implemented our algorithm (in C++) and present two sets of experimental results.
-For all the experimental results we use the tool Aspic~\cite{} for invariant generation, 
-and our algorithm requires linear-programming solver for which we use CPlex~\cite{}.
+For all the experimental results we use the tool Aspic~\cite{FG10:aspic} for 
+invariant generation, 
+and our algorithm requires linear-programming solver for which we use 
+CPlex~\cite{cplex}.
 All our experimental results were obtained on the following platform: 
 Ubuntu16.04, 7.7GB, Intel-Core i3-4130 CPU 3.40GHz QuadCore 64-bit.
-{\bf KRISH: TWO CITATIONS.}
+%{\bf KRISH: TWO CITATIONS.}
 

### 2017-07-06 paper/introduction.tex
re-structuring sec with compositionality

@@
 In applications such as stochastic network protocols~\cite{BaierBook,prism},
-randomized algorithms~\cite{RandBook,RandBook2}, machine learning~\cite{LearningSurvey,G15},
+randomized algorithms~\cite{RandBook,RandBook2}, 
+security~\cite{BGGHS16:diff-privacy-coupling,BGHP16:diff-privacy-siglog} 
+machine 
+learning~\cite{LearningSurvey,G15},
 the probabilistic behavior must be considered to faithfully model the underlying dynamic system.

### 2017-07-06 paper/prelims.tex
re-shuffled the proof of the main theorem a bit
re-structuring sec with compositionality

@@
 
-We use standard notation in the field of probabilistic program 
-analysis~\cite{POPL17PAPER}.
+We use a standard notation in the field of probabilistic program 
+analysis~\cite{CNZ17}.
 
@@
 To define this process, we first define so called 
-\emph{probabilistic control flow graphs}~\cite{HONGFEI-CAV16}.
+\emph{probabilistic control flow graphs}~\cite{CFG16}.
 
@@
 In the following, consider a \PP{} $P$ and its associated pCFG $\pCFG_P$. This 
-pCFG has a special location  $\loc_{P}^{\lout}$ corresponding to the value of 
+pCFG has a special location  $\locterm$ corresponding to the value of 
 the program counter after 

### 2017-07-06 paper/synthesis.tex
re-structuring sec with compositionality

@@
 \section{Algorithmic Aspects}
+\label{sec:algo}
 
-In this section we describe a polynomial-time algorithm for synthesizing linear $\eps$-LexRSM maps in affine probabilistic programs supported by a given linear invariant map $\inv$.
-The algorithm, based on iterative solving of linear constraints, is an adaptation of a similar algorithm for finding lexicographic ranking functions in non-probabilistic programs~\cite{ADFG10:lexicographic}.
-Hence, we provide only a high-level description, focusing on new aspects of our adaptation.
+In this section we describe a polynomial-time algorithm for synthesizing linear 
+$\eps$-LexRSM maps in affine probabilistic programs supported by a given linear 
+invariant map $\inv$. The algorithm, based on iterative solving of linear 
+constraints, is a generalization of an algorithm for finding 
+lexicographic ranking functions in non-probabilistic 
+programs~\cite{ADFG10:lexicographic}. Hence, we provide only a high-level 
+description, focusing on the new aspects.
 
@@
 \label{thm:algo}
+\label{THM:ALGO}
 	Suppose that Algorithm~\ref{algo:linlexrsm} is run on an \APP{} $\program$ 
@@
 	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
-	supported by $\inv$. The algorithm runs in time polynomial in size of 
-	$\program$ and $\inv$. \textbf{[PETR: POLYHEDRAL GUARDS]}
+	supported by $\inv$. If guards of all conditionals and loops in $\APP{}$ 
+	are linear assertion, then the
+	algorithm runs in time polynomial in size of 
+	$\program$ and $\inv$.
 \end{theorem}
@@
 
-Proving completeness is more intricate; as pointed out in~\cite{ADFG10:lexicographic}, one needs to show that the greedy strategy of selecting LEM that ranks maximal number of remaining transitions does not cut off some possible LinLexRSMs.
-In~\cite{ADFG10:lexicographic} the completeness of the greedy strategy is proved by using several geometric arguments that exploit the fact that the underlying programs are affine.
-The same geometric properties hold for our generalization to \APP{}s (all ranking conditions in Definitions~\ref{def:rank1} and~\ref{def:rank2} are linear in program variables), so the result is easily transferable.
+Proving completeness is more intricate; as pointed out 
+in~\cite{ADFG10:lexicographic}, one needs to show that the greedy strategy of 
+selecting LEM that ranks maximal number of remaining transitions does not cut 
+off some possible LinLexRSMs. In~\cite{ADFG10:lexicographic} the completeness 
+of the greedy strategy is proved by using several geometric arguments that 
+exploit the fact that the underlying programs are affine. The same geometric 
+properties hold for our generalization to \APP{}s (all ranking conditions in 
+Definitions~\ref{def:rank1} and~\ref{def:rank2} are linear in program 
+variables), so the result is easily transferable. The complexity argument is 
+rather standard and we present it in~\AppendixMaterial.
 \end{proof}

### 2017-07-06 paper/exp_appendix.tex
Examples added and tables changed in experimental section

@@
 
-   {Benchmark} & {Time (s)} & {Solution} & {Dimension} & {Prob. loops} & {Prob. Assignments
-} \\\hline
-{aaron2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{ax} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
-{complex} & {0.08} & {yes} & {1} & {yes} & {yes} \\\hline
-{counterex1b} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{cousot9} & {0.08} & {yes} & {3} & {yes} & {no} \\\hline
-{easy2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{loops} & {0.08} & {no} & {} & {no} & {no} \\\hline
-{nestedloop} & {0.09} & {yes} & {3} & {yes} & {no} \\\hline
-{perfect1} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{random1d} & {0.08} & {yes} & {2} & {yes} & {yes} \\\hline
-{realbubble} & {0.18} & {yes} & {3} & {yes} & {yes} \\\hline
-{realheapsort} & {0.18} & {no} & {} & {no} & {no} \\\hline
-{realheapsort\_step1} & {0.09} & {no} & {} & {no} & {no} \\\hline
-{realselect} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
-{rsd} & {0.08} & {yes} & {1} & {yes} & {yes} \\\hline
-{sipma91} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
-{speedDis1} & {0.07} & {no} & {} & {no} & {no} \\\hline
-{speedFails1} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{speedFails4} & {0.07} & {no} & {} & {no} & {no} \\\hline
-{speedNestedMultipleDep} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedpldi3} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedSimpleMultiple} & {0.08} & {no} & {} & {no} & {no} \\\hline
-{speedSingleSingle} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{terminate} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{wcet0} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{wcet2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{wise} & {0.07} & {no} & {} & {no} & {no} \\\hline 
+{Benchmark} & {Time (s)} & {Solution} & {Dimension} & {Prob. loops} & {Prob. Assignments} \\\hline \hline
+{aaron2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{ax} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
+{complex} & {0.1} & {yes} & {1} & {yes} & {yes} \\\hline
+{counterex1b} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{cousot9} & {0.09} & {no} & {3} & {yes} & {no} \\\hline
+{easy2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{loops} & {0.1} & {no} & {} & {no} & {no} \\\hline
+{\multirow{2}{*}{nestedloop}} & {0.16} & {yes} & {3} & {yes} & {no} \\\cline{2-6}
+{} & {0.23} & {no} & {} & {yes} & {yes} \\\hline
+{perfect1} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{random1d} & {0.1} & {yes} & {2} & {yes} & {yes} \\\hline
+{realbubble} & {0.22} & {yes} & {3} & {yes} & {yes} \\\hline
+{realheapsort} & {0.22} & {no} & {} & {no} & {no} \\\hline
+{realheapsort\_step1} & {0.11} & {no} & {} & {no} & {no} \\\hline
+{realheapsort\_step2} & {0.16} & {no} & {} & {yes} & {no} \\\hline
+{rsd} & {0.1} & {yes} & {1} & {yes} & {yes} \\\hline
+{sipma91} & {0.14} & {yes} & {2} & {yes} & {yes} \\\hline
+{speedDis1} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{\multirow{2}{*}{speedFails2}} & {0.14} & {yes} & {1} & {yes} & {no} \\\cline{2-6}
+{} & {0.08} & {no} & {} & {yes} & {yes} \\\hline
+{speedFails4} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{speedNestedMultipleDep} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedpldi3} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedSimpleMultiple} & {0.1} & {no} & {} & {no} & {no} \\\hline
+{speedSingleSingle} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
+{terminate} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{wcet0} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
+{wcet2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{wise} & {0.09} & {no} & {} & {no} & {no} \\\hline
     \end{tabular}

### 2017-07-06 paper/experiments.tex
Examples added and tables changed in experimental section

--- 2017-07-06 paper/experiments.tex
@@
 and our algorithm requires linear-programming solution for which we use CPlex~\cite{}.
-Our experimental results were obtained on [SPEC OF COMPUTER].
+Our experimental results were obtained on [Ubuntu16.04 LTS/7.7GiB/Intel® Core™ i3-4130 CPU @ 3.40GHz × 4 processor/64-bit].
 
@@
 
-   {Benchmark} & {Time (s)} & {Solution} & {Dimension} & {Prob. loops} & {Prob. Assignments} \\\hline
-{alain} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{catmouse} & {0.06} & {yes} & {2} & {yes} & {yes} \\\hline
-{counterex1a} & {0.08} & {no} & {} & {no} & {no} \\\hline
-{counterex1c} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
-{easy1} & {0.07} & {yes} & {1} & {yes} & {yes} \\\hline
-{exmini} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{insertsort} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{ndecr} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{perfect} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
-{perffect2} & {0.1} & {yes} & {3} & {yes} & {no} \\\hline
-{real2} & {0.07} & {no} & {} & {no} & {no} \\\hline
-{realheapsort\_step2} & {0.07} & {yes} & {1} & {no} & {no} \\\hline
-{realshellsort} & {0.07} & {yes} & {1} & {no} & {no} \\\hline
-{serpent} & {0.08} & {yes} & {1} & {yes} & {yes} \\\hline
-{sipmabubble} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedDis2} & {0.07} & {no} & {} & {no} & {no
-} \\\hline
-{speedFails2} & {0.06} & {yes} & {2} & {yes} & {no} \\\hline
-{speedNestedMultiple} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedpldi2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{speedpldi4} & {0.07} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedSimpleMultipleDep} & {0.07} & {no} & {} & {no} & {no} \\\hline
-{speedSingleSingle2} & {0.08} & {yes} & {2} & {yes} & {no} \\\hline
-{unperfect} & {0.08} & {yes} & {2} & {yes} & {no} \\\hline
-{wcet1} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{while2} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
+{Benchmark} & {Time (s)} & {Solution} & {Dimension} & {Prob. loops} & {Prob. Assignments} \\\hline \hline
+{alain} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
+{catmouse} & {0.08} & {yes} & {2} & {yes} & {yes} \\\hline
+{counterex1a} & {0.1} & {no} & {} & {no} & {no} \\\hline
+{counterex1c} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
+{easy1} & {0.09} & {yes} & {1} & {yes} & {yes} \\\hline
+{exmini} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{insertsort} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{ndecr} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{perfect} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
+{\multirow{2}{*}{perfect2}} & {0.1} & {yes} & {3} & {yes} & {no} \\\cline{2-6}
+{} & {0.11} & {no} & {} & {yes} & {yes} \\\hline
+{real2} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{realselect} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
+{realshellsort} & {0.09} & {no} & {} & {yes} & {no} \\\hline
+{serpent} & {0.1} & {yes} & {1} & {yes} & {yes} \\\hline
+{sipmabubble} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedDis2} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{speedFails1} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{speedNestedMultiple} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedpldi2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{speedpldi4} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedSimpleMultipleDep} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{\multirow{2}{*}{speedSingleSingle2}} & {0.12} & {yes} & {2} & {yes} & {no} \\\cline{2-6}
+{} & {0.1} & {no} & {} & {yes} & {yes} \\\hline
+{\multirow{2}{*}{unperfect}} & {0.1} & {yes} & {2} & {yes} & {no} \\\cline{2-6}
+{} & {0.16} & {no} & {} & {yes} & {yes} \\\hline
+{wcet1} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
+{while2} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
 
@@
 
-{Variables} & {LOC} & {Inv Time (s)} & {Our Time (s)} & {Total Time (s)
-} \\\hline
-{0} & {8} & {0.09} & {0.02} & {0.11} \\\hline
-{1} & {14} & {0.05} & {0.02} & {0.07} \\\hline
+{Variables} & {LOC} & {Inv Time (s)} & {Our Time (s)} & {Total Time (s)} \\\hline \hline
 {2} & {20} & {0.06} & {0.03} & {0.08} \\\hline

### 2017-07-06 paper/app.tex

@@
 \end{align}
-The individual steps in the above computation are justified as follows: in~\eqref{eq:lexrsm-ind-9} we use the facts that for all $\omega$'s whose level in step $k$ is $\fixn{j}$ it holds that $\noofdec_k(\omega)+1=\noofdec_{k+1}(\omega)$, and similarly, for $\omega$'s whose level in step $k$ is $>\fixn{j}$ it holds $\noofdec_k(\omega)=\noofdec_{k+1}(\omega)$.
-Moreover, for all $\omega\in D_{k,\ell}$ it holds that $\levelrank{k}{\omega}\geq \fixn{j} \Rightarrow U(\omega)\geq k+1$.
-Finally, in~\eqref{eq:lexrsm-ind-11} we use the fact that all $\omega\in D_{k+1,k+1-\fixn{i}}$ need to have level $\fixn{j}$ in step $k$, since otherwise such an $\omega$ would need to have level $\fixn{j}$ for at least $k+1-\fixn{i}$ times within steps $\{\fixn{i},\fixn{i}+1,\dots,k-1\}$, but there are $k-\fixn{i}$ such steps, a contradiction.
-This concludes the proof of~\eqref{eq:lexrsm-soundness-main}.
+The individual steps in the above computation are justified as follows: 
+in~\eqref{eq:lexrsm-ind-9} we use the facts that for all $\omega$'s whose level 
+in step $k$ is $\fixn{j}$ it holds that 
+$\noofdec_k(\omega)+1=\noofdec_{k+1}(\omega)$, and similarly, for $\omega$'s 
+whose level in step $k$ is $>\fixn{j}$ it holds 
+$\noofdec_k(\omega)=\noofdec_{k+1}(\omega)$. Moreover, for all $\omega\in 
+D_{k,\ell}$ it holds that $\levelrank{k}{\omega}\geq \fixn{j} \Rightarrow 
+U(\omega)\geq k+1$. Finally, in~\eqref{eq:lexrsm-ind-11} we use the fact that 
+all $\omega\in D_{k+1,k+1-\fixn{i}}$ need to have level $\fixn{j}$ in step $k$, 
+since otherwise such an $\omega$ would need to have level $\fixn{j}$ for at 
+least $k+1-\fixn{i}$ times within steps $\{\fixn{i},\fixn{i}+1,\dots,k-1\}$, 
+but there are $k-\fixn{i}$ such steps, a contradiction. This concludes the 
+proof of~\eqref{eq:lexrsm-soundness-main}.
+
+\section{Complexity Clarification for Theorem~\ref{THM:ALGO}}
+
+Since instances of linear programming is in~\textsc{PTIME}, it remains to show 
+that each system $\linsystem_{\tilde{\tau}}$ is constructible in polynomial 
+time. In~\cite{CFNH16:prob-termination} it shown that this can be done provided 
+that guard of each transition in pCFG is a propositionally linear predicate. 
+Now all transition guards in $\pCFG_\program$ are of the form $\phi$ or 
+$\neg\phi$, where $\phi$ is a guard of a conditional or of while-loop in 
+$\program$. If $\phi$ is a linear assertion, then $\neg\phi$ can be converted 
+into a propositionally linear predicate in polynomial time, after which the 
+construction of~\cite{CFNH16:prob-termination} can be used.
+
+\section{Additional Computation for Example~\ref{EX:UNIFORM}}
+
+ To see that $x$ is uniformly integrable in the left program and not in the 
+ right one (within the inner loop), 
+ imagine the inner loop as a stand-alone program and let $X_n$ be the value of 
+ variable $x$ after $n$ steps of this stand-alone program (i.e., when the loop 
+ terminates $x$ no longer changes). Solving a simple linear recurrence shows 
+ that in the right program $\E[X_n] \rightarrow 0$ as $n\rightarrow \infty$, 
+ which in particular shows uniform integrability of $X_0,X_1,X_2,\dots$. On the 
+ other hand, in the left program for each $K>0$ we have $\probm(X_K \geq 
+ 2^K\cdot{x_0})\geq \frac{1}{2^K}$, where $x_0$ is the value of $x$ upon 
+ entering the loop. Hence $\E[|X_K|cdot\indicator{X_K 
+ 	\geq K}]\geq 1$ for each $K$ sufficiently large, which is incompatible with 
+ uniform integrability.

### 2017-07-06 paper/conclu.tex
small changes

@@
 For probabilistic programs with nondeterminism, but restricted to discrete probabilistic
-choices, the termination problem was studied in~\cite{MM04,MM05}.
+choices, the termination problem was studied 
+in~\cite{MM04,MM05}.
 The RSM-based (ranking supermatingale-based) approach extending ranking functions was first 
@@
 but with real-valued variables, and its extension for probabilistic programs
-with non-determinism has been studied in~\cite{HolgerPOPL,CF17,CFNH16,CFG16,CNZ17}.
+with non-determinism has been studied 
+in~\cite{HolgerPOPL,CF17,CFNH16,CFG16,CNZ17,MM16:proofrule-arxiv}. 
+Supermartingales were also 
+considered for other liveness and safety 
+properties~\cite{CVS16:martingale-recurrence-persistence,BEFH16:doob}.
 While all these results deeply clarify the role of RSMs for probabilistic programs, 
@@
 
-\smallskip\noindent{\em Compositional a.s. termination proving.} A 
+\smallskip\noindent{\em Compositional a.s. termination proving.} 
+A 
 compositional rule for proving almost-sure termination was studied 

### 2017-07-06 paper/conclu.tex
some changes of bibliography

@@
 with non-determinism has been studied 
-in~\cite{HolgerPOPL,CF17,CFNH16,CFG16,CNZ17,MM16:proofrule-arxiv}. 
+in~\cite{HolgerPOPL,CF17,CFNH16:prob-termination,CFG16,CNZ17,MM16:proofrule-arxiv}. 
 Supermartingales were also 

### 2017-07-06 paper/experiments.tex
some changes of bibliography

@@
 We have implemented our algorithm (in C++) and present two sets of experimental results.
-For all the experimental results we use the tool Aspic~\cite{FG10:aspic} for 
+For all the experimental results we use the tool \textsc{Aspic}~\cite{FG10:aspic} for 
 invariant generation, 
 and our algorithm requires linear-programming solver for which we use 
-CPlex~\cite{cplex}.
+\textsc{CPlex}~\cite{cplex}.
 All our experimental results were obtained on the following platform: 

### 2017-07-06 paper/introduction.tex
some changes of bibliography

@@
 analysis of probabilistic programs, and algorithmic approaches for special cases such as 
-linear and polynomial RSMs have also been considered~\cite{SriramCAV,CFNH16,CFG16,CNZ17}.
+linear and polynomial RSMs have also been considered~\cite{SriramCAV,CFNH16:prob-termination,CFG16,CNZ17}.
 
@@
 While an impressive set of theoretical results related to RSMs has been
-established~\cite{SriramCAV,HolgerPOPL,CF17,CFNH16,CFG16,CNZ17}, 
+established~\cite{SriramCAV,HolgerPOPL,CF17,CFNH16:prob-termination,CFG16,CNZ17}, 
 for probabilistic programs with nondeterminism the current approaches  

### 2017-07-06 paper/exp_appendix.tex
small changes in experimental section

--- 2017-07-06 paper/exp_appendix.tex
@@
 
-   {Benchmark} & {Time (s)} & {Solution} & {Dimension} & {Prob. loops} & {Prob. Assignments
-} \\\hline
-{aaron2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{ax} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
-{complex} & {0.08} & {yes} & {1} & {yes} & {yes} \\\hline
-{counterex1b} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{cousot9} & {0.08} & {yes} & {3} & {yes} & {no} \\\hline
-{easy2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{loops} & {0.08} & {no} & {} & {no} & {no} \\\hline
-{nestedloop} & {0.09} & {yes} & {3} & {yes} & {no} \\\hline
-{perfect1} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{random1d} & {0.08} & {yes} & {2} & {yes} & {yes} \\\hline
-{realbubble} & {0.18} & {yes} & {3} & {yes} & {yes} \\\hline
-{realheapsort} & {0.18} & {no} & {} & {no} & {no} \\\hline
-{realheapsort\_step1} & {0.09} & {no} & {} & {no} & {no} \\\hline
-{realselect} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
-{rsd} & {0.08} & {yes} & {1} & {yes} & {yes} \\\hline
-{sipma91} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
-{speedDis1} & {0.07} & {no} & {} & {no} & {no} \\\hline
-{speedFails1} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{speedFails4} & {0.07} & {no} & {} & {no} & {no} \\\hline
-{speedNestedMultipleDep} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedpldi3} & {0.08} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedSimpleMultiple} & {0.08} & {no} & {} & {no} & {no} \\\hline
-{speedSingleSingle} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{terminate} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{wcet0} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{wcet2} & {0.07} & {yes} & {2} & {yes} & {yes} \\\hline
-{wise} & {0.07} & {no} & {} & {no} & {no} \\\hline 
+{Benchmark} & {Time (s)} & {Solution} & {Dimension} & {Prob. loops} & {Prob. Assignments} \\\hline \hline
+{aaron2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{ax} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
+{complex} & {0.1} & {yes} & {1} & {yes} & {yes} \\\hline
+{counterex1b} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{cousot9} & {0.09} & {no} & {3} & {yes} & {no} \\\hline
+{easy2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{loops} & {0.1} & {no} & {} & {no} & {no} \\\hline
+{\multirow{2}{*}{nestedloop}} & {0.16} & {yes} & {3} & {yes} & {no} \\\cline{2-6}
+{} & {0.23} & {no} & {} & {yes} & {yes} \\\hline
+{perfect1} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{random1d} & {0.1} & {yes} & {2} & {yes} & {yes} \\\hline
+%{realbubble} & {0.22} & {yes} & {3} & {yes} & {yes} \\\hline
+{realheapsort} & {0.22} & {no} & {} & {no} & {no} \\\hline
+{realheapsort\_step1} & {0.11} & {no} & {} & {no} & {no} \\\hline
+{realheapsort\_step2} & {0.16} & {no} & {} & {yes} & {no} \\\hline
+{rsd} & {0.1} & {yes} & {1} & {yes} & {yes} \\\hline
+{sipma91} & {0.14} & {yes} & {2} & {yes} & {yes} \\\hline
+{speedFails1} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{speedDis1} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{\multirow{2}{*}{speedFails2}} & {0.14} & {yes} & {1} & {yes} & {no} \\\cline{2-6}
+{} & {0.08} & {no} & {} & {yes} & {yes} \\\hline
+{speedFails4} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{speedNestedMultipleDep} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedpldi3} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedSimpleMultiple} & {0.1} & {no} & {} & {no} & {no} \\\hline
+{speedSingleSingle} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
+{terminate} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{wcet0} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
+{wcet2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+{wise} & {0.09} & {no} & {} & {no} & {no} \\\hline
     \end{tabular}

### 2017-07-06 paper/experiments.tex
small changes in experimental section

@@
 probability executing skip statements; and
-(b)~{\em probabilistic assignments} where the existing assignments are changed to 
-probabilistic assignments with variables generated uniformly in range $[-1,1]$ 
-(i.e., we consider an additional variable $r$ whose value is generated by 
-probabilistic assignment uniformly in the range $[-1,1]$ and we add the variable 
-to the existing assignment).
-We report our results on twentfy five benchmarks in Table~\ref{tab:exp1} (we 
+(b)~{\em probabilistic assignments} where the existing assignments are perturbed uniformly in range $[-1,1]$ 
+(i.e., we consider additional variables whose value is, in each loop iteration, generated by 
+probabilistic assignment uniformly in the range $[-1,1]$, and we add such variable 
+to the RHS of an existing assignment).
+We report our results on twenty five benchmarks in Table~\ref{tab:exp1} (we 
 consider around fifty benchmark examples and the results on the remaining ones 
-are presented in Appendix, Section~\ref{sec:app_ex}).
+are presented in \AppendixMaterial.
 The experimental results show that the time taken by our approach is always less 
@@
 or not), and if the solution exists we present the dimension of the lexicographic RSM
-we obtain. In all cases the dimension is quite small (at most~3).
+we obtain. 
 The final two columns of the table represent whether the non-probabilistic 
@@
 of large probabilistic programs generated as follows.
-Given $n$ Boolean variables, we consider probabilistic while loops,
-with some nondeterministic conditional branches, and all possible $2^n$ if conditions
+Given $n$ additional Boolean-like variables, we consider probabilistic while loops,
+with some nondeterministic conditional branches, and generate all possible $2^n$ if conditions
 based on the Boolean variables. 
@@
 {real2} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{realbubble} & {0.22} & {yes} & {3} & {yes} & {yes} \\\hline
 {realselect} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
@@
 {speedDis2} & {0.09} & {no} & {} & {no} & {no} \\\hline
-{speedFails1} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
+%{speedFails1} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
 {speedNestedMultiple} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline

### 2017-07-06 paper/app.tex
general edits

@@
 transition $(\ell^{\lin}_{P},\ell^{\lout}_P)$. In the first case, 
-$U(\ell^{\lin}_{P},\ell^{\lout}_P)=(j,E)$.
+$\updates(\ell^{\lin}_{P},\ell^{\lout}_P)=(j,E)$.
 \item {\em Probabilistic and Non-Deterministic Assignemnts}
@@
 and $\ell^{\lout}_P$ and a
-transition $\tau=(\ell^{\lin}_{P},\ell^{\lout}_P)$ with $U(\tau)=(j,d)$. For 
+transition $\tau=(\ell^{\lin}_{P},\ell^{\lout}_P)$ with $\updates(\tau)=(j,d)$. For 
 $P= 
@@
 the only transition being $\tau=(\ell^{\lin}_{P}\ell^{\lout}_P)$ and 
-$U(\tau)=(j,D)$, where 
+$\updates(\tau)=(j,D)$, where 
 $D$ is 
@@
 \label{eq:lexrsm-ind-1}
-\E[Y_{k+1}] &= \underbrace{\E[Y_{k+1}\cdot \indicator{\Omega\setminus D}]}_{=0=\E[Y_k\cdot\indicator{\Omega\setminus D}]} +  \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{U \leq k\}}]}_{=\E[Y_k\cdot\indicator{D\cap \{U\leq k\}}]} + \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{U > k\}}]}_{=\E[\vecseq{X}{k+1}{\fixn{j}}\cdot\indicator{D\cap \{U>k \} }]},
+\E[Y_{k+1}] &= \underbrace{\E[Y_{k+1}\cdot \indicator{\Omega\setminus D}]}_{=0=\E[Y_k\cdot\indicator{\Omega\setminus D}]} +  \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{F \leq k\}}]}_{=\E[Y_k\cdot\indicator{D\cap \{F\leq k\}}]} + \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{F > k\}}]}_{=\E[\vecseq{X}{k+1}{\fixn{j}}\cdot\indicator{D\cap \{F>k \} }]},
 \end{align}
-where the equality $\E[Y_{k+1}\cdot \indicator{D \cap \{U \leq k\}}] =\E[Y_k\cdot\indicator{D\cap \{U\leq k\}}] $ follows from the fact that $Y_{k+1}(\omega)=Y_k({\omega})=\vecseq{X}{U(\omega)}{\fixn{j}}(\omega)$ for $\omega\in \{U\leq k\}$, and similarly for the last term.
+where the equality $\E[Y_{k+1}\cdot \indicator{D \cap \{F \leq k\}}] =\E[Y_k\cdot\indicator{D\cap \{F\leq k\}}] $ follows from the fact that $Y_{k+1}(\omega)=Y_k({\omega})=\vecseq{X}{F(\omega)}{\fixn{j}}(\omega)$ for $\omega\in \{F\leq k\}$, and similarly for the last term.
 We prove that
@@
 \label{eq:lexrsm-ind-2}
-\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \}}] \leq \E[Y_k\cdot \indicator{D\cap \{U > k \}} -\eps\cdot \indicator{D\cap \{U>k\} \cap \{\levelrank{}{k}= \fixn{j}\} }] .
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{F > k \}}] \leq \E[Y_k\cdot \indicator{D\cap \{F > k \}} -\eps\cdot \indicator{D\cap \{F>k\} \cap \{\levelrank{}{k}= \fixn{j}\} }] .
 \end{equation}
@@
 \label{eq:lexrsm-ind-3}
-\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \}}] = \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }]  + \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j}\} }],
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{F > k \}}] = \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{F > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }]  + \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{F > k \} \cap \{\levelrank{}{k}>\fixn{j}\} }],
 \end{equation}
-since $\levelrank{\omega}{k}\geq \fixn{j}$ for all $\omega \in \{U>k\}$. Since the set $D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}$ is $\genfilt_k$-measurable, we get
+since $\levelrank{\omega}{k}\geq \fixn{j}$ for all $\omega \in \{F>k\}$. Since the set $D\cap \{F > k \} \cap \{\levelrank{}{k}=\fixn{j} \}$ is $\genfilt_k$-measurable, we get
 \begin{align}
-\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] &= \E[\E[\vecseq{X}{k+1}{\fixn{j}}\mid \genfilt_k]\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] \label{eq:lexrsm-ind-4}\\
-&\leq \E[(\vecseq{X}{k}{\fixn{j}} - \eps)\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-5}\\
-&=\E[(Y_k - \eps)\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-6},
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{F > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] &= \E[\E[\vecseq{X}{k+1}{\fixn{j}}\mid \genfilt_k]\cdot \indicator{D\cap \{F > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] \label{eq:lexrsm-ind-4}\\
+&\leq \E[(\vecseq{X}{k}{\fixn{j}} - \eps)\cdot \indicator{D\cap \{F > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-5}\\
+&=\E[(Y_k - \eps)\cdot \indicator{D\cap \{F > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-6},
 \end{align}
-where~\eqref{eq:lexrsm-ind-4} follows from the definition of conditional expectation~\eqref{eq:cond-exp},~\eqref{eq:lexrsm-ind-5} follows from the definition of $\{\levelrank{}{k}=\fixn{j}\}$, and~\eqref{eq:lexrsm-ind-6} holds since $Y_k(\omega)=\vecseq{X}{k}{\fixn{j}}(\omega)$ for $\omega$ with $U(\omega)> k$.
+where~\eqref{eq:lexrsm-ind-4} follows from the definition of conditional expectation~\eqref{eq:cond-exp},~\eqref{eq:lexrsm-ind-5} follows from the definition of $\{\levelrank{}{k}=\fixn{j}\}$, and~\eqref{eq:lexrsm-ind-6} holds since $Y_k(\omega)=\vecseq{X}{k}{\fixn{j}}(\omega)$ for $\omega$ with $F(\omega)> k$.
 Almost identical argument shows that
@@
 \label{eq:lexrsm-ind-7}
-\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j} \} }] \leq \E[Y_k\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j} \}}].
+\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{F > k \} \cap \{\levelrank{}{k}>\fixn{j} \} }] \leq \E[Y_k\cdot \indicator{D\cap \{F > k \} \cap \{\levelrank{}{k}>\fixn{j} \}}].
 \end{equation}
@@
 \begin{align}
-\E[Y_{k+1}]&\leq \E[Y_k] - \eps\cdot \E[\indicator{D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \} }] = \E[Y_k] - \eps\cdot \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \} ) \nonumber\\
+\E[Y_{k+1}]&\leq \E[Y_k] - \eps\cdot \E[\indicator{D \cap \{F>k\} \cap \{\levelrank{}{k} = \fixn{j} \} }] = \E[Y_k] - \eps\cdot \probm(D \cap \{F>k\} \cap \{\levelrank{}{k} = \fixn{j} \} ) \nonumber\\
 &\leq \fixn{B}\cdot \probm(D) - \eps\cdot\left(\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D 
-\cap \{U\geq k\} \cap \{\noofdec_k = \ell\})\right) -\eps\cdot \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \}),
+\cap \{F\geq k\} \cap \{\noofdec_k = \ell\})\right) -\eps\cdot \probm(D \cap \{F>k\} \cap \{\levelrank{}{k} = \fixn{j} \}),
 \end{align}
 where the last inequality follows from induction hypothesis. Hence, using $D_{k,\ell}$ as a shorthand for $D 
-\cap \{U\geq k\} \cap \{\noofdec_k = \ell\}$, to prove~\eqref{eq:lexrsm-soundness-main} it remains to show that
+\cap \{F\geq k\} \cap \{\noofdec_k = \ell\}$, to prove~\eqref{eq:lexrsm-soundness-main} it remains to show that
 \begin{equation}
 \label{eq:lexrsm-ind-8}
-\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}) + \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \}) = \sum_{\ell=0}^{k+1-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}).
+\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}) + \probm(D \cap \{F>k\} \cap \{\levelrank{}{k} = \fixn{j} \}) = \sum_{\ell=0}^{k+1-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}).
 \end{equation}
@@
 &\phantom{+}\;\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell} \cap \{\levelrank{}{k}=\fixn{j}\}) + \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} )\nonumber \\
-&+\sum_{\ell=0}^{k-\fixn{i}}\probm(\underbrace{D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j}\} \cap \{\noofdec_k = \ell\}}_{=D_{k,\ell} \cap \{\levelrank{}{k} = \fixn{j}\}})
+&+\sum_{\ell=0}^{k-\fixn{i}}\probm(\underbrace{D \cap \{F>k\} \cap \{\levelrank{}{k} = \fixn{j}\} \cap \{\noofdec_k = \ell\}}_{=D_{k,\ell} \cap \{\levelrank{}{k} = \fixn{j}\}})
 \nonumber \\
@@
 D_{k,\ell}$ it holds that $\levelrank{k}{\omega}\geq \fixn{j} \Rightarrow 
-U(\omega)\geq k+1$. Finally, in~\eqref{eq:lexrsm-ind-11} we use the fact that 
+F(\omega)\geq k+1$. Finally, in~\eqref{eq:lexrsm-ind-11} we use the fact that 
 all $\omega\in D_{k+1,k+1-\fixn{i}}$ need to have level $\fixn{j}$ in step $k$, 

### 2017-07-06 paper/compositional.tex
general edits

@@
 Given an invariant map $\inv$, we say that a 1-dimensional measurable map 
-$\lem$ for $\pCFG_{\program}$ is $\eps$-$\inv$-ranking in location $\loc$, if 
+$\lem$ for $\pCFG_{\program}$ is $\eps$-$\inv$-ranking/unaffecting in location $\loc$, if 
 for each $\vec{x}\in\inv(\loc)$  each generalized transition~$\tilde{\tau}$ 
-outgoing from $\loc$ is $\eps$-ranked by $\lem$.
+outgoing from $\loc$ is $\eps$-ranked/unaffected by $\lem$.
 
@@
 \begin{compactenum}
-\item  $\inv$-non-negative in each location of $\pCFG_{\program}$;
+\item  non-negative in each $(\loc,\vec{x})$ where $\loc$ is a location of $\pCFG_{\program}$ and $\vec{x}\in \inv(\loc)$;
 \item 
@@
 \item 
-  $\inv$-unaffected in each $\loc\in\loops(\program)$.
+  $\inv$-unaffecting in each $\loc\in\loops(\program)$.
 \end{compactenum}
@@
 \begin{proof}[Proof (Key Idea)]
-Assume that there is set of non-terminating runs $A$ of positive probability.
+%Assume that there is set of non-terminating runs $A$ of positive probability.
+%Let $\{X_i\}_{i=0}^{\infty}$ be a stochastic process returning the value of 
+%NCSM $\lem$ in step $i$. We show that for each $i$ it holds $\E[X_i] 
+%\leq \lem(\locinit,\vecinit) - \eps\cdot \E[\sharp_i]$, where $\sharp_i$ is the 
+%random variable returning the number of times a location from 
+%$\slice(\program)$ is visited. Since all sub-loops of $\program$ terminate 
+%a.s., $\lim_{i\rightarrow \infty}\sharp_i(\run)=\infty$ for $\run\in A$. Since 
+%$A$ has positive probability $\lim_{i\rightarrow 
+%\infty}\E[\sharp_i(\run)=\infty]$, it holds $\lim_{i\rightarrow 
+%\infty}\E[X_i] = -\infty$, a contradiction with non-negativity of $\lem$.
 Let $\{X_i\}_{i=0}^{\infty}$ be a stochastic process returning the value of 
-NCSM $\lem$ in step $i$. We show that for each $i$ it holds $\E[X_i] 
-\leq \lem(\locinit,\vecinit) - \eps\cdot \E[\sharp_i]$, where $\sharp_i$ is the 
-random variable returning the number of times a location from 
-$\slice(\program)$ is visited. Since all sub-loops of $\program$ terminate 
-a.s., $\lim_{i\rightarrow \infty}\sharp_i(\run)=\infty$ for $\run\in A$. Since 
-$A$ has positive probability $\lim_{i\rightarrow 
-\infty}\E[\sharp_i(\run)=\infty]$, it holds $\lim_{i\rightarrow 
-\infty}\E[X_i] = -\infty$, a contradiction with non-negativity of $\lem$.
+NCSM $\lem$ in step $i$.
+Then $\{X_i\}_{i=0}^{\infty}$ is a (non-strict) $1$-dimensional $\eps$-LexRSM for the termination time of the program, for some $\eps>0$.
+Since all sub-loops of $\program$ terminate, with probability one each run has level $<2$ in infinitely many steps.
+From theorem $\ref{thm:lexrsm-main}$ it follows that $\probm^\sigma_{\vecinit}(\ttime<\infty)=1$, for all $\sigma$ and $\vecinit$.
 \end{proof}
@@
 
-As noted previously, a fundamental advantage of LexRSMs over PVSMs is that LexRSMs can handle loops that go through several computational phases and as a result of this do not admit one-dimensional ranking supermartingales.
  Above, we defined NCSMs as one-dimensional objects, to make them analogous to PVSMs for better comparison.
-However, we can also define a multi-dimensional version of NCSMs.
-We say that, given an invariant map $\inv$, an $n$-dimensional measurable map is $\inv$-ranking in a location $\loc$ if for each $\vec{x}\in \inv(\loc)$ there exists $1\leq j \leq n$ such that $\preexp{\lem_j}(\loc,\vec{x}) \leq \lem_j(\loc,\vec{x})-\eps$ and for each $j'<j$ the map $\lem_{j'}$ is $\inv$-unaffected in $\loc$.
+However, we can also define a multi-dimensional version of NCSMs, to take advantage of the fact that LexRSMs can handle loops for which no 1-dimensional linear RSM exists.
+We say that, given an invariant map $\inv$, an $n$-dimensional measurable map is $\eps$-$\inv$-ranking in a location $\loc$ if for each $\vec{x}\in \inv(\loc)$ and each gen.
+transition $\tilde{\tau}$ outgoing from $\loc$ there exists $1\leq j \leq n$ such that $\tilde\tau$ is $\eps$-ranked by $\lem_j$ and for each $j'<j$ we have that  $\tilde{\tau}$ is unaffected by $\lem_{j'}$.
 
@@
 	\item 
-	$\vec{\lem}$ is $\eps$-ranking in each location $\loc\in\slice{(\program)}$; and
+	$\vec{\lem}$ is $\eps$-$\inv$-ranking in each location $\loc\in\slice{(\program)}$; and
 	\item 
@@
 \end{compactenum}
-An $n$-dimensional (propositionally) linear NCSM (LinNCSM) is an $n$-dimensional NCSM which is also a (propositionally) linear expression map.
+An $n$-dimensional linear NCSM (LinNCSM) is an $n$-dimensional NCSM which is also a linear expression map.
 \end{definition}
 
-The following theorem can be proved in the same way as Theorem~\ref{thm:nonneg-comp}.
+The following theorem can be proved in essentially the same way as Theorem~\ref{thm:nonneg-comp}.
 
@@
 
-For \APP{}s, we can generalize Algorithm~\ref{algo:compositional} by changing line~\ref{algoline:one} to ``check existence of a multi-dimensional LinNCSM for $\program'$'' and line~\ref{algoline:two} to ``\textbf{if} a multi-dimensional LinNCSM does not exist.'' The check of existence of a multi-dimensional LinNCSM for $\program'$ can be done by algorithm presented in Section~\ref{xxx}, modified so as to only pursue ranking for transitions outgoing from locations belonging to $\slice(\program')$.
-(I.e., only these transitions have $\eps_{\tau}$ included in the objective function, and algorithm terminates once all such transitions are ranked.) This further expands the usage of the algorithm.
+For \APP{}s, we can generalize Algorithm~\ref{algo:compositional} by changing line~\ref{algoline:one} to ``check existence of a multi-dimensional LinNCSM for $\program'$'' and line~\ref{algoline:two} to ``\textbf{if} a multi-dimensional LinNCSM does not exist.'' The check of existence of a multi-dimensional LinNCSM for $\program'$ can be done by algorithm presented in Section~\ref{sec:algo}, modified so as to only pursue ranking for generalized transitions outgoing from locations belonging to $\slice(\program')$.
+(I.e., only these gen.
+transitions have $\eps_{\tau}$ included in the objective function, and algorithm terminates once all such transitions are ranked.) %This further expands the usage of the algorithm.
 
-There are several advantages of using LexRSMs in a compositonal way.
+%There are several advantages of using LexRSMs in a compositonal way.
 First, using Algorithm~\ref{algo:compositional} (modified for multi-dimensional LinNCSMs) we opt for solving a larger number of smaller linear systems rather than a smaller number of larger ones, which may be sometimes advantageous for performance reasons.

### 2017-07-06 paper/experiments.tex
general edits

@@
 \section{Experimental Results}\label{sec:experiments}
-We have implemented our algorithm (in C++) and present two sets of experimental results.
+We have implemented the algorithm of Section~\ref{sec:algo} (in C++) and present two sets of experimental results.
 For all the experimental results we use the tool \textsc{Aspic}~\cite{FG10:aspic} for 

### 2017-07-06 paper/extensions.tex
general edits

@@
 As in the case of a.s. termination we start with general mathematical statement 
-about LexRSMs. We define a restricted class of LexRSMs with \emph{bounded 
+about LexRSMs. We define a restricted class of strict LexRSMs with \emph{bounded 
 expected conditional increase} property. Recall from 
-definition~\ref{def:lexrsm} that 
+definition~\ref{def:lexrsm} that strict
 LexRSM for a stopping time $\stime$ is characterized by the possibility to 
 a.s. partition, for each $i\in \Nset_0$, the set $\{\omega\in \Omega\mid 
-\stime(\omega)>i \}$ into $d$ sets $L^i_1,\dots,L^i_d$ such that, intuitively, 
+\stime(\omega)>i \}$ into $n$ sets $L^i_1,\dots,L^i_d$ such that, intuitively, 
 on $L^i_j$ the conditional expectation of $\vec{X}_{i+1}[j]$ given $\genfilt_i$ 
@@
 expectation of $\vec{X}_{i+1}[j]$ being larger than $\vec{X}_i[j]$ on 
-$L^i_{j''}$ with $j''$. The conditional expected increase property bounds the 
+$L^i_{j''}$ with $j''>j$. The conditional expected increase property bounds the 
 possibility of this increase.
@@
 \begin{definition}
-Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be a 
-$d$-dimensional LexRSM for some stopping time $\stime$, defined w.r.t. some 
+	\label{def:lexrsm-eci}
+Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be an 
+$n$-dimensional strict LexRSM for some stopping time $\stime$, defined w.r.t. some 
 filtration $\{\genfilt_i\}_{i=0}^{\infty}$. We say that 
@@
 increase (ECI),} for 
-some non-negative vector $\vec{c}\in\Rset^d$, if for each $i\in \Nset_0 $ and 
+some non-negative vector $\vec{c}\in\Rset^d$, if there exists an instance $(\{\vec{X}_{i=0}^{\infty},\{L_1^i,\dots,L_{n+1}^i\}_{i=0}^{\infty})$ of the strict LexRSM (i.e.
+$L^i_{n+1}=\emptyset$ for all $i$) such that for each $i\in \Nset_0 $ and
 each $1\leq 
-j \leq d $ it holds 
+j \leq n $ it holds 
 $\E[\vecseq{X}{i+1}{j}\mid \genfilt_i]\leq \vecseq{X}{i}{j}+\vec{c}[j]$ on 
@@
 for all $j''>j$ 
-(here $L^i_1,\dots,L^i_d$ are as in Definition~\ref{def:lexrsm}).
+(here $L^i_1,\dots,L^i_n$ are as in Definition~\ref{def:lexrsm}).
 \end{definition}
 
-For LexRSMs with $\vec{c}$-bounded ECI we have the following result. For 
+For strict LexRSMs with $\vec{c}$-bounded ECI we have the following result. For 
 simplicity, 
@@
 \label{thm:runtime-bound}
-Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be a 
-$d$-dimensional LexRSM with $c$-bounded ECI for some stopping time $\stime$. 
+Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be an 
+$n$-dimensional strict LexRSM with $c$-bounded ECI for some stopping time $\stime$. 
 Then $\E[\stime]\leq  
-\sum_{j=1}^{d}\E[\vecseq{X}{0}{j}]\cdot(\vec{c}[j]+1)^{d-j}$.
+\sum_{j=1}^{n}\E[\vecseq{X}{0}{j}]\cdot(\vec{c}[j]+1)^{n-j}$.
 \end{theorem}
 \begin{proof}
+Fix an instance $(\{\vec{X}_{i=0}^{\infty},\{L_1^i,\dots,L_{n+1}^i\}_{i=0}^{\infty})$ satisfying Definition~\ref{def:lexrsm-eci}.
 Denote $\noofdecrank_j(\omega)$ the number of steps $i$ in which $\omega\in
-L_j^i$. Since by Theorem~\ref{thm:lexrsm-main} the existence of LexRSM entails 
-$\probm(\stime<\infty)=1$, the value $\noofdecrank_j(\omega)$ is a.s. finite. 
-We prove that for each $1\leq j \leq d$ it holds $\E[\noofdecrank_j]\leq 
+L_j^i$. Since by Theorem~\ref{thm:lexrsm-main} the existence of strict LexRSM entails 
+$\probm(\stime<\infty)=1$, the value $\noofdecrank_j(\omega)$ is a.s. finite for all $1\leq j \leq n$. 
+We prove that for each $1\leq j \leq n$ it holds $\E[\noofdecrank_j]\leq 
 \vec{c}[j]\cdot\left(\sum_{j'<j}\E[\noofdecrank_{j'}]\right) + 
-\E[\vecseq{X}{0}{j}].$ Since $\stime(\omega)=\sum_{1\leq j \leq d} 
+\E[\vecseq{X}{0}{j}].$ Since $\stime(\omega)=\sum_{1\leq j \leq n} 
 \noofdecrank_j(\omega)$, for each $\omega\in \Omega$ (and hence, due to 
-linearity of expectation $\E[\stime]=\sum_{1\leq j \leq d} 
+linearity of expectation $\E[\stime]=\sum_{1\leq j \leq n} 
 E[\noofdecrank_j]$), the statement of the 
@@
 $\program$ be a \PP{} with a pCFG $\pCFG_\program$ and let 
-$\vec{\lem}=(\lem_1,\dots,\lem_d)$ be a 
-$d$-dimensional $1$-LexRSM map for $\program$. Consider a $d$-dimensional 
+$\vec{\lem}=(\lem_1,\dots,\lem_n)$ be an 
+$n$-dimensional $1$-LexRSM map for $\program$. Consider an $n$-dimensional 
 vector 
-$\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_d)$ whose each component is an 
+$\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_n)$ whose each component is an 
 expression over variables of 
@@
 \label{col:runtime-progs}
-Let $\program$ be a probabilistic program. Assume that there exists a 
-$d$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_d)$ for 
+Let $\program$ be a probabilistic program. Assume that there exists an 
+$n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_d)$ for 
 $\program$ supported 
@@
 (w.r.t. $\inv$) 
-for some vector of expressions $\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_d)$. 
+for some vector of expressions $\vec{\bar{c}}=(\bar{c}_1,\dots,\bar{c}_n)$. 
 Then under each scheduler $\sigma$ and for each initial valuation of program 
 variables $\vecinit\in\vecinitset$ it holds $\E^{\sigma}_{\vecinit}[\ttime]\leq 
-\sum_{j=1}^{d}\lem_j(\locinit,\vecinit)\cdot(\bar{c}_j(\vecinit))^{d-j}$.
+\sum_{j=1}^{n}\lem_j(\locinit,\vecinit)\cdot(\bar{c}_j(\vecinit))^{n-j}$.
 \end{corollary}

### 2017-07-06 paper/lexicographic-programs.tex
minor changes
finished sec on expected runtime
re-shuffled the proof of the main theorem a bit
general edits

@@
 \section{Applying Lexicographic Supermartingales to Probabilistic Programs}
+\label{sec:lex-programs}
 
@@
 
-\textbf{[PETR: DEFINE LOCATION TERMINATION]}
+%\textbf{[PETR: DEFINE LOCATION TERMINATION]}
 
@@
 \lem_j(\cfg{\sigma}{i}(\run))$. We claim that $\{\vec{X}_{i}\}_{i=0}^{\infty}$ 
-is an $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
+is a strict $n$-dimensional $\eps$-LexRSM for the termination time $\ttime$ of $\program$. Clearly 
 the process is real-valued, componentwise non-negative, and adapted to the 
@@
 definition of an $\eps$-LexRSM map such a $j$ exists for all 
-$\run\in\{\ttime>i\}$ and hence we indeed have a partition. It remains to prove 
+$\run\in\{\ttime>i\}$ and hence we indeed have a partition (so $L^i_{n+1}=\emptyset$ for all $i$). It remains to prove 
 that irrespective of the initial choice of $\sigma$ and $\vecinit$ it holds, 
@@
 is an $\eps$-LexRSM for $\ttime$, from Theorem~\ref{thm:lexrsm-programs} it 
-follows that $\probm^{\sigma}_{\vecinit}\ttime <\infty) =1$, irrespective of 
+follows that $\probm^{\sigma}_{\vecinit}(\ttime <\infty) =1$, irrespective of 
 $\sigma$ and $\vecinit$.
@@
 \begin{example}
+\label{ex:infinite-time}
 Consider the program in Figure~\ref{fig:inftime}. It is easy to see that it 

### 2017-07-06 paper/lexicographic.tex
small changes
general edits

@@
 $\vec{X}_i (\omega)[j]\geq {0}$. % and $\vec{X}_i(\omega)[j]\neq \infty$.
-\item For each $i\in \Nset_0$ there exists a partition of the set $\{\stime>i\}$ into $n$ subsets $L^i_1,\dots,L^i_n$, all of them $\genfilt_i$-measurable, such that for each $1\leq j \leq n$:
+\item For each $i\in \Nset_0$ there exists a partition of the set $\{\stime>i\}$ into $n+1$ subsets $L^i_1,\dots,L^i_n,L^i_{n+1}$, all of them $\genfilt_i$-measurable, such that for each $1\leq j \leq n$:
 \begin{compactitem}
@@
 	\genfilt_i]\leq\vecseq{X}{i}{j} - 
-	\eps$ on $L_j^i$; and
-	\item for all $1 \leq j' < j$ we have $\E[\vecseq{X}{i+1}{j'}\mid 
-	\genfilt_i]=\vecseq{X}{i}{j'}$ on $L_{j}^i$.
+	\eps$ on $L_j^i$; 
+	\item 
+	 for all $1 \leq j' < j$ we have $\E[\vecseq{X}{i+1}{j'}\mid 
+	 \genfilt_i]=\vecseq{X}{i}{j'}$ on $L_{j}^i$; and
+	\item $\E[\vecseq{X}{i+1}{j}\mid 
+	\genfilt_i]\leq\vecseq{X}{i}{j}$ on $L_{n+1}^i$. 
 \end{compactitem}
 \end{compactenum}
+The $n$-dimensional LexRSM is \emph{strict} if $L^i_{n+1}=\emptyset$ for each $i$.
 \end{definition}
 
-The 1-dimensional lexicographic $\eps$-ranking supermartingale is, to a large extent, equivalent to the notion of a ranking supermartingale as studied in~\cite{HolgerPOPL,CFNH16:prob-termination}.
+An \emph{instance} of an $n$-dimensional LexRSM $\{\vec{X} \}_{i=0}^{\infty}$ is a tuple $(\{\vec{X}_{i=0}^{\infty},\{L_1^i,\dots,L_{n+1}^i\}_{i=0}^{\infty})$ where the second component is a sequence of partitions of $\Omega$ satisfying the condition in Definition~\ref{def:lexrsm}.
+Intuitively, the sets $ L_j^{i}$ for $1\leq j \leq n$ represent the lexicographic ranking condition, i.e.
+for strict LexRSMs, we are in each step able to partition $\Omega$ into subsets such that on $j$-th subset the $j$-th component of $\vec{X}$ is expected to decrease while the previous components are not expected to increase.
+On additional sets $L_{n+1}^i$, none of the components is expected to increase, but decrease is not required: this will become handy later when we deal with compositional LexRSM-based proofs (Section~\ref{sec:compositional}) -- we will not require decrease in every step as long as decrease happens at least once on each cycle in the pCFG.
+We say that $\omega\in \Omega$ has level $j$ in step $i$ of instance $(\{\vec{X}_{i=0}^{\infty},\{L_1^i,\dots,L_{n+1}^i\}_{i=0}^{\infty})$ if $\omega\in L^i_j$.
+We also say that $\omega$ has level $0$ in step $i$ if $\stime(\omega)\leq i$.
+
+The strict 1-dimensional lexicographic $\eps$-ranking supermartingale is, to a large extent, equivalent to the notion of a ranking supermartingale as studied in~\cite{HolgerPOPL,CFNH16:prob-termination}.
 There is one significant difference: in these works there is an additional \emph{integrability} condition imposed on the one-dimensional process $\{X_i\}_{i=0}^{\infty}$, which requires that for each $i\geq 0$ it holds $\E[|X_i|]<\infty$ (or equivalently $\E[X_i]<\infty$, as the process is required to be non-negative).
@@
 %We do not impose this condition, and indeed, it is possible to define a 3-dimensional LexRSM $\{\vec{X}_{i}\}_{i=0}^{\infty}$ such that $\E[X[3]_i]=\infty$ for some $i$ (example in \AppendixMaterial).
-The reason why integrability condiiton can be dropped is that it is only needed in the previous works to ensure that the conditional expectations exist and are well-defined.
+The reason why integrability condition can be dropped is that it is only needed in the previous works to ensure that the conditional expectations exist and are well-defined.
 However, the existence of conditional expectations is also guaranteed for random variables that are real-valued non-negative, see Proposition~\ref{prop:conditional-exp-existence}.
-This is exactly the case in both the original 1-dimensional ranking supermartingales and our generalization to LexRSMs.
+%This is exactly the case in both the original 1-dimensional ranking supermartingales and our generalization to LexRSMs.
 
@@
 Let $(\Omega,\genfilt,\probm)$ be a probability space, 
-$\{\genfilt_i\}_{i=0}^{\infty}$ a filtration of $\genfilt$, and $\stime$ a 
+$\{\genfilt_i\}_{i=0}^{\infty}$ a filtration of $\genfilt$, $\stime$ a 
 stopping 
-time w.r.t. that filtration. If there exists an $n$-dimensional $\eps$-LexRSM for $\stime$, 
-then $\probm(\stime<\infty)=1$.
+time w.r.t.
+that filtration, and $\eps>0$.
+Assume there exists an $n$-dimensional $\eps$-LexRSM for $\stime$ and its instance $(\{\vec{X}_{i=0}^{\infty},\{L_1^i,\dots,L_{n+1}^i\}_{i=0}^{\infty})$ such that $\probm(\{\omega\in \Omega \mid \text{level of $\omega$ is $<n+1$ in infinitely many steps}\})=1$.
+Then $\probm(\stime<\infty)=1$.
+In particular, if there exists a strict $\eps$-LexRSM for $\stime$, then $\probm(\stime<\infty)=1$.
 \end{theorem}
@@
 The proof proceeds by contradiction, i.e. we assume that an $\eps$-LexRSM for 
-$\stime$ exists and $\probm(\stime=\infty)>0$. For succinctness we denote the 
+$\stime$ satisfying the above conditions exists and $\probm(\stime=\infty)>0$. For succinctness we denote the 
 set $\{\omega\mid \stime(\omega)=\infty\}$ by $\genRunSet_{\infty}$.
 
-For $\omega\in \Omega$ we define the level of $\omega$ at step $i$ to be the 
-number $\levelrank{\omega}{i}$ equal to either $0$, if $\stime(\omega)\leq i$, 
-or otherwise to the $1\leq j \leq n$ such that $\omega\in L^i_j$. By the definition of 
-$\eps$-LexRSM the value $\levelrank{\omega}{i}$ is well-defined for all 
+For $\omega\in \Omega$ we denote the level of $\omega$ at step $i$ by $\levelrank{\omega}{i}$.
+%equal to either $0$, if $\stime(\omega)\leq i$, 
+%or otherwise to the $1\leq j \leq n$ such that $\omega\in L^i_j$. 
+The value $\levelrank{\omega}{i}$ is well-defined for all 
 $\omega$ and moreover, the random variable $\levelrank{}{i}$ is $\genfilt_i$-measurable. We denote by $\minlev(\omega)$ the smallest $0 \leq j \leq n$ such 
@@
 
-Now since $\genRunSet_{\infty}=M_1,\dots,M_n$, there must be $1\leq \fixn{j} \leq n$ 
+Now $\genRunSet_{\infty}=M_1\cup\dots\cup M_n\cup M_{n+1}$ and $\probm(M_{n+1})=0$ (as the measure of $\omega$'s that have level $<n+1$ in only finitely many steps is zero, per Theorem's assumption), there must be $1\leq \fixn{j} \leq n$
 s.t. $\probm(M_{\fixn{j}})>0$, i.e. with positive probability the smallest level appearing infinitely often is $\fixn{j}$. For each $\omega\in M_{\fixn{j}}$ there is the smallest number
@@
 \vecseq{X}{k}{\fixn{j}}(\omega) & \text{if $\omega \in D$, $k\geq \fixn{i}$ and 
-$U(\omega)>k$ 
+$F(\omega)>k$ 
 }\\
@@
 First, clearly for all $k\geq 0$, $Y_k(\omega)\geq 0$.
-Second, for each $k\geq \fixn{i}$, the variable $Y_k$ is $\genfilt_k$-measurable, as $D\in \genfilt_{\fixn{i}}$, $\{\vecseq{X}{i}{\fixn{j}}(\omega)\}_{i=0}^{\infty}$ is adapted to the filtration $\{\genfilt_i\}_{i=0}^{\infty}$ and $U$ is a stopping time w.r.t.
+Second, for each $k\geq \fixn{i}$, the variable $Y_k$ is $\genfilt_k$-measurable, as $D\in \genfilt_{\fixn{i}}$, $\{\vecseq{X}{i}{\fixn{j}}(\omega)\}_{i=0}^{\infty}$ is adapted to the filtration $\{\genfilt_i\}_{i=0}^{\infty}$ and $F$ is a stopping time w.r.t.
 this filtration.
@@
 Now according to~\eqref{eq:lexrsm-soundness-main} it holds  $\E[Y_k]\leq \fixn{B}\cdot \probm(D) - \eps\cdot\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D 
-\cap \{U\geq k\} \cap \{\noofdec_k 
+\cap \{F\geq k\} \cap \{\noofdec_k 
 = 
@@
 \E[Y_{\ell_0}]&\leq \fixn{B}\cdot \probm(D) - \eps\cdot\sum_{\ell=0}^{\ell_0-\fixn{i}} \ell\cdot\probm(D 
-\cap \{U\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
+\cap \{F\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
 =
@@
 &= \fixn{B}\cdot \probm(D) - \eps\cdot \sum_{\ell=1}^{\ell_0-\fixn{i}} \probm(D 
-\cap \{U\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
+\cap \{F\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
 \geq 
@@
 & \leq \fixn{B}\cdot \probm(D) - \eps\cdot \sum_{\ell=1}^{m} \probm(D 
-\cap \{U\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
+\cap \{F\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
 \geq 
 \ell\}) \\&\leq \fixn{B}\cdot \probm(D) - \eps\cdot \sum_{\ell=1}^{m} \probm(D 
-\cap \{U\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
+\cap \{F\geq \ell_0\} \cap \{\noofdec_{\ell_0} 
 \geq 
@@
 \geq 
-m \} \subseteq \noofdec_{\ell_0} 
+m \} \subseteq\{ \noofdec_{\ell_0} 
 \geq 
-\ell$ for each $\ell \leq m$, the first inequality on the last line follows by  using $\probm(M)/2\leq \probm(M \cap \{\noofdec_{\ell_0}\geq m\}) \leq \probm(D\cap \{F\geq \ell_0 \} \cap \{\noofdec_{\ell_0}\geq m \})$, and
-the last inequality follows by expanding the definition of $m$. But for each $k$ the random variable $Y_k$ is non-negative, so it must also have a non-negative expectation, a contradiction.
-%\begin{enumerate}
-%\
-%\end{itemize}
-%\end{enumerate}
-
-%\begin{fact}
-
-%\end{fact}
-
-%\[
-%\levelrank{\omega}{i} = \begin{dcases}
-%0 & \text{if $\stime(\omega)\leq i$} \\
-% 
-%\end{dcases}
-%\]
+\ell\}$ for each $\ell \leq m$, the first inequality on the last line follows 
+by  using $\probm(M)/2\leq \probm(M \cap \{\noofdec_{\ell_0}\geq m\}) \leq 
+\probm(D\cap \{F\geq \ell_0 \} \cap \{\noofdec_{\ell_0}\geq m \})$, and
+the last inequality follows by expanding the definition of $m$.
+But for each $k$ the random variable $Y_k$ is non-negative, so it must also have a non-negative expectation, a contradiction.
+Finally, note that for strict $n$-dimensional LexRSMs the condition of level $<n+1$ appearing infinitely many times is trivially satisfied.
 \end{proof}

### 2017-07-06 paper/exp_appendix.tex

@@
 {counterex1b} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
-{cousot9} & {0.09} & {no} & {3} & {yes} & {no} \\\hline
+{cousot9} & {0.09} & {no} & {} & {yes} & {no} \\\hline
 {easy2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline

### 2017-07-07 paper/app.tex
added example for runtime bound

@@
  uniform integrability.
+
+\section{Computations for the proof of Theorem~\ref{THM:RUNTIME-BOUND}}
+
+We 
+prove, by induction on $k$, that for each $k$ it holds 
+$\E[\nodecrank{k}{j}]\leq 
+\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k}{j'}] \right)+ 
+\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k}{j}]$.
+
+The base case $k=0$ is simple as both sides of the inequality are zero. Assume 
+that the inequality holds for some $k\geq 0$. We have 
+$\E[\nodecrank{k+1}{j}]=\E[\nodecrank{k}{j}]+\probm{(L_j^{k})}$, so from 
+induction hypothesis we get 
+\begin{equation}
+\label{eq:time1}
+\E[\nodecrank{k+1}{j}]\leq 
+\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k}{j'}] \right)+ 
+\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k}{j}] + \probm{(L_j^{k})}.\end{equation} 
+Now denote 
+$L^k_{<j} = L^k_1 \cup \dots\cup L^k_{j-1}$ and $L^k_{>j}= 
+L^k_{j+1}\cup\dots\cup L^k_{d}$. We have  
+$\E[\vecseq{X}{k}{j}] = \E[\vecseq{X}{k}{j}\cdot 
+\indicator{ L_{<j}^k}] + \E[\vecseq{X}{k}{j}\cdot 
+\indicator{ L_j^k}] + \E[\vecseq{X}{k}{j}\cdot 
+	\indicator{L_{>j}^k}] \geq 
+\E[\vecseq{X}{k+1}{j}\cdot 
+	\indicator{ L_{<j}^k}] -\vec{c}[j]\cdot \probm(L_{<j}^k) + 
+	\E[\vecseq{X}{k+1}{j}\cdot 
+	\indicator{ L_{j}^k}] + \probm(L_{j}^k) + \E[\vecseq{X}{k+1}{j}\cdot 
+	\indicator{ L_{>j}^k}]= \E[\vecseq{X}{k+1}{j}] -\vec{c}[j]\cdot 
+	\probm(L_{<j}^k)+ \probm(L_{j}^k)$. Plugging this 
+	into~\ref{eq:time1} yields
+\begin{align*}
+\E[\nodecrank{k+1}{j}]&\leq 
+\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k}{j'}] \right)+ 
+\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k+1}{j}] + \vec{c}[j]\cdot 
+\probm(L_{<j}^k)\\
+&=\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k+1}{j'}] \right)+ 
+\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k+1}{j}].
+\end{align*}

### 2017-07-07 paper/compositional.tex
added example for runtime bound

@@
 	\begin{lstlisting}[mathescape]
-while $x\geq 0 $ do
-	$y:=x$;
-	while $y\geq 1$ do
+while $x\geq 0$ and $z \geq 0 $ do
+	$y:=z$;
+	while $y\geq 3$ do
 		$y:=y+$sample$(\mathit{Uniform[-3,1]})$
@@
 		\node[det] (fin) at (0,-1) {$\loc^{\lout}$};
-		\node[det] (as1) at (1,0) {$\loc_1$};
+		\node[det] (as1) at (1.2,0) {$\loc_1$};
 		\node[det] (inner) at (2,0) {$\loc_2$};
@@
 		\draw[tran] (init) to node[font=\scriptsize,draw, fill=white, 
-		rectangle,pos=0.5] {$x<0$} (fin);
+		rectangle,pos=0.6] {$x<0 \vee z<0$} (fin);
 		\draw[tran] (init) to node[font=\scriptsize,draw, fill=white, 
-		rectangle,pos=0.5] {$x\geq 0$} (as1);
-		\draw[tran] (as1) -- node[auto,font=\scriptsize] {$y:=x$} (inner);
+		rectangle,pos=0.5,text width=0.8cm] {$x\geq 0$\\$\wedge$\\$z\geq 0$} 
+		(as1);
+		\draw[tran] (as1) -- node[auto,font=\scriptsize] {$y:=z$} (inner);
 		\draw[tran] (inner) to node[font=\scriptsize,draw, fill=white, 
-		rectangle,pos=0.5] {$y<1$} (as3);
+		rectangle,pos=0.5] {$y<3$} (as3);
 		\draw[tran] (inner) to node[font=\scriptsize,draw, fill=white, 
-		rectangle,pos=0.5] {$y\geq 1$} (as2);
+		rectangle,pos=0.5] {$y\geq 3$} (as2);
 		\draw[tran,rounded corners] (as2) -- (1.25,-1) -- 
@@
 The soundness of the algorithm follows from Theorems~\ref{thm:nonneg-comp} and 
-\ref{thm:holger-comp}. Note that we use the PVSM-based algorithm 
+\ref{thm:holger-comp}. Note that we can use the PVSM-based algorithm 
 of~\cite{HolgerPOPL} as a back-up sub-procedure for the case when LinNCSM-based 

### 2017-07-07 paper/extensions.tex
added example for runtime bound

@@
 \label{thm:runtime-bound}
+\label{THM:RUNTIME-BOUND}
 Let $\{\vec{X}_{i}\}_{i=0}^{\infty}$ be an 
@@
 Then $\E[\stime]\leq  
-\sum_{j=1}^{n}\E[\vecseq{X}{0}{j}]\cdot(\vec{c}[j]+1)^{n-j}$.
+\sum_{j=1}^{n}\E[\vecseq{X}{0}{j}]\cdot(1+\sum_{j'= j+1}^{n}\prod_{j'}^{n} 
+\vec{c}[j'])$.
 \end{theorem}
@@
 $\lim_{k\rightarrow \infty}\E[\nodecrank{k}{j}] = \E[\noofdecrank_j]$ and 
-$\lim_{k\rightarrow \infty}\E[\vecseq{X}{k}{j}] \geq 0$.
+$\lim_{k\rightarrow \infty}\E[\vecseq{X}{k}{j}] \geq 0$. The inductive proof is 
+somewhat technical and deferred to \AppendixMaterial.
 
-The base case $k=0$ is simple as both sides of the inequality are zero. Assume 
-that the inequality holds for some $k\geq 0$. We have 
-$\E[\nodecrank{k+1}{j}]=\E[\nodecrank{k}{j}]+\probm{(L_j^{k})}$, so from 
-induction hypothesis we get 
-\begin{equation}
-\label{eq:time1}
-\E[\nodecrank{k+1}{j}]\leq 
-\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k}{j'}] \right)+ 
-\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k}{j}] + \probm{(L_j^{k})}.\end{equation} 
-Now denote 
-$L^k_{<j} = L^k_1 \cup \dots\cup L^k_{j-1}$ and $L^k_{>j}= 
-L^k_{j+1}\cup\dots\cup L^k_{d}$. We have  
-$\E[\vecseq{X}{k}{j}] = \E[\vecseq{X}{k}{j}\cdot 
-\indicator{ L_{<j}^k}] + \E[\vecseq{X}{k}{j}\cdot 
-\indicator{ L_j^k}] + \E[\vecseq{X}{k}{j}\cdot 
-	\indicator{L_{>j}^k}] \geq 
-\E[\vecseq{X}{k+1}{j}\cdot 
-	\indicator{ L_{<j}^k}] -\vec{c}[j]\cdot \probm(L_{<j}^k) + 
-	\E[\vecseq{X}{k+1}{j}\cdot 
-	\indicator{ L_{j}^k}] + \probm(L_{j}^k) + \E[\vecseq{X}{k+1}{j}\cdot 
-	\indicator{ L_{>j}^k}]= \E[\vecseq{X}{k+1}{j}] -\vec{c}[j]\cdot 
-	\probm(L_{<j}^k)+ \probm(L_{j}^k)$. Plugging this 
-	into~\ref{eq:time1} yields
-\begin{align*}
-\E[\nodecrank{k+1}{j}]&\leq 
-\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k}{j'}] \right)+ 
-\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k+1}{j}] + \vec{c}[j]\cdot 
-\probm(L_{<j}^k)\\
-&=\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k+1}{j'}] \right)+ 
-\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k+1}{j}].
-\end{align*}
+%The base case $k=0$ is simple as both sides of the inequality are zero. Assume 
+%that the inequality holds for some $k\geq 0$. We have 
+%$\E[\nodecrank{k+1}{j}]=\E[\nodecrank{k}{j}]+\probm{(L_j^{k})}$, so from 
+%induction hypothesis we get 
+%\begin{equation}
+%\label{eq:time1}
+%\E[\nodecrank{k+1}{j}]\leq 
+%\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k}{j'}] \right)+ 
+%\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k}{j}] + \probm{(L_j^{k})}.\end{equation} 
+%Now denote 
+%$L^k_{<j} = L^k_1 \cup \dots\cup L^k_{j-1}$ and $L^k_{>j}= 
+%L^k_{j+1}\cup\dots\cup L^k_{d}$. We have  
+%$\E[\vecseq{X}{k}{j}] = \E[\vecseq{X}{k}{j}\cdot 
+%\indicator{ L_{<j}^k}] + \E[\vecseq{X}{k}{j}\cdot 
+%\indicator{ L_j^k}] + \E[\vecseq{X}{k}{j}\cdot 
+%	\indicator{L_{>j}^k}] \geq 
+%\E[\vecseq{X}{k+1}{j}\cdot 
+%	\indicator{ L_{<j}^k}] -\vec{c}[j]\cdot \probm(L_{<j}^k) + 
+%	\E[\vecseq{X}{k+1}{j}\cdot 
+%	\indicator{ L_{j}^k}] + \probm(L_{j}^k) + \E[\vecseq{X}{k+1}{j}\cdot 
+%	\indicator{ L_{>j}^k}]= \E[\vecseq{X}{k+1}{j}] -\vec{c}[j]\cdot 
+%	\probm(L_{<j}^k)+ \probm(L_{j}^k)$. Plugging this 
+%	into~\ref{eq:time1} yields
+%\begin{align*}
+%\E[\nodecrank{k+1}{j}]&\leq 
+%\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k}{j'}] \right)+ 
+%\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k+1}{j}] + \vec{c}[j]\cdot 
+%\probm(L_{<j}^k)\\
+%&=\vec{c}[j]\cdot\left(\sum_{j'<j}\E[\nodecrank{k+1}{j'}] \right)+ 
+%\E[\vecseq{X}{0}{j}] - \E[\vecseq{X}{k+1}{j}].
+%\end{align*}
 %
@@
 variables are periodically reset to a value related to the initial variable 
-values, such as variable $z$ in Figure~\ref{fig:uniint2}. To this end, let 
+values, such as variable $y$ in Figure~\ref{fig:uniint4}. To this end, let 
 $\program$ be a \PP{} with a pCFG $\pCFG_\program$ and let 
@@
 variables $\vecinit\in\vecinitset$ it holds $\E^{\sigma}_{\vecinit}[\ttime]\leq
-\sum_{j=1}^{n}\lem_j(\locinit,\vecinit)\cdot(\bar{c}_j(\vecinit))^{n-j}$.
+\sum_{j=1}^{n}\lem_j(\locinit,\vecinit)\cdot(1+\sum_{j'= j+1}^{n}\prod_{j'}^{n} 
+\bar{c}_{j'}(\vecinit)).$
+%\sum_{j=1}^{n}\lem_j(\locinit,\vecinit)\cdot(\bar{c}_j(\vecinit))^{n-j}$.
 \end{corollary}
+
+\begin{example}
+Consider the program in Figure~\ref{fig:uniint4}. For the program we have an 
+invariant map $\inv$ with $\inv(\loc_0)=x\geq -1$, $\inv(\loc_1)=\inv(\loc_4) = 
+x\geq 0 
+\wedge z\geq 0$, $\inv(\loc_2)=\inv(\loc_3)= x\geq 0 \wedge y \geq 0
+\wedge z\geq 0$, and $\inv(\loc^{\lout})=\mathit{true}$. Further, there exists 
+the following 
+$3$-dimensional 
+$1$-LinLexRSM map $\vec{\lem}$ supported by $\inv$: 
+$\vec{\lem}(\loc_0)=(2x+3,0,2)$, $\vec{\lem}(\loc_1)=(2x+3,0,1)$, 
+$\vec{\lem}(\loc_2)=(2x+2,y,1)$, $\vec{\lem}(\loc_3)=(2x+2,y,0)$, 
+$\vec{\lem}(\loc_4)=(2x+2,0,0)$, and $\vec{\lem}(\loc^{\lout})=(0,0,0)$. It is 
+easy to check that $\vec{\lem}$ is has $(0,z,2)$-bounded ECI. Applying 
+Corollary~\ref{col:runtime-progs} yields $\E[\ttime]\leq (2x_{\mathit{init}} + 
+3)\cdot(1+3z_{\mathit{init}})+0+2= 6x_{\mathit{init}}z_{\mathit{init}} + 
+2x_{\mathit{init}} + 9z_{\mathit{init}}+5$. Note that this quadratic bound is 
+asymptotically optimal for the given program.
+\end{example}
 

### 2017-07-07 paper/introduction.tex
incorporated Krish's comments

@@
 approach. This is because to prove termination, a witness in the form of a 
-ranking function has to be computed: to this automatically, ranking functions 
-of a restricted shape (such as linear ranking functions) have to be considered, 
+ranking function has to be computed: to do this automatically, ranking 
+functions 
+of a restricted form (such as linear ranking functions) have to be considered, 
 and 1-dimensional ranking functions of a restricted type can only prove 

### 2017-07-07 paper/lexicographic-programs.tex
incorporated Krish's comments

@@
 We now discuss how to leverage the mathematical results of the previous section 
-to provide a sound proof rule for almost-sure termination of probabilistic 
+to provide a sound method for almost-sure termination of probabilistic 
 programs. Hence, for the rest of this section we fix a \PP{} $\program$ and the 

### 2017-07-07 paper/synthesis.tex
incorporated Krish's comments

@@
 Both soundness and relative completeness of Algorithm~\ref{algo:linlexrsm} 
-follow by arguments identical to those presented 
+can be proved by generalization of arguments presented 
 in~\cite{ADFG10:lexicographic}. 

### 2017-07-07 paper/compositional.tex
fixed issue with \vec{\lem}

@@
 \begin{definition}
-An $n$-dimensional measurable map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ is an NC supermartingale (NCSM) for a program $\program$ supported by an invariant map $\inv$ if there exists $\eps>0$ such that:
+An $n$-dimensional measurable map $\overrightarrow{\lem}=(\lem_1,\dots,\lem_n)$ 
+is an NC supermartingale (NCSM) for a program $\program$ supported by an 
+invariant map $\inv$ if there exists $\eps>0$ such that:
 \begin{compactenum}
@@
 	\item 
-	$\vec{\lem}$ is $\eps$-$\inv$-ranking in each location $\loc\in\slice{(\program)}$; and
+	$\overrightarrow{\lem}$ is $\eps$-$\inv$-ranking in each location 
+	$\loc\in\slice{(\program)}$; and
 	\item 

### 2017-07-07 paper/extensions.tex

@@
 $\program$ be a \PP{} with a pCFG $\pCFG_\program$ and let 
-$\vec{\lem}=(\lem_1,\dots,\lem_n)$ be an 
+$\overrightarrow{\lem}=(\lem_1,\dots,\lem_n)$ be an 
 $n$-dimensional $1$-LexRSM map for $\program$. Consider an $n$-dimensional 
@@
 $\program$. We say 
-that $\vec{\lem}$ has 
+that $\overrightarrow{\lem}$ has 
 $\vec{\bar{c}}$-bounded ECI w.r.t. invariant map $\inv$ if the following holds 
@@
 Let $\program$ be a probabilistic program. Assume that there exists an 
-$n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_d)$ for 
+$n$-dimensional $\eps$-LexRSM map $\overrightarrow{\lem}=(\lem_1,\dots,\lem_d)$ 
+for 
 $\program$ supported 
 by some 
-invariant map $\inv$, scuh that $\vec{\lem}$ has $\vec{\bar{c}}$-bounded ECI 
+invariant map $\inv$, such that $\overrightarrow{\lem}$ has 
+$\vec{\bar{c}}$-bounded ECI 
 (w.r.t. $\inv$) 
@@
 $3$-dimensional 
-$1$-LinLexRSM map $\vec{\lem}$ supported by $\inv$: 
-$\vec{\lem}(\loc_0)=(2x+3,0,2)$, $\vec{\lem}(\loc_1)=(2x+3,0,1)$, 
-$\vec{\lem}(\loc_2)=(2x+2,y,1)$, $\vec{\lem}(\loc_3)=(2x+2,y,0)$, 
-$\vec{\lem}(\loc_4)=(2x+2,0,0)$, and $\vec{\lem}(\loc^{\lout})=(0,0,0)$. It is 
-easy to check that $\vec{\lem}$ is has $(0,z,2)$-bounded ECI. Applying 
+$1$-LinLexRSM map $\overrightarrow{\lem}$ supported by $\inv$: 
+$\overrightarrow{\lem}(\loc_0)=(2x+3,0,2)$, 
+$\overrightarrow{\lem}(\loc_1)=(2x+3,0,1)$, 
+$\overrightarrow{\lem}(\loc_2)=(2x+2,y,1)$, 
+$\overrightarrow{\lem}(\loc_3)=(2x+2,y,0)$, 
+$\overrightarrow{\lem}(\loc_4)=(2x+2,0,0)$, and 
+$\overrightarrow{\lem}(\loc^{\lout})=(0,0,0)$. It is 
+easy to check that $\overrightarrow{\lem}$ is has $(0,z,2)$-bounded ECI. 
+Applying 
 Corollary~\ref{col:runtime-progs} yields $\E[\ttime]\leq (2x_{\mathit{init}} + 
@@
 
-Now assume that we have synthesized a $1$-LexRSM map $\vec{\lem}$ and we want 
-to check if there exists $\bar{\vec{c}}$ such that $\vec{\lem}$ has 
+Now assume that we have synthesized a $1$-LexRSM map $\overrightarrow{\lem}$ 
+and we want 
+to check if there exists $\bar{\vec{c}}$ such that $\overrightarrow{\lem}$ has 
 $\bar{\vec{c}}$-bounded ECI. In the linear setting (i.e. the program is an 
@@
 constraints, which includes the unknown coefficients as variables, encoding the 
-fact that $\vec{\lem}$ has $\bar{\vec{c}}$-bounded ECI. Details are provided 
+fact that $\overrightarrow{\lem}$ has $\bar{\vec{c}}$-bounded ECI. Details are 
+provided 
 in~\AppendixMaterial. In this way, we can check in polynomial time if 
-Corollary~\ref{col:runtime-progs} can be applied to $\vec{\lem}$, and if yes, 
+Corollary~\ref{col:runtime-progs} can be applied to $\overrightarrow{\lem}$, 
+and if yes, 
 we can synthesize the witness vector $\bar{\vec{c}}$. Since $\bar{\vec{c}}$ 

### 2017-07-07 paper/lexicographic-programs.tex

@@
 As a special case, if all the functions $\lem(\loc)$ are affine, then we call $\lem$ a 1-dimensional \emph{linear expression map (LEM)}.
-Am $n$-dimensional measurable/linear expression map is a vector $\vec{\lem}=(\lem_1,\dots,\lem_n)$ of 1-dimensional measurable/linear expression maps. 
+Am $n$-dimensional measurable/linear expression map is a vector 
+$\overrightarrow{\lem}=(\lem_1,\dots,\lem_n)$ of 1-dimensional 
+measurable/linear expression maps. 
 \end{definition}
@@
 
-We now have all the ingredients needed to define the notion of LexRSM maps for probabilistic programs.
-For notational convenience, we extend the function $\guards$ (which assigns guards to deterministic transitions) to the set of all generalized transition: for a generalized transition $\tau'$ which is not a standard transition outgoing from deterministic location, we put $\guards(\tau')=0\leq 0 \equiv \mathit{true}$.
+We now have all the ingredients needed to define the notion of LexRSM maps for 
+probabilistic programs. For notational convenience, we extend the function 
+$\guards$ (which assigns guards to deterministic transitions) to the set of all 
+generalized transition: for a generalized transition $\tau'$ which is not a 
+standard transition outgoing from deterministic location, we put 
+$\guards(\tau')=0\leq 0 \equiv \mathit{true}$.
 
 \begin{definition}[Lexicographic Ranking Supermartingale Map]
-Let $\eps>0$.
-An $n$-dimensional \emph{lexicographic $\eps$-ranking supermartingale map} ($\eps$-LexRSM map) for a program $\program$ supported by an invariant map $\inv$ is an $n$-dimensional measurable map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for $\program$ such that for each configuration $(\loc,\vec{x})$ where $\loc\neq \locterm$ and $\vec{x}\in \inv(\loc)$ the following conditions are satisfied:
+Let $\eps>0$. An $n$-dimensional \emph{lexicographic $\eps$-ranking 
+supermartingale map} ($\eps$-LexRSM map) for a program $\program$ supported by 
+an invariant map $\inv$ is an $n$-dimensional measurable map 
+$\overrightarrow{\lem}=(\lem_1,\dots,\lem_n)$ for $\program$ such that for each 
+configuration $(\loc,\vec{x})$ where $\loc\neq \locterm$ and $\vec{x}\in 
+\inv(\loc)$ the following conditions are satisfied:
  \begin{compactitem}
@@
 Let $\program$ be a probabilistic program. Assume that there exists an $\eps>0$ 
-and an $n$-dimensional $\eps$-LexRSM map $\vec{\lem}=(\lem_1,\dots,\lem_n)$ for 
+and an $n$-dimensional $\eps$-LexRSM map 
+$\overrightarrow{\lem}=(\lem_1,\dots,\lem_n)$ for 
 $\program$ supported 

### 2017-07-07 paper/prelims.tex

@@
 consider in our analysis. We consider two classes of probabilistic programs: 
-general probabilistic programs (\PP{}s ) with arbitrary (measurable) 
+general probabilistic programs (\PP{}s) with arbitrary (measurable) 
 expressions and 
@@
 for this dual view is that our work also has two main points of focus: a 
-theoretical one, where we introduce new proof rules that can be used to prove 
+theoretical one, where we introduce new concepts that can be used to prove 
 properties of general probabilistic programs; and an algorithmic one, where we 
 aim to prove properties of probabilistic programs automatically, using the 
-aforementioned proof rules. As already testified in the non-probabilistic 
+aforementioned concepts. As already testified in the non-probabilistic 
 world, programs that contain only affine expressions allow for more efficient 
@@
 values of the variables are real numbers. 
-We also assume that assume that each \PP{} $\program$ is preceded by a preamble 
+We also assume that each \PP{} $\program$ is preceded by a preamble 
 specifying possible initial values of program variables: the preamble consists 
@@
 distribution specified by a cumulative distribution 
-function~\cite{Billingsley:book}. We denote by $\mathcal{D}$ be a set of 
+function~\cite{Billingsley:book}. We denote by $\mathcal{D}$ the set of 
 probability distributions on 
@@
 pCFG we assume the following features are known: expected value $\expv[d]$ of 
-$d$ and a set $SP_d$ containing the \emph{support} of $d$. Tthe support is the  
+$d$ and a set $SP_d$ containing the \emph{support} of $d$. The support is the  
 smallest 
 closed set of real numbers whose complement has probability zero 
-under $d$\footnote{In particular, a support of a \emph{discrete} probability 
+under $d$.\footnote{In particular, a support of a \emph{discrete} probability 
 	distribution $d$ is simply the at most countable set of all points on a 
 	real 
-	line that have positive probability under $d$.}) Finally, we assume that 
+	line that have positive probability under $d$.} Finally, we assume that 
 	each 

### 2017-07-07 paper/synthesis.tex

@@
 	together with linear invariant map $\inv$. If the algorithm returns a 
-	$d$-dimensional LEM $\vec{\lem}=(\lem_1,\dots,\lem_d)$, then $\vec{\lem}$ 
+	$d$-dimensional LEM $\overrightarrow{\lem}=(\lem_1,\dots,\lem_d)$, then 
+	$\overrightarrow{\lem}$ 
 	is a $1$-LinLexRSM map for $\program$ supported by $\inv$. Conversely, if 
@@
 \begin{proof}[Proof (Key Ideas)]
-For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the $i$-th iteration of the while loop.
-Then each $\tilde{\tau}\in U_{i}\setminus U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$.
-Since $U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 1-ranked by some component of $\vec{\lem}$.
-Non-negativity of each $\lem_i$ is ensured directly by $\lp_{U_i}$.
-Hence, it remains to show that each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$.
-But this follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the fact that each gen.
-transition in $U_j$ is unaffected by $\lem_j$.
+For soundness, let $U_i$ and $U_{i}$ denote the content of $U$ just before the 
+$i$-th iteration of the while loop. Then each $\tilde{\tau}\in U_{i}\setminus 
+U_{i+1}$ is $1$-ranked by $\lem_i$, for each $1\leq i \leq d$. Since 
+$U_{d}=\emptyset$, each generalized transition of $\pCFG_{\program}$ is 
+1-ranked by some component of $\overrightarrow{\lem}$. Non-negativity of each 
+$\lem_i$ is ensured directly by $\lp_{U_i}$. Hence, it remains to show that 
+each $\tilde{\tau}\in U_i$ is unaffected by $\lem_1,\dots,\lem_{i-1}$. But this 
+follows from $U_i \subseteq U_{i-1}\subseteq \cdots\subseteq U_1$ and from the 
+fact that each gen. transition in $U_j$ is unaffected by $\lem_j$.
 

### 2017-07-07 paper/compositional.tex
added one more example

@@
 		\draw[tran] (init) to node[font=\scriptsize,draw, fill=white, 
-		rectangle,pos=0.5,text width=0.8cm] {$x\geq 0$\\$\wedge$\\$z\geq 0$} 
+		rectangle,pos=0.5,text width=0.8cm, text centered] {$x\geq 
+		0$\\$\wedge$\\$z\geq 0$} 
 		(as1);
@@
 
- Above, we defined NCSMs as one-dimensional objects, to make them analogous to PVSMs for better comparison.
-However, we can also define a multi-dimensional version of NCSMs, to take advantage of the fact that LexRSMs can handle loops for which no 1-dimensional linear RSM exists.
-We say that, given an invariant map $\inv$, an $n$-dimensional measurable map is $\eps$-$\inv$-ranking in a location $\loc$ if for each $\vec{x}\in \inv(\loc)$ and each gen.
-transition $\tilde{\tau}$ outgoing from $\loc$ there exists $1\leq j \leq n$ such that $\tilde\tau$ is $\eps$-ranked by $\lem_j$ and for each $j'<j$ we have that  $\tilde{\tau}$ is unaffected by $\lem_{j'}$.
+ Above, we defined NCSMs as one-dimensional objects, to make them analogous to 
+ PVSMs for better comparison. However, we can also define a multi-dimensional 
+ version of NCSMs, to take advantage of the fact that LexRSMs can handle loops 
+ for which no 1-dimensional linear RSM exists (see Example~\ref{ex:oneloop}). 
+ We say that, given an invariant map $\inv$, an $n$-dimensional measurable map 
+ is $\eps$-$\inv$-ranking in a location $\loc$ if for each $\vec{x}\in 
+ \inv(\loc)$ and each gen. transition $\tilde{\tau}$ outgoing from $\loc$ there 
+ exists $1\leq j \leq n$ such that $\tilde\tau$ is $\eps$-ranked by $\lem_j$ 
+ and for each $j'<j$ we have that  $\tilde{\tau}$ is unaffected by $\lem_{j'}$.
 

### 2017-07-07 paper/extensions.tex
added one more example

@@
 somewhat technical and deferred to \AppendixMaterial.
-
+%
 %The base case $k=0$ is simple as both sides of the inequality are zero. Assume 
@@
 \begin{example}
-Consider the program in Figure~\ref{fig:uniint4}. For the program we have an 
+Consider the program in Figure~\ref{fig:uniint4}, with initial condition 
+$x_{\mathit{init}}\geq 0$. For the program we have an 
 invariant map $\inv$ with $\inv(\loc_0)=x\geq -1$, $\inv(\loc_1)=\inv(\loc_4) = 
@@
 Applying 
-Corollary~\ref{col:runtime-progs} yields $\E[\ttime]\leq (2x_{\mathit{init}} + 
+Corollary~\ref{col:runtime-progs} yields $\E^{\sigma}_{\vecinit}[\ttime]\leq 
+(2x_{\mathit{init}} + 
 3)\cdot(1+3z_{\mathit{init}})+0+2= 6x_{\mathit{init}}z_{\mathit{init}} + 

### 2017-07-07 paper/lexicographic-programs.tex
added one more example

@@
 However, a.s. termination of the program is proved by a 2-dimensional LinLexRSM 
+map
 $\lem$ pictured in the figure.
 \end{example}
+
+Also, similarly to non-probabilistic programs, LexRSMs are more powerful than 
+1-dimensional LexRSMs already on 
+single-loop programs.
+
+\newcommand{\olem}{\overrightarrow{\lem}}
+
+\begin{example}
+\label{ex:oneloop}
+Consider the program in Figure~\ref{fig:invariant-running}, and for simplicity, 
+assume that initial values of variables are non-negative. One can rather 
+straightforwardly check that there is no 1-dimensional Lin(Lex)RSM map for the 
+program, given any invariant. Intuitively, this is because any such RSM map has 
+to 
+decrease as $x$ decreases, since $x$ is the only variable modified in the 
+``if'' branch of non-deterministic choice, but $x$ can increase unboundedly in 
+the ``else'' branch, and this cannot be compensated by a bounded decrease of 
+$y$. However, there is the following $3$-dimensional $\frac{1}{2}$-LinLexRSM 
+map 
+$\overrightarrow{\lem}$: $\olem(\loc_0)=(y+1,x+2,3)$, $\olem(\loc_1) = 
+(y+1,x+2,2)$, $\olem(\loc_2)=\olem(\loc_3)=(y+1,x+2,1)$, 
+$\olem(\loc_4)=(y+1,x+3,0)$, $\olem(\loc_5)=(y+1,x+1,0)$, 
+$\olem(\loc_6)=(y+\frac{1}{2},x+2,1)$, $\olem(\loc_7)=(0,0,0)$. (Here $\olem$ 
+is supported by invariant $\inv$ s.t. $\inv(\loc_0)=x\geq -2 \wedge y\geq -1 $, 
+$\inv(\loc_i)=x\geq 0 \wedge y\geq 0$ for $i=1,\dots,6$, and 
+$\inv(\loc_7)=\mathit{true}$.) A bit more intricate LinLexRSM map exists also 
+without the assumption that initial valuation is non-negative.
+\end{example}

### 2017-07-07 paper/prelims.tex
added one more example

@@
 \begin{lstlisting}[mathescape]
-$x:=10$
-while $x\geq 1$ do
-	if prob(0.75) then $x:=x-1$	else $x:=x+1$ 
+while $x\geq 0$ and $y\geq 0$ do
+	if * then
+		if prob(0.5) then 
+			$x:=x-2$ 
+		else skip 
+		fi
+	else 
+		$x:=2x$;
+		$y:=y-1$
 	fi
@@
 \begin{figure}[t]
+	\begin{subfigure}{0.45\textwidth}
 \centering
 \usebox{\invrun}
-
-\begin{tikzpicture}[x = 1.8cm]
+%\caption{}
+\end{subfigure}
+\begin{subfigure}{0.45\textwidth}
+	\centering
+\begin{tikzpicture}
 %\node[det] (det1) at (0,0) {$\loc_0$};
@@
 
-\node[det] (while) at (1.5,0)  {$\loc_0$};
+\node[det] (while) at (0,0)  {$\loc_0$};
+%\node[det] (while) at (0,0)  {$\loc_0$};
 %%\node[det] (fin) at (0,-3) {};
-\node[ran] (prob) at (3,0) {$\loc_1$};
+\node[dem] (demon) at (2,0) {$\loc_1$};
 %\node[ang] (angel) at (3,0) {$\loc_3$};
@@
 %\node[det] (while) at (1.5,-1.5)  {};
-\node[det] (fin) at (0,0) {$\loc_2$};
+\node[det] (br1) at (3.5,0)  {$\loc_2$};
+\node[det] (br11) at (2,1)	{$\loc_6$};
+\node[ran] (br2) at (2,-1.5)  {$\loc_3$};
+\node[det] (pr1) at (3.5,-1.5)  {$\loc_4$};
+\node[det] (pr2) at (3.5,-3)  {$\loc_5$};
+\node[det] (fin) at (0,-1.5) {$\loc_7$};
 %\node[ran] (prob) at (3,-1.5) {};
@@
 \draw[tran] (while) to node[font=\scriptsize,draw, fill=white, 
-rectangle,pos=0.5] {$x<1$} (fin);
+rectangle,pos=0.5] {$x<0 \vee y<0$} (fin);
 \draw[tran, loop, looseness = 5, in =-65, out = -115] (fin) to (fin);
 \draw[tran] (while) to node[font=\scriptsize,draw, fill=white, 
-rectangle,pos=0.5] {$x\geq 1$} (prob);
-%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white, 
-%rectangle,pos=0.3, inner sep = 1pt] {$\frac{6}{10}$} (angel);
-%\draw[tran] (prob) to node[font=\scriptsize,draw, fill=white, 
-%rectangle,pos=0.3, inner sep = 1pt] {$\frac{4}{10}$} (demon);
+rectangle,pos=0.5, text width=0.8cm, text centered] {$x\geq0$ \\$\wedge$ 
+\\$y\geq0$} (demon);
+\draw[tran] (demon) to (br1);
+\draw[tran] (demon) to (br2);
+\draw[tran] (br2) to node[font=\scriptsize,draw, fill=white, 
+rectangle,pos=0.5, inner sep = 1pt] {$\frac{1}{2}$} (pr1);
+\draw[tran] (br2) to node[font=\scriptsize,draw, fill=white, 
+rectangle,pos=0.5, inner sep = 1pt] {$\frac{1}{2}$} (pr2);
+\draw[tran, rounded corners] (br1) -- (3.5,1) -- node[font=\scriptsize, 
+label={[font=\scriptsize, yshift=-0.5cm] {$x:=2x$}},swap] {}  
+(br11);
+\draw[tran, rounded corners] (br11) -- node[font=\scriptsize, 
+label={[font=\scriptsize, yshift=-0.55cm] {$y:=y-1$}}] {} (while.70|-br11) -- 
+(while.70);
+\draw[tran, rounded corners] (pr1) -- (3.9,-1.5) -- (3.9,1.5) -- 
+(while|-3.9,1.5) -- 
+(while);
+\draw[tran, rounded corners] (pr2) -- (4.1,-3) -- node[font=\scriptsize, 
+label={[font=\scriptsize,xshift=0.6cm] {$x:=x-2$}}] {} (4.1,1.7) -- 
+(while.110|-4.1,1.7) -- 
+(while.110);
 %\draw[tran] (demon) -- node[auto] {x:=x+1} (demon-|while) -- (while);
@@
 
-\node (dum1) at (0,0.8) {};
-\node (dum2) at (0,-0.8) {};
-
-\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
-rectangle,pos=0.5, inner sep = 1pt] {$\frac{3}{4}$} (prob|-dum1) -- node[auto] 
-{x:=x-1}
-(while|-dum1)--(while);
-\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
-rectangle,pos=0.5, inner sep = 1pt] {$\frac{1}{4}$} (prob|-dum2) 
--- node[auto,swap] {x:=x+1} (while|-dum2)--(while);
+%\node (dum1) at (0,0.8) {};
+%\node (dum2) at (0,-0.8) {};
+
+%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
+%rectangle,pos=0.5, inner sep = 1pt] {$\frac{3}{4}$} (prob|-dum1) -- node[auto] 
+%{x:=x-1}
+%(while|-dum1)--(while);
+%\draw[tran] (prob) -- node[font=\scriptsize,draw, fill=white, 
+%rectangle,pos=0.5, inner sep = 1pt] {$\frac{1}{4}$} (prob|-dum2) 
+%-- node[auto,swap] {x:=x+1} (while|-dum2)--(while);
 %\draw[tran] (dif) --  (demon-|while) -- node[auto, pos=0.2] {x:=x+1} (while);
@@
 \end{tikzpicture}
-\caption{An \APP{} modelling an asymmetric 1-D random walk and the associated 
+%\caption{}
+\end{subfigure}
+\caption{An example \APP{} and its associated 
 pCFG. Probabilistic locations are depicted by circles, with probabilities given 
 on outgoing 
-transitions. Transitions are labelled by their effects. Location $\loc_0$ is 
-initial and $\loc_2$ is terminal.}
+transitions, demonic locations are denoted by diamonds. Transitions are 
+labelled by their effects. Location $\loc_0$ is 
+initial and $\loc_7$ is terminal.}
 \label{fig:invariant-running}

### 2017-07-07 paper/compositional.tex
Several small changes

@@
 \label{sec:compositional}
+In this section we discuss compositional aspects. We start with the 
+one-dimensional case, and then present the multidimensional case.
 
@@
 a length of an array, which can be doubled/halved in some sub-program do to (de)allocation, merging, or splitting.
-To overcome the rather strict typechecking, we use the results on LexRSMs to define a new notion of compositional ranking supermartingales, which we call \emph{nob-negative compositional (NC)} supermartingales.
+To overcome the rather strict typechecking, we use the results on LexRSMs to define a new notion of compositional ranking supermartingales, which we call \emph{non-negative compositional (NC)} supermartingales.
 For the sake of generality, we allow NC martingales to be general measurable maps, not necessarily propositionally linear.

### 2017-07-07 paper/exp_appendix.tex
Several small changes

@@
 We now present additional result for other twenty seven benchmarks in Table~\ref{tab:exp3}.
+
+\smallskip\noindent{\em Additional description of table.}
+For the benchmark results reported in Table~\ref{tab:exp1} and Table~\ref{tab:exp3}, 
+we explain the possibilities of Yes/No in the last two column entries.
+\begin{compactitem}
+\item If for the non-probabilistic program no lexicographic RSM exists, then the last two 
+column entries are no and no.
+
+\item If extending the probabilistic program with both probabilistic loop and assignment,
+a lexicographic RSM exists, then the last two column entries are yes and yes.
+
+\item If there exist lexicographic RSM for the non-probabilistic program, but there is no
+lexicographic RSM for the program extended with probabilistic loops, then the second-last 
+entry is yes and the last entry is no, and no RSM exists (i.e., the third entry is no).
+
+\item If there exists lexicographic RSM for extending the non-probabilistic program 
+with probabilistic loop, but no lexicographic RSM for extending the non-probabilistic programs
+with both probabilistic loops and assignments, then for such benchmarks there are two rows 
+explaining the results. 
+Hence while in all previous cases there is a single row for the benchmarks, for this case,
+there are two rows presenting the results.
+\end{compactitem}
 
@@
     \end{tabular}
-    \caption{Additonal experimental results for benchmarks from~\cite{ADFG10:lexicographic} extended with probabilistic loops and/or probabilistic assignments.}
+    \caption{Additonal experimental results for benchmarks from~\cite{ADFG10:lexicographic} extended with probabilistic loops/probabilistic assignments.}
     \label{tab:exp3}

### 2017-07-07 paper/experiments.tex
Several small changes

@@
 We report our results on twenty five benchmarks in Table~\ref{tab:exp1} (we 
-consider around fifty benchmark examples and the results on the remaining ones 
-are presented in \AppendixMaterial.
-The experimental results show that the time taken by our approach is always less 
+consider around fifty benchmark examples and the results on the remaining ones,
+along with detailed description of the table, are presented in \AppendixMaterial).
+The experimental results show that the time taken by our approach is almost always less 
 than $1/10$-th of second.
@@
 of large probabilistic programs generated as follows.
-Given $n$ additional Boolean-like variables, we consider probabilistic while loops,
+Given $n$ Boolean variables, we consider probabilistic while loops,
 with some nondeterministic conditional branches, and generate all possible $2^n$ if conditions
@@
 {catmouse} & {0.08} & {yes} & {2} & {yes} & {yes} \\\hline
-{counterex1a} & {0.1} & {no} & {} & {no} & {no} \\\hline
+{counterex1a} & {0.1} & {no} & {-} & {no} & {no} \\\hline
 {counterex1c} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
@@
 {\multirow{2}{*}{perfect2}} & {0.1} & {yes} & {3} & {yes} & {no} \\\cline{2-6}
-{} & {0.11} & {no} & {} & {yes} & {yes} \\\hline
-{real2} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{} & {0.11} & {no} & {-} & {yes} & {yes} \\\hline
+{real2} & {0.09} & {no} & {-} & {no} & {no} \\\hline
 {realbubble} & {0.22} & {yes} & {3} & {yes} & {yes} \\\hline
 {realselect} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
-{realshellsort} & {0.09} & {no} & {} & {yes} & {no} \\\hline
+{realshellsort} & {0.09} & {no} & {-} & {yes} & {no} \\\hline
 {serpent} & {0.1} & {yes} & {1} & {yes} & {yes} \\\hline
 {sipmabubble} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedDis2} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{speedDis2} & {0.09} & {no} & {-} & {no} & {no} \\\hline
 %{speedFails1} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
@@
 {\multirow{2}{*}{speedSingleSingle2}} & {0.12} & {yes} & {2} & {yes} & {no} \\\cline{2-6}
-{} & {0.1} & {no} & {} & {yes} & {yes} \\\hline
+{} & {0.1} & {no} & {-} & {yes} & {yes} \\\hline
 {\multirow{2}{*}{unperfect}} & {0.1} & {yes} & {2} & {yes} & {no} \\\cline{2-6}
-{} & {0.16} & {no} & {} & {yes} & {yes} \\\hline
+{} & {0.16} & {no} & {-} & {yes} & {yes} \\\hline
 {wcet1} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
@@
 \end{tabular}
-\caption{Experimental results for benchmarks from~\cite{ADFG10:lexicographic} extended with probabilistic loops and/or probabilistic assignments.} \label{tab:exp1}
+\caption{Experimental results for benchmarks from~\cite{ADFG10:lexicographic} extended with probabilistic loops/assignments.} \label{tab:exp1}
+\vspace{-1em}
 \end{table}
@@
 \caption{Experimental results for synthetic examples.}\label{tab:exp2}
+\vspace{-1em}
 \end{table}

### 2017-07-07 paper/extensions.tex
Several small changes

@@
 is infinite. However, it is often desirable to obtain bounds on expected 
-runtime of a program. In this section, we present a LexRSM-based proof rule for 
+runtime of a program. In this section, we present a LexRSM-based method for 
 obtaining bounds on expected runtime, and we show how to automatize the usage 
-of this proof rule to obtain bounds on expected runtime in a 
+of this method to obtain bounds on expected runtime in a 
 subclass of \PP{}s.
@@
 
-Now assume that we have synthesized a $1$-LexRSM map $\overrightarrow{\lem}$ 
+\begin{remark}[Polynomial bounds]\label{rem:poly}
+Now consider that we have synthesized a $1$-LexRSM map $\overrightarrow{\lem}$ 
 and we want 
@@
 initial variable valuation) upper bound on expected runtime.
+\end{remark}
 

### 2017-07-07 paper/lexicographic-programs.tex
Several small changes

@@
-\section{Applying Lexicographic Supermartingales to Probabilistic Programs}
+\section{Lexicographic Supermartingales for Probabilistic Programs}
 \label{sec:lex-programs}
@@
 As a special case, if all the functions $\lem(\loc)$ are affine, then we call $\lem$ a 1-dimensional \emph{linear expression map (LEM)}.
-Am $n$-dimensional measurable/linear expression map is a vector 
+An $n$-dimensional measurable/linear expression map is a vector 
 $\overrightarrow{\lem}=(\lem_1,\dots,\lem_n)$ of 1-dimensional 
@@
 
-We conclude this section by showing that (Lin)LexRSMs can, unlike 1-dimensional 
-RSMs, prove a.s. 
-termination of programs whose expected termination time is infinite.
+For 1-dimensional RSMs, their existence imply both a.s. termination as 
+well as positive termination.
+In contrast, we show that (Lin)LexRSMs can prove a.s. termination of programs 
+whose expected termination time is infinite.
 \lstset{language=affprob}

### 2017-07-07 paper/prelims.tex
Several small changes

@@
 depending on the type of $\up$:
-\begin{compactitem}
-\item if $\up$ is a Borel-measurable function, then $a=\up(\vec{x})$;
-\item if $\up$ is an integrable\footnote{A distribution on some numerical 
+%\begin{compactitem}
+%\item 
+(a)~if $\up$ is a Borel-measurable function, then $a=\up(\vec{x})$;
+%\item 
+(b)~if $\up$ is an integrable\footnote{A distribution on some numerical 
 domain 
@@
 Dirac distribution is integrable.} distribution $d$, then 
-$a\in \support(d)$;
-\item if $\up$ is a set, then $a$ is some element of $\up$.
-\end{compactitem}
+$a\in \support(d)$; and
+%\item
+(c)~if $\up$ is a set, then $a$ is some element of $\up$.
+%\end{compactitem}
 \end{compactitem}

### 2017-07-07 paper/exp_appendix.tex
Changes in table

@@
 {ax} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
-{complex} & {0.1} & {yes} & {1} & {yes} & {yes} \\\hline
-{counterex1b} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
-{cousot9} & {0.09} & {no} & {} & {yes} & {no} \\\hline
+{complex} & {0.10} & {yes} & {1} & {yes} & {yes} \\\hline
+{counterex1b} & {0.10} & {yes} & {3} & {yes} & {yes} \\\hline
+{cousot9} & {0.09} & {no} & {-} & {yes} & {no} \\\hline
 {easy2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{loops} & {0.1} & {no} & {} & {no} & {no} \\\hline
+{loops} & {0.10} & {no} & {-} & {no} & {no} \\\hline
 {\multirow{2}{*}{nestedloop}} & {0.16} & {yes} & {3} & {yes} & {no} \\\cline{2-6}
-{} & {0.23} & {no} & {} & {yes} & {yes} \\\hline
-{perfect1} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
-{random1d} & {0.1} & {yes} & {2} & {yes} & {yes} \\\hline
+{} & {0.23} & {no} & {-} & {yes} & {yes} \\\hline
+{perfect1} & {0.10} & {yes} & {3} & {yes} & {yes} \\\hline
+{random1d} & {0.10} & {yes} & {2} & {yes} & {yes} \\\hline
 %{realbubble} & {0.22} & {yes} & {3} & {yes} & {yes} \\\hline
-{realheapsort} & {0.22} & {no} & {} & {no} & {no} \\\hline
-{realheapsort\_step1} & {0.11} & {no} & {} & {no} & {no} \\\hline
-{realheapsort\_step2} & {0.16} & {no} & {} & {yes} & {no} \\\hline
-{rsd} & {0.1} & {yes} & {1} & {yes} & {yes} \\\hline
+{realheapsort} & {0.22} & {no} & {-} & {no} & {no} \\\hline
+{realheapsort\_step1} & {0.11} & {no} & {-} & {no} & {no} \\\hline
+{realheapsort\_step2} & {0.16} & {no} & {-} & {yes} & {no} \\\hline
+{rsd} & {0.10} & {yes} & {1} & {yes} & {yes} \\\hline
 {sipma91} & {0.14} & {yes} & {2} & {yes} & {yes} \\\hline
 {speedFails1} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{speedDis1} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{speedDis1} & {0.09} & {no} & {-} & {no} & {no} \\\hline
 {\multirow{2}{*}{speedFails2}} & {0.14} & {yes} & {1} & {yes} & {no} \\\cline{2-6}
-{} & {0.08} & {no} & {} & {yes} & {yes} \\\hline
-{speedFails4} & {0.09} & {no} & {} & {no} & {no} \\\hline
-{speedNestedMultipleDep} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedpldi3} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedSimpleMultiple} & {0.1} & {no} & {} & {no} & {no} \\\hline
+{} & {0.08} & {no} & {-} & {yes} & {yes} \\\hline
+{speedFails4} & {0.09} & {no} & {-} & {no} & {no} \\\hline
+{speedNestedMultipleDep} & {0.10} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedpldi3} & {0.10} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedSimpleMultiple} & {0.10} & {no} & {-} & {no} & {no} \\\hline
 {speedSingleSingle} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
@@
 {wcet2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{wise} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{wise} & {0.09} & {no} & {-} & {no} & {no} \\\hline
     \end{tabular}

### 2017-07-07 paper/experiments.tex
Changes in table

@@
 {catmouse} & {0.08} & {yes} & {2} & {yes} & {yes} \\\hline
-{counterex1a} & {0.1} & {no} & {-} & {no} & {no} \\\hline
+{counterex1a} & {0.10} & {no} & {-} & {no} & {no} \\\hline
 {counterex1c} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
@@
 {exmini} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{insertsort} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{insertsort} & {0.10} & {yes} & {3} & {yes} & {yes} \\\hline
 {ndecr} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
 {perfect} & {0.11} & {yes} & {3} & {yes} & {yes} \\\hline
-{\multirow{2}{*}{perfect2}} & {0.1} & {yes} & {3} & {yes} & {no} \\\cline{2-6}
+{\multirow{2}{*}{perfect2}} & {0.10} & {yes} & {3} & {yes} & {no} \\\cline{2-6}
 {} & {0.11} & {no} & {-} & {yes} & {yes} \\\hline
@@
 {realshellsort} & {0.09} & {no} & {-} & {yes} & {no} \\\hline
-{serpent} & {0.1} & {yes} & {1} & {yes} & {yes} \\\hline
-{sipmabubble} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{serpent} & {0.10} & {yes} & {1} & {yes} & {yes} \\\hline
+{sipmabubble} & {0.10} & {yes} & {3} & {yes} & {yes} \\\hline
 {speedDis2} & {0.09} & {no} & {-} & {no} & {no} \\\hline
 %{speedFails1} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
-{speedNestedMultiple} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{speedNestedMultiple} & {0.10} & {yes} & {3} & {yes} & {yes} \\\hline
 {speedpldi2} & {0.09} & {yes} & {2} & {yes} & {yes} \\\hline
 {speedpldi4} & {0.09} & {yes} & {3} & {yes} & {yes} \\\hline
-{speedSimpleMultipleDep} & {0.09} & {no} & {} & {no} & {no} \\\hline
+{speedSimpleMultipleDep} & {0.09} & {no} & {-} & {no} & {no} \\\hline
 {\multirow{2}{*}{speedSingleSingle2}} & {0.12} & {yes} & {2} & {yes} & {no} \\\cline{2-6}
-{} & {0.1} & {no} & {-} & {yes} & {yes} \\\hline
-{\multirow{2}{*}{unperfect}} & {0.1} & {yes} & {2} & {yes} & {no} \\\cline{2-6}
+{} & {0.10} & {no} & {-} & {yes} & {yes} \\\hline
+{\multirow{2}{*}{unperfect}} & {0.10} & {yes} & {2} & {yes} & {no} \\\cline{2-6}
 {} & {0.16} & {no} & {-} & {yes} & {yes} \\\hline
 {wcet1} & {0.11} & {yes} & {2} & {yes} & {yes} \\\hline
-{while2} & {0.1} & {yes} & {3} & {yes} & {yes} \\\hline
+{while2} & {0.10} & {yes} & {3} & {yes} & {yes} \\\hline
 
@@
 {5} & {104} & {0.14} & {0.06} & {0.19} \\\hline
-{6} & {200} & {0.36} & {0.1} & {0.46} \\\hline
+{6} & {200} & {0.36} & {0.10} & {0.46} \\\hline
 {7} & {392} & {1.31} & {0.3} & {1.61} \\\hline

### 2017-07-07 paper/lexicographic.tex
Several small changes

@@
 
+\begin{remark}
+\label{rem:canfilt}
 In context of probabilistic programs we work with probability 
@@
 $\natfilt$ such that all the functions $\cfg{\sigma}{j}$, $0\leq j \leq 
-i$, are $\natfilt_i$-measurable. Intuitively, each set $A$ belonging to such 
-an $\natfilt_i$ consists of runs whose first $i$ steps satisfy some 
-property, and the probability space $(\Omega,\natfilt_i,\probm^\sigma)$ 
-allows us to reason about probabilities of certain events happening in the 
-first 
-$i$ steps of program execution. 
+i$, are $\natfilt_i$-measurable. 
+Intuitively, each set $A\in\natfilt_i$ consists of runs whose first $i$ steps 
+satisfy some property.
+%property, and the probability space $(\Omega,\natfilt_i,\probm^\sigma)$ 
+%allows us to reason about probabilities of certain events happening in the 
+%first 
+%$i$ steps of program execution. 
 %(note that since $\natfilt_i \subseteq 
@@
 filtration of $\natfilt$, which we call a \emph{canonical filtration}.
+\end{remark}
 
@@
 
+While lexicographic ranking supermartingales provide a multidimensional extension of ranking supermartingales, 
+there is an important difference even in the single dimension, which we explain below.
 The strict 1-dimensional lexicographic $\eps$-ranking supermartingale is, to a large extent, equivalent to the notion of a ranking supermartingale as studied in~\cite{HolgerPOPL,CFNH16:prob-termination}.

### 2017-07-07 paper/synthesis.tex
Several small changes

@@
 
-The main idea is to iteratively synthesize 1-dimensional linear expression map that $1$-rank a subset of generalized transitions.
+The main idea is to iteratively synthesize 1-dimensional {\em linear expression map (LEM)} that $1$-rank a subset of generalized transitions.
 These maps form the individual components of the sought-after $1$-LinLexRSM map.
@@
 	$\eps>0$ there is no $d'$-dimensional $\eps$-LinLexRSM for $\program$  
-	supported by $\inv$. If guards of all conditionals and loops in $\APP{}$ 
-	are linear assertion, then the
+	supported by $\inv$. If guards of all conditionals and loops in $\program$
+	are linear assertions, then the
 	algorithm runs in time polynomial in size of 

### 2017-07-07 paper/conclu.tex
Several small changes
OK

@@
+\vspace{-1em}
 \section{Related Work}
-In this section we discuss the related works.
+\vspace{-0.5em}
+%%In this section we discuss the related works.
 
-\smallskip\noindent{\em Probabilistic programs and termination.}
+%%\smallskip
+\noindent{\em Probabilistic programs and termination.}
 In early works the termination for concurrent probabilistic programs was studied 
@@
 
+\vspace{-1em}
 \section{Conclusion and Future Work}
+\vspace{-0.5em}
 In this work we considered lexicographic RSMs for termination analysis of probabilistic

### 2017-07-07 paper/compositional.tex
small changes

@@
 
-Hence, NCSMs effectively trade the uniform integrability condition for 
-non-negativity over the whole program. We believe that the latter condition is 
-substantially easier to impose and check automatically, especially in the case 
-of affine probabilistic programs and (propositionally) linear NCSMs: for 
-\APP{}s, synthesizing NCSM entails synthesizing sufficient program invariants 
-(for which there is a good automated tool support~\cite{FG10:aspic})  encoding 
+%Hence, NCSMs effectively trade the uniform integrability condition for 
+%non-negativity over the whole program. We believe that the latter condition is 
+%substantially easier to impose and check automatically, especially in the case 
+%of affine probabilistic programs and (propositionally) linear NCSMs: 
+For 
+\APP{}s, synthesizing linear NCSM entails synthesizing sufficient program 
+invariants 
+(for which there is a good automated tool support~\cite{FG10:aspic}) and  
+encoding 
 the ranking, 
@@
 \begin{example}
-For all programs in Figures~\ref{fig:uniint} and~\ref{fig:uniint2} there are 
-LinNCSMs for all the loops in the program, which shows that the programs 
+All programs in Figures~\ref{fig:uniint} and~\ref{fig:uniint2} have 
+LinNCSMs for all their loops, which shows that the programs 
 terminate 
@@
 \Input{An \APP{} $\program$ together with an invariant map $\inv$.}
-$d\leftarrow \text{ depth of loop nesting in $\program$}$\;
-\For{$i \leftarrow d$ \KwTo $0$}{
+%$d\leftarrow \text{ depth of loop nesting in $\program$}$\;
+\For{$i \leftarrow \text{ depth of loop nesting in $\program$}$ \KwTo $0$}{
 \ForEach{sub-loop $\program'$ of $\program$ nested $i$ levels below the main loop}{

### 2017-07-07 paper/extensions.tex
small changes

@@
 runtime of a program. In this section, we present a LexRSM-based method for 
-obtaining bounds on expected runtime, and we show how to automatize the usage 
-of this method to obtain bounds on expected runtime in a 
-subclass of \PP{}s.
+obtaining bounds on expected runtime using a restricted class of LexRSMs.
 
@@
 expectation of $\vec{X}_{i+1}[j]$ being larger than $\vec{X}_i[j]$ on 
-$L^i_{j''}$ with $j''>j$. The conditional expected increase property bounds the 
+$L^i_{j''}$ with $j''>j$. The expected conditional increase (ECI) property 
+bounds the 
 possibility of this increase.
@@
 
+Previous work testifies that analogous ``bounded 
+expected change'' constraints naturally emerge when one aims to obtain 
+additional information about runtime of programs 
+~\cite{HolgerPOPL,CFNH16:prob-termination,CNZ17,CF17}. The major novelty of our 
+bounded ECI condition is that it can be used to obtain polynomial bounds on 
+expected runtime using \emph{linear} supermartingales.
+
 \begin{example}

### 2017-07-07 paper/prelims.tex
small changes

@@
 if $f(x)\leq g(x)$ for all $x$ in the domain. 
-
-\smallskip\noindent{\em Variables.}
+%\smallskip\noindent{\em Variables.}
 Throughout the paper we fix a countable set of variables $\vars$. 
@@
 In this subsection we define the form of probabilistic programs that we 
-consider in our analysis. We consider two classes of probabilistic programs: 
+consider in our analysis. 
+We consider two classes of probabilistic programs: 
 general probabilistic programs (\PP{}s) with arbitrary (measurable) 
@@
 %\end{enumerate}
-
+%
 From measure theory it is known that these conditions hold in particular for programs where expressions are build using the standard arithmetic operators of addition, subtraction, multiplication, and division (provided that expressions evaluating to zero are not allowed as divisors).
@@
 
-\smallskip\noindent{\em Probabilistic Control Flow Graphs.}
+\noindent{\em Probabilistic Control Flow Graphs.}
 %There are several ways in which one can express the semantics of \APP's with
@@
 
-\smallskip
 \begin{definition}
@@
 \end{definition}
-
-\smallskip\noindent{\em Configurations.}
+\noindent{\em Configurations.}
 A \emph{configuration} of a pCFG $\pCFG$ is a tuple $(\ell,\vec{x})$,
@@
 
-\smallskip\noindent{\em Executions and reachable configurations.}
+\noindent{\em Executions and reachable configurations.}
 We say that a configuration $(\loc',\vec{x}')$ is a \emph{successor} of a 
@@
 \end{subfigure}
+%\vspace{-3em}
 \caption{An example \APP{} and its associated 

### 2017-07-07 paper/prelims.tex
small changes

@@
 $\probm$ is a \emph{probability measure} on $\mathcal{F}$, i.e., a function
-$\probm\colon \mathcal{F}\rightarrow[0,1]$ such that
-\begin{compactitem}
-\item $\probm(\emptyset)=0$,
-\item for all $A\in \mathcal{F}$ it holds $\probm(\Omega\smallsetminus
+$\probm\colon \mathcal{F}\rightarrow[0,1]$ such that:
+%\begin{compactitem}
+(1) $\probm(\emptyset)=0$,
+(2) for all $A\in \mathcal{F}$ it holds $\probm(\Omega\smallsetminus
 A)=1-\probm(A)$, and
-\item for all pairwise disjoint countable set sequences $A_1,A_2,\dots \in
+(3) for all pairwise disjoint countable set sequences $A_1,A_2,\dots \in
 \mathcal{F}$ (i.e., $A_i \cap A_j = \emptyset$ for all $i\neq j$)
 we have $\sum_{i=1}^{\infty}\probm(A_i)=\probm(\bigcup_{i=1}^{\infty} A_i)$.
-\end{compactitem}
+%\end{compactitem}
 
@@
 \vspace{-0.5em}
-Termination is the basic liveness property of \PP{}s.
+%Termination is the basic liveness property of \PP{}s.
 
@@
 %\end{example}
-Since a probabilistic program may exhibit more than one run, we are interested 
-in probabilities of runs that terminate or reach some set of configurations. 
-This gives rise to the following fundamental computational problems regarding 
+%Since a probabilistic program may exhibit more than one run, we are interested 
+%in probabilities of runs that terminate or reach some set of configurations. 
+We consider the following fundamental computational problems regarding 
 termination:

### 2017-07-07 paper/conclu.tex

@@
-\vspace{-1em}
+\vspace{-4em}
 \section{Related Work}
@@
 
-\smallskip\noindent{\em Compositional a.s. termination proving.} 
+\noindent{\em Compositional a.s. termination proving.} 
 A 
@@
 
-\smallskip\noindent{\em Other approaches.}
+\noindent{\em Other approaches.}
 Besides RSMs, other approaches has also been considered for probabilistic programs.
@@
 
-\smallskip\noindent{\em Non-probabilistic programs.}
+\noindent{\em Non-probabilistic programs.}
 Termination analysis of non-probabilistic programs has also been extensively 
@@
 
-\vspace{-1em}
+\vspace{-0.5em}
 \section{Conclusion and Future Work}

### 2017-07-07 paper/experiments.tex

@@
 We consider benchmarks from abstraction of real-world non-probabilistic programs 
-in~\cite{ADFG10:lexicographic}.
+in~\cite{ADFG10:lexicographic} (see also~\cite{CCGSY08:ranking-abstractions}).
 Note that there are no abstraction tools available for probabilistic programs 

### 2017-07-07 paper/synthesis.tex

@@
 
-Each system $\linsystem_{\tilde{\tau}}$ is constructed in such a way that if it admits a solution with some $\eps_{\tilde{\tau}}$ positive, then decreasing the value of $\eps_{\tilde{\tau}}$ in that solution to any non-negative value still yield a valid solution (this corresponds to the fact that if some transition is $\eps$-ranked by $\lem$, than it is $\eps'$-ranked by $\lem$ for each $0\leq \eps'\leq \eps$).
-Morever, each solution where $\eps_{\tilde{\tau}}$ is positive can be rescaled into another solution in which $\eps_{\tilde{\tau}}$ is at least $1$.
-It follows that of $\lp_{U}$ has at least one feasible solution, then it has an optimal solution in which each $\eps_{\tilde{\tau}}$ is either $0$ or $1$.
-If the system does not have a solution or all the $\eps_{\tilde{\tau}}$ are equal to zero then there is no LEM satisfying (1)--(3).
-Otherwise, the optimal solution of $\lp_{U}$ yields a LEM $\lem$ which satisfies (1)--(3) and maximizes the number of 1-ranked elements of $U$.
+Each system $\linsystem_{\tilde{\tau}}$ is constructed in such a way that if it 
+admits a solution with some $\eps_{\tilde{\tau}}$ positive, then decreasing the 
+value of $\eps_{\tilde{\tau}}$ in that solution to any non-negative value still 
+yield a valid solution (this corresponds to the fact that if some transition is 
+$\eps$-ranked by $\lem$, than it is $\eps'$-ranked by $\lem$ for each $0\leq 
+\eps'\leq \eps$). Moreover, each solution where $\eps_{\tilde{\tau}}$ is 
+positive can be rescaled into another solution in which $\eps_{\tilde{\tau}}$ 
+is at least $1$. It follows that of $\lp_{U}$ has at least one feasible 
+solution, then it has an optimal solution in which each $\eps_{\tilde{\tau}}$ 
+is either $0$ or $1$. If the system does not have a solution or all the 
+$\eps_{\tilde{\tau}}$ are equal to zero then there is no LEM satisfying 
+(1)--(3). Otherwise, the optimal solution of $\lp_{U}$ yields a LEM $\lem$ 
+which satisfies (1)--(3) and maximizes the number of 1-ranked elements of $U$.
 

### 2017-07-07 paper/app.tex
small edits

@@
 
+\emph{Measurable Schedulers} 
+First we need to clarify what are measurable sets of configurations 
+and histories. We define a sigma-algebra $\mathcal{F}_{\Conf}$ of measurable 
+sets of configurations to be the sigma-algebra 
+over $\Conf$ 
+generated\footnote{In general, it is known that for each set $\Omega$ and each 
+collection of its subsets $F\subseteq 2^{\Omega}$ there exists at least one 
+sigma-algebra $\mathcal{F}$ s.t. $F\subseteq \mathcal{F}$ and the intersection 
+of all such sigma-algebras is again a sigma algebra --  so called 
+sigma-algebra 
+generated by $F$~\cite{Billingsley:book}.} by 
+all sets of the form $\{\loc\}\times B$, where $\loc$ is a location of $\pCFG$ 
+and  
+$B$ is a Borel-measurable subset 
+of $\Rset^{|\pvars|}$. Next, the set of finite paths $\Fpath$ can be viewed as 
+a subset of $\Conf\cup \Conf\times\Conf \cup \Conf\times\Conf\times\Conf \cup 
+\dots$. Hence, we define the sigma-algebra $\mathcal{H}$ of measurable sets of 
+finite paths 
+to be the sigma algebra generated by all sets of the form $\confset_1\times 
+\confset_2 \times 
+\cdots \times \confset_k \subseteq \Fpath$ such that $k\in \Nset$ and 
+$\confset_i \in 
+\mathcal{F}_{\Conf}$ for 
+all $1\leq i \leq k$. Now we can define the measurability of schedulers. 
+Recall 
+that for each finite path $\genpath$ ending in a non-deterministic 
+configuration 
+we have that $\sigma(\genpath)$ is a probability distribution on $\Conf$. For 
+each measurable set of configurations $\confset$ we denote by 
+$\sigma(\genpath)(\confset)$ the probability that the random draw from 
+distribution $\sigma(\genpath)$ selects an element of $\confset$. We say that 
+a 
+scheduler $\sigma$ is measurable if for each $\confset \in 
+\mathcal{F}_{\Conf}$ 
+and each $p\in[0,1]$ the set $\{\pi \in \Fpath \mid 
+\sigma(\genpath)(\confset)\leq p\}$ belongs to $\mathcal{H}$, i.e., it is a 
+measurable set of paths.
+
 \def\xx{\ref{prop:conditional-exp-existence}}

### 2017-07-07 paper/experiments.tex
small edits

@@
 \section{Experimental Results}\label{sec:experiments}
-We have implemented the algorithm of Section~\ref{sec:algo} (in C++) and present two sets of experimental results.
-For all the experimental results we use the tool \textsc{Aspic}~\cite{FG10:aspic} for 
+We have implemented the algorithm of Section~\ref{sec:algo} (in C++) and 
+present two sets of experimental results.
+For all the experimental results we use the tool 
+\textsc{Aspic}~\cite{FG10:aspic} for 
 invariant generation, 
@@
 \caption{Experimental results for benchmarks from~\cite{ADFG10:lexicographic} extended with probabilistic loops/assignments.} \label{tab:exp1}
-\vspace{-1em}
+\vspace{-1.5em}
 \end{table}
@@
 \caption{Experimental results for synthetic examples.}\label{tab:exp2}
-\vspace{-1em}
+\vspace{-2 em}
 \end{table}
+%\vskip-1cm
 \end{center}

### 2017-07-07 paper/lexicographic.tex
small edits

@@
 \Rset$ an $\mathcal{F}$-measurable function, and $\mathcal{F}'\subseteq 
-\mathcal{F}$ sub-sigma-algebra of $\mathcal{F}$. Assume that one of the following conditions hold:
-\begin{itemize}
-\item $\E[|X|]<\infty$; or
-\item $X$ is real-valued and non-negative.
-\end{itemize}
+\mathcal{F}$ sub-sigma-algebra of $\mathcal{F}$. Assume that one of the 
+following conditions hold: (a)
+%\begin{itemize}
+$\E[|X|]<\infty$, or (b)
+$X$ is real-valued and non-negative. 
+%\end{itemize}
 Then there exists a conditional expectation of $X$ given $\mathcal{F}'$ and it is almost-surely unique, i.e.
@@
 $$
+\vspace{-3mm}
 

### 2017-07-07 paper/prelims.tex
small edits

@@
 mathematically 
-sound way. First we need to clarify what are measurable sets of configurations 
-and histories. We define a sigma-algebra $\mathcal{F}_{\Conf}$ of measurable 
-sets of configurations to be the sigma-algebra 
-over $\Conf$ 
-generated\footnote{In general, it is known that for each set $\Omega$ and each 
-collection of its subsets $F\subseteq 2^{\Omega}$ there exists at least one 
-sigma-algebra $\mathcal{F}$ s.t. $F\subseteq \mathcal{F}$ and the intersection 
-of all such sigma-algebras is again a sigma algebra --  so called sigma-algebra 
-generated by $F$~\cite{Billingsley:book}.} by 
-all sets of the form $\{\loc\}\times B$, where $\loc$ is a location of $\pCFG$ 
-and  
-$B$ is a Borel-measurable subset 
-of $\Rset^{|\pvars|}$. Next, the set of finite paths $\Fpath$ can be viewed as 
-a subset of $\Conf\cup \Conf\times\Conf \cup \Conf\times\Conf\times\Conf \cup 
-\dots$. Hence, we define the sigma-algebra $\mathcal{H}$ of measurable sets of 
-finite paths 
-to be the sigma algebra generated by all sets of the form $\confset_1\times 
-\confset_2 \times 
-\cdots \times \confset_k \subseteq \Fpath$ such that $k\in \Nset$ and 
-$\confset_i \in 
-\mathcal{F}_{\Conf}$ for 
-all $1\leq i \leq k$. Now we can define the measurability of schedulers. Recall 
-that for each finite path $\genpath$ ending in a non-deterministic 
-configuration 
-we have that $\sigma(\genpath)$ is a probability distribution on $\Conf$. For 
-each measurable set of configurations $\confset$ we denote by 
-$\sigma(\genpath)(\confset)$ the probability that the random draw from 
-distribution $\sigma(\genpath)$ selects an element of $\confset$. We say that a 
-scheduler $\sigma$ is measurable if for each $\confset \in \mathcal{F}_{\Conf}$ 
-and each $p\in[0,1]$ the set $\{\pi \in \Fpath \mid 
-\sigma(\genpath)(\confset)\leq p\}$ belongs to $\mathcal{H}$, i.e., it is a 
-measurable set of paths.
-
-While the definition of a measurable scheduler might seem somewhat technical, 
-it is natural from measure-theoretic point of view and analogous definitions 
-inevitably emerge in works dealing with systems that exhibit both probabilistic 
+sound way. The definition of a measurable scheduler that we use is the standard 
+one used when dealing with systems that exhibit both probabilistic 
 and non-deterministic behaviour over a continuous state 
-space~\cite{NSK:CTMDP-delayed,NK:CTMDP-bisimulation}. In particular, if all the 
-variables in a program range over a discrete set (such as the integers), 
+space~\cite{NSK:CTMDP-delayed,NK:CTMDP-bisimulation}. The definition is 
+somewhat 
+technical, and not substantial to 
+our results, so we defer it to~\AppendixMaterial. In the rest of this work, we 
+refer to measurable schedulers simply as ``schedulers.'' In particular, if the 
+set of all reachable variable valuations is discrete (such as the integers), 
 then each scheduler in the associated pCFG is measurable.
+
+%The definition
+%First we need to clarify what are measurable sets of configurations 
+%and histories. We define a sigma-algebra $\mathcal{F}_{\Conf}$ of measurable 
+%sets of configurations to be the sigma-algebra 
+%over $\Conf$ 
+%generated\footnote{In general, it is known that for each set $\Omega$ and each 
+%collection of its subsets $F\subseteq 2^{\Omega}$ there exists at least one 
+%sigma-algebra $\mathcal{F}$ s.t. $F\subseteq \mathcal{F}$ and the intersection 
+%of all such sigma-algebras is again a sigma algebra --  so called 
+%sigma-algebra 
+%generated by $F$~\cite{Billingsley:book}.} by 
+%all sets of the form $\{\loc\}\times B$, where $\loc$ is a location of $\pCFG$ 
+%and  
+%$B$ is a Borel-measurable subset 
+%of $\Rset^{|\pvars|}$. Next, the set of finite paths $\Fpath$ can be viewed as 
+%a subset of $\Conf\cup \Conf\times\Conf \cup \Conf\times\Conf\times\Conf \cup 
+%\dots$. Hence, we define the sigma-algebra $\mathcal{H}$ of measurable sets of 
+%finite paths 
+%to be the sigma algebra generated by all sets of the form $\confset_1\times 
+%\confset_2 \times 
+%\cdots \times \confset_k \subseteq \Fpath$ such that $k\in \Nset$ and 
+%$\confset_i \in 
+%\mathcal{F}_{\Conf}$ for 
+%all $1\leq i \leq k$. Now we can define the measurability of schedulers. 
+%Recall 
+%that for each finite path $\genpath$ ending in a non-deterministic 
+%configuration 
+%we have that $\sigma(\genpath)$ is a probability distribution on $\Conf$. For 
+%each measurable set of configurations $\confset$ we denote by 
+%$\sigma(\genpath)(\confset)$ the probability that the random draw from 
+%distribution $\sigma(\genpath)$ selects an element of $\confset$. We say that 
+%a 
+%scheduler $\sigma$ is measurable if for each $\confset \in 
+%\mathcal{F}_{\Conf}$ 
+%and each $p\in[0,1]$ the set $\{\pi \in \Fpath \mid 
+%\sigma(\genpath)(\confset)\leq p\}$ belongs to $\mathcal{H}$, i.e., it is a 
+%measurable set of paths.
+
+%While the definition of a measurable scheduler might seem somewhat technical, 
+%it is natural from measure-theoretic point of view and analogous definitions 
+%inevitably emerge in works dealing with systems that exhibit both 
+%probabilistic 
+%and non-deterministic behaviour over a continuous state 
+%space~\cite{NSK:CTMDP-delayed,NK:CTMDP-bisimulation}. In particular, if all 
+%the 
+%variables in a program range over a discrete set (such as the integers), 
+%then each scheduler in the associated pCFG is measurable.
 %Formally, let 
@@
 set of configurations $\confset\in\mathcal{F}_{\Conf}$ it holds 
-$\{\run\mid\cfg{\sigma}{i}(\run)\in \confset\}\in \natfilt$). Equivalently, 
-$\natfilt$ can be defined as a sigma algebra generated by all set of runs of 
-the form $F\times \Conf^{\infty}$, where $F\in \mathcal{H}$ is a measurable set 
-of finite paths. The 
+$\{\run\mid\cfg{\sigma}{i}(\run)\in \confset\}\in \natfilt$). 
+%Equivalently, 
+%$\natfilt$ can be defined as a sigma algebra generated by all set of runs of 
+%the form $F\times \Conf^{\infty}$, where $F\in \mathcal{H}$ is a measurable 
+%set 
+%of finite paths. 
+The 
 probability 

### 2017-07-07 paper/conclu.tex
small edits
added Bradley Manna Sipma citations to a couple of points

@@
 this work. 
-
-\noindent{\em Compositional a.s. termination proving.} 
+%
+%\noindent{\em Compositional a.s. termination proving.} 
 A 
@@
 ranking function has emerged as one of the most efficient and practical approaches
-for termination analysis~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal}, 
+for termination 
+analysis~\cite{DBLP:conf/cav/BradleyMS05,CSZ13,ADFG10:lexicographic,GMR15:rank-extremal},
+
 being used e.g. in 

### 2017-07-07 paper/introduction.tex
Several small changes
added Bradley Manna Sipma citations to a couple of points

@@
 notion of lexicographic ranking functions has been widely 
-studied~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal,BCIKP16:T2}. 
+studied~\cite{CSZ13,ADFG10:lexicographic,GMR15:rank-extremal,BCIKP16:T2, 
+DBLP:conf/cav/BradleyMS05}.
+
 Algorithmic approaches for linear lexicographic ranking functions allow the 
@@
 We describe our main contributions below:
-\begin{enumerate}
+\begin{compactenum}
 
@@
 Corollary~\ref{col:runtime-progs}).
+Moreover, the asymptotic bounds we obtain establish polynomial bounds on the expected termination 
+time (see Remark~\ref{rem:poly}).
 
@@
 
-\end{enumerate}
+\end{compactenum}

### 2017-07-07 paper/lexicographic-programs.tex
small typos
small correction

@@
 $\tau=(\loc,\loc')$ be a 
-transition outgoing from $\loc$. For an $\eps\geq 0$ we say that $\tau$ is 
+transition outgoing from $\loc$. For an $\eps\in \Rset$ we say that $\tau$ is 
 \emph{$\eps$-ranked by $\lem$ from $(\loc,\vec{x})$} if the following 
@@
 Let $\lem$ be a measurable map, and let $(\loc,\vec{x})$ be a configuration with
-$\loc\in \locsPB$. For an $\eps\geq 0$ we say that $\loc$ is $\eps$-ranked by 
+$\loc\in \locsPB$. For an $\eps\in \Rset$ we say that $\loc$ is $\eps$-ranked 
+by 
 $\lem$ from $(\loc,\vec{x})$ if 
@@
  \end{compactitem}
-If additionally $\lem$ is a linear expression map, then we call it a linear $\eps$-LexRSM map ($\eps$-LinLexRSM).
+If additionally $\lem$ is a linear expression map, then we call it a linear 
+$\eps$-LexRSM map ($\eps$-LinLexRSM).
 \end{definition}
@@
 the process is real-valued, componentwise non-negative, and adapted to the 
-canonical filtration of $\natfilt$. It remains to prove that condition (3) in 
+canonical filtration of $\natfilt$ (see Remark~\ref{rem:canfilt}). It remains 
+to prove that condition 
+(3) in 
 Definition~\ref{def:lexrsm} is satisfied. To this end, for each $i\geq 0$ we 

### 2017-07-08 paper/extensions.tex
small correction in Remark 2
OK

@@
 expected conditional increase} property. Recall from 
-definition~\ref{def:lexrsm} that strict
+Definition~\ref{def:lexrsm} that strict
 LexRSM for a stopping time $\stime$ is characterized by the possibility to 
@@
 
-Previous work testifies that analogous ``bounded 
-expected change'' constraints naturally emerge when one aims to obtain 
-additional information about runtime of programs 
-~\cite{HolgerPOPL,CFNH16:prob-termination,CNZ17,CF17}. The major novelty of our 
-bounded ECI condition is that it can be used to obtain polynomial bounds on 
-expected runtime using \emph{linear} supermartingales.
+\begin{remark}
+Previous works testify that analogous ``bounded expected change'' constraints 
+naturally emerge when one aims to obtain additional information about termination 
+time of probabilistic programs~\cite{HolgerPOPL,CFNH16:prob-termination,CNZ17}.
+However all previous works consider bounded expected change to use well-known 
+tail-bound inequalities for martingales such as Azuma's inequality to obtain 
+probability bounds on termination time, but not asymptotic bounds on expected
+termination time.
+In contrast, the major novelty of our bounded ECI condition is that it we show it can be used to 
+obtain asymptotic bounds on expected termination time, and in particular, they 
+can be used to obtain polynomial bounds on expected runtime using \emph{linear} 
+supermartingales.
+\end{remark}
 

### 2017-07-08 paper/lexicographic-programs.tex
OK

--- 2017-07-07 paper/lexicographic-programs.tex
@@
  \end{compactitem}
-If additionally $\lem$ is a linear expression map, then we call it a linear $\eps$-LexRSM map ($\eps$-LinLexRSM).
+If additionally $\lem$ is a linear expression map, then we call it a linear 
+$\eps$-LexRSM map ($\eps$-LinLexRSM).
 \end{definition}
 
+\begin{example}\label{ex:lexrsmbasic} 
+Consider the program shown in Figure~\ref{fig:inftime}.
+A 2-dimensional $1$-LinLexRSM map for the program is given on the right, along 
+with the supporting invariants in square brackets. The invariants and a LinLexRSM 
+on each line belong to the program location in which the program is \emph{before} 
+executing the command on that line. The function is indeed a $1$-LinLexRSM, 
+since in the probabilistic branching location $\loc$ we have 
+$\preexp{\lem}(\loc,(x,c))=3c+3$ and $3c+3\leq 6c+1-1$ for all $c\geq 1$.
+\end{example}
 %\textbf{[PETR: DEFINE LOCATION TERMINATION]}
@@
 	\usebox{\infastinv}
-\caption{An a.s. terminating program with infinite expected termination time. A 
-2-dimensional $1$-LinLexRSM map for the program is given on the right, along 
-with the 
-supporting invariants in square brackets. The invariants and a LinLexRSM on 
-each 
-line belong to the program location in which the program is \emph{before} 
-executing the command on that line. The function is indeed a $1$-LinLexRSM, 
-since in the probabilistic branching location $\loc$ we have 
-$\preexp{\lem}(\loc,(x,c))=3c+3$ and $3c+3\leq 6c+1-1$ for all $c\geq 1$.} 
+\caption{An example program that is a.s. terminating but with infinite expected termination time.} 
 \label{fig:inftime}
@@
 \label{ex:infinite-time}
-Consider the program in Figure~\ref{fig:inftime}. It is easy to see that it 
-terminates a.s., but the expected termination time is infinite: to see this, 
+Consider the program in Figure~\ref{fig:inftime}. 
+By Example~\ref{ex:lexrsmbasic} and Theorem~\ref{thm:lexrsm-programs} it follows that 
+the program terminates a.s., but the expected termination time is infinite: to see this, 
 note that that the expected value of variable $x$ upon reaching the second loop 
@@
 get out of the second loop is equal to the value of $x$ upon entering the loop. 
-However, a.s. termination of the program is proved by a 2-dimensional LinLexRSM 
-map
-$\lem$ pictured in the figure.
+%However, a.s. termination of the program is proved by a 2-dimensional LinLexRSM map
+%$\lem$ pictured in the figure.
 \end{example}

### 2017-07-08 paper/conclu.tex
OK

@@
 analysis~\cite{DBLP:conf/cav/BradleyMS05,CSZ13,ADFG10:lexicographic,GMR15:rank-extremal},
-
 being used e.g. in 

### 2017-07-08 paper/experiments.tex
correcting \eps-ranking

@@
 In all these examples lexicographic RSMs exist, and has dimension at most~3.
-Even for programs with around 12K lines of codes the total time taken is around 
+Even for programs with around 12K lines of code the total time taken is around 
 one hour, where the invariant generation (i.e., running Aspic) takes the maximum time, 

### 2017-07-08 paper/extensions.tex
correcting \eps-ranking

@@
 2x_{\mathit{init}} + 9z_{\mathit{init}}+5$. Note that this quadratic bound is 
-asymptotically optimal for the given program.
+asymptotically optimal for the program.
 \end{example}
@@
 a vector of affine expressions) we can encode the existence of $\vec{\bar{c}}$ 
-into a system of linear inequalities in the similar way as the existence of a 
-linear LexRSM maps was encoded in Section~\ref{sec:lex-programs}. That is, we 
+into a system of linear inequalities in a similar way as the existence of a 
+linear LexRSM maps was encoded in Section~\ref{sec:lex-programs}. I.e., we 
 set up a linear template with unknown coefficients for each component of 
@@
 constraints, which includes the unknown coefficients as variables, encoding the 
-fact that $\overrightarrow{\lem}$ has $\bar{\vec{c}}$-bounded ECI. Details are 
-provided 
-in~\AppendixMaterial. In this way, we can check in polynomial time if 
+fact that $\overrightarrow{\lem}$ has $\bar{\vec{c}}$-bounded ECI. In this way, 
+we can check in polynomial time if 
 Corollary~\ref{col:runtime-progs} can be applied to $\overrightarrow{\lem}$, 

### 2017-07-08 paper/compositional.tex
small edits
final edit

@@
 \label{sec:compositional}
-In this section we discuss compositional aspects. We start with the 
+In this section we discuss compositional aspects. 
+We start with the 
 one-dimensional case, and then present the multidimensional case.
@@
 Also, in both cases, a LEM of the form $x+d_{\loc}'$, again for some suitable 
-location-specific constants $d_{\loc}'$, is a PVSM for the outer loop, as $x$ 
-decreases and is non-negative within the outer loop and its expected change is 
-non-negative within the inner loop (more precisely, the inner-loop expected 
-change of $x$ zero in the left program and $-\frac{x}{4} - \frac{3}{4}$ in the 
-right 
-program). 
+location-specific constants $d_{\loc}'$, is a PVSM for the outer loop.
+%, as $x$ 
+%decreases and is non-negative within the outer loop and its expected change is 
+%non-negative within the inner loop (more precisely, the inner-loop expected 
+%change of $x$ is zero in the left program and $-\frac{x}{4} - \frac{3}{4}$ in 
+%the 
+%right 
+%program). 
 However, the variable $x$ is uniformly integrable within the inner loop of the 
@@
 \end{compactenum}
-An $n$-dimensional linear NCSM (LinNCSM) is an $n$-dimensional NCSM which is also a linear expression map.
+%An $n$-dimensional linear NCSM (LinNCSM) is an $n$-dimensional NCSM which is 
+%also a linear expression map.
 \end{definition}
 
+An $n$-dimensional linear NCSM (LinNCSM) is an $n$-dimensional NCSM which is 
+also a linear expression map.
 The following theorem can be proved in essentially the same way as Theorem~\ref{thm:nonneg-comp}.

### 2017-07-08 paper/experiments.tex
final edit

@@
+%\vspace{-2em}
 \section{Experimental Results}\label{sec:experiments}
-We have implemented the algorithm of Section~\ref{sec:algo} (in C++) and 
+We implemented the algorithm of Section~\ref{sec:algo} (in C++) and 
 present two sets of experimental results.
@@
 \textsc{CPlex}~\cite{cplex}.
-All our experimental results were obtained on the following platform: 
+The results were obtained on the following platform: 
 Ubuntu16.04, 7.7GB, Intel-Core i3-4130 CPU 3.40GHz QuadCore 64-bit.
@@
 \smallskip\noindent{\em Abstraction of real-world programs.}
-We consider benchmarks from abstraction of real-world non-probabilistic programs 
+We consider benchmarks from abstraction of real-world non-probabilistic 
+programs 
 in~\cite{ADFG10:lexicographic} (see also~\cite{CCGSY08:ranking-abstractions}).

### 2017-07-08 paper/lexicographic-programs.tex
final edit

@@
 $\tau=(\loc,\loc')$ be a 
-transition outgoing from $\loc$. For an $\eps\geq 0$ we say that $\tau$ is 
+transition outgoing from $\loc$. For an $\eps\in\Rset$ we say that $\tau$ is 
 \emph{$\eps$-ranked by $\lem$ from $(\loc,\vec{x})$} if the following 
